### 2025-04-30 docs/progress-report/sections/benchmark.tex
small changes and fixes; add docs folder

\documentclass[../ppG48.tex]{subfiles}
\begin{document}

The benchmark is designed with a modular architecture, separating the view and model layers from the controller layer, which allows for easy extension and integration of new template engines and frameworks. It also includes a set of tests to ensure the correctness of implementations and to validate the HTML output.

The benchmark includes two different data models, defined as follows: 
\begin{lstlisting}
data class Presentation(
    val id: Long,
    val title: String,
    val speakerName: String,
    val summary: String,
)

data class Stock(
    val name: String,
    val name2: String,
    val url: String,
    val symbol: String,
    val price: Double,
    val change: Double,
    val ratio: Double,
)
\end{lstlisting}

The application's repository contains a list of 10 instances of the `Presentation` class and 20 instances of the `Stock` class. Each list is used to generate a respective HTML view. Although the instances are kept in memory, the repository uses the `Observable` class from the RxJava library to interleave list items with a delay of 1 millisecond. This delay promotes context switching and frees up the calling thread to handle other requests in non-blocking scenarios, mimicking actual I/O operations.

By using the \textbf{blockingIterable()} method of the `Observable` class, we provide a blocking interface for template engines that do not support asynchronous data models, while still simulating the asynchronous nature of the data source to enable PSSR. Template engines that do not support non-blocking I/O for PSSR include KotlinX, Rocker, JStachio, Pebble, Freemarker, Trimou, and Velocity. HtmlFlow supports non-blocking I/O through suspendable templates and asynchronous rendering, while Thymeleaf enables it using the \textbf{ReactiveDataDriverContextVariable} in conjunction with a non-blocking Spring `ViewResolver`.

The Spring WebFlux implementation uses Project Reactor to support a reactive programming model: each method returns a `Flux<String>` as the response body, which acts as a publisher that progressively streams the HTML content to the client. It also includes methods using Kotlin coroutines and other asynchronous mechanisms supported by template engines, such as the \textbf{writeAsync()} method from HtmlFlow, which enables non-blocking I/O using continuations. The aforementioned blocking template engines are used in the context of Virtual Threads or alternative coroutine dispatchers, allowing the handler thread to be released and reused for other requests.

The Spring MVC implementation uses handlers based solely on the blocking interface of the `Observable` class. To enable PSSR in this context, we utilize the \textbf{StreamingResponseBody} interface, which allows the application to write directly to the response `OutputStream` without blocking the servlet container thread. According to the Spring documentation, this class is "a controller method return value type for asynchronous request processing where the application can write directly to the response OutputStream without holding up the Servlet container thread".

In Spring MVC, `StreamingResponseBody` enables asynchronous writing relative to the request-handling thread, but the underlying I/O remains blocking—specifically the writes to the `OutputStream`. When using Virtual Threads, the I/O operations are more efficient, as they are executed in the context of a lightweight thread. Most of the computation is done in a separate thread from the one that receives each request; we use a thread pool `TaskExecutor` to process requests, allowing the application to scale and handle multiple clients more efficiently as opposed to the default 'TaskExecutor' implementation, which tries to create a thread for each request.

However, the Spring MVC implementation does not effectively support PSSR for these templates, as HTML content is not streamed progressively to the client. This is because the response is only sent once the content written to the `OutputStream` exceeds the output buffer size, which defaults to 8KB. As a result, the client receives the response only after the entire HTML content is rendered, defeating the purpose of PSSR in this context. We attempted to reduce the buffer size by configuring it via an `HttpFilter`, but this did not affect the observed behavior.

The Quarkus implementation also uses handlers based on the blocking interface of the `Observable` class. It implements the \textbf{StreamingOutput} interface from the JAX-RS specification to enable PSSR, allowing HTML content to be streamed to the client. While `StreamingOutput` also uses blocking I/O, it operates on Vert.x worker threads, which prevents blocking of the event loop. When Virtual Threads are used, the I/O operations are handled efficiently, as they are executed in lightweight threads.

The Quarkus implementation supports PSSR for these templates by configuring the response buffer size in the \textit{application.properties} file. The default buffer size is 8KB, but we reduced it to 512 bytes, which allows the response to be sent to the client progressively as the HTML content is rendered.

\end{document}

### 2025-04-30 docs/progress-report/sections/conclusion.tex
small changes and fixes; add docs folder

\documentclass[../apG48.tex]{subfiles}

\begin{document}

Through benchmarking across Spring WebFlux, Spring MVC, and Quarkus, we evaluated eight different template engines and found that non-blocking implementations—especially those using virtual threads—consistently deliver performance on par with traditional blocking approaches under high concurrency. 
These results highlight virtual threads as a promising alternative to complex asynchronous programming models, offering a simpler development experience without sacrificing scalability or responsiveness.

The next steps include a deeper analysis of the benchmark results in the final report, as well as expanded coverage of the current state of non-blocking PSSR and asynchronous programming techniques. Additionally, the project may be extended to support other frameworks such as Vert.x.

\end{document}

### 2025-04-30 docs/progress-report/sections/introduction.tex
small changes and fixes; add docs folder

\documentclass[../ppG48.tex]{subfiles}

\begin{document}

Non-blocking Progressive Server-Side Rendering (PSSR) is a technique that combines some of the advantages of both server-side and client-side rendering to improve web application performance and user experience. By leveraging non-blocking I/O operations, PSSR allows for efficient handling of multiple concurrent requests while progressively streaming HTML content to the client.

This project consists of a benchmark that evaluates the performance of different PSSR strategies on the Java Virtual Machine (JVM). The benchmark focuses on evaluating the performance of non-blocking PSSR implementations using approaches such as Kotlin coroutines, reactive-style programming, and Java virtual threads. Its main objective is to evaluate virtual threads as an alternative that ensures a non-blocking approach to PSSR while offering a more straightforward programming model than traditional asynchronous programming techniques.

This project aims to extend the previous work done in \textit{spring-webflux-comparing-template-engines} \cite{spring-webflux-comparing-template-engines}, which compares the performance of various template engines in a Spring WebFlux application using different approaches to PSSR. It adds support for Java virtual threads and evaluates non-blocking PSSR implementations in other frameworks compared to Spring WebFlux.

Currently, this project features a working benchmark in Spring WebFlux, Spring MVC, and Quarkus, comparing the performance of eight different template engines: Rocker, JStachio, Pebble, Freemarker, Trimou, Thymeleaf, HtmlFlow, and KotlinX. It includes Java Microbenchmark Harness (JMH) benchmarks for template engine performance testing, as well as Apache Bench and Apache JMeter scripts for evaluating throughput and scalability. In the future, the project could be extended to include other frameworks such as Vert.x, if time allows. However, the next objective is the completion of the written report and article submission. The written report will include a more detailed analysis of the benchmark results, along with a contextualization of the state of the art in non-blocking PSSR and asynchronous programming.

This report will focus on detailing relevant aspects of the benchmark implementation and methodology, and will also provide a brief overview of the results obtained from the JMeter and Apache Bench tests, which evaluate the throughput and scalability of the different implementations.

\end{document}

### 2025-04-30 docs/progress-report/apG48.tex
small changes and fixes; add docs folder

\documentclass[a4paper,twoside,11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{graphicx}
\usepackage{url}
\usepackage{listings}
\usepackage{subfiles}
\usepackage{placeins}

\setlength{\textheight}{24.00cm}
\setlength{\textwidth}{15.50cm}
\setlength{\topmargin}{0.35cm}
\setlength{\headheight}{0cm}
\setlength{\headsep}{0cm}
\setlength{\oddsidemargin}{0.25cm}
\setlength{\evensidemargin}{0.25cm}
\setlength{\parskip}{0.1cm}

\title{
    \includegraphics[width=80mm]{logoISEL.png}\\[12pt]
    {\fontsize{14pt}{16pt}\selectfont Instituto Superior de Engenharia de Lisboa}\\[12pt]
    {\fontsize{16pt}{18pt}\selectfont \textbf{Non-blocking Progressive Server-side Rendering (PSSR) Benchmark}}\\[6pt]
    {\fontsize{11pt}{16pt}\selectfont \textbf{Progress Report}}\\[6pt]
    {\fontsize{11pt}{16pt}\selectfont Bsc in Computer Science and Engineering}\\[10pt]
}

\author{
\begin{tabular}{c}
    {\fontsize{14pt}{16pt}\selectfont Bernardo Pereira, A50493@alunos.isel.pt, n.º 50493}
\end{tabular}}

\date{
\begin{tabular}{ll}
  {\fontsize{14pt}{15pt}\selectfont Supervisor:} & {\fontsize{14pt}{16pt}\selectfont Miguel Carvalho, miguel.gamboa@isel.pt}
\end{tabular}\\
\vspace{5mm}
April, 2025}

\begin{document}

\maketitle

\vspace{-1cm}

\section{Introduction}

\subfile{sections/introduction.tex}

\section{Benchmark Implementation}

\subfile{sections/benchmark.tex}

\section{Benchmark Results}

\subfile{sections/results.tex}

\section{Conclusion}

\subfile{sections/conclusion.tex}

\bibliographystyle{unsrt}
\bibliography{referencias}

\end{document}

### 2025-04-30 docs/progress-report/sections/results.tex
small changes and fixes; add docs folder

\documentclass[../ppG48.tex]{subfiles}
\begin{document}
All the following benchmarks were conducted on a local machine running Ubuntu 22.04 LTS with a 6-core, 12-thread CPU and 32GB of RAM. All tests were conducted on the OpenJDK VM Corretto 21. The JVM was configured with a minimum heap size of 1024MB and a maximum heap size of 16GB.
In the final version of the report, the tests will be conducted on a GitHub-hosted virtual machine under GitHub Actions, running on 4 cores and 8 threads with 16GB of RAM.

For both the Apache Bench and JMeter tests, we simulate a 1000-request warmup period for each route with a concurrent user load of 32 users. The warmup period is followed by the actual test period, during which we simulate 256 requests per user, scaling in increments up to 128 concurrent users.
\subsection{Presentations Results}

\begin{figure}[h]
\centering \includegraphics[width=0.8\textwidth]{../Graphs/presentations-webflux-ab.png} \caption{Presentation Benchmark Results in Spring WebFlux with Apache Bench} \label{fig:presentations-webflux-ab} \end{figure}

\begin{figure}[h] \centering \includegraphics[width=0.8\textwidth]{../Graphs/presentations-webflux-jmeter.png} \caption{Presentation Benchmark Results in Spring WebFlux with JMeter} \label{fig:presentations-webflux-jmeter} \end{figure}

The results in Figures \ref{fig:presentations-webflux-ab} and \ref{fig:presentations-webflux-jmeter} depict the throughput (number of requests per second) for each template engine, with concurrent users ranging from 1 to 128, from left to right. The benchmarks include HtmlFlow using suspendable web templates (HtmlFlow-Susp), Jstachio using Virtual Threads (Jstachio-Virtual), and Thymeleaf using the reactive View Resolver driver (Thymeleaf-Rx). "Sync" and "Virtual" represent the average throughput of the blocking approaches (i.e., KotlinX, Rocker, Jstachio, Pebble, Freemarker, Trimou, HtmlFlow, and Thymeleaf) when run in the context of a separate coroutine dispatcher or Virtual Threads, respectively.

The results show that when using the blocking template engines with a separate coroutine dispatcher, the engines are unable to scale effectively beyond 16 concurrent users. In contrast, the non-blocking engines scale effectively up to 128 concurrent users, with HtmlFlow achieving approximately 11,000 requests per second. When using the blocking approaches in the context of Virtual Threads (achieving non-blocking I/O), the engines scale effectively up to 128 concurrent users, with Jstachio matching HtmlFlow's performance at approximately 11,000 requests per second. Thymeleaf, using the reactive View Resolver driver, also scales to 128 users, albeit less effectively, achieving around 8,500 requests per second.

Additionally, the results show no significant difference between the JMeter and Apache Bench benchmarks, indicating that both tools effectively measure the throughput of the different implementations.

\begin{figure}[h]
\centering \includegraphics[width=0.8\textwidth]{../Graphs/presentations-springmvc-ab.png} \caption{Presentation Benchmark Results in Spring MVC with Apache Bench} \label{fig:presentations-springmvc-ab} \end{figure}

\begin{figure}[h] \centering \includegraphics[width=0.8\textwidth]{../Graphs/presentations-springmvc-jmeter.png} \caption{Presentation Benchmark Results in Spring MVC with JMeter} \label{fig:presentations-springmvc-jmeter} \end{figure}

The results for the Spring MVC implementation are shown in Figures \ref{fig:presentations-springmvc-ab} and \ref{fig:presentations-springmvc-jmeter}. "Sync" represents an approach using platform threads with StreamingResponseBody, while "Virtual" uses Virtual Threads.

The results show that both approaches scale effectively up to 128 concurrent users, with the Virtual Thread approach achieving a slightly higher throughput of 9,000 requests per second. These values are slightly lower than those seen in the Spring WebFlux implementation.

There is a slight difference in measured throughput between the JMeter and Apache Bench benchmarks, with Apache Bench achieving a marginally higher throughput for the Sync approach, while the Virtual approach shows similar results across both tools.

\begin{figure}[h]
\centering
 \includegraphics[width=0.8\textwidth]{../Graphs/presentations-quarkus-ab.png} 
 \caption{Presentation Benchmark Results in Quarkus with Apache Bench}
 \label{fig:presentations-quarkus-ab}
\end{figure}

\begin{figure}[h]
     \centering 
     \includegraphics[width=0.8\textwidth]{../Graphs/presentations-quarkus-jmeter.png} 
     \caption{Presentation Benchmark Results in Quarkus with JMeter} 
     \label{fig:presentations-quarkus-jmeter} 
\end{figure}

The results for the Quarkus implementation are shown in Figures \ref{fig:presentations-quarkus-ab} and \ref{fig:presentations-quarkus-jmeter}. "Sync" represents an approach using platform threads with StreamingOutput, while "Virtual" uses Virtual Threads.

The results show that Quarkus handles blocking approaches more effectively than Spring WebFlux, with the blocking engines able to scale up to 128 concurrent users, achieving 10,000 requests per second. Using Virtual Threads results in a slightly higher throughput, but the difference is not significant, with both approaches achieving similar performance.

Again, the difference between the JMeter and Apache Bench benchmarks is negligible.

\subsection{Stocks Results}

\vspace{1cm}

\begin{figure}[h]
\centering 
\includegraphics[width=0.8\textwidth]{../Graphs/stocks-webflux-ab.png} 
\caption{Stocks Benchmark Results in Spring WebFlux with Apache Bench}
 \label{fig:stocks-webflux-ab}
\end{figure}

\begin{figure}[h] \centering \includegraphics[width=0.8\textwidth]{../Graphs/stocks-webflux-jmeter.png} \caption{Stocks Benchmark Results in Spring WebFlux with JMeter} \label{fig:stocks-webflux-jmeter} \end{figure}

The results in Figures \ref{fig:stocks-webflux-ab} and \ref{fig:stocks-webflux-jmeter} involve the same template engines and approaches as the previous benchmark, but with the data model replaced by the Stock class.

The results show that the scalability of the engines is not significantly affected by the increased complexity of the data model or the increased number of instances (20 instances of the Stock class). However, the throughput is reduced by approximately 50 percent for all engines, with HtmlFlow achieving 6,000 requests per second. The Thymeleaf implementation with the reactive View Resolver driver achieves 5,000 requests per second, indicating that there isn’t a significant performance drop compared to the Presentation benchmark.

JMeter and Apache Bench benchmarks, once again, show similar results.

\begin{figure}[h]
    \centering
    \includegraphics[width=0.8\textwidth]{../Graphs/stocks-springmvc-ab.png} 
    \caption{Stocks Benchmark Results in Spring MVC with Apache Bench}
    \label{fig:stocks-springmvc-ab} 
\end{figure}

\begin{figure}[h] \centering \includegraphics[width=0.8\textwidth]{../Graphs/stocks-springmvc-jmeter.png} \caption{Stocks Benchmark Results in Spring MVC with JMeter} \label{fig:stocks-springmvc-jmeter} \end{figure}

\vspace{4cm}

The results for the Spring MVC implementation are shown in Figures \ref{fig:stocks-springmvc-ab} and \ref{fig:stocks-springmvc-jmeter}.

The results indicate that for the Stock data model, both approaches scale effectively up to 128 concurrent users, with similar performance outcomes.

\begin{figure}[hbt!]
\centering 
\includegraphics[width=0.8\textwidth]{../Graphs/stocks-quarkus-ab.png}
 \caption{Stocks Benchmark Results in Quarkus with Apache Bench}
  \label{fig:stocks-quarkus-ab} 
\end{figure}

\begin{figure}[hbt!]
     \centering 
     \includegraphics[width=0.8\textwidth]{../Graphs/stocks-quarkus-jmeter.png} 
     \caption{Stocks Benchmark Results in Quarkus with JMeter}
     \label{fig:stocks-quarkus-jmeter}
\end{figure}

\vspace{2cm}

The results depicted in Figures \ref{fig:stocks-quarkus-ab} and \ref{fig:stocks-quarkus-jmeter} show that the Quarkus implementation scales effectively up to 128 concurrent users, achieving performance comparable to the Spring WebFlux implementation. The blocking engines reach 6,000 requests per second. Again, there is no significant difference between the Virtual Threads and platform threads approaches, with both achieving similar results

\end{document}

### 2025-04-30 docs/project-proposal/ppG48.tex
small changes and fixes; add docs folder

\documentclass[a4paper,twoside,11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{graphicx}
\usepackage{url}

\setlength{\textheight}{24.00cm}
\setlength{\textwidth}{15.50cm}
\setlength{\topmargin}{0.35cm}
\setlength{\headheight}{0cm}
\setlength{\headsep}{0cm}
\setlength{\oddsidemargin}{0.25cm}
\setlength{\evensidemargin}{0.25cm}
\setlength{\parskip}{0.1cm}

\title{
    \includegraphics[width=80mm]{logoISEL.png}\\[12pt]
    {\fontsize{14pt}{16pt}\selectfont Instituto Superior de Engenharia de Lisboa}\\[12pt]
    {\fontsize{16pt}{18pt}\selectfont \textbf{Non-blocking Progressive Server-side Rendering (PSSR) Benchmark}}\\[6pt]
    {\fontsize{11pt}{16pt}\selectfont Bsc in Computer Science and Engineering}\\[10pt]
}

\author{
\begin{tabular}{c}
    {\fontsize{14pt}{16pt}\selectfont Bernardo Pereira, A50493@alunos.isel.pt, n.º 50493}
\end{tabular}}

\date{
\begin{tabular}{ll}
  {\fontsize{14pt}{15pt}\selectfont Supervisor:} & {\fontsize{14pt}{16pt}\selectfont Miguel Carvalho, miguel.gamboa@isel.pt}
\end{tabular}\\
\vspace{5mm}
March, 2025}

\begin{document}

\maketitle

\vspace{-1cm}

\section{Introduction}

Modern web applications rely on different rendering strategies to optimize performance, user experience, and \textit{scalability}. The two most dominant approaches are \textit{Server-Side Rendering (SSR)} and \textit{Client-Side Rendering (CSR)}.

SSR generates HTML content on the server before sending it to the client, resulting in faster initial page loads and better Search Engine Optimization (SEO) \cite{prismic}. However, SSR can increase server load and reduce \textit{throughput}\footnote{The number of requests the server can handle per second (RPS)} since each request requires additional processing before responding.

In contrast, CSR shifts the rendering workload to the browser. The server initially sends a minimal HTML document with JavaScript, which dynamically loads the page content. While CSR reduces the server’s burden, it can lead to slower initial load times, as users must wait for JavaScript execution before meaningful content appears \cite{nextjs}.

\textit{Progressive Server-Side Rendering (PSSR)} combines benefits from both SSR and CSR by streaming HTML content progressively. This technique enhances user-perceived performance by allowing immediate rendering while data loads asynchronously.

Efficient thread utilization is crucial for low-thread servers, necessitating asynchronous, non-blocking request handling \cite{webist23}. Non-blocking PSSR is particularly advantageous in event-driven architectures, where asynchronous I/O operations (e.g., database queries, and API calls) allow servers to manage multiple concurrent requests efficiently. By reducing idle time during I/O operations, non-blocking PSSR enhances a server's capacity to handle high request volumes with limited threads and immediately send partial responses to clients.

This work will focus on exploring Kotlin Coroutines and Structured Concurrency, as well as Java Virtual Threads as an alternative to asynchronous programming techniques, to evaluate their performance in non-blocking PSSR \cite{kotlin_coroutines, virtual_threads}.

The benchmark will be built on Spring WebFlux and analyze methods such as those presented in \textit{Progressive Server-Side Rendering with Suspendable Web Templates} \cite{pssr-susp}. Additionally, it will build upon and extend the \textit{Spring WebFlux Comparing Template Engines} repository \cite{spring-comparing-template-engines} with the addition of approaches using Java Virtual Threads.

To achieve this, benchmarking tools such as JMH, Apache Bench, and Apache JMeter will be utilized to assess performance metrics, including scalability, throughput, and latency \cite{benchmarking_web, jmh_benchmark}. The accuracy and setup complexity of these tools will also be compared.

\section{Deliverables}

This project consists of three main components: the development of the benchmark, the creation of a report analyzing the state-of-the-art and benchmark results, and the submission of an article to a technology journal.

The main deliverables include:

\begin{itemize}
  \item \textbf{A PSSR Benchmark in the JVM}
  \begin{itemize}
    \item Evaluate PSSR strategies with the use of currently existing libraries and frameworks (e.g., HtmlFlow, Thymeleaf, etc.).
    \item Compare the performance of these implementations in terms of throughput, latency, and scalability.
    \item Assess the feasibility of using Apache JMeter instead of Apache Bench for performance testing.
    \item Compare the ease of setup and accuracy of results between the two tools.
    \item Evaluate the performance and viability of using Java Virtual Threads for non-blocking PSSR.
  \end{itemize}

  \item \textbf{A Research Report}
  \begin{itemize}
    \item Provide an introduction and contextualization of progressive rendering and asynchronous request handling.
    \item Analyze existing literature and state-of-the-art approaches in non-blocking web rendering.
    \item Present an analysis of benchmark results, highlighting trade-offs between different PSSR implementations.
  \end{itemize}

  \item \textbf{An Article submission}
  \begin{itemize}
    \item Written article for submission in a technology journal.
    \item The article will present the benchmark results and analysis, providing insights into the performance of different PSSR strategies in the JVM.
  \end{itemize}
\end{itemize}

\section{Planning}

\begin{table}[h]
  \centering
  \renewcommand{\arraystretch}{1.3}
  \begin{tabular}{|p{8cm}|c|c|p{6cm}|}
      \hline
      \textbf{Task} & \textbf{Duration} & \textbf{Deadline} \\
      \hline
      Research on PSSR, reactive data models, Spring WebFlux & 2 weeks & March 3 \\
      \hline
      Project proposal & 1 week & March 10 \\
      \hline
      Proposal presentation & 1 week & March 17 \\
      \hline
      Benchmark development & 3 weeks & April 7 \\
      \hline
      Initial writing phase – Introduction, contextualization & 3 weeks & April 28 \\
      \hline
      Project status presentation & 1 week & May 5 \\
      \hline
      Final writing phase – Refinements, result analysis, conclusions & 3 weeks & May 26 \\
      \hline
      Draft article, beta submission & 1 week & June 2 \\
      \hline
      Writing quality improvements & 3 weeks & June 30 \\
      \hline
      Final article version & 1 week & July 5 \\
      \hline
      Final submission & 1 week & July 12 \\
      \hline
  \end{tabular}
  \label{tab:planning}
\end{table}

\bibliographystyle{unsrt}
\bibliography{referencias}

\end{document}

### 2025-04-30 benches/ab/bench-mvc.sh

@@ -1,6 +1,6 @@
 cd ../../code || exit
 
-./gradlew runMVC -DbenchTimeout=1 > ../benches/ab/spring-mvc.log &
+./gradlew runMVC -DbenchTimeout=1 -DXms=1024m -DXmx=16G > ../benches/ab/spring-mvc.log &
 PID_GRADLE=$!
 
 cd ../benches/ab || exit
@@ -22,7 +22,7 @@ ROUTES=(
    presentations/pebble
    presentations/freemarker
    presentations/trimou
-   presentations/velocity
+#   presentations/velocity
    presentations/thymeleaf
    presentations/htmlFlow
    presentations/kotlinx
@@ -31,7 +31,7 @@ ROUTES=(
    stocks/pebble
    stocks/freemarker
    stocks/trimou
-   stocks/velocity
+#   stocks/velocity
    stocks/thymeleaf
    stocks/htmlFlow
 )
@@ -68,7 +68,7 @@ echo ":::::::::::::::::::::::::::::::     Sync Bench Done"
 
 cd ../../code || exit
 
-./gradlew runMVCVirtual -DbenchTimeout=1 > ../benches/ab/spring-mvc.log &
+./gradlew runMVCVirtual -DbenchTimeout=1 -DXms=1024m -DXmx=16G > ../benches/ab/spring-mvc.log &
 PID_GRADLE=$!
 
 cd ../benches/ab || exit

### 2025-04-30 benches/ab/bench-quarkus.sh

@@ -1,6 +1,6 @@
 cd ../../code || exit
 
-./gradlew runQuarkus -DbenchTimeout=1 -Djdk.tracePinnedThreads > ../benches/ab/quarkus.log &
+./gradlew runQuarkus -DbenchTimeout=1 -DXms=1024m -DXmx=16G -Djdk.tracePinnedThreads > ../benches/ab/quarkus.log &
 PID_GRADLE=$!
 
 cd ../benches/ab || exit
@@ -24,7 +24,7 @@ ROUTES=(
   presentations/pebble
   presentations/freemarker
   presentations/trimou
-  presentations/velocity
+#  presentations/velocity
   presentations/thymeleaf
   presentations/htmlFlow
   presentations/kotlinx
@@ -33,7 +33,7 @@ ROUTES=(
   stocks/pebble
   stocks/freemarker
   stocks/trimou
-  stocks/velocity
+#  stocks/velocity
   stocks/thymeleaf
   stocks/htmlFlow
 )
@@ -71,7 +71,7 @@ echo ":::::::::::::::::::::::::::::::     Sync Bench Done"
 
 cd ../../code || exit
 
-./gradlew runQuarkusVirtual -DbenchTimeout=1 > ../benches/ab/quarkus.log &
+./gradlew runQuarkusVirtual -DbenchTimeout=1 -DXms=1024m -DXmx=16G > ../benches/ab/quarkus.log &
 PID_GRADLE=$!
 
 cd ../benches/ab || exit

### 2025-04-30 benches/ab/bench-webflux.sh

@@ -6,7 +6,7 @@
 
 cd ../../code || exit
 
-./gradlew runWebflux -DbenchTimeout=1 > ../benches/ab/spring-webflux.log &
+./gradlew runWebflux -DbenchTimeout=1 -DXms=1024m -DXmx=16G > ../benches/ab/spring-webflux.log &
 PID_GRADLE=$!
 
 cd ../benches/ab || exit
@@ -23,7 +23,6 @@ echo ":::::::::::::::::::::::::::::::     Gradle running PID = $PID_GRADLE"
 
 #
 # Define routes for benchmark
-#
 #
 ROUTES=(
    presentations/thymeleaf
@@ -46,16 +45,17 @@ ROUTES=(
    presentations/freemarker/virtualSync
    presentations/trimou/sync
    presentations/trimou/virtualSync
-   presentations/velocity/sync
-   presentations/velocity/virtualSync
+#   presentations/velocity/sync
+#   presentations/velocity/virtualSync
    presentations/thymeleaf
+    stocks/thymeleaf
    stocks/thymeleaf/sync
    stocks/thymeleaf/virtualSync
-   stocks/htmlFlow
+#   stocks/htmlFlow
    stocks/htmlFlow/suspending
    stocks/htmlFlow/sync
    stocks/htmlFlow/virtualSync
-   stocks/kotlinx
+#   stocks/kotlinx
    stocks/kotlinx/sync
    stocks/kotlinx/virtualSync
    stocks/rocker/sync
@@ -68,8 +68,8 @@ ROUTES=(
    stocks/freemarker/virtualSync
    stocks/trimou/sync
    stocks/trimou/virtualSync
-   stocks/velocity/sync
-   stocks/velocity/virtualSync
+#   stocks/velocity/sync
+#   stocks/velocity/virtualSync
 )
 
 #

### 2025-04-30 benches/ab/run-ab.sh

@@ -1,7 +1,7 @@
 #!/bin/bash
 
 REQUESTS=256
-THREADS=(1 2 4 8 16 32 64 128 256)
+THREADS=(1 2 4 8 16 32 64 128)
 ITERATIONS=2
 
 for ths in "${THREADS[@]}"; do   # For each number of threads

### 2025-04-30 benches/jmeter/bench-mvc.sh

@@ -1,6 +1,6 @@
 cd ../../code || exit
 
-./gradlew runMVC -DbenchTimeout=1 > ../benches/jmeter/spring-mvc.log &
+./gradlew runMVC -DbenchTimeout=1 -DXms=1024m -DXmx=16G > ../benches/jmeter/spring-mvc.log &
 
 PID_GRADLE=$!
 
@@ -22,7 +22,7 @@ ROUTES=(
    presentations/pebble
    presentations/freemarker
    presentations/trimou
-   presentations/velocity
+#   presentations/velocity
    presentations/thymeleaf
    presentations/htmlFlow
    presentations/kotlinx
@@ -31,7 +31,7 @@ ROUTES=(
    stocks/pebble
    stocks/freemarker
    stocks/trimou
-   stocks/velocity
+#   stocks/velocity
    stocks/thymeleaf
    stocks/htmlFlow
 )
@@ -51,36 +51,36 @@ if [ "$GH" != "true" ]; then
   wait $PID_GRADLE
 fi
 
-cd ../../code || exit
-
-./gradlew runMVCVirtual -DbenchTimeout=1 > ../benches/jmeter/spring-mvc.log &
-
-PID_GRADLE=$!
-
-cd ../benches/jmeter || exit
-
-sleep 1
-while ! grep -m1 'Tomcat started on port 8080' < spring-mvc.log; do
-    sleep 1
-done
-
-PID_MVC=$(grep -oP 'with PID \K[0-9]+' spring-mvc.log)
-
-echo ":::::::::::::::::::::::::::::::     Spring running PID = $PID_MVC"
-echo ":::::::::::::::::::::::::::::::     Gradle running PID = $PID_GRADLE"
-
-echo "##########################################"
-echo "############# RUN BENCH ##################"
-echo "##########################################"
-./run-jmeter.sh "${ROUTES[@]}" | tee spring-mvc-virtual-results.log
-
-# Gracefully terminate the Spring Boot application when running on local machine.
-# It will send a SIGTERM corresponding to Exit code 143.
-if [ "$GH" != "true" ]; then
-  kill $PID_GRADLE
-  kill "$PID_MVC"
-
-  # Wait for the process to exit
-  wait $PID_GRADLE
-fi
+#cd ../../code || exit
+#
+#./gradlew runMVCVirtual -DbenchTimeout=1 -DXms=1024m -DXmx=16G > ../benches/jmeter/spring-mvc.log &
+#
+#PID_GRADLE=$!
+#
[…]

### 2025-04-30 benches/jmeter/bench-quarkus.sh

@@ -1,6 +1,6 @@
 cd ../../code || exit
 
-./gradlew runQuarkus -DbenchTimeout=1 -Djdk.tracePinnedThreads > ../benches/jmeter/quarkus.log &
+./gradlew runQuarkus -DbenchTimeout=1 -DXms=1024m -DXmx=16G -Djdk.tracePinnedThreads > ../benches/jmeter/quarkus.log &
 PID_GRADLE=$!
 
 cd ../benches/jmeter || exit
@@ -16,23 +16,23 @@ echo ":::::::::::::::::::::::::::::::     Gradle running PID = $PID_GRADLE"
 echo ":::::::::::::::::::::::::::::::     Quarkus running PID = $PID_QUARKUS"
 
 ROUTES=(
-#  presentations/rocker
-#  presentations/jstachio
-#  presentations/pebble
-#  presentations/freemarker
-#  presentations/trimou
+  presentations/rocker
+  presentations/jstachio
+  presentations/pebble
+  presentations/freemarker
+  presentations/trimou
 #  presentations/velocity
-#  presentations/thymeleaf
-#  presentations/htmlFlow
-#  presentations/kotlinx
-#  stocks/rocker
-#  stocks/jstachio
-#  stocks/pebble
-#  stocks/freemarker
-#  stocks/trimou
+  presentations/thymeleaf
+  presentations/htmlFlow
+  presentations/kotlinx
+  stocks/rocker
+  stocks/jstachio
+  stocks/pebble
+  stocks/freemarker
+  stocks/trimou
 #  stocks/velocity
-#  stocks/thymeleaf
-#  stocks/htmlFlow
+  stocks/thymeleaf
+  stocks/htmlFlow
 )
 
 #
@@ -58,7 +58,7 @@ echo ":::::::::::::::::::::::::::::::     Sync Bench Done"
 
 cd ../../code || exit
 
-./gradlew runQuarkusVirtual -DbenchTimeout=1 > ../benches/jmeter/quarkus.log &
+./gradlew runQuarkusVirtual -DbenchTimeout=1 -DXms=1024m -DXmx=16G > ../benches/jmeter/quarkus.log &
 PID_GRADLE=$!
 
 cd ../benches/jmeter || exit

### 2025-04-30 benches/jmeter/bench-webflux.sh

@@ -1,6 +1,6 @@
 cd ../../code || exit
 
-./gradlew runWebflux -DbenchTimeout=1 > ../benches/jmeter/spring-webflux.log &
+./gradlew runWebflux -DbenchTimeout=1 -DXms=1024m -DXmx=16G > ../benches/jmeter/spring-webflux.log &
 
 PID_GRADLE=$!
 
@@ -32,16 +32,16 @@ ROUTES=(
   presentations/freemarker/virtualSync
   presentations/trimou/sync
   presentations/trimou/virtualSync
-  presentations/velocity/sync
-  presentations/velocity/virtualSync
-  presentations/thymeleaf
+#  presentations/velocity/sync
+#  presentations/velocity/virtualSync
+  stocks/thymeleaf
   stocks/thymeleaf/sync
   stocks/thymeleaf/virtualSync
-  stocks/htmlFlow
+#  stocks/htmlFlow
   stocks/htmlFlow/suspending
   stocks/htmlFlow/sync
   stocks/htmlFlow/virtualSync
-  stocks/kotlinx
+#  stocks/kotlinx
   stocks/kotlinx/sync
   stocks/kotlinx/virtualSync
   stocks/rocker/sync
@@ -54,8 +54,8 @@ ROUTES=(
   stocks/freemarker/virtualSync
   stocks/trimou/sync
   stocks/trimou/virtualSync
-  stocks/velocity/sync
-  stocks/velocity/virtualSync
+#  stocks/velocity/sync
+#  stocks/velocity/virtualSync
 )
 
 PID_WEBFLUX=$(grep -oP 'with PID \K[0-9]+' spring-webflux.log)

### 2025-05-14 docs/article/Sections/benchmark.tex
article 50%~ completed

\section{Benchmark Implementation}

The benchmark is designed with a modular architecture, separating the
\textit{view} and \textit{model} layers from the \textit{controller} layer
\citeauthor{Bucanek2009}, which allows for easy extension and integration of new
template engines and frameworks. It also includes a set of tests to ensure the
correctness of implementations and to validate the \textit{HTML} output.

The benchmark includes two different data models, defined as follows:

\lstdefinestyle{listingstyle}{
  backgroundcolor=\color{lightgray!20},
  basicstyle=\ttfamily\small,
  breaklines=true,
  frame=single,
  captionpos=b,
  numbers=left,
  numberstyle=\tiny,
  keywordstyle=\color{blue},
  commentstyle=\color{gray},
  stringstyle=\color{orange},
}

\lstset{style=listingstyle}

\begin{lstlisting}[language=Kotlin, caption={Data Models}, label={lst:data-models}]
data class Presentation(
    val id: Long,
    val title: String, 
    val speakerName: String,
    val summary: String
)

data class Stock(
    val name: String,
    val name2: String,
    val url: String,
    val symbol: String,
    val price: Double, 
    val change: Double, 
    val ratio: Double
)
\end{lstlisting}

The application's repository contains a list of 10 instances of the
\texttt{Presentation} class and 20 instances of the \texttt{Stock} class. Each list is used
to generate a respective HTML view. Although the instances are kept in memory,
the repository uses the \texttt{Observable} class from the \textit{RxJava}
library to interleave list items with a delay of 1 millisecond. This delay
promotes context switching and frees up the calling thread to handle other
requests in non-blocking scenarios, mimicking actual I/O operations.

By using the \texttt{blockingIterable} method of the \texttt{Observable}
class, we provide a blocking interface for template engines that do not support
asynchronous data models, while still simulating the asynchronous nature of the
data source to enable PSSR\@. Template engines that do not support non-blocking
I/O for PSSR include KotlinX, Rocker, JStachio, Pebble, Freemarker, Trimou, and
Velocity. HtmlFlow supports non-blocking I/O through suspendable templates and
asynchronous rendering, while Thymeleaf enables it using the
\texttt{ReactiveDataDriverContextVariable} in conjunction with a non-blocking
Spring \texttt{ViewResolver}.

The Spring WebFlux implementation uses Project Reactor to support a reactive
programming model: each method returns a \texttt{Flux<String>} as the response
body, which acts as a publisher that progressively streams the HTML content to
the client. It also includes methods using Kotlin coroutines and other
asynchronous mechanisms supported by template engines, such as the
\texttt{writeAsync} method from HtmlFlow, which enables non-blocking I/O
using \textit{continuations}. The aforementioned blocking template engines are used in
the context of Virtual Threads or alternative coroutine dispatchers, allowing
the handler thread to be released and reused for other requests.

The Spring MVC implementation uses handlers based solely on the blocking
interface of the \texttt{Observable} class. To enable PSSR in this context, we utilize
the \texttt{StreamingResponseBody} interface, which allows the application to
write directly to the response \texttt{OutputStream} without blocking the servlet
container thread. According to the Spring documentation, this class is \textit{a
controller method return value type for asynchronous request processing where
the application can write directly to the response OutputStream without holding
up the Servlet container thread} \citeauthor{Spring-StreamingResponseBody}.

In Spring MVC, \texttt{StreamingResponseBody} enables asynchronous writing relative to
the request-handling thread, but the underlying I/O remains
blocking—specifically the writes to the \texttt{OutputStream}. When using Virtual
Threads, the I/O operations are more efficient, as they are executed in the
context of a lightweight thread. Most of the computation is done in a separate
thread from the one that receives each request; we use a thread pool
\texttt{TaskExecutor} to process requests, allowing the application to scale and
handle multiple clients more efficiently as opposed to the default
\texttt{TaskExecutor} implementation, which tries to create a thread for each request.

However, the Spring MVC implementation does not effectively support PSSR for
these templates, as HTML content is not streamed progressively to the client.
This is because the response is only sent once the content written to the
\texttt{OutputStream} exceeds the output buffer size, which defaults to 8KB\@. As a
result, the client receives the response only after the entire HTML content is
rendered, defeating the purpose of PSSR in this context. We attempted to reduce
the buffer size by configuring it via an \texttt{HttpFilter}, but this did not affect
the observed behavior.

The Quarkus implementation also uses handlers based on the blocking interface
of the \texttt{Observable} class. It implements the \textbf{StreamingOutput} interface
from the JAX-RS specification to enable PSSR, allowing HTML content to be
streamed to the client. While \texttt{StreamingOutput} also uses blocking I/O, it
operates on Vert.x worker threads, which prevents blocking of the event loop.
When Virtual Threads are used, the I/O operations are handled efficiently, as
they are executed in lightweight threads.

The Quarkus implementation supports PSSR for these templates by configuring the
response buffer size in the \textit{application.properties} file. The default
buffer size is 8KB, but we reduced it to 512 bytes, which allows the response
to be sent to the client progressively as the HTML content is rendered.

### 2025-05-14 docs/article/Sections/introduction.tex
article 50%~ completed

\section{Introduction}

% Non-blocking Progressive Server-Side Rendering (PSSR) is a technique that
% combines some of the advantages of both server-side and client-side rendering
% to improve web application performance and user experience. By leveraging
% non-blocking I/O operations, PSSR allows for efficient handling of multiple
% concurrent requests while progressively streaming HTML content to the client.
% 
% This project consists of a benchmark that evaluates the performance of
% different PSSR strategies on the Java Virtual Machine (JVM). The benchmark
% focuses on evaluating the performance of non-blocking PSSR implementations
% using approaches such as Kotlin coroutines, reactive-style programming, and
% Java virtual threads. Its main objective is to evaluate virtual threads as an
% alternative that ensures a non-blocking approach to PSSR while offering a more
% straightforward programming model than traditional asynchronous programming
% techniques.
% 
% This project aims to extend the previous work done in
% \textit{spring-webflux-comparing-template-engines}
% \citeauthor{spring-webflux-comparing-template-engines}, which compares the
% performance of various template engines in a Spring WebFlux application using
% different approaches to PSSR\@. It adds support for Java virtual threads and
% evaluates non-blocking PSSR implementations in other frameworks compared to
% Spring WebFlux.
% 
% Currently, this project features a working benchmark in Spring WebFlux, Spring
% MVC, and Quarkus, comparing the performance of eight different template
% engines: Rocker, JStachio, Pebble, Freemarker, Trimou, Thymeleaf, HtmlFlow, and
% KotlinX. It includes Java Microbenchmark Harness (JMH) benchmarks for template
% engine performance testing, as well as Apache Bench and Apache JMeter scripts
% for evaluating throughput and scalability. In the future, the project could be
% extended to include other frameworks such as Vert.x, if time allows. However,
% the next objective is the completion of the written report and article
% submission. The written report will include a more detailed analysis of the
% benchmark results, along with a contextualization of the state of the art in
% non-blocking PSSR and asynchronous programming.
% 
% This report will focus on detailing relevant aspects of the benchmark
% implementation and methodology, and will also provide a brief overview of the
% results obtained from the JMeter and Apache Bench tests, which evaluate the
% throughput and scalability of the different implementations.

### 2025-05-14 docs/article/Sections/results.tex
article 50%~ completed

\section{Results}

All the following benchmarks were conducted on a local machine running Ubuntu
22.04 LTS with a 6-core, 12-thread CPU and 32GB of RAM\@. All tests were
conducted on the OpenJDK VM Corretto 21. The JVM was configured with a minimum
heap size of 1024MB and a maximum heap size of 16GB\@.

For both the Apache Bench and JMeter tests, we simulate a 1000-request warmup
period for each route with a concurrent user load of 32 users. The warmup
period is followed by the actual test period, during which we simulate 256
requests per user, scaling in increments up to 128 concurrent users.

The results are presented in the form of throughput (number of requests per
second) for each template engine, with the x-axis representing the number of
concurrent users and the y-axis representing the throughput in requests per
second.

Since the obtained results for JMeter and Apache Bench show no significant
differences, only the JMeter results will be presented.

\subsection{Presentations Results}

The results in Figure~\ref{fig:presentations-webflux-jmeter} depict the
throughput (number of requests per second) for each template engine, with
concurrent users ranging from 1 to 128, from left to right. The benchmarks
include HtmlFlow using suspendable web templates (HtmlFlow-Susp), Jstachio
using Virtual Threads (Jstachio-Virtual), and Thymeleaf using the reactive View
Resolver driver (Thymeleaf-Rx). \textit{Sync} and \textit{Virtual} represent
the average throughput of the blocking approaches (i.e., KotlinX, Rocker,
Jstachio, Pebble, Freemarker, Trimou, HtmlFlow, and Thymeleaf) when run in the
context of a separate coroutine dispatcher or Virtual Threads, respectively.

We show the HtmlFlow-Susp, Jstachio-Virtual, and Thymeleaf-Rx engines
separately to observe the performance of the non-blocking engines when using
the Reactive, Virtual Thread, and Suspending approaches. The \textit{Sync} and
\textit{Virtual} are aggregated due to the similar performance of different
engines when using those approaches.

\begin{figure}[h]
     \centering
     \includegraphics[width=0.8\textwidth]{./Graphs/presentations-webflux-jmeter.png}
     \caption{Presentation Benchmark Results in Spring WebFlux with JMeter}\label{fig:presentations-webflux-jmeter}
\end{figure}

The results show that when using the blocking template engines with a separate
coroutine dispatcher, the engines are unable to scale effectively beyond 16
concurrent users. In contrast, the non-blocking engines scale effectively up to
128 concurrent users, with HtmlFlow achieving approximately 11,000 requests per
second. When using the blocking approaches in the context of Virtual Threads
(achieving non-blocking I/O), the engines scale effectively up to 128
concurrent users, with Jstachio matching HtmlFlow`s performance at
approximately 11,000 requests per second. Thymeleaf, using the reactive View
Resolver driver, also scales to 128 users, albeit less effectively, achieving
around 8,500 requests per second.

\begin{figure}[h]
     \centering
     \includegraphics[width=0.8\textwidth]{./Graphs/presentations-springmvc-jmeter.png}
     \caption{Presentation Benchmark Results in Spring MVC with JMeter}\label{fig:presentations-springmvc-jmeter}
\end{figure}

The results for the Spring MVC implementation, shown in
Figure~\ref{fig:presentations-springmvc-jmeter}, compare two approaches:
\textit{Sync}, which uses platform threads with \texttt{StreamingResponseBody},
and \textit{Virtual}, which uses Virtual Threads. Both approaches scale
effectively up to 128 concurrent users, with the Virtual Thread approach
achieving a slightly higher throughput of 9,000 requests per second. However,
these values are slightly lower than those observed in the Spring WebFlux
implementation.

\begin{figure}[h]
     \centering
     \includegraphics[width=0.8\textwidth]{./Graphs/presentations-quarkus-jmeter.png}
     \caption{Presentation Benchmark Results in Quarkus with JMeter}\label{fig:presentations-quarkus-jmeter}
\end{figure}

The results for the Quarkus implementation, shown in
Figure~\ref{fig:presentations-quarkus-jmeter} demonstrate that Quarkus handles
blocking approaches more effectively than Spring WebFlux, with the blocking
engines scaling up to 128 concurrent users and achieving 10,000 requests per
second. While the use of Virtual Threads results in a slightly higher
throughput, the difference is not significant, as both approaches deliver
similar performance.

\subsection{Stocks Results}

\begin{figure}[h]
     \centering
     \includegraphics[width=0.8\textwidth]{./Graphs/stocks-webflux-jmeter.png}
     \caption{Stocks Benchmark Results in Spring WebFlux with JMeter}\label{fig:stocks-webflux-jmeter}
\end{figure}

The results in Figure~\ref{fig:stocks-webflux-jmeter} use the same template
engines and approaches as the previous benchmark, but replace the data model
with the more complex Stock class, including 20 instances. Despite the
increased complexity and number of instances, the scalability of the engines
remains largely unaffected. However, throughput is reduced by approximately 50
percent across all engines, with HtmlFlow achieving 6,000 requests per second.
The Thymeleaf implementation using the reactive View Resolver driver reaches
5,000 requests per second, suggesting that the performance drop compared to the
Presentation benchmark is not substantial.

\begin{figure}[h]
     \centering
     \includegraphics[width=0.8\textwidth]{./Graphs/stocks-springmvc-jmeter.png}
     \caption{Stocks Benchmark Results in Spring MVC with JMeter}\label{fig:stocks-springmvc-jmeter}
\end{figure}

The results observed in Figure~\ref{fig:stocks-springmvc-jmeter} show that the
Spring MVC implementation using the blocking approach with
\texttt{StreamingResponseBody} achieves a throughput of up to 5500 requests per
second, while no significant change is observed when using Virtual Threads. As
such, both approaches scale effectively up to 128 concurrent users.

\begin{figure}[h]
     \centering
     \includegraphics[width=0.8\textwidth]{./Graphs/stocks-quarkus-jmeter.png}
     \caption{Stocks Benchmark Results in Quarkus with JMeter}\label{fig:stocks-quarkus-jmeter}
\end{figure}

The results depicted in Figure~\ref{fig:stocks-quarkus-jmeter} show that the
Quarkus implementation scales effectively up to 128 concurrent users, achieving
performance comparable to the Spring WebFlux implementation. The blocking
engines reach 6,000 requests per second. Again, there is no significant
difference between the Virtual Threads and platform threads approaches, with
both achieving similar results.

The results of the benchmarks show that non-blocking engines, through the use
of reactive programming, Kotlin coroutines, or Java virtual threads, are able
to scale effectively up to 128 concurrent users. Out of all the tested
frameworks, Spring Webflux showed itself the most effective at enabling PSSR,
mostly due to its native support for Publish and Subscriber interfaces, which
allow for HTML content to be progressively streamed to the client. Quarkus also
enabled PSSR effectively, but it required additional configuration of the
\texttt{OutputBuffer} size to achieve the same results as Spring Webflux. The
Spring MVC implementation, on the other hand, did not enable PSSR for the
tested templates.

Additionally, the results show that approaches using Virtual Threads are able
to scale as effectively as those using reactive programming or Kotlin
coroutines, with the advantage of being easier to implement and understand for 
developers.

### 2025-05-14 docs/article/Sections/state-of-the-art.tex
article 50%~ completed

\section{State of the Art}

\subsection{Related Work}

\subsection{Web Framework Architectures and Approaches to PSSR}

\subsubsection{Spring WebFlux}

\subsubsection{Spring MVC}

\subsubsection{Quarkus}

### 2025-05-14 docs/article/Sections/state-of-the-art.tex
Chapter 2 - Introduce State of Art divided in Related Work on web template engines and Web Servers approaches. Start Related Work and Web templates basics. Introduce the main characteristics of web templates.

@@
 
+In this section, we first present the main properties that characterize each web
+template technology approach, along with the advantages and drawbacks resulting
+from these characteristics.
+
+Next, we delve into the different design choices adopted by web servers in their
+internal architectures, and how these choices impact the behavior of the web
+template engines described in the previous subsection.
+
 \subsection{Related Work}
+
+\textit{Web templates} have been the most widely adopted approach for
+constructing dynamic HTML pages.
+Web templates (such as JSP, Handlebars, or Thymeleaf), also known as a
+\textit{web pages} or \textit{web views}~\cite{Fowler02,Alur01}, are based on
+HTML documents augmented with template-specific markers (e.g., \texttt{<\%>},
+\texttt{\{\{\}\}}, or \texttt{\$\{\}}), which represent \textit{dynamic}
+information to be replaced at runtime with the results of corresponding
+computations, producing the final HTML page.
+The process of parsing and replacing these markers---i.e.,
+\textit{resolution}---is the primary responsibility of the \textit{template
+engine}~\cite{Parr04}.
+
+One key characteristic of \textit{web templates} is their ability to receive a 
+\textit{context object}---equivalent to the \textit{model} in the model-view 
+design pattern~\cite{mvc88,Parr04}---which provides the data used to fill 
+template placeholders at runtime.
+
+Web templates can be distinguished by several properties, namely:
+\begin{enumerate}
+    \item Templating idiom
+    \item Supported data model APIs
+    \item Asynchronous support
+    \item Type safety and HTML safety
+    \item Progressive rendering
+\end{enumerate}
+
+Although some of the aforementioned characteristics apply to both server-side 
+and client-side approaches, we focus solely on web template technologies for 
+server-side rendering, as our work is centered on that approach.
+
+\subsubsection{Templating idiom}
+
+\subsubsection{Supported data model APIs}
+
+Restricted or unrestricted.
+
+\subsubsection{Asynchronous support}
+
+Explicar Observable e modelos reactivos
+
+Distinguish between template engines that support and do not support asynchronous data models.
+
+1. Nem se quer Existe API para iterar sobre um Async data model. 
+   E.g JStachio, .... apenas suportam interface Iterable. => Compilation Error.
+2. Consegue usar qq API e.g Kotlinx.Html and Groovy  => usa mal => produzir HTML out of order
+3. Sim suporta => well-formed HTML => Thymleaf e o HtmlFlow
+
+\subsubsection{Type safety and HTML safety}
+
+\subsubsection{Progressive rendering}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%-------------------------------------------------------------------
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 

### 2025-05-14 docs/article/Sections/benchmark.tex
small improvements and fixes to chapters 4 and 5 of the article; changed journal to Software.

@@
 \textit{view} and \textit{model} layers from the \textit{controller} layer
-\citeauthor{Bucanek2009}, which allows for easy extension and integration of new
+\cite{Bucanek2009}, which allows for easy extension and integration of new
 template engines and frameworks. It also includes a set of tests to ensure the
@@
 
-The benchmark includes two different data models, defined as follows:
+The benchmark includes two different data models, defined as \texttt{Presentation} and 
+\texttt{Stock} shown in \autoref{lst:data-models}.
 
@@
 
-The Spring WebFlux implementation uses Project Reactor to support a reactive
+The aforementioned blocking template engines are used in
+the context of Virtual Threads or alternative coroutine dispatchers, allowing
+the handler thread to be released and reused for other requests.
+
+The Spring WebFlux core implementation uses Project Reactor to support a reactive
 programming model: each method returns a \texttt{Flux<String>} as the response
@@
 \texttt{writeAsync} method from HtmlFlow, which enables non-blocking I/O
-using \textit{continuations}. The aforementioned blocking template engines are used in
-the context of Virtual Threads or alternative coroutine dispatchers, allowing
-the handler thread to be released and reused for other requests.
-
+using \textit{continuations}. 
 The Spring MVC implementation uses handlers based solely on the blocking
@@
 the application can write directly to the response OutputStream without holding
-up the Servlet container thread} \citeauthor{Spring-StreamingResponseBody}.
+up the Servlet container thread} \cite{Spring-StreamingResponseBody}.
 
@@
 blocking—specifically the writes to the \texttt{OutputStream}. When using Virtual
-Threads, the I/O operations are more efficient, as they are executed in the
-context of a lightweight thread. Most of the computation is done in a separate
+Threads, the I/O operations are more efficient when compared to platform threads,
+as they are executed in the context of a lightweight thread. Most of the computation is done in a separate
 thread from the one that receives each request; we use a thread pool
@@
 result, the client receives the response only after the entire HTML content is
-rendered, defeating the purpose of PSSR in this context. We attempted to reduce
-the buffer size by configuring it via an \texttt{HttpFilter}, but this did not affect
-the observed behavior.
+rendered, defeating the purpose of PSSR in this context.
 
 The Quarkus implementation also uses handlers based on the blocking interface
-of the \texttt{Observable} class. It implements the \textbf{StreamingOutput} interface
+of the \texttt{Observable} class. It implements the \texttt{StreamingOutput} interface
 from the JAX-RS specification to enable PSSR, allowing HTML content to be

### 2025-05-14 docs/article/Sections/results.tex
small improvements and fixes to chapters 4 and 5 of the article; changed journal to Software.

@@
 include HtmlFlow using suspendable web templates (HtmlFlow-Susp), Jstachio
-using Virtual Threads (Jstachio-Virtual), and Thymeleaf using the reactive View
+using Virtual Threads with the Iterable interface (Jstachio-Virtual), and Thymeleaf using the reactive View
 Resolver driver (Thymeleaf-Rx). \textit{Sync} and \textit{Virtual} represent
@@
 separately to observe the performance of the non-blocking engines when using
-the Reactive, Virtual Thread, and Suspending approaches. The \textit{Sync} and
+the Suspending, Virtual Thread, and Reactive approaches. The \textit{Sync} and
 \textit{Virtual} are aggregated due to the similar performance of different

### 2025-05-15 docs/article/Sections/state-of-the-art.tex
Chapter 2 - Distinguish template idioms based on DSL classification: external versus internal. Examples of external DSLs. Describe internal DSL and its advantages.

@@
 constructing dynamic HTML pages.
-Web templates (such as JSP, Handlebars, or Thymeleaf), also known as a
-\textit{web pages} or \textit{web views}~\cite{Fowler02,Alur01}, are based on
+Web templates (such as JSP, Handlebars, or Thymeleaf), also known as
+\textit{web views}~\cite{Fowler02,Alur01}, are based on
 HTML documents augmented with template-specific markers (e.g., \texttt{<\%>},
@@
 \subsubsection{Templating idiom}
+
+Web templates are based on a \textit{domain-specific language}
+(DSL)~\cite{landin1966next}, which defines a language tailored to a specific
+\textit{domain}~\cite{evans2004domain}—in this case, HTML for expressing web
+documents.
+The DSL constrains the template's syntax and semantics to match the structure
+and purpose of HTML.
+
+DSLs can be divided in two types: \textit{external} or
+\textit{internal}\cite{dslbook}.
+\textit{External} DSLs are languages created without any affiliation to a
+concrete programming language.
+An example of an \textit{external} DSL is the regular expressions search
+pattern\cite{thompson1968}, since it defines its own syntax without any
+dependency of programming languages. 
+On the other hand an \textit{internal} DSL is defined within a host programming
+language as a library and tends to be limited to the syntax of the host language,
+such as Java.
+JQuery\cite{resig2007pro} is one of the most well-known examples of an internal
+DSL in Javascript, designed to simplify HTML DOM\cite{dom} tree traversal and
+manipulation.
+
+Traditionally, web template technologies use an \textit{external} DSL to define
+control flow constructs and data binding primitives. Early web template engines
+such as JSP, ASP, Velocity, PHP, and others adopted this \textit{external} DSL
+approach as their templating dialect.
+For example, a \texttt{foreach} loop can be expressed in each technology using its 
+own DSL:
+\\\texttt{<\% for(String item : items) \%>} in JSP, \texttt{<\% For Each item 
+In items \%>} in legacy ASP with VBScript, \texttt{\#foreach(\$item in \$items)} in 
+Velocity, or \texttt{<?php foreach (\$items as \$item):?>} in PHP.
+
+On the other hand, \textit{internal} DSLs for HTML allow templates to be defined 
+directly within the \emph{host} language (such as Java, Kotlin, Groovy, Scala, or 
+other general-purpose programming languages), rather than using text-based 
+template files~\cite{carvalho2019hot}. In this case, a web template is not limited 
+to templating constructs but may instead leverage any available primitive of the 
+host language or any external API.
+
+For example, in Java, you may iterate over data using a \texttt{for} statement,
+or using the \texttt{java.util.streams} API, or any other JVM library such as
+Guava, Varv, StreamEx, or others—according to your preferences.
+
+Using an internal DSL can have several benefits over using textual templates:
+\begin{enumerate}
+  \item \emph{Type safety}: Because the templates are defined with the host
+  programming language, the compiler can check the syntax and types of the
+  templates at compile time, which can help catch errors earlier in the
+  development process.
+
+  \item \emph{IDE support}: Many modern IDEs provide code completion, syntax
+  highlighting, and other features, which can make it easier to write and
+  maintain templates.
+
+  \item \emph{Flexibility}: Use all the features of the host programming language
+  to generate HTML, can make it easier to write complex templates and reuse code.
+
+  \item \emph{Integration}: Because the templates are defined in Java code, for
+  example, you can easily integrate them with other Java code in your
+  application, such as controllers, services, repositories and models.
+
+\end{enumerate}
 

### 2025-05-15 docs/article/Sections/state-of-the-art.tex
Chapter 2 - Describe the 3 different idioms in of internal DSL: function sequence, nested and chain. Describe type Safety.

@@
 
+DSLs for HTML provide an API where methods or functions correspond to the names of
+available HTML elements. 
+These methods, also known as \textit{builders}, can be combined in a chain of
+calls to mimic the construction of an HTML document in a fluent manner.
+Martin Fowler\cite{dslbook} identifies three different patterns for combining
+functions to create a DSL: 1) \textit{function sequence}; 2) \textit{nested
+function}, and 3) \textit{method chaining}, which are illustrated in the snippets of
+Figure~\ref{fig:dsl-idioms}.
+\begin{figure}[h]
+\centering
+
+% Subfloat 1
+\begin{minipage}[c]{0.33\linewidth}
+  \centering
+  \begin{lstlisting}[
+    language=java,
+    basicstyle=\scriptsize\ttfamily,
+    numbers=none
+  ]
+html();
+ head();
+  title();text("JT");end();
+ end();
+ body();
+  p();
+   text("Hi JATL");
+  end();
+ end();
+end();
+  \end{lstlisting}
+  \caption*{(a) Function sequence}
+\end{minipage}
+\hfil
+% Subfloat 2
+\begin{minipage}[c]{0.23\linewidth}
+  \centering
+  \begin{lstlisting}[
+    language=java,
+    basicstyle=\scriptsize\ttfamily,
+    numbers=none
+  ]
+html(
+ head(
+  title("ST")
+ ),
+ body(
+  p("Hi ScalaTags")
+ )
+);
+  \end{lstlisting}
+  \caption*{(b) Nested function}
+\end{minipage}
+\hfil
+% Subfloat 3
+\begin{minipage}[c]{0.32\linewidth}
+  \centering
+  \begin{lstlisting}[
+    language=java,
+    basicstyle=\scriptsize\ttfamily,
+    numbers=none
+  ]
+html()
+ .head()
+  .title().text("HF").__()
+ .__()
+ .body()
+  .p()
+   .text("Hi HtmlFlow")
+  .__()
+ .__()
+.__();
+  \end{lstlisting}
+  \caption*{(c) Method chaining}
+\end{minipage}
+
+\caption{Utilizing DSL for HTML libraries with JATL, ScalaTags, and HtmlFlow.}
+\label{fig:dsl-idioms}
+\end{figure}
+
+From the three examples depicted in Figure~\ref{fig:dsl-idioms}, the \textit{nested
+function} idiom used by \textbf{ScalaTags}, as shown in the snippet of
+Figure~\ref{fig:dsl-idioms}.b), is the least verbose, requiring fewer statements than
+the other two cases.
+This approach of combining function calls is also utilized by \textbf{Hiccup}
+Clojure Library and \textbf{j2html} Java library.
+One reason for verbosity in \textit{function sequence} and \textit{method
+chaining} approaches is the requirement of a dedicated HTML builder to emit the
+closing tag, exemplified by \texttt{end()} in \textbf{JATL} (Figure~\ref{fig:dsl-idioms}.a) and
+\texttt{\_\_()} in \textbf{HtmlFlow} (Figure~\ref{fig:dsl-idioms}.c).
+But the \textit{nested function} approach comes with a notable drawback: \textbf{they
+do not support PSSR} because the sequence of nested functions is evaluated
+backward to the order in which they are written. 
+In other words, arguments are evaluated before the functions are invoked. 
+Taking the example in Figure~\ref{fig:dsl-idioms}.b), the \texttt{title()} function is
+first evaluated, and its resulting paragraph becomes the argument for the
+\texttt{head()} call, which, in turn, becomes the argument for \texttt{html()},
+and so on.
+If HTML is emitted as functions are called, it will print tags in reverse order. 
+The aforementioned DSLs must collect resulting nodes into an internal data
+structure, which is later traversed to produce the HTML output. Therefore, they
+cannot progressively emit the output as builders are called.
+
+Two other JVM libraries, \textbf{Groovy MarkupBuilder} and
+\textbf{KotlinX.html}, also adopt a \textit{nested function} approach, but they
+address the backward evaluation issue by implementing \textit{lazy evaluation}
+of arguments~\cite{Landin65}, expressed in lambda expressions (also known as
+function literals).
+Due to the concise form of expressing lambdas with brackets (i.e.,
+\texttt{\{\}}) in both Groovy and Kotlin, translating the template shown in
+Figure~\ref{fig:dsl-idioms}.b) to use Groovy MarkupBuilder or KotlinX.html only
+requires replacing parentheses with brackets for parent elements.
+\textbf{HtmlFlow} also adopts this approach and provides a Kotlin-idiomatic API 
+as an alternative to its original Java API.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%-------------------------------------------------------------------
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsubsection{Supported data model APIs}
@@
 Restricted or unrestricted.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%-------------------------------------------------------------------
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%-------------------------------------------------------------------
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \subsubsection{Type safety and HTML safety}
+
+Another key characteristic of such DSLs is \textbf{HTML safety}, which refers to
+whether they produce only valid HTML conforming to a well-formed document.
+To ensure HTML safety, a DSL API should only allow combining calls to builders
+that result in valid HTML. 
+However, most HTML DSLs using the \textit{function sequence} or \textit{nested
+function} approach cannot ensure HTML safety at compile time.
+
+The \textit{function sequence} approach, as illustrated in the snippet of
+Figure~\ref{fig:dsl-idioms}.a), involves combining function calls as a sequence of
+statements, making it challenging to restrict the order of statements. 
+In the \textit{nested function} approach shown in
+Figure~\ref{fig:dsl-idioms}.b), variable-length arguments are often used to
+allow an undefined number of child elements, which cannot be strongly typed in
+every programming language.
+
+KotlinX.html, which utilizes the \textit{nested function} idiom with lazy
+evaluation, mitigates this issue through function types with a receiver. 
+In this approach, the receiver (i.e., \texttt{this} within the lambda) is
+strongly typed and provides a set of methods corresponding to legal child
+elements.
+This enables KotlinX.html to enforce HTML safety by restricting the available
+methods during compile time.
+
+To achieve this KotlinX.html provides HTML builders using \textit{function
+literals with receiver}~\cite{kotlinlang}.
+In Kotlin, a block of code enclosed in curly braces \texttt{\{...\}} is known as
+a \emph{lambda}, and can be used as an argument to
+a function that expects a \emph{function literal}.
+When we write, for example, \texttt{body \{ div \{ hr() \} \}}, we are
+invoking the \texttt{body} function with a lambda as its argument. This
+lambda, in turn, calls the \texttt{div} function with another lambda as an
+argument that creates a horizontal row (i.e. \texttt{hr}).
+Each call to an HTML builder (e.g., \texttt{body}, \texttt{div}, \texttt{hr})
+creates the child element within the element generated by the outer function
+call.
+
+HtmlFlow provides two APIs: one in idiomatic Kotlin, similar to the \texttt{KotlinX.html} 
+API, and another that employs the \textit{method chaining} idiom, as illustrated 
+in the snippet of Figure~\ref{fig:dsl-idioms}.c).
+In this approach, the receiver object is implicitly passed as an argument to
+each method call, enabling subsequent methods to be invoked on the result of the
+preceding one.
+This facilitates the composition of methods, with each call building upon the
+other.
+Similar to KotlinX.html, HtmlFlow ensures HTML safety, restricting the available
+HTML builders and attributes after the dot (\texttt{.}) operator.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%-------------------------------------------------------------------
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 

### 2025-05-16 docs/article/Sections/state-of-the-art.tex
chapter 2 - include text about Web Servers approaches and Async APIs Support.

@@
-\section{State of the Art}
+\section{Background and Related Work}
 
-In this section, we first present the main properties that characterize each web
-template technology approach, along with the advantages and drawbacks resulting
-from these characteristics.
+In this section, we first examine the different design choices adopted by web 
+servers in their internal architectures, and how these choices impact the 
+behavior of the web template engines.
+Then, in Subsection~2.2, we present the main properties that characterize each 
+web template technology approach, along with the advantages and drawbacks that 
+result from these characteristics.
 
-Next, we delve into the different design choices adopted by web servers in their
-internal architectures, and how these choices impact the behavior of the web
-template engines described in the previous subsection.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%-------------------------------------------------------------------
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\subsection{Related Work}
+\subsection{Web Framework Architectures and Approaches to PSSR}
+
+In traditional thread-per-request architectures, each incoming request is
+handled by a dedicated thread. The web server maintains a thread pool from
+which a thread is assigned to handle each request. As the load increases, the
+number of active threads can grow rapidly, potentially exhausting the pool.
+This can lead to performance and scalability issues, as the system may become
+bogged down by context switching and thread management
+overhead~\cite{kant2000scalable}. Spring MVC is one of the most widely used
+frameworks that follows this model.
+
+On the other hand, in modern \textit{low-thread} architectures, the server uses
+a small number of threads to handle a large number of requests.
+\textit{Low-thread} servers, also known as
+\textit{event-driven}~\cite{event-driven-servers}, offer a significant
+advantage in efficiently managing a high number of concurrent I/O operations
+with minimal resource usage.
+
+The \textit{non-blocking} I/O model employed in low-thread servers is
+well-suited for handling large volumes of data asynchronously~\cite{Meijer12}.
+This combination of low-thread servers and asynchronous data models has
+facilitated the development of highly scalable, responsive, and resilient web
+applications capable of managing substantial data loads~\cite{Jin15}.
+The prominence of this concept increased with the advent of Node.js in
+2009, and subsequently, various technologies adopted this approach
+in the Java ecosystem, including Netty, Akka HTTP, Vert.X, and Spring WebFlux.
+% Among these, Spring WebFlux stands out as the most widely used web framework in
+% Java, according to surveys like JetBrains' State of Java report (2021) and
+% community metrics from platforms like Github and StackOverflow.
+The \textit{non-blocking} I/O model in low-thread servers functions optimally
+only when HTTP handlers avoid blocking.
+Therefore, HTML templates need to be proficient in dealing with the asynchronous
+APIs provided by data models.
+While most legacy web templates struggle with asynchronous models, DSLs for HTML
+face no such limitations, leveraging all constructions available in the host
+programming language.
+However, the unexpected intertwining of asynchronous handlers' completion and
+HTML builders' execution may potentially lead to malformed HTML with an
+unexpected layout, as demonstrated in the subsequent subsection~\ref{sec:async-support}.
+
+An alternative approach involves utilizing user-level threads, while maintaining
+a blocking I/O and a synchronous programming paradigm.
+However, this approach still requires a user-level I/O subsystem capable of
+mitigating system-level blocking, which is crucial for the performance of
+I/O-intensive applications. This technique offers a lightweight solution for
+efficiently managing a larger number of concurrent sessions by minimizing
+per-thread overhead.
+In 2020, Karsten~\cite{karsten2020} demonstrated how this strategy supports a
+synchronous programming style, effectively abstracting away the complexities
+associated with managing continuations in asynchronous programming.
+A similar idea has been followed in Kotlin with
+\textit{coroutines}~\cite{elizarov2021coroutines}, and most recently in the Java
+standard library with \textit{virtual threads} released in Java 21.
+% Both Spring WebFlux and Vert.X web frameworks provide native support for Kotlin
+% \textit{coroutines}, enabling developers to utilize a specific
+% \texttt{Dispatcher} to switch the execution thread context to a suitable
+% subsystem and reduce \textit{system-level} blocking.
+However, no work has yet leveraged this mechanism to enable legacy template
+engines to provide PSSR while preserving non-blocking progress.
+% We will demonstrate this capability by comparing it with our proposal of
+% \textit{suspendable web templates} in a benchmark using Spring WebFlux.
+
+% Determining the appropriate size of \emph{N} for the I/O susbsystem may hinge on
+% factors such as the nature of the workload and the frequency of
+% \textit{Blocking} I/O operations.
+
+\subsubsection{Spring WebFlux}
+
+\subsubsection{Spring MVC}
+
+\subsubsection{Quarkus}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%-------------------------------------------------------------------
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\subsection{Web Templates}
 
@@
 server-side rendering, as our work is centered on that approach.
+
+Before detailing each of the aforementioned characteristics, Table~11 presents 
+a breakdown of state-of-the-art template engines, classified according to the 
+identified properties.
 
@@
 \subsubsection{Asynchronous support}
+\label{sec:async-support}
 
@@
    E.g JStachio, .... apenas suportam interface Iterable. => Compilation Error.
+
 2. Consegue usar qq API e.g Kotlinx.Html and Groovy  => usa mal => produzir HTML out of order
+
 3. Sim suporta => well-formed HTML => Thymleaf e o HtmlFlow
+
+One of the reasons for legacy web templates not supporting asynchronous APIs
+is the absence of a unified standard calling convention for asynchronous calls.
+While there is a single, straightforward way to use a synchronous API with a
+direct style, where the result of a method call corresponds to its returned
+value, there is no equivalent standard in the asynchronous approach. 
+Instead, we may encounter various asynchronous conventions depending on the
+programming language and runtime environment.
+Some of these approaches include \emph{continuation-passing style}
+(CPS)~\cite{scheme}, \textit{promises}~\cite{promise} async/await
+idiom~\cite{async_await}, reactive streams \cite{ReactiveStreams}, Kotlin
+Flow~\cite{kotlinlang}, and others.
+
+Many \textit{general-purpose languages} (GPLs) have embraced the
+\texttt{async}/\texttt{await} feature~\cite{async_await} enabling non-blocking
+routines to mimic the structure of synchronous ones, allowing developers to
+reason about instruction flow sequentially.
+The simplicity and broad adoption of this programming model have led to its
+incorporation into mainstream languages like C\#, JavaScript, Python, Perl,
+Swift, Kotlin, and others, excluding Java.
+However, implementing \texttt{async}/\texttt{a}wait requires compiler support to
+translate \textit{suspension points} (i.e., \texttt{await} statements) into state
+machines. Most template engines operate using an external DSL with their own
+templating dialect (e.g., Thymeleaf, JSP, Jade, Handlebars, and others), which do not
+inherently leverage asynchronous capabilities from their host GPLs.
 
@@
 \subsubsection{Progressive rendering}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%-------------------------------------------------------------------
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{Web Framework Architectures and Approaches to PSSR}
-
-\subsubsection{Spring WebFlux}
-
-\subsubsection{Spring MVC}
-
-\subsubsection{Quarkus}

### 2025-05-20 docs/article/Sections/state-of-the-art.tex
chapter 2 - clearly distinguish between Kotlin coroutines and Java virtual threads.

@@
 associated with managing continuations in asynchronous programming.
-A similar idea has been followed in Kotlin with
-\textit{coroutines}~\cite{elizarov2021coroutines}, and most recently in the Java
-standard library with \textit{virtual threads} released in Java 21.
-% Both Spring WebFlux and Vert.X web frameworks provide native support for Kotlin
-% \textit{coroutines}, enabling developers to utilize a specific
-% \texttt{Dispatcher} to switch the execution thread context to a suitable
-% subsystem and reduce \textit{system-level} blocking.
-However, no work has yet leveraged this mechanism to enable legacy template
-engines to provide PSSR while preserving non-blocking progress.
-% We will demonstrate this capability by comparing it with our proposal of
-% \textit{suspendable web templates} in a benchmark using Spring WebFlux.
 
-% Determining the appropriate size of \emph{N} for the I/O susbsystem may hinge on
-% factors such as the nature of the workload and the frequency of
-% \textit{Blocking} I/O operations.
+Among mainstream technologies, the Kotlin programming language introduces a new 
+abstraction for managing coroutines and provides structured concurrency, which 
+ensures that coroutines are \textit{scoped}, \textit{cancellable}, and 
+\textit{coordinated} within a well-defined 
+lifecycle~\cite{elizarov2021coroutines}.
+Although this model embraces the \texttt{async}/\texttt{await} 
+feature~\cite{async_await}, enabling non-blocking routines to mimic the 
+structure of synchronous ones and allowing developers to reason about control 
+flow sequentially, it still lacks a user-level I/O subsystem and relies on an 
+explicit I/O dispatcher tied to a dedicated thread pool that frees worker 
+threads for handling processing tasks.
+Moreover, web templates using their own templating dialects, such as JSP, 
+Thymeleaf, or Handlebars, are unable to take advantage of such constructs, as 
+they typically support only a limited subset of the host library's API—most 
+commonly just the \texttt{Iterable} interface.
+
+Only Java virtual threads, introduced in
+JEP~444~\footnote{\url{https://openjdk.org/jeps/444}}, adopt a similar approach
+to Karsten's proposal, using user-mode threads to preserve a synchronous
+programming model while still allowing blocking I/O operations without tying up
+platform threads. When a virtual thread performs a blocking I/O operation, the
+JVM intercepts the call and transparently parks the virtual thread, freeing the
+underlying platform thread to perform other tasks. Once the I/O operation
+completes, the virtual thread is rescheduled on a platform thread and resumes
+execution. This mechanism is enabled by the JVM's integration with non-blocking
+system calls at the OS level, allowing developers to write traditional
+synchronous code while benefiting from the scalability of non-blocking I/O.
+
+\subsubsection{Spring MVC}
 
 \subsubsection{Spring WebFlux}
-
-\subsubsection{Spring MVC}
 

### 2025-05-21 docs/article/Sections/state-of-the-art.tex
chapter 2 - Minor fix. Add link to Web tech survey. Simplify async/await description.

@@
 bogged down by context switching and thread management
-overhead~\cite{kant2000scalable}. Spring MVC is one of the most widely used
-frameworks that follows this model.
+overhead~\cite{kant2000scalable}.
+In the Java ecosystem, Spring Boot is one of the most widely used frameworks that
+follow this model, according to several reports such as the Stack Overflow 2024
+Developer
+Survey~\footnote{~\url{https://survey.stackoverflow.co/2024/technology\#1-web-frameworks-and-technologies}}.
 
@@
 feature~\cite{async_await}, enabling non-blocking routines to mimic the
-structure of synchronous ones and allowing developers to reason about control 
-flow sequentially, it still lacks a user-level I/O subsystem and relies on an 
-explicit I/O dispatcher tied to a dedicated thread pool that frees worker 
-threads for handling processing tasks.
+structure of synchronous ones, it still lacks a user-level I/O subsystem and
+relies on an explicit I/O dispatcher tied to a dedicated thread pool that frees
+worker threads for handling processing tasks.
 Moreover, web templates using their own templating dialects, such as JSP, 

### 2025-05-21 docs/article/Sections/benchmark.tex
article: added Abstract, Introduction, State of the Art, Problem Statement sections; improved benchmark and results sections.
Issues in Problem statement section (!)

@@
 template engines and frameworks. It also includes a set of tests to ensure the
-correctness of implementations and to validate the \textit{HTML} output.
+correctness of implementations and to validate the \textit{HTML} output. It
+includes two different data models, defined as \texttt{Presentation} and
+\texttt{Stock}, shown in \autoref{lst:stock-class} and
+\autoref{lst:presentation-class}, respectively. The \texttt{Presentation} class
+represents a presentation with a title, speaker name, and summary, while the
+\texttt{Stock} class represents a stock with a name, URL, symbol, price,
+change, and ratio.
 
-The benchmark includes two different data models, defined as \texttt{Presentation} and 
-\texttt{Stock} shown in \autoref{lst:data-models}.
-
-\lstdefinestyle{listingstyle}{
-  backgroundcolor=\color{lightgray!20},
-  basicstyle=\ttfamily\small,
-  breaklines=true,
-  frame=single,
-  captionpos=b,
-  numbers=left,
-  numberstyle=\tiny,
-  keywordstyle=\color{blue},
-  commentstyle=\color{gray},
-  stringstyle=\color{orange},
-}
+\vspace{0.5cm}
 
 \lstset{style=listingstyle}
-
-\begin{lstlisting}[language=Kotlin, caption={Data Models}, label={lst:data-models}]
-data class Presentation(
-    val id: Long,
-    val title: String, 
-    val speakerName: String,
-    val summary: String
-)
-
+\begin{minipage}{0.5\textwidth}
+  \begin{lstlisting}[
+  language=Kotlin,
+  numbers=none,
+  basicstyle=\scriptsize\ttfamily,
+  caption={\texttt{Stock} class},
+  label={lst:stock-class},
+]
 data class Stock(
@@
 \end{lstlisting}
+\end{minipage}
+\hfill
+\begin{minipage}{0.4\textwidth}
+  \begin{lstlisting}[
+    language=Kotlin,
+    basicstyle=\scriptsize\ttfamily,
+    numbers=none,
+    caption={\texttt{Presentation} class},
+    label={lst:presentation-class},
+]
+data class Presentation(
+    val id: Long,
+    val title: String, 
+    val speakerName: String,
+    val summary: String
+)
+\end{lstlisting}
+\end{minipage}
 
 The application's repository contains a list of 10 instances of the
-\texttt{Presentation} class and 20 instances of the \texttt{Stock} class. Each list is used
-to generate a respective HTML view. Although the instances are kept in memory,
-the repository uses the \texttt{Observable} class from the \textit{RxJava}
-library to interleave list items with a delay of 1 millisecond. This delay
-promotes context switching and frees up the calling thread to handle other
-requests in non-blocking scenarios, mimicking actual I/O operations.
+\texttt{Presentation} class and 20 instances of the \texttt{Stock} class. Each
+list is used to generate a respective HTML view. Although the instances are
+kept in memory, the repository uses the \texttt{Observable} class from the
+\textit{RxJava} library to interleave list items with a delay of 1 millisecond.
+This delay promotes context switching and frees up the calling thread to handle
+other requests in non-blocking scenarios, mimicking actual I/O operations.
 
-By using the \texttt{blockingIterable} method of the \texttt{Observable}
-class, we provide a blocking interface for template engines that do not support
+By using the \texttt{blockingIterable} method of the \texttt{Observable} class,
+we provide a blocking interface for template engines that do not support
 asynchronous data models, while still simulating the asynchronous nature of the
@@
 
-The aforementioned blocking template engines are used in
-the context of Virtual Threads or alternative coroutine dispatchers, allowing
-the handler thread to be released and reused for other requests.
+The aforementioned blocking template engines are used in the context of Virtual
+Threads or alternative coroutine dispatchers, allowing the handler thread to be
+released and reused for other requests.
 
-The Spring WebFlux core implementation uses Project Reactor to support a reactive
-programming model: each method returns a \texttt{Flux<String>} as the response
-body, which acts as a publisher that progressively streams the HTML content to
-the client. It also includes methods using Kotlin coroutines and other
-asynchronous mechanisms supported by template engines, such as the
-\texttt{writeAsync} method from HtmlFlow, which enables non-blocking I/O
-using \textit{continuations}. 
+The Spring WebFlux core implementation uses Project Reactor to support a
+reactive programming model: each method returns a \texttt{Flux<String>} as the
+response body, which acts as a publisher that progressively streams the HTML
+content to the client. The implementation includes three main approaches to
+PSSR:
+\begin{itemize}
+  \item \textbf{Reactive}: The template engine is used in a reactive context,
+        where the HTML content rendered using the reactive programming model.
+        An example of this approach is the Thymeleaf template engine
+        when using the \texttt{ReactiveDataDriverContextVariable} in conjunction with a
+        non-blocking Spring \texttt{ViewResolver}.
+  \item \textbf{Virtual}: The template engine is used in a non-blocking
+        context, where the HTML content is rendered within the context of Virtual
+        Threads. This method is used for the template engines that do not traditionally
+        support non-blocking I/O, be it either because they use external DSLs and in
+        consequence only support the blocking \texttt{Iterable} interface, or
+        because they do not support the asynchronous rendering of HTML content.
+  \item \textbf{Suspendable}: The template engine is used in a suspendable
+        context, where the HTML content is rendered within the context of a suspending function
+        An example of this approach is the HtmlFlow template engine, which
+        supports suspendable templates with the use of the \texttt{Flow} class from the
+        Kotlin standard library.
+\end{itemize}
+% It also includes methods using Kotlin coroutines and other
+% asynchronous mechanisms supported by template engines, such as the
+% \texttt{writeAsync} method from HtmlFlow, which enables non-blocking I/O
+% using \textit{continuations}. 
 The Spring MVC implementation uses handlers based solely on the blocking
-interface of the \texttt{Observable} class. To enable PSSR in this context, we utilize
-the \texttt{StreamingResponseBody} interface, which allows the application to
-write directly to the response \texttt{OutputStream} without blocking the servlet
-container thread. According to the Spring documentation, this class is \textit{a
-controller method return value type for asynchronous request processing where
-the application can write directly to the response OutputStream without holding
-up the Servlet container thread} \cite{Spring-StreamingResponseBody}.
+interface of the \texttt{Observable} class. To enable PSSR in this context, we
+utilize the \texttt{StreamingResponseBody} interface, which allows the
+application to write directly to the response \texttt{OutputStream} without
+blocking the servlet container thread. According to the Spring documentation,
+this class is \textit{a controller method return value type for asynchronous
+  request processing where the application can write directly to the response
+  OutputStream without holding up the Servlet container thread}
+\cite{Spring-StreamingResponseBody}.
 
-In Spring MVC, \texttt{StreamingResponseBody} enables asynchronous writing relative to
-the request-handling thread, but the underlying I/O remains
-blocking—specifically the writes to the \texttt{OutputStream}. When using Virtual
-Threads, the I/O operations are more efficient when compared to platform threads,
-as they are executed in the context of a lightweight thread. Most of the computation is done in a separate
-thread from the one that receives each request; we use a thread pool
-\texttt{TaskExecutor} to process requests, allowing the application to scale and
-handle multiple clients more efficiently as opposed to the default
-\texttt{TaskExecutor} implementation, which tries to create a thread for each request.
+In Spring MVC, \texttt{StreamingResponseBody} enables asynchronous writing
+relative to the request-handling thread, but the underlying I/O remains
+blocking—specifically the writes to the \texttt{OutputStream}. When using
+Virtual Threads, the I/O operations are more efficient when compared to
+platform threads, as they are executed in the context of a lightweight thread.
+Most of the computation is done in a separate thread from the one that receives
+each request; we use a thread pool \texttt{TaskExecutor} to process requests,
+allowing the application to scale and handle multiple clients more efficiently
+as opposed to the default \texttt{TaskExecutor} implementation, which tries to
+create a thread for each request.
 
@@
 This is because the response is only sent once the content written to the
-\texttt{OutputStream} exceeds the output buffer size, which defaults to 8KB\@. As a
-result, the client receives the response only after the entire HTML content is
-rendered, defeating the purpose of PSSR in this context.
+\texttt{OutputStream} exceeds the output buffer size, which defaults to 8KB\@.
+As a result, the client receives the response only after the entire HTML
+content is rendered, defeating the purpose of PSSR in this context.
 
 The Quarkus implementation also uses handlers based on the blocking interface
-of the \texttt{Observable} class. It implements the \texttt{StreamingOutput} interface
-from the JAX-RS specification to enable PSSR, allowing HTML content to be
-streamed to the client. While \texttt{StreamingOutput} also uses blocking I/O, it
-operates on Vert.x worker threads, which prevents blocking of the event loop.
-When Virtual Threads are used, the I/O operations are handled efficiently, as
-they are executed in lightweight threads.
+of the \texttt{Observable} class. It implements the \texttt{StreamingOutput}
+interface from the JAX-RS specification to enable PSSR, allowing HTML content
+to be streamed to the client. While \texttt{StreamingOutput} also uses blocking
+I/O, it operates on Vert.x worker threads, which prevents blocking of the event
+loop. When Virtual Threads are used, the I/O operations are handled
+efficiently, as they are executed in lightweight threads.
 

### 2025-05-21 docs/article/Sections/introduction.tex
article: added Abstract, Introduction, State of the Art, Problem Statement sections; improved benchmark and results sections.
Issues in Problem statement section (!)

@@
 
-% Non-blocking Progressive Server-Side Rendering (PSSR) is a technique that
-% combines some of the advantages of both server-side and client-side rendering
-% to improve web application performance and user experience. By leveraging
-% non-blocking I/O operations, PSSR allows for efficient handling of multiple
-% concurrent requests while progressively streaming HTML content to the client.
-% 
-% This project consists of a benchmark that evaluates the performance of
-% different PSSR strategies on the Java Virtual Machine (JVM). The benchmark
-% focuses on evaluating the performance of non-blocking PSSR implementations
-% using approaches such as Kotlin coroutines, reactive-style programming, and
-% Java virtual threads. Its main objective is to evaluate virtual threads as an
-% alternative that ensures a non-blocking approach to PSSR while offering a more
-% straightforward programming model than traditional asynchronous programming
-% techniques.
-% 
-% This project aims to extend the previous work done in
-% \textit{spring-webflux-comparing-template-engines}
-% \citeauthor{spring-webflux-comparing-template-engines}, which compares the
-% performance of various template engines in a Spring WebFlux application using
-% different approaches to PSSR\@. It adds support for Java virtual threads and
-% evaluates non-blocking PSSR implementations in other frameworks compared to
-% Spring WebFlux.
-% 
-% Currently, this project features a working benchmark in Spring WebFlux, Spring
-% MVC, and Quarkus, comparing the performance of eight different template
-% engines: Rocker, JStachio, Pebble, Freemarker, Trimou, Thymeleaf, HtmlFlow, and
-% KotlinX. It includes Java Microbenchmark Harness (JMH) benchmarks for template
-% engine performance testing, as well as Apache Bench and Apache JMeter scripts
-% for evaluating throughput and scalability. In the future, the project could be
-% extended to include other frameworks such as Vert.x, if time allows. However,
-% the next objective is the completion of the written report and article
-% submission. The written report will include a more detailed analysis of the
-% benchmark results, along with a contextualization of the state of the art in
-% non-blocking PSSR and asynchronous programming.
-% 
-% This report will focus on detailing relevant aspects of the benchmark
-% implementation and methodology, and will also provide a brief overview of the
-% results obtained from the JMeter and Apache Bench tests, which evaluate the
-% throughput and scalability of the different implementations.
+% Falta adicionar referências
+
+Modern web applications rely on different rendering strategies to optimize
+performance, user experience, and \textit{scalability}. The two most dominant
+approaches are \textit{Server-Side Rendering (SSR)} and \textit{Client-Side
+    Rendering (CSR)}.
+
+SSR generates HTML content on the server before sending it to the client,
+resulting in faster initial page loads and better Search Engine Optimization
+(SEO). However, SSR can increase server load and reduce
+\textit{throughput}\footnote{The number of requests the server can handle per
+    second (RPS)} since each request requires additional processing before
+responding.
+
+In contrast, CSR shifts the rendering workload to the browser. The server
+initially sends a minimal HTML document with JavaScript, which dynamically
+loads the page content. While CSR reduces the server’s burden, it can lead to
+slower initial load times, as users must wait for JavaScript execution before
+meaningful content appears.
+
+\textit{Progressive Server-Side Rendering (PSSR)} combines benefits from both SSR and CSR
+by streaming HTML content progressively. This technique enhances user-perceived performance by
+allowing immediate rendering while data loads asynchronously. In this respect,
+PSSR is similar to CSR in that the server initially sends a minimal HTML
+document to the client and subsequently streams additional HTML fragments as data becomes available.
+However, unlike CSR, PSSR retains all rendering responsibilities on the server side,
+thereby reducing the load on the client. As a result, the client does
+not need to execute JavaScript or make additional requests to retrieve page content.
+To enable non-blocking PSSR, template engines require support for asynchronous
+data models and non-blocking I/O operations. While some template engines, such
+as HtmlFlow, and Thymeleaf, have support for these features, traditional
+template engines that use \textit{external} domain-specific languages (DSLs)
+often rely on blocking interfaces, such as \texttt{Iterable}, to define the
+data model. This reliance on blocking interfaces leads to the system thread
+being blocked until the entire HTML content is rendered, which can
+significantly impact performance and scalability.
+
+Efficient thread utilization is crucial for low-thread servers, necessitating
+asynchronous, non-blocking request handling. In these architectures,
+asynchronous I/O operations (e.g., database queries, and API calls) allow
+servers to manage multiple concurrent requests efficiently by avoiding blocking
+threads while waiting for I/O operations to complete. In these architectures,
+where idle time during I/O operations is reduced, non-blocking PSSR enhances a
+server's capacity to handle high request volumes with limited threads and
+immediately send partial responses to clients.
+
+With the introduction of \textit{Virtual Threads} in Java 21, it is now possible to execute blocking
+I/O operations in a non-blocking manner, allowing for the use of these 
+traditional template engines in non-blocking contexts without the need for
+complex asynchronous programming models. This advancement enables the 
+implementation of PSSR using familiar HTML syntax, simplifying template
+development and adoption.
+
+This work will focus on exploring existing approaches to non-blocking PSSR,
+which include the use of \textit{reactive programming} and \textit{coroutines}
+to implement non-blocking I/O operations. As an alternative to these
+approaches, we will investigate the viability of \textit{Virtual Threads} in Java 21 
+for implementing non-blocking PSSR\@. Section 2 provides an overview
+of the state-of-the-art in non-blocking PSSR and HTML template engines. Section
+3 provides a more detailed description of the issues associated with
+traditional template engines for non-blocking PSSR, and how this work aims to
+address them. Section 4 describes the benchmark methodology and implementation.
+Finally, Sections 5 and 6 present the results and conclusions of this work,
+respectively.

### 2025-05-21 docs/article/Sections/problem-statement.tex
article: added Abstract, Introduction, State of the Art, Problem Statement sections; improved benchmark and results sections.
Issues in Problem statement section (!)

\section{Problem Statement}

In this section we discuss the problems associated with progressive server-side
rendering (PSSR) in modern web applications, focusing on the challenges and
limitations of existing approaches. We then present the objective of our work,
which aims to increase the available options for PSSR in web applications,
particularly in the context of JVM based frameworks.

\textit As we outlined in the previous section, \textit{external} DSLs for HTML such as Thymeleaf, Jstachio, Handlebars, and others define HTML
templates within HTML documents with markers specific to the engine. Template engines that use \textit{internal} DSLs, such as KotlinX or HtmlFlow,
define the HTML templates in the programming language itself, leveraging the language's
syntax and features to create the templates.

\textit{Internal} DSLs can more easily support asynchronous data models since
they can leverage the language's existing features and libraries, thus enabling
the progressive streaming of HTML content to the client, with \textit{external}
DSLs supporting simpler blocking data models such as \texttt{Iterable} or
\texttt{List}. While on the previous section we outlined several advantages of
using \textit{internal} DSLs over textual templates for HTML, there are also
advantages to using \textit{external} DSLs:

\begin{enumerate}
    \item \emph{Separation of Concerns}: \textit{External} DSLs allow for a clear separation between
          the HTML structure and the application logic, making it easier to maintain and update
          templates without modifying the underlying code. This ''
          separation'' can be beneficial, as it allows front-end developers to work on the HTML
          templates without needing to understand the underlying application logic or programming language.
    \item \emph{Cross-Language Compatibility}: \textit{External} DSLs can be used across different programming languages
          and frameworks, making them more versatile and easier to integrate into existing projects.
          This is particularly useful in environments where multiple languages are used.
    \item \emph{Familiarity}: Many developers are already familiar with HTML and its syntax, making
          it easier to work with \textit{external} DSLs that use HTML-like syntax.
\end{enumerate}

Due to these advantages, many developers prefer to use \textit{external} DSLs
for HTML templates, even if it means sacrificing some of the advantages of
using \textit{internal} DSLs. However, this preference can lead to challenges
when it comes to progressive server-side rendering (PSSR) and non-blocking I/O,
as many \textit{external} DSLs do not support asynchronous data models.

\autoref{lst:presentation-jstachio} shows the HTML template for a \texttt{Presentation} class using the JStachio template engine, 
while \autoref{lst:presentation-htmlflow} presents the equivalent template using HtmlFlow.

\lstset{style=listingstyle}
\begin{lstlisting}[
  language=Kotlin,
  numbers=none,
  basicstyle=\scriptsize\ttfamily,
  caption={Presentation HTML template using \textit{JStachio}}
  label={lst:presentation-jstachio}
]
<!DOCTYPE html>
<html lang="en-US">
<head>
    <meta charset="utf-8">
    <meta name="viewport" content="width=device-width, initial-scale=1.0">
    <meta http-equiv="X-UA-compatible" content="IE=Edge">
    <title>
        JFall 2013 Presentations - JStachio
    </title>
    <link rel="Stylesheet" href="/webjars/bootstrap/5.3.0/css/bootstrap.min.css">
</head>
<body>
<div class="container">
    <div class="pb-2 mt-4 mb-3 border-bottom">
        <h1>
            JFall 2013 Presentations - JStachio
        </h1>
    </div>
    {{#presentationItems}}
    <div class="card mb-3 shadow-sm rounded">
        <div class="card-header">
            <h5 class="card-title">
                {{title}} - {{speakerName}}
            </h5>
        </div>
        <div class="card-body">
            {{summary}}
        </div>
    </div>
    {{/presentationItems}}
</div>
</body>
</html>
\end{lstlisting}
\vspace{0.5cm}
\begin{lstlisting}[
  language=Kotlin,
  numbers=none,
  basicstyle=\scriptsize\ttfamily,
  caption={Presentation HTML template using \textit{HtmlFlow}}
  label={lst:presentation-htmlflow}
]
 val htmlFlowTemplate: HtmlViewAsync<Observable<Presentation>> =
    HtmlFlow.viewAsync<Observable<Presentation>?> { view ->
        view.html {
            attrLang("en-us")
            head {
                meta { attrCharset("UTF-8") }
                meta { attrName("viewport").attrContent("width=device-width, initial-scale=1.0") }
                meta { addAttr("http-equiv", "X-UA-Compatible").attrContent("IE=Edge") }
                title { text("JFall 2013 Presentations - HtmlFlow") }
                link { attrRel(EnumRelType.STYLESHEET).attrHref("/webjars/bootstrap/5.3.0/css/bootstrap.min.css") }
            }
            body {
                div {
                    attrClass("container")
                    div {
                        attrClass("pb-2 mt-4 mb-3 border-bottom")
                        h1 { text("JFall 2013 Presentations - HtmlFlow") }
                    }
                    await { div, model: Observable<Presentation>, onCompletion ->
                        model
                            .doOnNext { 
                                presentationFragmentAsync
                                    .renderAsync(it)
                                    .thenApply { frag -> div.raw(frag) }
                            }
                            .doOnComplete { onCompletion.finish() }
                            .subscribe()
                    }
                }
            }
        }
    }.threadSafe()
\end{lstlisting}
\lstset{style=listingstyle}

HtmlFlow is an example of an \textit{internal} DSL that uses calls to HTML
builders as a way to emit HTML\@. It allows for the use of the
\texttt{Observable} class as a data model, which in turn allows for the definition of a
\texttt{Consumer<Presentation>} through the \texttt{doOnNext} method. Every
time a new value is emitted to the \texttt{Observable}, the \texttt{doOnNext}
method is called, emitting the HTML fragment for the \texttt{Presentation}
object to the client.

The Jstachio template engine, on the other hand, uses an \textit{external} DSL,
where the HTML template is defined in a separate file. The template engine uses
the \texttt{presentationItems} variable to define the data model, which is a
list of \texttt{Presentation} objects. This allows the template engine to
iterate over the list of \texttt{Presentation} objects and render the HTML for
each one.

As demonstrated, the HtmlFlow template enables PSSR by directly supporting the
use of the \texttt{Observable} class as a data model. However, not all
programmers are familiar with asynchronous programming models, which can make
implementing this solution challenging. In contrast, Jstachio defines templates
using standard HTML syntax with specific tags for the template engine,
resulting in a simpler and more accessible template definition. However,
because Jstachio templates rely on an \texttt{Iterable} interface,
they block the calling thread until the entire HTML content is rendered and
therefore do not support non-blocking PSSR\@.

% Mencionar que certas DSLs internas não suportam PSSR por 
% causa de HTML mal formado? (KotlinX)

In previous work \cite{PSSR-WISE2024}, Carvalho developed a benchmark
using Spring Webflux to evaluate the performance of HtmlFlow with suspendable
web templates. In this benchmark, HtmlFlow's performance was compared to that
of Thymeleaf using a reactive approach, as well as other template engines such
as the one shown in \autoref{lst:presentation-JStachio}. The results
demonstrated that HtmlFlow scaled efficiently up to 128 concurrent users,
achieving a throughput of up to 4,000 requests per second. In contrast, the
blocking template engines ceased to scale beyond 4 concurrent users, reaching a
maximum throughput of approximately 400 requests per second. These findings
indicate that engines which block the calling thread scale poorly and are not
suitable for PSSR\@.

% Expandir sobre virtual threads aqui?
With the advent of Java 21 and the introduction of \textit{Virtual Threads}, it
is now possible to execute code that would otherwise block the calling thread
in a scalable, non-blocking manner. This advancement enables the use of these
\textit{external} DSLs in non-blocking contexts. The objective of this work is
to evaluate the performance of these template engines when leveraging Virtual
Threads, and to determine whether they can be made viable for PSSR\@. We will
also compare the performance of this approach with existing solutions based on
reactive programming and Kotlin coroutines.

If the performance of these engines proves comparable to that of current
approaches, it would enable PSSR to be implemented using more familiar HTML
syntax, simplifying template development and adoption. Furthermore, template
engine developers seeking to support PSSR would no longer need to implement
complex asynchronous programming models to achieve non-blocking I/O; instead,
they could rely on the traditional approaches used for Server Side Rendering
(SSR).

### 2025-05-21 docs/article/Sections/results.tex
article: added Abstract, Introduction, State of the Art, Problem Statement sections; improved benchmark and results sections.
Issues in Problem statement section (!)

@@
 
+Both the Quarkus and Spring MVC implementations were configured with an output
+buffer size of 8KB, despite the Quarkus implementation not enabling PSSR at
+this size. This configuration was chosen to ensure consistency across Spring
+MVC and Quarkus, as the Spring MVC implementation does not support PSSR for the
+tested templates.
+
 Since the obtained results for JMeter and Apache Bench show no significant
@@
 include HtmlFlow using suspendable web templates (HtmlFlow-Susp), Jstachio
-using Virtual Threads with the Iterable interface (Jstachio-Virtual), and Thymeleaf using the reactive View
-Resolver driver (Thymeleaf-Rx). \textit{Sync} and \textit{Virtual} represent
-the average throughput of the blocking approaches (i.e., KotlinX, Rocker,
-Jstachio, Pebble, Freemarker, Trimou, HtmlFlow, and Thymeleaf) when run in the
-context of a separate coroutine dispatcher or Virtual Threads, respectively.
+using Virtual Threads with the \texttt{Iterable} interface (Jstachio-Virtual),
+and Thymeleaf using the reactive View Resolver driver (Thymeleaf-Rx).
+\textit{Sync} and \textit{Virtual} represent the average throughput of the
+blocking approaches (i.e., KotlinX, Rocker, Jstachio, Pebble, Freemarker,
+Trimou, HtmlFlow, and Thymeleaf) when run in the context of a separate
+coroutine dispatcher or Virtual Threads, respectively.
 
@@
 \begin{figure}[h]
-     \centering
+     \isPreprints{\centering}{}
      \includegraphics[width=0.8\textwidth]{./Graphs/presentations-webflux-jmeter.png}
@@
 \begin{figure}[h]
-     \centering
+     \isPreprints{\centering}{}
      \includegraphics[width=0.8\textwidth]{./Graphs/presentations-springmvc-jmeter.png}
@@
 \begin{figure}[h]
-     \centering
+     \isPreprints{\centering}{}
      \includegraphics[width=0.8\textwidth]{./Graphs/presentations-quarkus-jmeter.png}
@@
 \begin{figure}[h]
-     \centering
+     \isPreprints{\centering}{}
      \includegraphics[width=0.8\textwidth]{./Graphs/stocks-webflux-jmeter.png}
@@
 percent across all engines, with HtmlFlow achieving 6,000 requests per second.
-The Thymeleaf implementation using the reactive View Resolver driver reaches
-5,000 requests per second, suggesting that the performance drop compared to the
-Presentation benchmark is not substantial.
+Since the \texttt{Stock} class\ contains more data properties than the
+\texttt{Presentation} class, the data binding to the template engine adds
+overhead, which explains the reduced performance. The Thymeleaf implementation
+using the reactive View Resolver driver reaches 5,000 requests per second,
+suggesting that the performance drop compared to the Presentation benchmark is
+not substantial.
 
 \begin{figure}[h]
-     \centering
+     \isPreprints{\centering}{}
      \includegraphics[width=0.8\textwidth]{./Graphs/stocks-springmvc-jmeter.png}
@@
 \begin{figure}[h]
-     \centering
+     \isPreprints{\centering}{}
      \includegraphics[width=0.8\textwidth]{./Graphs/stocks-quarkus-jmeter.png}
@@
 to scale as effectively as those using reactive programming or Kotlin
-coroutines, with the advantage of being easier to implement and understand for 
-developers.
+coroutines, allowing \textit{external} DSLs to be used in a non-blocking context, 
+effectively enabling PSSR\@.

### 2025-05-21 docs/project-proposal/ppG48.tex
article: added Abstract, Introduction, State of the Art, Problem Statement sections; improved benchmark and results sections.
Issues in Problem statement section (!)

@@
 
-Modern web applications rely on different rendering strategies to optimize performance, user experience, and \textit{scalability}.
-The two most dominant approaches are \textit{Server-Side Rendering (SSR)} and \textit{Client-Side Rendering (CSR)}.
+Modern web applications rely on different rendering strategies to optimize
+performance, user experience, and \textit{scalability}. The two most dominant
+approaches are \textit{Server-Side Rendering (SSR)} and \textit{Client-Side
+  Rendering (CSR)}.
 
-SSR generates HTML content on the server before sending it to the client, resulting in faster initial page loads and better Search Engine Optimization (SEO) \cite{prismic}.
-However, SSR can increase server load and reduce \textit{throughput}\footnote{The number of requests the server can handle per second (RPS)} since each request requires additional processing before responding.
+SSR generates HTML content on the server before sending it to the client,
+resulting in faster initial page loads and better Search Engine Optimization
+(SEO). However, SSR can increase server load and reduce
+\textit{throughput}\footnote{The number of requests the server can handle per
+  second (RPS)} since each request requires additional processing before
+responding.
 
-In contrast, CSR shifts the rendering workload to the browser.
-The server initially sends a minimal HTML document with JavaScript, which dynamically loads the page content.
-While CSR reduces the server’s burden, it can lead to slower initial load times, as users must wait for JavaScript execution before meaningful content appears \cite{nextjs}.
+In contrast, CSR shifts the rendering workload to the browser. The server
+initially sends a minimal HTML document with JavaScript, which dynamically
+loads the page content. While CSR reduces the server’s burden, it can lead to
+slower initial load times, as users must wait for JavaScript execution before
+meaningful content appears.
 
@@
 
-Efficient thread utilization is crucial for low-thread servers, necessitating asynchronous, non-blocking request handling \cite{webist23}.
-Non-blocking PSSR is particularly advantageous in event-driven architectures, where asynchronous I/O operations (e.g., database queries, and API calls) allow servers to manage multiple concurrent requests efficiently.
-By reducing idle time during I/O operations, non-blocking PSSR enhances a server's capacity to handle high request volumes with limited threads and immediately send partial responses to clients.
+Efficient thread utilization is crucial for low-thread servers, necessitating
+asynchronous, non-blocking request handling \cite{webist23}. Non-blocking PSSR
+is particularly advantageous in event-driven architectures, where asynchronous
+I/O operations (e.g., database queries, and API calls) allow servers to manage
+multiple concurrent requests efficiently. By reducing idle time during I/O
+operations, non-blocking PSSR enhances a server's capacity to handle high
+request volumes with limited threads and immediately send partial responses to
+clients.
 
-This work will focus on exploring Kotlin Coroutines and Structured Concurrency, as well as Java Virtual Threads as an alternative to asynchronous programming techniques, to evaluate their performance in non-blocking PSSR \cite{kotlin_coroutines, virtual_threads}.
+This work will focus on exploring Kotlin Coroutines and Structured Concurrency,
+as well as Java Virtual Threads as an alternative to asynchronous programming
+techniques, to evaluate their performance in non-blocking PSSR
+\cite{kotlin_coroutines, virtual_threads}.
 
-The benchmark will be built on Spring WebFlux and analyze methods such as those presented in \textit{Progressive Server-Side Rendering with Suspendable Web Templates} \cite{pssr-susp}.
-Additionally, it will build upon and extend the \textit{Spring WebFlux Comparing Template Engines} repository \cite{spring-comparing-template-engines} with the addition of approaches using Java Virtual Threads.
+The benchmark will be built on Spring WebFlux and analyze methods such as those
+presented in \textit{Progressive Server-Side Rendering with Suspendable Web
+  Templates} \cite{pssr-susp}. Additionally, it will build upon and extend the
+\textit{Spring WebFlux Comparing Template Engines} repository
+\cite{spring-comparing-template-engines} with the addition of approaches using
+Java Virtual Threads.
 
-To achieve this, benchmarking tools such as JMH, Apache Bench, and Apache JMeter will be utilized to assess performance metrics, including scalability, throughput, and latency \cite{benchmarking_web, jmh_benchmark}.
-The accuracy and setup complexity of these tools will also be compared.
+To achieve this, benchmarking tools such as JMH, Apache Bench, and Apache
+JMeter will be utilized to assess performance metrics, including scalability,
+throughput, and latency \cite{benchmarking_web, jmh_benchmark}. The accuracy
+and setup complexity of these tools will also be compared.
 
@@
 
-This project consists of three main components: the development of the benchmark, the creation of a report analyzing the state-of-the-art and benchmark results, and the submission of an article to a technology journal.
+This project consists of three main components: the development of the
+benchmark, the creation of a report analyzing the state-of-the-art and
+benchmark results, and the submission of an article to a technology journal.
 
@@
         \begin{itemize}
-    \item Evaluate PSSR strategies with the use of currently existing libraries and frameworks (e.g., HtmlFlow, Thymeleaf, etc.).
-    \item Compare the performance of these implementations in terms of throughput, latency, and scalability.
-    \item Assess the feasibility of using Apache JMeter instead of Apache Bench for performance testing.
+          \item Evaluate PSSR strategies with the use of currently existing libraries and
+                frameworks (e.g., HtmlFlow, Thymeleaf, etc.).
+          \item Compare the performance of these implementations in terms of throughput,
+                latency, and scalability.
+          \item Assess the feasibility of using Apache JMeter instead of Apache Bench for
+                performance testing.
           \item Compare the ease of setup and accuracy of results between the two tools.
-    \item Evaluate the performance and viability of using Java Virtual Threads for non-blocking PSSR.
+          \item Evaluate the performance and viability of using Java Virtual Threads for
+                non-blocking PSSR.
         \end{itemize}
@@
         \begin{itemize}
-    \item Provide an introduction and contextualization of progressive rendering and asynchronous request handling.
-    \item Analyze existing literature and state-of-the-art approaches in non-blocking web rendering.
-    \item Present an analysis of benchmark results, highlighting trade-offs between different PSSR implementations.
+          \item Provide an introduction and contextualization of progressive rendering and
+                asynchronous request handling.
+          \item Analyze existing literature and state-of-the-art approaches in non-blocking web
+                rendering.
+          \item Present an analysis of benchmark results, highlighting trade-offs between
+                different PSSR implementations.
         \end{itemize}
@@
           \item Written article for submission in a technology journal.
-    \item The article will present the benchmark results and analysis, providing insights into the performance of different PSSR strategies in the JVM.
+          \item The article will present the benchmark results and analysis, providing insights
+                into the performance of different PSSR strategies in the JVM.
         \end{itemize}

### 2025-05-21 docs/article/Sections/state-of-the-art.tex
article: added Abstract, Introduction, State of the Art, Problem Statement sections; improved benchmark and results sections.
Issues in Problem statement section (!)
chapter 2 - resuming table with a characteristics comparison among Java template engines

@@
 constructing dynamic HTML pages.
-Web templates (such as JSP, Handlebars, or Thymeleaf), also known as
-\textit{web views}~\cite{Fowler02,Alur01}, are based on
-HTML documents augmented with template-specific markers (e.g., \texttt{<\%>},
-\texttt{\{\{\}\}}, or \texttt{\$\{\}}), which represent \textit{dynamic}
-information to be replaced at runtime with the results of corresponding
-computations, producing the final HTML page.
+Web templates or \textit{web views}~\cite{Fowler02,Alur01} (e.g. JSP, Handlebars,
+or Thymeleaf), are based on HTML documents augmented with template-specific
+markers (e.g., \texttt{<\%>}, \texttt{\{\{\}\}}, or \texttt{\$\{\}}), which
+represent \textit{dynamic} information to be replaced at runtime with the
+results of corresponding computations, producing the final HTML page.
 The process of parsing and replacing these markers---i.e.,
@@
 engine}~\cite{Parr04}.
-
 One key characteristic of \textit{web templates} is their ability to receive a 
@@
 template placeholders at runtime.
-
 Web templates can be distinguished by several properties, namely:
 \begin{enumerate}
-    \item Templating idiom
+    \item \textit{Domain-specific language} idiom
     \item Supported data model APIs
@@
 
-Although some of the aforementioned characteristics apply to both server-side 
-and client-side approaches, we focus solely on web template technologies for 
-server-side rendering, as our work is centered on that approach.
+Although some of the aforementioned characteristics apply to both \textit{server-side}
+and \textit{client-side} approaches, we focus solely on web template technologies for
+\textit{server-side rendering}, as our work is centered on that approach.
+Before getting deep on each of the aforementioned characteristics,
+Table~\ref{table:cmplibs} presents a breakdown of mainstream template engines,
+classified according to the identified properties.
 
-Before detailing each of the aforementioned characteristics, Table~11 presents 
-a breakdown of state-of-the-art template engines, classified according to the 
-identified properties.
+\begin{table}[h]
+  \small
+	\tabcolsep=0.1cm
+	\def\arraystretch{1.2}
+	\begin{tabular}{|c|c|c|c|c|c|c|}
+		\hline                                                                  
+		\textbf{Library}
+		&\textbf{DSL idiom}
+		&\textbf{Data Model APIs}
+    &\shortstack{\textbf{Asynchronous}\\\textbf{Support}}
+    &\shortstack{\textbf{Type}\\\textbf{Safety}}
+    &\shortstack{\textbf{HTML}\\\textbf{Safety}}
+		&\shortstack{\textbf{Progressive}\\\textbf{Rendering}}
+		\\
+		\hline
+		\textbf{Freemarker}
+		&External DSL
+		&\texttt{Iterable}
+		&\large{$\textcolor{red}{\times}$}
+    &\large{$\textcolor{red}{\times}$}
+    &\large{$\textcolor{red}{\times}$}
+		&\large{$\textcolor{PineGreen}{\checkmark}$}
+		\\
+    \hline
+		\textbf{JSP}
+		&External DSL
+		&\texttt{Iterable}
+		&\large{$\textcolor{red}{\times}$}
+    &\large{$\textcolor{red}{\times}$}
+    &\large{$\textcolor{red}{\times}$}
+		&\large{$\textcolor{red}{\times}$}
+    \\
+    \hline
+		\textbf{JStachio}
+		&External DSL
+		&\texttt{Iterable}
+		&\large{$\textcolor{red}{\times}$}
+    &\large{$\textcolor{red}{\times}$}
+    &\large{$\textcolor{red}{\times}$}
+		&\large{$\textcolor{PineGreen}{\checkmark}$}
+		\\\hline
+		\textbf{Pebble}
+		&External DSL
+		&\texttt{Iterable}
+		&\large{$\textcolor{red}{\times}$}
+    &\large{$\textcolor{red}{\times}$}
+    &\large{$\textcolor{red}{\times}$}
+		&\large{$\textcolor{PineGreen}{\checkmark}$}
+		\\
+    \hline
+		\textbf{Rocker}
+		&External DSL
+		&\texttt{Iterable}
+		&\large{$\textcolor{red}{\times}$}
+    &\large{$\textcolor{red}{\times}$}
+    &\large{$\textcolor{red}{\times}$}
+		&\large{$\textcolor{PineGreen}{\checkmark}$}
+		\\
+    \hline
+		\textbf{Thymeleaf}
+		&External DSL
+		&\shortstack{\texttt{Iterable}\\\texttt{Publisher}}
+		&\texttt{Publisher}$^{(1)}$
+    &\large{$\textcolor{red}{\times}$}
+    &\large{$\textcolor{red}{\times}$}
+		&\large{$\textcolor{PineGreen}{\checkmark}$}
+		\\
+    \hline
+		\textbf{Trimou}
+		&External DSL
+		&\texttt{Iterable}
+		&\large{$\textcolor{red}{\times}$}
+    &\large{$\textcolor{red}{\times}$}
+    &\large{$\textcolor{red}{\times}$}
+		&\large{$\textcolor{PineGreen}{\checkmark}$}
+		\\
+    \hline
+		\textbf{Velocity}
+		&External DSL
+		&\texttt{Iterable}
+		&\large{$\textcolor{red}{\times}$}
+    &\large{$\textcolor{red}{\times}$}
+    &\large{$\textcolor{red}{\times}$}
+		&\large{$\textcolor{PineGreen}{\checkmark}$}
+		\\
+    \hline
+    \shortstack{\textbf{Groovy }\\\textbf{MarkupBuilder}}
+		&Nested Lazy
+		&All
+    &\large{$\textcolor{red}{\times}$}
+    &\large{$\textcolor{PineGreen}{\checkmark}$}
+    &\large{$\textcolor{red}{\times}$}
+		&\large{$\textcolor{PineGreen}{\checkmark}$}
+		\\
+    \hline
+    \textbf{HtmlFlow}
+		&Chaining
+		&All
+    &\large{$\textcolor{PineGreen}{\checkmark}$}
+    &\large{$\textcolor{PineGreen}{\checkmark}$}
+    &\large{$\textcolor{PineGreen}{\checkmark}$}
+		&\large{$\textcolor{PineGreen}{\checkmark}$}
+		\\
+    \hline
+		\textbf{j2html}
+		&Nested Eager
+		&All
+    &\large{$\textcolor{red}{\times}$}
+    &\large{$\textcolor{PineGreen}{\checkmark}$}
+    &\large{$\textcolor{red}{\times}$}
+    &\large{$\textcolor{red}{\times}$}
+		\\
+    \hline
+		\textbf{KotlinX}
+		&Nested Lazy
+		&All
+    &\large{$\textcolor{red}{\times}$}
+    &\large{$\textcolor{PineGreen}{\checkmark}$}
+    &\textcolor{PineGreen}{\checkmark}$^{(2)}$
+		&\large{$\textcolor{PineGreen}{\checkmark}$}
+		\\
+		\hline                                                      
+	\end{tabular}
+	\caption{
+    Comparison of web template technologies in the Java ecosystem.
+    \\$^{(1)}$ Limited to a single model per web template and compatible with
+    the reactive stream \texttt{Publisher}.
+    \\$^{(2)}$ Non-safety for HTML attributes.
+  }
+	\label{table:cmplibs}
+\end{table}
+
+The first half of Table~\ref{table:cmplibs} lists template engines that use their
+own templating dialects, referred to as \textit{external DSLs}. The second half
+lists Java libraries that provide an \textit{internal DSL}, typically using a
+\textit{nested} or \textit{chaining} style to build HTML.
+Note that, by leveraging the host language (e.g., Groovy, Java, or Kotlin) as
+the templating idiom, the latter impose no restrictions on the data model and
+fully support all styles of data access APIs.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%-------------------------------------------------------------------
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 

### 2025-05-21 docs/article/Sections/state-of-the-art.tex
chapter 2 - Add Clojure Hiccup and ScalaTags to the resuming table.

--- 2025-05-21 docs/article/Sections/state-of-the-art.tex
@@
 
-Although some of the aforementioned characteristics apply to both server-side 
-and client-side approaches, we focus solely on web template technologies for 
-server-side rendering, as our work is centered on that approach.
+Although some of the aforementioned characteristics apply to both \textit{server-side}
+and \textit{client-side} approaches, we focus solely on web template technologies for
+\textit{server-side rendering}, as our work is centered on that approach.
+Before getting deep on each of the aforementioned characteristics,
+Table~\ref{table:cmplibs} presents a breakdown of mainstream template engines,
+classified according to the identified properties.
 
-Before detailing each of the aforementioned characteristics, Table~11 presents 
-a breakdown of state-of-the-art template engines, classified according to the 
-identified properties.
+\begin{table}[h]
+  \small
+	\tabcolsep=0.1cm
+	\def\arraystretch{1.2}
+	\begin{tabular}{|c|c|c|c|c|c|c|}
+		\hline                                                                  
+		\textbf{Library}
+		&\textbf{DSL idiom}
+		&\textbf{Data Model APIs}
+    &\shortstack{\textbf{Asynchronous}\\\textbf{Support}}
+    &\shortstack{\textbf{Type}\\\textbf{Safety}}
+    &\shortstack{\textbf{HTML}\\\textbf{Safety}}
+		&\shortstack{\textbf{Progressive}\\\textbf{Rendering}}
+		\\
+		\hline
+		\textbf{Freemarker}
+		&External DSL
+		&\texttt{Iterable}
+		&\large{$\textcolor{red}{\times}$}
+    &\large{$\textcolor{red}{\times}$}
+    &\large{$\textcolor{red}{\times}$}
+		&\large{$\textcolor{PineGreen}{\checkmark}$}
+		\\
+    \hline
+		\textbf{JSP}
+		&External DSL
+		&\texttt{Iterable}
+		&\large{$\textcolor{red}{\times}$}
+    &\large{$\textcolor{red}{\times}$}
+    &\large{$\textcolor{red}{\times}$}
+		&\large{$\textcolor{red}{\times}$}
+    \\
+    \hline
+		\textbf{JStachio}
+		&External DSL
+		&\texttt{Iterable}
+		&\large{$\textcolor{red}{\times}$}
+    &\large{$\textcolor{red}{\times}$}
+    &\large{$\textcolor{red}{\times}$}
+		&\large{$\textcolor{PineGreen}{\checkmark}$}
+		\\\hline
+		\textbf{Pebble}
+		&External DSL
+		&\texttt{Iterable}
+		&\large{$\textcolor{red}{\times}$}
+    &\large{$\textcolor{red}{\times}$}
+    &\large{$\textcolor{red}{\times}$}
+		&\large{$\textcolor{PineGreen}{\checkmark}$}
+		\\
+    \hline
+		\textbf{Rocker}
+		&External DSL
+		&\texttt{Iterable}
+		&\large{$\textcolor{red}{\times}$}
+    &\large{$\textcolor{red}{\times}$}
+    &\large{$\textcolor{red}{\times}$}
+		&\large{$\textcolor{PineGreen}{\checkmark}$}
+		\\
+    \hline
+		\textbf{Thymeleaf}
+		&External DSL
+		&\shortstack{\texttt{Iterable}\\\texttt{Publisher}}
+		&\texttt{Publisher}$^{(1)}$
+    &\large{$\textcolor{red}{\times}$}
+    &\large{$\textcolor{red}{\times}$}
+		&\large{$\textcolor{PineGreen}{\checkmark}$}
+		\\
+    \hline
+		\textbf{Trimou}
+		&External DSL
+		&\texttt{Iterable}
+		&\large{$\textcolor{red}{\times}$}
+    &\large{$\textcolor{red}{\times}$}
+    &\large{$\textcolor{red}{\times}$}
+		&\large{$\textcolor{PineGreen}{\checkmark}$}
+		\\
+    \hline
+		\textbf{Velocity}
+		&External DSL
+		&\texttt{Iterable}
+		&\large{$\textcolor{red}{\times}$}
+    &\large{$\textcolor{red}{\times}$}
+    &\large{$\textcolor{red}{\times}$}
+		&\large{$\textcolor{PineGreen}{\checkmark}$}
+		\\
+    \hline
+		\textbf{Clojure Hiccup}
+		&Nested Eager
+		&All
+    &\large{$\textcolor{red}{\times}$}
+    &\large{$\textcolor{PineGreen}{\checkmark}$}
+    &\large{$\textcolor{red}{\times}$}
+    &\large{$\textcolor{red}{\times}$}
+		\\
+    \hline
+    \shortstack{\textbf{Groovy }\\\textbf{MarkupBuilder}}
+		&Nested Lazy
+		&All
+    &\large{$\textcolor{red}{\times}$}
+    &\large{$\textcolor{PineGreen}{\checkmark}$}
+    &\large{$\textcolor{red}{\times}$}
+		&\large{$\textcolor{PineGreen}{\checkmark}$}
+		\\
+    \hline
+    \textbf{HtmlFlow}
+		&Chaining
+		&All
+    &\large{$\textcolor{PineGreen}{\checkmark}$}
+    &\large{$\textcolor{PineGreen}{\checkmark}$}
+    &\large{$\textcolor{PineGreen}{\checkmark}$}
+		&\large{$\textcolor{PineGreen}{\checkmark}$}
+		\\
+    \hline
+		\textbf{j2html}
+		&Nested Eager
+		&All
+    &\large{$\textcolor{red}{\times}$}
+    &\large{$\textcolor{PineGreen}{\checkmark}$}
+    &\large{$\textcolor{red}{\times}$}
+    &\large{$\textcolor{red}{\times}$}
+		\\
+    \hline
+		\textbf{KotlinX}
+		&Nested Lazy
+		&All
+    &\large{$\textcolor{red}{\times}$}
+    &\large{$\textcolor{PineGreen}{\checkmark}$}
+    &\textcolor{PineGreen}{\checkmark}$^{(2)}$
+		&\large{$\textcolor{PineGreen}{\checkmark}$}
+		\\
+    \hline
+		\textbf{ScalaTags}
+		&Nested Eager
+		&All
+    &\large{$\textcolor{red}{\times}$}
+    &\large{$\textcolor{PineGreen}{\checkmark}$}
+    &\large{$\textcolor{red}{\times}$}
+    &\large{$\textcolor{red}{\times}$}
+		\\
+		\hline                                                      
+	\end{tabular}
+	\caption{
+    Comparison of web template technologies in the Java ecosystem.
+    \\$^{(1)}$ Limited to a single model per web template and compatible with
+    the reactive stream \texttt{Publisher}.
+    \\$^{(2)}$ Non-safety for HTML attributes.
+  }
+	\label{table:cmplibs}
+\end{table}
+
+The first half of Table~\ref{table:cmplibs} lists template engines that use their
+own templating dialects, referred to as \textit{external DSLs}. The second half
+lists Java libraries that provide an \textit{internal DSL}, typically using a
+\textit{nested} or \textit{chaining} style to build HTML.
+Note that, by leveraging the host language (e.g., Clojure, Groovy, Java, Kotlin,
+or Scala) as the templating idiom, the latter impose no restrictions on the data
+model and fully support all styles of data access APIs. They also benefit from
+static type checking, which helps ensure type safety.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%-------------------------------------------------------------------
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 host language or any external API.
-
 For example, in Java, you may iterate over data using a \texttt{for} statement,

### 2025-05-21 docs/article/Sections/state-of-the-art.tex
chapter 2 - Explain Supported Data Models APIs

@@
 host language or any external API.
-For example, in Java, you may iterate over data using a \texttt{for} statement,
-or using the \texttt{java.util.streams} API, or any other JVM library such as
-Guava, Varv, StreamEx, or others—according to your preferences.
 
@@
 
-Restricted or unrestricted.
+Template engines with dedicated templating dialects typically rely on specific 
+control flow constructs—such as a \texttt{foreach}-like statement—to iterate 
+over elements in a data source. In the Java ecosystem, \texttt{Iterable} is the 
+common supertype implemented by most data structures and serves as the standard 
+API for iteration. However, other standard interfaces, such as 
+\texttt{java.util.Stream}, also represent sequences of elements but are not 
+compatible with \texttt{Iterable}.
+This fragmentation becomes more evident when considering other JVM languages, 
+such as Kotlin, which introduces additional abstractions like the 
+\texttt{Sequence} interface—also incompatible with \texttt{Iterable}. As a result, 
+template engines based on \textit{external} DSLs must explicitly support each of these 
+interfaces to accommodate the various iteration protocols across languages and 
+libraries.
+
+On the other hand, template engines that employ an \textit{internal} DSL do not 
+face such limitations, as they can leverage any available API within the host 
+programming language. This flexibility extends beyond the standard library to 
+include third-party APIs as well.
+For example, in Java, developers can iterate over data using a traditional 
+\texttt{for} loop, the \texttt{java.util.stream} API, or external libraries such 
+as Guava, Vavr, StreamEx, and others—depending on their specific needs or preferences.
 

### 2025-05-21 docs/article/Sections/state-of-the-art.tex
chapter 2 - added progressive rendering

@@
 \subsubsection{Progressive rendering}
+
+Pursuing the optimal user experience for web users has been a consistent goal
+since the inception of the World Wide Web.
+Consequently, various approaches have been explored to deliver the initial
+meaningful content to the end-user as promptly as possible, while the remaining
+content \textit{progressively} (or incrementally) loads as the server streams
+the HTML content.
+
+\textit{\textbf{Progressive rendering}} and \textit{\textbf{progressive loading}}
+encompass different concepts. The former pertains to the \textit{dynamic} content of a
+\textit{dynamic web page}, encompassing elements with logic and placeholders
+that are fulfilled by data from an object \textit{model} constructed at runtime. 
+On the other hand, the latter is associated with \textit{render-blocking
+resources} such as scripts, stylesheets, and HTML imports, which may hinder the
+browser from rendering page content to the screen~\cite{progressive-2022}.
+The notion of \textit{progressive rendering} has typically been aligned with
+\textit{client-side rendering} (CSR), where a single HTML page with static
+content is delivered upfront, while the dynamic content is fetched as the data
+becomes available to complete the web page.
+However, despite being overlooked, HTTP and browsers were designed from their
+inception to also support this feature in the context of SSR approaches,
+which offers the advantage of not being dependent on
+\textit{rendering-blocking resources}, such as the required JavaScript for CSR.
+% \textit{Progressive server-side rendering} (PSSR) is the
+% ability of streaming HTML content to the client in chunks or data frames, as it
+% gets resolved on the server at runtime. 
+% This lets the user-interface to be rendered incrementally (i.e.
+% \textit{progressively}) by the browser in line with the availability of the data. 
+% However, most SSR template engines do not inherently support this feature at the
+% template level, requiring developers to manually break down web pages into 
+% fragments or partials to enable PSSR. 
+An example of this limitation in SSR web templates
+was highlighted by Jeff Atwood in 2005, who criticized Microsoft ASP.NET for loading the
+entire web page into memory before sending any data to the browser~\cite{pssr2005}.
+% This practice delays the display of content to the user until the server
+% completes its processing, contradicting the original design intent of HTML to
+% render progressively as content is received. 
+% Notably, both Netscape and Internet Explorer were capable of rendering partial
+% HTML content from their inception.
+Despite historical critiques and HTML's inherent capabilities, most Web application
+frameworks, such as ASP.Net, Express.js, Spring, and others, 
+persistently lack support for progressive rendering, leading to the
+appearance of alternative techniques leveraging client-side JavaScript.
+Since 2007, various patents have addressed the PSSR issue, with Microsoft's
+patent enabling the infinite scrolling technique by displaying a single page of
+results~\cite{scroll2007}, and Yahoo's patent focusing on differentiating
+elements based on their position relative to the visible
+portion~\cite{schiller2007progressive}.
+
+Former techniques all describe methods of progressively adding content to a web
+page depending on client-side JavaScript.
+
+Introduce Thymeleaf
+
+Spring resolver does ot natively
+
+Carvalho~\cite{carvalho2023async} introduced an SSR solution that manages
+multiple data models and asynchronous APIs while ensuring well-formed HTML,
+PSSR, and non-blocking template resolution.
+That solution builds upon HtmlFlow, a Java DSL for HTML, with an internal
+processing mechanism that streams HTML in chunks as data items become available
+from data sources, without requiring client-side JavaScript.
+Like Thymeleaf, their proposal avoids the use of client-side JavaScript;
+however, unlike Thymeleaf, it supports a wide range of asynchronous APIs and
+multiple data sources, not limited to the \texttt{Publisher} API.
+However, the main counter-argument is the non-trivial management of the
+\emph{resume} callback in continuations~\cite{von2003events,callbackhell}, which
+is used to linearize the execution flow between asynchronous calls.

### 2025-05-23 docs/article/Sections/introduction.tex
Chapter 1: improve clarity, add references

--- 2025-05-14 docs/article/Sections/introduction.tex
@@
 
-% Non-blocking Progressive Server-Side Rendering (PSSR) is a technique that
-% combines some of the advantages of both server-side and client-side rendering
-% to improve web application performance and user experience. By leveraging
-% non-blocking I/O operations, PSSR allows for efficient handling of multiple
-% concurrent requests while progressively streaming HTML content to the client.
-% 
-% This project consists of a benchmark that evaluates the performance of
-% different PSSR strategies on the Java Virtual Machine (JVM). The benchmark
-% focuses on evaluating the performance of non-blocking PSSR implementations
-% using approaches such as Kotlin coroutines, reactive-style programming, and
-% Java virtual threads. Its main objective is to evaluate virtual threads as an
-% alternative that ensures a non-blocking approach to PSSR while offering a more
-% straightforward programming model than traditional asynchronous programming
-% techniques.
-% 
-% This project aims to extend the previous work done in
-% \textit{spring-webflux-comparing-template-engines}
-% \citeauthor{spring-webflux-comparing-template-engines}, which compares the
-% performance of various template engines in a Spring WebFlux application using
-% different approaches to PSSR\@. It adds support for Java virtual threads and
-% evaluates non-blocking PSSR implementations in other frameworks compared to
-% Spring WebFlux.
-% 
-% Currently, this project features a working benchmark in Spring WebFlux, Spring
-% MVC, and Quarkus, comparing the performance of eight different template
-% engines: Rocker, JStachio, Pebble, Freemarker, Trimou, Thymeleaf, HtmlFlow, and
-% KotlinX. It includes Java Microbenchmark Harness (JMH) benchmarks for template
-% engine performance testing, as well as Apache Bench and Apache JMeter scripts
-% for evaluating throughput and scalability. In the future, the project could be
-% extended to include other frameworks such as Vert.x, if time allows. However,
-% the next objective is the completion of the written report and article
-% submission. The written report will include a more detailed analysis of the
-% benchmark results, along with a contextualization of the state of the art in
-% non-blocking PSSR and asynchronous programming.
-% 
-% This report will focus on detailing relevant aspects of the benchmark
-% implementation and methodology, and will also provide a brief overview of the
-% results obtained from the JMeter and Apache Bench tests, which evaluate the
-% throughput and scalability of the different implementations.
+% Falta adicionar referências
+
+Modern web applications rely on different rendering strategies to optimize
+performance, user experience, and \textit{scalability}. The two most dominant
+approaches are \textit{Server-Side Rendering (SSR)} and \textit{Client-Side
+    Rendering (CSR)}. SSR generates HTML content on the server before sending it to
+the client, resulting in a faster \textit{First Contentful Paint (FCP)}
+\cite{Edgar2024-FCP} and better Search Engine Optimization (SEO). However, SSR
+can increase server load and reduce \textit{throughput}\footnote{The number of
+    requests the server can handle per second (RPS)} since each request requires
+additional processing before responding. In contrast, CSR shifts the rendering
+workload to the browser—the server initially sends a minimal HTML document with
+JavaScript, which dynamically loads the page content. While CSR reduces the
+server’s burden, it can lead to a slower \textit{FCP}, as users must wait for
+JavaScript execution before meaningful content appears.
+
+\textit{Progressive Server-Side Rendering (PSSR)} combines benefits from both SSR and CSR
+by streaming HTML content progressively. This technique enhances user-perceived performance by
+allowing progressive rendering as data becomes available. In this respect,
+PSSR is similar to CSR in that the server initially sends a minimal HTML
+document to the client and subsequently streams additional HTML fragments.
+However, unlike CSR, PSSR retains all rendering responsibilities on the server side,
+thereby reducing the load on the client. As a result, the client does
+not need to execute JavaScript or make additional requests to retrieve page content.
+
+Efficient thread utilization is crucial for low-thread servers. These servers
+must handle many concurrent requests without being overwhelmed, which
+necessitates asynchronous, non-blocking request handling
+\cite{carvalho2023async}. By leveraging asynchronous I/O operations—such as
+database queries and API calls, servers can avoid blocking threads while
+waiting for data, thereby maximizing throughput and responsiveness. To support
+this non-blocking architecture, PSSR implementations require template engines
+that are compatible with asynchronous data models and I/O. Some modern template
+engines, such as HtmlFlow \cite{htmlflow} and Thymeleaf \cite{thymeleaf}, have
+been designed with these capabilities in mind. However, many traditional
+template engines—particularly those using external domain-specific languages
+\cite{Fowler03} (DSLs)—still depend on blocking interfaces like
+\texttt{Iterable} for data processing. This blocking behavior forces server
+threads to remain idle until the entire HTML output is ready, undermining the
+performance benefits of non-blocking I/O and limiting scalability in
+high-concurrency environments.
+
+With the introduction of \textit{Virtual Threads} \cite{oracle-virtual-threads}
+in Java 21, it is now possible to execute blocking I/O operations in a
+scalable, lightweight manner. This capability allows traditional template
+engines—often reliant on blocking interfaces—to operate efficiently in
+high-concurrency, non-blocking environments without requiring complex
+asynchronous programming models. As a result, Progressive Server-Side Rendering
+(PSSR) can now be implemented using familiar HTML templates, 
+simplifying development and improving maintainability.
+
+This work explores the current landscape of non-blocking PSSR, focusing on two
+primary paradigms: \textit{reactive programming} and \textit{coroutines}, both
+of which have traditionally been used to achieve asynchronous I/O. As an
+alternative, we investigate whether Java’s Virtual Threads can offer comparable
+performance while preserving the simplicity of synchronous code. Section 2
+reviews the state-of-the-art in PSSR and template engine design. Section 3
+outlines the limitations of conventional engines in asynchronous settings and
+presents our proposed approach. Section 4 details the benchmark methodology,
+followed by the results and analysis in Section 5. Conclusions are presented in Section 6.

### 2025-05-23 docs/article/Sections/problem-statement.tex
Chapter 3: start with reactive approaches in internal DSLs, contrast with synchronous external DSLs, introduce virtual threads as possible solution.

@@
 
-In this section we discuss the problems associated with progressive server-side
-rendering (PSSR) in modern web applications, focusing on the challenges and
-limitations of existing approaches. We then present the objective of our work,
-which aims to increase the available options for PSSR in web applications,
-particularly in the context of JVM based frameworks.
+In this section, we examine the challenges of implementing Progressive
+Server-Side Rendering (PSSR) in modern web applications, with a focus on the
+limitations of current template engine designs. Our goal is to broaden the
+range of options available for PSSR, particularly within JVM-based frameworks.
 
-\textit As we outlined in the previous section, \textit{external} DSLs for HTML such as Thymeleaf, Jstachio, Handlebars, and others define HTML
-templates within HTML documents with markers specific to the engine. Template engines that use \textit{internal} DSLs, such as KotlinX or HtmlFlow,
-define the HTML templates in the programming language itself, leveraging the language's
-syntax and features to create the templates.
+Reactive streams, such as those based on \texttt{Observable} (from RxJava) and
+\texttt{Flow} (from the Java 9+ standard library), are abstractions that enable
+progressive server-side rendering (PSSR) in environments utilizing
+\textit{internal} DSLs. These abstractions provide a non-blocking, asynchronous
+way to handle data streams, making them ideal for generating and sending HTML
+content incrementally as data becomes available. In PSSR, the server does not
+wait for the entire data model to be ready before beginning to render HTML.
+Instead, it processes and streams each piece of data to the client as soon as
+it arrives. Reactive types like \texttt{Observable<T>} or \texttt{Flow<T>}
+facilitate this by representing data as a sequence of asynchronous events.
+These streams emit values over time and support backpressure, allowing the
+rendering engine to consume data at a manageable rate without overwhelming
+system resources. A reactive stream might emit a sequence of
+\texttt{Presentation} objects, each representing a talk in a conference
+schedule. As each \texttt{Presentation} is emitted by the \texttt{Observable}
+or \texttt{Flow}, the internal DSL-based engine—such as HtmlFlow—can render the
+corresponding HTML fragment and immediately flush it to the client. This
+approach is demonstrated in \autoref{lst:presentation-observable}, where each
+presentation is rendered asynchronously as it is emitted by an
+\texttt{Observable}, and in \autoref{lst:presentation-flow}, which shows an
+equivalent coroutine-based implementation using Kotlin's \texttt{Flow}. Both
+examples highlight how internal DSLs can natively integrate with reactive types
+to enable non-blocking, progressive rendering on the server side.
 
-\textit{Internal} DSLs can more easily support asynchronous data models since
-they can leverage the language's existing features and libraries, thus enabling
-the progressive streaming of HTML content to the client, with \textit{external}
-DSLs supporting simpler blocking data models such as \texttt{Iterable} or
-\texttt{List}. While on the previous section we outlined several advantages of
-using \textit{internal} DSLs over textual templates for HTML, there are also
-advantages to using \textit{external} DSLs:
+\lstset{style=listingstyle}
+\begin{lstlisting}[
+    language=Kotlin,
+    basicstyle=\scriptsize\ttfamily,
+    numbers=none,
+    caption={\textit{HtmlFlow} presentation template with \texttt{Observable}},
+    label={lst:presentation-observable}
+]
+await { div, model: Observable<Presentation>, onCompletion ->
+    model
+        .doOnNext { 
+            presentationFragmentAsync
+                .renderAsync(it)
+                .thenApply { frag -> div.raw(frag) }
+        }
+        .doOnComplete { onCompletion.finish() }
+        .subscribe()
+}
+\end{lstlisting}
+\hfil
+\begin{lstlisting}[
+    language=Kotlin,
+    basicstyle=\scriptsize\ttfamily,
+    numbers=none,
+    label={lst:presentation-flow},
+    caption={\textit{HtmlFlow} presentation template with \texttt{Flow}},
+]
+suspending { model: Flow<Presentation> ->
+    model.collect {
+        presentationFragmentAsync
+            .renderAsync(it)
+            .thenApply { frag -> raw(frag) }
+    }
+}
+\end{lstlisting}
 
-\begin{enumerate}
-    \item \emph{Separation of Concerns}: \textit{External} DSLs allow for a clear separation between
-          the HTML structure and the application logic, making it easier to maintain and update
-          templates without modifying the underlying code. This ''
-          separation'' can be beneficial, as it allows front-end developers to work on the HTML
-          templates without needing to understand the underlying application logic or programming language.
-    \item \emph{Cross-Language Compatibility}: \textit{External} DSLs can be used across different programming languages
-          and frameworks, making them more versatile and easier to integrate into existing projects.
-          This is particularly useful in environments where multiple languages are used.
-    \item \emph{Familiarity}: Many developers are already familiar with HTML and its syntax, making
-          it easier to work with \textit{external} DSLs that use HTML-like syntax.
-\end{enumerate}
+By contrast, template engines that use \textit{external} DSLs—such as JStachio,
+Thymeleaf, or Handlebars—typically define templates within static HTML
+documents using custom markers, and rely on blocking interfaces like
+\texttt{Iterable} or \texttt{List}. These interfaces require the entire data
+model to be materialized before rendering can begin, which blocks server
+threads during template expansion and significantly limits scalability under
+high concurrency. These interfaces require the entire data model to be
+materialized before rendering can begin, which blocks server threads during
+template expansion and significantly limits scalability under high concurrency.
+Some reactive libraries, such as RxJava, provide bridging mechanisms like
+\texttt{Observable.blockingIterable()}, which allows asynchronous data sources
+to be exposed as Iterable by blocking the thread until new items are available.
+While useful for compatibility with traditional APIs, this approach
+reintroduces blocking behavior and undermines the benefits of non-blocking
+I/O—especially under high concurrency. \autoref{lst:presentation-jstachio}
+illustrates this model using a JStachio template, where the engine performs a
+blocking loop over \texttt{presentationItems}.
 
-Due to these advantages, many developers prefer to use \textit{external} DSLs
-for HTML templates, even if it means sacrificing some of the advantages of
-using \textit{internal} DSLs. However, this preference can lead to challenges
-when it comes to progressive server-side rendering (PSSR) and non-blocking I/O,
-as many \textit{external} DSLs do not support asynchronous data models.
-
-\autoref{lst:presentation-jstachio} shows the HTML template for a \texttt{Presentation} class using the JStachio template engine, 
-while \autoref{lst:presentation-htmlflow} presents the equivalent template using HtmlFlow.
+\vspace{1cm}
 
@@
   basicstyle=\scriptsize\ttfamily,
-  caption={Presentation HTML template using \textit{JStachio}}
+  caption={Presentation HTML template using \textit{JStachio}},
   label={lst:presentation-jstachio}
 ]
-<!DOCTYPE html>
-<html lang="en-US">
-<head>
-    <meta charset="utf-8">
-    <meta name="viewport" content="width=device-width, initial-scale=1.0">
-    <meta http-equiv="X-UA-compatible" content="IE=Edge">
-    <title>
-        JFall 2013 Presentations - JStachio
-    </title>
-    <link rel="Stylesheet" href="/webjars/bootstrap/5.3.0/css/bootstrap.min.css">
-</head>
-<body>
-<div class="container">
-    <div class="pb-2 mt-4 mb-3 border-bottom">
-        <h1>
-            JFall 2013 Presentations - JStachio
-        </h1>
-    </div>
 {{#presentationItems}}
@@
 {{/presentationItems}}
-</div>
-</body>
-</html>
 \end{lstlisting}
-\vspace{0.5cm}
-\begin{lstlisting}[
-  language=Kotlin,
-  numbers=none,
-  basicstyle=\scriptsize\ttfamily,
-  caption={Presentation HTML template using \textit{HtmlFlow}}
-  label={lst:presentation-htmlflow}
-]
- val htmlFlowTemplate: HtmlViewAsync<Observable<Presentation>> =
-    HtmlFlow.viewAsync<Observable<Presentation>?> { view ->
-        view.html {
-            attrLang("en-us")
-            head {
-                meta { attrCharset("UTF-8") }
-                meta { attrName("viewport").attrContent("width=device-width, initial-scale=1.0") }
-                meta { addAttr("http-equiv", "X-UA-Compatible").attrContent("IE=Edge") }
-                title { text("JFall 2013 Presentations - HtmlFlow") }
-                link { attrRel(EnumRelType.STYLESHEET).attrHref("/webjars/bootstrap/5.3.0/css/bootstrap.min.css") }
-            }
-            body {
-                div {
-                    attrClass("container")
-                    div {
-                        attrClass("pb-2 mt-4 mb-3 border-bottom")
-                        h1 { text("JFall 2013 Presentations - HtmlFlow") }
-                    }
-                    await { div, model: Observable<Presentation>, onCompletion ->
-                        model
-                            .doOnNext { 
-                                presentationFragmentAsync
-                                    .renderAsync(it)
-                                    .thenApply { frag -> div.raw(frag) }
-                            }
-                            .doOnComplete { onCompletion.finish() }
-                            .subscribe()
-                    }
-                }
-            }
-        }
-    }.threadSafe()
-\end{lstlisting}
-\lstset{style=listingstyle}
 
-HtmlFlow is an example of an \textit{internal} DSL that uses calls to HTML
-builders as a way to emit HTML\@. It allows for the use of the
-\texttt{Observable} class as a data model, which in turn allows for the definition of a
-\texttt{Consumer<Presentation>} through the \texttt{doOnNext} method. Every
-time a new value is emitted to the \texttt{Observable}, the \texttt{doOnNext}
-method is called, emitting the HTML fragment for the \texttt{Presentation}
-object to the client.
+Despite these performance limitations, external DSLs remain popular due to
+several advantages:
 
-The Jstachio template engine, on the other hand, uses an \textit{external} DSL,
-where the HTML template is defined in a separate file. The template engine uses
-the \texttt{presentationItems} variable to define the data model, which is a
-list of \texttt{Presentation} objects. This allows the template engine to
-iterate over the list of \texttt{Presentation} objects and render the HTML for
-each one.
+\begin{enumerate}
+    \item \emph{Separation of Concerns}: HTML templates are decoupled from application logic, enabling front-end developers to contribute without modifying back-end code.
+    \item \emph{Cross-Language Compatibility}: External DSLs are portable across languages and frameworks, easing integration in multi-language environments.
+    \item \emph{Familiarity}: Many developers are comfortable with HTML syntax, lowering the barrier to entry and improving maintainability.
+\end{enumerate}
 
-As demonstrated, the HtmlFlow template enables PSSR by directly supporting the
-use of the \texttt{Observable} class as a data model. However, not all
-programmers are familiar with asynchronous programming models, which can make
-implementing this solution challenging. In contrast, Jstachio defines templates
-using standard HTML syntax with specific tags for the template engine,
-resulting in a simpler and more accessible template definition. However,
-because Jstachio templates rely on an \texttt{Iterable} interface,
-they block the calling thread until the entire HTML content is rendered and
-therefore do not support non-blocking PSSR\@.
+These strengths make external DSLs appealing—even when they come at the cost of
+blocking, synchronous rendering. However, this tradeoff becomes critical under
+high concurrency, where blocking threads severely degrades throughput. Emerging
+features in the Java ecosystem, particularly \textit{virtual threads}
+introduced in Java 21 as part of Project Loom, offer a promising solution to
+this challenge. Virtual threads drastically reduce the overhead of blocking
+operations by decoupling thread execution from OS-level threads. As a result,
+engines that rely on blocking interfaces—like those used in
+external DSLs—can potentially achieve scalability levels that approach those of
+non-blocking, asynchronous engines.
 
-% Mencionar que certas DSLs internas não suportam PSSR por 
-% causa de HTML mal formado? (KotlinX)
+% Incluir este parágrafo?
+% In previous work \cite{PSSR-WISE2024}, Carvalho benchmarked the performance of
+% HtmlFlow (an internal DSL) using suspendable templates within a Spring WebFlux
+% application. HtmlFlow scaled efficiently to 128 concurrent users and delivered
+% up to 4,000 requests per second. In contrast, blocking engines such as JStachio
+% saturated at just 4 concurrent users, maxing out at around 400 requests per
+% second—highlighting the scalability limitations of blocking models.This work
+% builds upon those findings by evaluating whether virtual threads can close this
+% performance gap for blocking template engines. Specifically, we aim to
+% determine whether external DSLs—traditionally unsuitable for PSSR due to their
+% reliance on blocking APIs—can become viable for progressive rendering when
+% powered by virtual threads.
 
-In previous work \cite{PSSR-WISE2024}, Carvalho developed a benchmark
-using Spring Webflux to evaluate the performance of HtmlFlow with suspendable
-web templates. In this benchmark, HtmlFlow's performance was compared to that
-of Thymeleaf using a reactive approach, as well as other template engines such
-as the one shown in \autoref{lst:presentation-JStachio}. The results
-demonstrated that HtmlFlow scaled efficiently up to 128 concurrent users,
-achieving a throughput of up to 4,000 requests per second. In contrast, the
-blocking template engines ceased to scale beyond 4 concurrent users, reaching a
-maximum throughput of approximately 400 requests per second. These findings
-indicate that engines which block the calling thread scale poorly and are not
-suitable for PSSR\@.
-
-% Expandir sobre virtual threads aqui?
-With the advent of Java 21 and the introduction of \textit{Virtual Threads}, it
-is now possible to execute code that would otherwise block the calling thread
-in a scalable, non-blocking manner. This advancement enables the use of these
-\textit{external} DSLs in non-blocking contexts. The objective of this work is
-to evaluate the performance of these template engines when leveraging Virtual
-Threads, and to determine whether they can be made viable for PSSR\@. We will
-also compare the performance of this approach with existing solutions based on
-reactive programming and Kotlin coroutines.
-
-If the performance of these engines proves comparable to that of current
-approaches, it would enable PSSR to be implemented using more familiar HTML
-syntax, simplifying template development and adoption. Furthermore, template
-engine developers seeking to support PSSR would no longer need to implement
-complex asynchronous programming models to achieve non-blocking I/O; instead,
-they could rely on the traditional approaches used for Server Side Rendering
-(SSR).
+If successful, this approach would enable developers to reap the ergonomic
+benefits of external DSLs while maintaining scalable performance, without
+adopting complex asynchronous paradigms. It would also simplify the
+implementation of PSSR across a broader range of frameworks and engines, making
+the technique more accessible to mainstream web development.

### 2025-05-23 docs/article/Sections/introduction.tex
Chapter 2: add Spring MVC subsection

@@
 the client, resulting in a faster \textit{First Contentful Paint (FCP)}
-\cite{Edgar2024-FCP} and better Search Engine Optimization (SEO). However, SSR
+~\cite{Edgar2024-FCP} and better Search Engine Optimization (SEO). However, SSR
 can increase server load and reduce \textit{throughput}\footnote{The number of
@@
 necessitates asynchronous, non-blocking request handling
-\cite{carvalho2023async}. By leveraging asynchronous I/O operations—such as
+~\cite{carvalho2023async}. By leveraging asynchronous I/O operations—such as
 database queries and API calls, servers can avoid blocking threads while
@@
 that are compatible with asynchronous data models and I/O. Some modern template
-engines, such as HtmlFlow \cite{htmlflow} and Thymeleaf \cite{thymeleaf}, have
+engines, such as HtmlFlow~\cite{htmlflow} and Thymeleaf~\cite{thymeleaf}, have
 been designed with these capabilities in mind. However, many traditional
 template engines—particularly those using external domain-specific languages
-\cite{Fowler03} (DSLs)—still depend on blocking interfaces like
+~\cite{Fowler03} (DSLs)—still depend on blocking interfaces like
 \texttt{Iterable} for data processing. This blocking behavior forces server
@@
 
-With the introduction of \textit{Virtual Threads} \cite{oracle-virtual-threads}
+With the introduction of \textit{Virtual Threads}~\footnote{\url{https://openjdk.org/jeps/444}}
 in Java 21, it is now possible to execute blocking I/O operations in a

### 2025-05-23 docs/article/Sections/state-of-the-art.tex
Chapter 2: add Spring MVC subsection

@@
 servers in their internal architectures, and how these choices impact the
-behavior of the web template engines.
-Then, in Subsection~2.2, we present the main properties that characterize each 
-web template technology approach, along with the advantages and drawbacks that 
-result from these characteristics.
+behavior of the web template engines. Then, in Subsection~2.2, we present the
+main properties that characterize each web template technology approach, along
+with the advantages and drawbacks that result from these characteristics.
 
@@
 bogged down by context switching and thread management
-overhead~\cite{kant2000scalable}.
-In the Java ecosystem, Spring Boot is one of the most widely used frameworks that
-follow this model, according to several reports such as the Stack Overflow 2024
-Developer
+overhead~\cite{kant2000scalable}. In the Java ecosystem, Spring Boot is one of
+the most widely used frameworks that follow this model, according to several
+reports such as the Stack Overflow 2024 Developer
 Survey~\footnote{~\url{https://survey.stackoverflow.co/2024/technology\#1-web-frameworks-and-technologies}}.
@@
 facilitated the development of highly scalable, responsive, and resilient web
-applications capable of managing substantial data loads~\cite{Jin15}.
-The prominence of this concept increased with the advent of Node.js in
-2009, and subsequently, various technologies adopted this approach
-in the Java ecosystem, including Netty, Akka HTTP, Vert.X, and Spring WebFlux.
+applications capable of managing substantial data loads~\cite{Jin15}. The
+prominence of this concept increased with the advent of Node.js in 2009, and
+subsequently, various technologies adopted this approach in the Java ecosystem,
+including Netty, Akka HTTP, Vert.X, and Spring WebFlux.
 % Among these, Spring WebFlux stands out as the most widely used web framework in
@@
 The \textit{non-blocking} I/O model in low-thread servers functions optimally
-only when HTTP handlers avoid blocking.
-Therefore, HTML templates need to be proficient in dealing with the asynchronous
-APIs provided by data models.
-While most legacy web templates struggle with asynchronous models, DSLs for HTML
-face no such limitations, leveraging all constructions available in the host
-programming language.
-However, the unexpected intertwining of asynchronous handlers' completion and
-HTML builders' execution may potentially lead to malformed HTML with an
-unexpected layout, as demonstrated in the subsequent subsection~\ref{sec:async-support}.
+only when HTTP handlers avoid blocking. Therefore, HTML templates need to be
+proficient in dealing with the asynchronous APIs provided by data models. While
+most legacy web templates struggle with asynchronous models, DSLs for HTML face
+no such limitations, leveraging all constructions available in the host
+programming language. However, the unexpected intertwining of asynchronous
+handlers' completion and HTML builders' execution may potentially lead to
+malformed HTML with an unexpected layout, as demonstrated in the subsequent
+subsection~\ref{sec:async-support}.
 
-An alternative approach involves utilizing user-level threads, while maintaining
-a blocking I/O and a synchronous programming paradigm.
-However, this approach still requires a user-level I/O subsystem capable of
-mitigating system-level blocking, which is crucial for the performance of
-I/O-intensive applications. This technique offers a lightweight solution for
-efficiently managing a larger number of concurrent sessions by minimizing
-per-thread overhead.
-In 2020, Karsten~\cite{karsten2020} demonstrated how this strategy supports a
-synchronous programming style, effectively abstracting away the complexities
-associated with managing continuations in asynchronous programming.
+An alternative approach involves utilizing user-level threads, while
+maintaining a blocking I/O and a synchronous programming paradigm. However,
+this approach still requires a user-level I/O subsystem capable of mitigating
+system-level blocking, which is crucial for the performance of I/O-intensive
+applications. This technique offers a lightweight solution for efficiently
+managing a larger number of concurrent sessions by minimizing per-thread
+overhead. In 2020, Karsten~\cite{karsten2020} demonstrated how this strategy
+supports a synchronous programming style, effectively abstracting away the
+complexities associated with managing continuations in asynchronous
+programming.
 
@@
 \textit{coordinated} within a well-defined
-lifecycle~\cite{elizarov2021coroutines}.
-Although this model embraces the \texttt{async}/\texttt{await}
-feature~\cite{async_await}, enabling non-blocking routines to mimic the
-structure of synchronous ones, it still lacks a user-level I/O subsystem and
-relies on an explicit I/O dispatcher tied to a dedicated thread pool that frees
-worker threads for handling processing tasks.
+lifecycle~\cite{elizarov2021coroutines}. Although this model embraces the
+\texttt{async}/\texttt{await} feature~\cite{async_await}, enabling non-blocking
+routines to mimic the structure of synchronous ones, it still lacks a
+user-level I/O subsystem and relies on an explicit I/O dispatcher tied to a
+dedicated thread pool that frees worker threads for handling processing tasks.
 Moreover, web templates using their own templating dialects, such as JSP,
@@
 \subsubsection{Spring MVC}
+
+\subsubsection{Spring MVC}
+
+Spring MVC is a classical, synchronous web framework built on the
+thread-per-request model. When an HTTP request is received, it is handled by a
+dedicated thread from the server’s thread pool (e.g., Tomcat, Jetty, or
+Undertow). This thread is responsible for executing the full request lifecycle
+synchronously, including invoking controllers, executing business logic, and
+rendering the response view. Because the model is blocking, the thread remains
+occupied throughout the request—even during I/O operations such as database
+queries or external API calls—which can lead to performance bottlenecks under
+high concurrency.
+
+Controllers in Spring MVC typically return Java objects or models that are
+processed by view resolvers in conjunction with template engines like JSP,
+Thymeleaf, or FreeMarker. These templates generate the full HTML response in
+one pass, which is then sent to the client only after all necessary data has
+been gathered and rendered. By default, this architecture limits the ability to
+perform \textit{Progressive Server-Side Rendering (PSSR)}, as traditional
+rendering waits for all data before sending any output. This results in higher
+latency for clients and less responsive page loads, particularly for
+data-intensive views. However, Spring MVC provides a mechanism for response
+streaming via the \texttt{StreamingResponseBody} interface, introduced in
+Spring 4.2. When a controller method returns a \texttt{StreamingResponseBody},
+Spring writes directly to the \texttt{HttpServletResponse} output stream,
+allowing developers to send parts of the response incrementally as they become
+available. This is particularly useful for:
+\begin{itemize}
+  \item Sending large responses without buffering the entire output in memory.
+  \item Writing dynamic HTML in fragments as data is fetched or computed.
+  \item Reducing Time to First Byte (TTFB) and improving perceived performance.
+\end{itemize}
+
+A regular usage might look like the following:
+\begin{lstlisting}[language=Java]
+  @GetMapping("/stream")
+  fun handleStream(): StreamingResponseBody? {
+      return StreamingResponseBody { outputStream ->
+          for (i in 0..9) {
+              val htmlFragment = "<p>Chunk " + i + "</p>\n"
+              outputStream.write(htmlFragment.toByteArray())
+              outputStream.flush() // ensure partial response is sent
+              Thread.sleep(500) // simulate delay
+          }
+      }
+  }
+\end{lstlisting}
+
+In this example, HTML content is written and flushed in discrete chunks,
+allowing the client to progressively render the response as it arrives, while
+the server continues processing additional data. Although
+\texttt{StreamingResponseBody} does not eliminate the blocking nature inherent
+in Spring MVC—since the handling thread remains active during the streaming
+process—it does enable a basic form of PSSR by facilitating incremental content
+delivery. This approach is particularly useful for long-running operations or
+large response bodies, where providing early partial output can enhance
+perceived performance.
+
+However, a key limitation we encountered with this method involves the servlet
+response buffer. By default, most servlet containers (e.g., Tomcat) use a
+response buffer size of approximately 8KB. Data written to the response output
+stream is held in this buffer until it reaches capacity. Consequently, if the
+total size of the initial HTML content is less than the buffer threshold, the
+server will not transmit any data to the client until the buffer is filled.
+This buffering behavior introduces a delay in sending the first chunk of
+content, thereby negating the benefits of progressive rendering. In scenarios
+where the rendered HTML template does not produce sufficient data to exceed the
+buffer threshold early, PSSR becomes ineffective, and the client experiences a
+delayed initial response.
 
@@
 
-Although some of the aforementioned characteristics apply to both \textit{server-side}
-and \textit{client-side} approaches, we focus solely on web template technologies for
-\textit{server-side rendering}, as our work is centered on that approach.
-Before getting deep on each of the aforementioned characteristics,
-Table~\ref{table:cmplibs} presents a breakdown of mainstream template engines,
-classified according to the identified properties.
+Although some of the aforementioned characteristics apply to both
+\textit{server-side} and \textit{client-side} approaches, we focus solely on
+web template technologies for \textit{server-side rendering}, as our work is
+centered on that approach. Before getting deep on each of the aforementioned
+characteristics, Table~\ref{table:cmplibs} presents a breakdown of mainstream
+template engines, classified according to the identified properties.
 
@@
     \textbf{Library}
-		&\textbf{DSL idiom}
-		&\textbf{Data Model APIs}
-    &\shortstack{\textbf{Asynchronous}\\\textbf{Support}}
-    &\shortstack{\textbf{Type}\\\textbf{Safety}}
-    &\shortstack{\textbf{HTML}\\\textbf{Safety}}
-		&\shortstack{\textbf{Progressive}\\\textbf{Rendering}}
+     & \textbf{DSL idiom}
+     & \textbf{Data Model APIs}
+     & \shortstack{\textbf{Asynchronous}           \\\textbf{Support}}
+     & \shortstack{\textbf{Type}                   \\\textbf{Safety}}
+     & \shortstack{\textbf{HTML}                   \\\textbf{Safety}}
+     & \shortstack{\textbf{Progressive}            \\\textbf{Rendering}}
     \\
@@
     \textbf{Freemarker}
-		&External DSL
-		&\texttt{Iterable}
-		&\large{$\textcolor{red}{\times}$}
-    &\large{$\textcolor{red}{\times}$}
-    &\large{$\textcolor{red}{\times}$}
-		&\large{$\textcolor{PineGreen}{\checkmark}$}
+     & External DSL
+     & \texttt{Iterable}
+     & \large{$\textcolor{red}{\times}$}
+     & \large{$\textcolor{red}{\times}$}
+     & \large{$\textcolor{red}{\times}$}
+     & \large{$\textcolor{PineGreen}{\checkmark}$}
     \\
@@
     \textbf{JSP}
-		&External DSL
-		&\texttt{Iterable}
-		&\large{$\textcolor{red}{\times}$}
-    &\large{$\textcolor{red}{\times}$}
-    &\large{$\textcolor{red}{\times}$}
-		&\large{$\textcolor{red}{\times}$}
+     & External DSL
+     & \texttt{Iterable}
+     & \large{$\textcolor{red}{\times}$}
+     & \large{$\textcolor{red}{\times}$}
+     & \large{$\textcolor{red}{\times}$}
+     & \large{$\textcolor{red}{\times}$}
     \\
@@
     \textbf{JStachio}
-		&External DSL
-		&\texttt{Iterable}
-		&\large{$\textcolor{red}{\times}$}
-    &\large{$\textcolor{red}{\times}$}
-    &\large{$\textcolor{red}{\times}$}
-		&\large{$\textcolor{PineGreen}{\checkmark}$}
+     & External DSL
+     & \texttt{Iterable}
+     & \large{$\textcolor{red}{\times}$}
+     & \large{$\textcolor{red}{\times}$}
+     & \large{$\textcolor{red}{\times}$}
+     & \large{$\textcolor{PineGreen}{\checkmark}$}
     \\\hline
     \textbf{Pebble}
-		&External DSL
-		&\texttt{Iterable}
-		&\large{$\textcolor{red}{\times}$}
-    &\large{$\textcolor{red}{\times}$}
-    &\large{$\textcolor{red}{\times}$}
-		&\large{$\textcolor{PineGreen}{\checkmark}$}
+     & External DSL
+     & \texttt{Iterable}
+     & \large{$\textcolor{red}{\times}$}
+     & \large{$\textcolor{red}{\times}$}
+     & \large{$\textcolor{red}{\times}$}
+     & \large{$\textcolor{PineGreen}{\checkmark}$}
     \\
@@
     \textbf{Rocker}
-		&External DSL
-		&\texttt{Iterable}
-		&\large{$\textcolor{red}{\times}$}
-    &\large{$\textcolor{red}{\times}$}
-    &\large{$\textcolor{red}{\times}$}
-		&\large{$\textcolor{PineGreen}{\checkmark}$}
+     & External DSL
+     & \texttt{Iterable}
+     & \large{$\textcolor{red}{\times}$}
+     & \large{$\textcolor{red}{\times}$}
+     & \large{$\textcolor{red}{\times}$}
+     & \large{$\textcolor{PineGreen}{\checkmark}$}
     \\
@@
     \textbf{Thymeleaf}
-		&External DSL
-		&\shortstack{\texttt{Iterable}\\\texttt{Publisher}}
-		&\texttt{Publisher}$^{(1)}$
-    &\large{$\textcolor{red}{\times}$}
-    &\large{$\textcolor{red}{\times}$}
-		&\large{$\textcolor{PineGreen}{\checkmark}$}
+     & External DSL
+     & \shortstack{\texttt{Iterable}               \\\texttt{Publisher}}
+     & \texttt{Publisher}$^{(1)}$
+     & \large{$\textcolor{red}{\times}$}
+     & \large{$\textcolor{red}{\times}$}
+     & \large{$\textcolor{PineGreen}{\checkmark}$}
     \\
@@
     \textbf{Trimou}
-		&External DSL
-		&\texttt{Iterable}
-		&\large{$\textcolor{red}{\times}$}
-    &\large{$\textcolor{red}{\times}$}
-    &\large{$\textcolor{red}{\times}$}
-		&\large{$\textcolor{PineGreen}{\checkmark}$}
+     & External DSL
+     & \texttt{Iterable}
+     & \large{$\textcolor{red}{\times}$}
+     & \large{$\textcolor{red}{\times}$}
+     & \large{$\textcolor{red}{\times}$}
+     & \large{$\textcolor{PineGreen}{\checkmark}$}
     \\
@@
     \textbf{Velocity}
-		&External DSL
-		&\texttt{Iterable}
-		&\large{$\textcolor{red}{\times}$}
-    &\large{$\textcolor{red}{\times}$}
-    &\large{$\textcolor{red}{\times}$}
-		&\large{$\textcolor{PineGreen}{\checkmark}$}
+     & External DSL
+     & \texttt{Iterable}
+     & \large{$\textcolor{red}{\times}$}
+     & \large{$\textcolor{red}{\times}$}
+     & \large{$\textcolor{red}{\times}$}
+     & \large{$\textcolor{PineGreen}{\checkmark}$}
     \\
@@
     \textbf{Clojure Hiccup}
-		&Nested Eager
-		&All
-    &\large{$\textcolor{red}{\times}$}
-    &\large{$\textcolor{PineGreen}{\checkmark}$}
-    &\large{$\textcolor{red}{\times}$}
-    &\large{$\textcolor{red}{\times}$}
+     & Nested Eager
+     & All
+     & \large{$\textcolor{red}{\times}$}
+     & \large{$\textcolor{PineGreen}{\checkmark}$}
+     & \large{$\textcolor{red}{\times}$}
+     & \large{$\textcolor{red}{\times}$}
     \\
     \hline
-    \shortstack{\textbf{Groovy }\\\textbf{MarkupBuilder}}
-		&Nested Lazy
-		&All
-    &\large{$\textcolor{red}{\times}$}
-    &\large{$\textcolor{PineGreen}{\checkmark}$}
-    &\large{$\textcolor{red}{\times}$}
-		&\large{$\textcolor{PineGreen}{\checkmark}$}
+    \shortstack{\textbf{Groovy }                   \\\textbf{MarkupBuilder}}
+     & Nested Lazy
+     & All
+     & \large{$\textcolor{red}{\times}$}
+     & \large{$\textcolor{PineGreen}{\checkmark}$}
+     & \large{$\textcolor{red}{\times}$}
+     & \large{$\textcolor{PineGreen}{\checkmark}$}
     \\
@@
     \textbf{HtmlFlow}
-		&Chaining
-		&All
-    &\large{$\textcolor{PineGreen}{\checkmark}$}
-    &\large{$\textcolor{PineGreen}{\checkmark}$}
-    &\large{$\textcolor{PineGreen}{\checkmark}$}
-		&\large{$\textcolor{PineGreen}{\checkmark}$}
+     & Chaining
+     & All
+     & \large{$\textcolor{PineGreen}{\checkmark}$}
+     & \large{$\textcolor{PineGreen}{\checkmark}$}
+     & \large{$\textcolor{PineGreen}{\checkmark}$}
+     & \large{$\textcolor{PineGreen}{\checkmark}$}
     \\
@@
     \textbf{j2html}
-		&Nested Eager
-		&All
-    &\large{$\textcolor{red}{\times}$}
-    &\large{$\textcolor{PineGreen}{\checkmark}$}
-    &\large{$\textcolor{red}{\times}$}
-    &\large{$\textcolor{red}{\times}$}
+     & Nested Eager
+     & All
+     & \large{$\textcolor{red}{\times}$}
+     & \large{$\textcolor{PineGreen}{\checkmark}$}
+     & \large{$\textcolor{red}{\times}$}
+     & \large{$\textcolor{red}{\times}$}
     \\
@@
     \textbf{KotlinX}
-		&Nested Lazy
-		&All
-    &\large{$\textcolor{red}{\times}$}
-    &\large{$\textcolor{PineGreen}{\checkmark}$}
-    &\textcolor{PineGreen}{\checkmark}$^{(2)}$
-		&\large{$\textcolor{PineGreen}{\checkmark}$}
+     & Nested Lazy
+     & All
+     & \large{$\textcolor{red}{\times}$}
+     & \large{$\textcolor{PineGreen}{\checkmark}$}
+     & \textcolor{PineGreen}{\checkmark}$^{(2)}$
+     & \large{$\textcolor{PineGreen}{\checkmark}$}
     \\
@@
     \textbf{ScalaTags}
-		&Nested Eager
-		&All
-    &\large{$\textcolor{red}{\times}$}
-    &\large{$\textcolor{PineGreen}{\checkmark}$}
-    &\large{$\textcolor{red}{\times}$}
-    &\large{$\textcolor{red}{\times}$}
+     & Nested Eager
+     & All
+     & \large{$\textcolor{red}{\times}$}
+     & \large{$\textcolor{PineGreen}{\checkmark}$}
+     & \large{$\textcolor{red}{\times}$}
+     & \large{$\textcolor{red}{\times}$}
     \\
@@
 
-The first half of Table~\ref{table:cmplibs} lists template engines that use their
-own templating dialects, referred to as \textit{external DSLs}. The second half
-lists Java libraries that provide an \textit{internal DSL}, typically using a
-\textit{nested} or \textit{chaining} style to build HTML.
-Note that, by leveraging the host language (e.g., Clojure, Groovy, Java, Kotlin,
-or Scala) as the templating idiom, the latter impose no restrictions on the data
-model and fully support all styles of data access APIs. They also benefit from
-static type checking, which helps ensure type safety.
+The first half of Table~\ref{table:cmplibs} lists template engines that use
+their own templating dialects, referred to as \textit{external DSLs}. The
+second half lists Java libraries that provide an \textit{internal DSL},
+typically using a \textit{nested} or \textit{chaining} style to build HTML.
+Note that, by leveraging the host language (e.g., Clojure, Groovy, Java,
+Kotlin, or Scala) as the templating idiom, the latter impose no restrictions on
+the data model and fully support all styles of data access APIs. They also
+benefit from static type checking, which helps ensure type safety.
 
@@
 \textit{domain}~\cite{evans2004domain}—in this case, HTML for expressing web
-documents.
-The DSL constrains the template's syntax and semantics to match the structure
-and purpose of HTML.
+documents. The DSL constrains the template's syntax and semantics to match the
+structure and purpose of HTML.
 
 DSLs can be divided in two types: \textit{external} or
-\textit{internal}\cite{dslbook}.
-\textit{External} DSLs are languages created without any affiliation to a
-concrete programming language.
-An example of an \textit{external} DSL is the regular expressions search
+\textit{internal}\cite{dslbook}. \textit{External} DSLs are languages created
+without any affiliation to a concrete programming language. An example of an
+\textit{external} DSL is the regular expressions search
 pattern\cite{thompson1968}, since it defines its own syntax without any
-dependency of programming languages. 
-On the other hand an \textit{internal} DSL is defined within a host programming
-language as a library and tends to be limited to the syntax of the host language,
-such as Java.
+dependency of programming languages. On the other hand an \textit{internal} DSL
+is defined within a host programming language as a library and tends to be
+limited to the syntax of the host language, such as Java.
 JQuery\cite{resig2007pro} is one of the most well-known examples of an internal
@@
 such as JSP, ASP, Velocity, PHP, and others adopted this \textit{external} DSL
-approach as their templating dialect.
-For example, a \texttt{foreach} loop can be expressed in each technology using its 
-own DSL:
-\\\texttt{<\% for(String item : items) \%>} in JSP, \texttt{<\% For Each item 
-In items \%>} in legacy ASP with VBScript, \texttt{\#foreach(\$item in \$items)} in 
-Velocity, or \texttt{<?php foreach (\$items as \$item):?>} in PHP.
+approach as their templating dialect. For example, a \texttt{foreach} loop can
+be expressed in each technology using its own DSL: \\\texttt{<\% for(String
+  item : items) \%>} in JSP, \texttt{<\% For Each item In items \%>} in legacy
+ASP with VBScript, \texttt{\#foreach(\$item in \$items)} in Velocity, or
+\texttt{<?php foreach (\$items as \$item):?>} in PHP.
 
-On the other hand, \textit{internal} DSLs for HTML allow templates to be defined 
-directly within the \emph{host} language (such as Java, Kotlin, Groovy, Scala, or 
-other general-purpose programming languages), rather than using text-based 
-template files~\cite{carvalho2019hot}. In this case, a web template is not limited 
-to templating constructs but may instead leverage any available primitive of the 
-host language or any external API.
+On the other hand, \textit{internal} DSLs for HTML allow templates to be
+defined directly within the \emph{host} language (such as Java, Kotlin, Groovy,
+Scala, or other general-purpose programming languages), rather than using
+text-based template files~\cite{carvalho2019hot}. In this case, a web template
+is not limited to templating constructs but may instead leverage any available
+primitive of the host language or any external API.
 
@@
 
-DSLs for HTML provide an API where methods or functions correspond to the names of
-available HTML elements. 
-These methods, also known as \textit{builders}, can be combined in a chain of
-calls to mimic the construction of an HTML document in a fluent manner.
-Martin Fowler\cite{dslbook} identifies three different patterns for combining
-functions to create a DSL: 1) \textit{function sequence}; 2) \textit{nested
-function}, and 3) \textit{method chaining}, which are illustrated in the snippets of
-Figure~\ref{fig:dsl-idioms}.
+DSLs for HTML provide an API where methods or functions correspond to the names
+of available HTML elements. These methods, also known as \textit{builders}, can
+be combined in a chain of calls to mimic the construction of an HTML document
+in a fluent manner. Martin Fowler\cite{dslbook} identifies three different
+patterns for combining functions to create a DSL: 1) \textit{function
+  sequence}; 2) \textit{nested function}, and 3) \textit{method chaining}, which
+are illustrated in the snippets of Figure~\ref{fig:dsl-idioms}.
 \begin{figure}[h]
@@
 
-From the three examples depicted in Figure~\ref{fig:dsl-idioms}, the \textit{nested
-function} idiom used by \textbf{ScalaTags}, as shown in the snippet of
-Figure~\ref{fig:dsl-idioms}.b), is the least verbose, requiring fewer statements than
-the other two cases.
-This approach of combining function calls is also utilized by \textbf{Hiccup}
-Clojure Library and \textbf{j2html} Java library.
-One reason for verbosity in \textit{function sequence} and \textit{method
-chaining} approaches is the requirement of a dedicated HTML builder to emit the
-closing tag, exemplified by \texttt{end()} in \textbf{JATL} (Figure~\ref{fig:dsl-idioms}.a) and
-\texttt{\_\_()} in \textbf{HtmlFlow} (Figure~\ref{fig:dsl-idioms}.c).
-But the \textit{nested function} approach comes with a notable drawback: \textbf{they
-do not support PSSR} because the sequence of nested functions is evaluated
-backward to the order in which they are written. 
-In other words, arguments are evaluated before the functions are invoked. 
-Taking the example in Figure~\ref{fig:dsl-idioms}.b), the \texttt{title()} function is
-first evaluated, and its resulting paragraph becomes the argument for the
-\texttt{head()} call, which, in turn, becomes the argument for \texttt{html()},
-and so on.
-If HTML is emitted as functions are called, it will print tags in reverse order. 
-The aforementioned DSLs must collect resulting nodes into an internal data
-structure, which is later traversed to produce the HTML output. Therefore, they
-cannot progressively emit the output as builders are called.
+From the three examples depicted in Figure~\ref{fig:dsl-idioms}, the
+\textit{nested function} idiom used by \textbf{ScalaTags}, as shown in the
+snippet of Figure~\ref{fig:dsl-idioms}.b), is the least verbose, requiring
+fewer statements than the other two cases. This approach of combining function
+calls is also utilized by \textbf{Hiccup} Clojure Library and \textbf{j2html}
+Java library. One reason for verbosity in \textit{function sequence} and
+\textit{method chaining} approaches is the requirement of a dedicated HTML
+builder to emit the closing tag, exemplified by \texttt{end()} in \textbf{JATL}
+(Figure~\ref{fig:dsl-idioms}.a) and \texttt{\_\_()} in \textbf{HtmlFlow}
+(Figure~\ref{fig:dsl-idioms}.c). But the \textit{nested function} approach
+comes with a notable drawback: \textbf{they do not support PSSR} because the
+sequence of nested functions is evaluated backward to the order in which they
+are written. In other words, arguments are evaluated before the functions are
+invoked. Taking the example in Figure~\ref{fig:dsl-idioms}.b), the
+\texttt{title()} function is first evaluated, and its resulting paragraph
+becomes the argument for the \texttt{head()} call, which, in turn, becomes the
+argument for \texttt{html()}, and so on. If HTML is emitted as functions are
+called, it will print tags in reverse order. The aforementioned DSLs must
+collect resulting nodes into an internal data structure, which is later
+traversed to produce the HTML output. Therefore, they cannot progressively emit
+the output as builders are called.
 
@@
 of arguments~\cite{Landin65}, expressed in lambda expressions (also known as
-function literals).
-Due to the concise form of expressing lambdas with brackets (i.e.,
-\texttt{\{\}}) in both Groovy and Kotlin, translating the template shown in
-Figure~\ref{fig:dsl-idioms}.b) to use Groovy MarkupBuilder or KotlinX.html only
-requires replacing parentheses with brackets for parent elements.
+function literals). Due to the concise form of expressing lambdas with brackets
+(i.e., \texttt{\{\}}) in both Groovy and Kotlin, translating the template shown
+in Figure~\ref{fig:dsl-idioms}.b) to use Groovy MarkupBuilder or KotlinX.html
+only requires replacing parentheses with brackets for parent elements.
 \textbf{HtmlFlow} also adopts this approach and provides a Kotlin-idiomatic API
@@
 \texttt{java.util.Stream}, also represent sequences of elements but are not
-compatible with \texttt{Iterable}.
-This fragmentation becomes more evident when considering other JVM languages, 
-such as Kotlin, which introduces additional abstractions like the 
-\texttt{Sequence} interface—also incompatible with \texttt{Iterable}. As a result, 
-template engines based on \textit{external} DSLs must explicitly support each of these 
-interfaces to accommodate the various iteration protocols across languages and 
-libraries.
+compatible with \texttt{Iterable}. This fragmentation becomes more evident when
+considering other JVM languages, such as Kotlin, which introduces additional
+abstractions like the \texttt{Sequence} interface—also incompatible with
+\texttt{Iterable}. As a result, template engines based on \textit{external}
+DSLs must explicitly support each of these interfaces to accommodate the
+various iteration protocols across languages and libraries.
 
@@
 programming language. This flexibility extends beyond the standard library to
-include third-party APIs as well.
-For example, in Java, developers can iterate over data using a traditional 
-\texttt{for} loop, the \texttt{java.util.stream} API, or external libraries such 
-as Guava, Vavr, StreamEx, and others—depending on their specific needs or preferences.
+include third-party APIs as well. For example, in Java, developers can iterate
+over data using a traditional \texttt{for} loop, the \texttt{java.util.stream}
+API, or external libraries such as Guava, Vavr, StreamEx, and others—depending
+on their specific needs or preferences.
 
@@
 
-Distinguish between template engines that support and do not support asynchronous data models.
+Distinguish between template engines that support and do not support
+asynchronous data models.
 
@@
 
-One of the reasons for legacy web templates not supporting asynchronous APIs
-is the absence of a unified standard calling convention for asynchronous calls.
+One of the reasons for legacy web templates not supporting asynchronous APIs is
+the absence of a unified standard calling convention for asynchronous calls.
 While there is a single, straightforward way to use a synchronous API with a
 direct style, where the result of a method call corresponds to its returned
-value, there is no equivalent standard in the asynchronous approach. 
-Instead, we may encounter various asynchronous conventions depending on the
-programming language and runtime environment.
-Some of these approaches include \emph{continuation-passing style}
-(CPS)~\cite{scheme}, \textit{promises}~\cite{promise} async/await
-idiom~\cite{async_await}, reactive streams \cite{ReactiveStreams}, Kotlin
-Flow~\cite{kotlinlang}, and others.
+value, there is no equivalent standard in the asynchronous approach. Instead,
+we may encounter various asynchronous conventions depending on the programming
+language and runtime environment. Some of these approaches include
+\emph{continuation-passing style} (CPS)~\cite{scheme},
+\textit{promises}~\cite{promise} async/await idiom~\cite{async_await}, reactive
+streams \cite{ReactiveStreams}, Kotlin Flow~\cite{kotlinlang}, and others.
 
@@
 routines to mimic the structure of synchronous ones, allowing developers to
-reason about instruction flow sequentially.
-The simplicity and broad adoption of this programming model have led to its
-incorporation into mainstream languages like C\#, JavaScript, Python, Perl,
-Swift, Kotlin, and others, excluding Java.
-However, implementing \texttt{async}/\texttt{a}wait requires compiler support to
-translate \textit{suspension points} (i.e., \texttt{await} statements) into state
-machines. Most template engines operate using an external DSL with their own
-templating dialect (e.g., Thymeleaf, JSP, Jade, Handlebars, and others), which do not
-inherently leverage asynchronous capabilities from their host GPLs.
+reason about instruction flow sequentially. The simplicity and broad adoption
+of this programming model have led to its incorporation into mainstream
+languages like C\#, JavaScript, Python, Perl, Swift, Kotlin, and others,
+excluding Java. However, implementing \texttt{async}/\texttt{a}wait requires
+compiler support to translate \textit{suspension points} (i.e., \texttt{await}
+statements) into state machines. Most template engines operate using an
+external DSL with their own templating dialect (e.g., Thymeleaf, JSP, Jade,
+Handlebars, and others), which do not inherently leverage asynchronous
+capabilities from their host GPLs.
 
@@
 
-Another key characteristic of such DSLs is \textbf{HTML safety}, which refers to
-whether they produce only valid HTML conforming to a well-formed document.
+Another key characteristic of such DSLs is \textbf{HTML safety}, which refers
+to whether they produce only valid HTML conforming to a well-formed document.
 To ensure HTML safety, a DSL API should only allow combining calls to builders
-that result in valid HTML. 
-However, most HTML DSLs using the \textit{function sequence} or \textit{nested
-function} approach cannot ensure HTML safety at compile time.
+that result in valid HTML. However, most HTML DSLs using the \textit{function
+  sequence} or \textit{nested function} approach cannot ensure HTML safety at
+compile time.
 
 The \textit{function sequence} approach, as illustrated in the snippet of
-Figure~\ref{fig:dsl-idioms}.a), involves combining function calls as a sequence of
-statements, making it challenging to restrict the order of statements. 
-In the \textit{nested function} approach shown in
-Figure~\ref{fig:dsl-idioms}.b), variable-length arguments are often used to
-allow an undefined number of child elements, which cannot be strongly typed in
-every programming language.
+Figure~\ref{fig:dsl-idioms}.a), involves combining function calls as a sequence
+of statements, making it challenging to restrict the order of statements. In
+the \textit{nested function} approach shown in Figure~\ref{fig:dsl-idioms}.b),
+variable-length arguments are often used to allow an undefined number of child
+elements, which cannot be strongly typed in every programming language.
 
 KotlinX.html, which utilizes the \textit{nested function} idiom with lazy
-evaluation, mitigates this issue through function types with a receiver. 
-In this approach, the receiver (i.e., \texttt{this} within the lambda) is
-strongly typed and provides a set of methods corresponding to legal child
-elements.
-This enables KotlinX.html to enforce HTML safety by restricting the available
+evaluation, mitigates this issue through function types with a receiver. In
+this approach, the receiver (i.e., \texttt{this} within the lambda) is strongly
+typed and provides a set of methods corresponding to legal child elements. This
+enables KotlinX.html to enforce HTML safety by restricting the available
 methods during compile time.
@@
 To achieve this KotlinX.html provides HTML builders using \textit{function
-literals with receiver}~\cite{kotlinlang}.
-In Kotlin, a block of code enclosed in curly braces \texttt{\{...\}} is known as
-a \emph{lambda}, and can be used as an argument to
-a function that expects a \emph{function literal}.
-When we write, for example, \texttt{body \{ div \{ hr() \} \}}, we are
-invoking the \texttt{body} function with a lambda as its argument. This
-lambda, in turn, calls the \texttt{div} function with another lambda as an
-argument that creates a horizontal row (i.e. \texttt{hr}).
-Each call to an HTML builder (e.g., \texttt{body}, \texttt{div}, \texttt{hr})
-creates the child element within the element generated by the outer function
-call.
+  literals with receiver}~\cite{kotlinlang}. In Kotlin, a block of code enclosed
+in curly braces \texttt{\{...\}} is known as a \emph{lambda}, and can be used
+as an argument to a function that expects a \emph{function literal}. When we
+write, for example, \texttt{body \{ div \{ hr() \} \}}, we are invoking the
+\texttt{body} function with a lambda as its argument. This lambda, in turn,
+calls the \texttt{div} function with another lambda as an argument that creates
+a horizontal row (i.e. \texttt{hr}). Each call to an HTML builder (e.g.,
+\texttt{body}, \texttt{div}, \texttt{hr}) creates the child element within the
+element generated by the outer function call.
 
-HtmlFlow provides two APIs: one in idiomatic Kotlin, similar to the \texttt{KotlinX.html} 
-API, and another that employs the \textit{method chaining} idiom, as illustrated 
-in the snippet of Figure~\ref{fig:dsl-idioms}.c).
-In this approach, the receiver object is implicitly passed as an argument to
-each method call, enabling subsequent methods to be invoked on the result of the
-preceding one.
-This facilitates the composition of methods, with each call building upon the
-other.
-Similar to KotlinX.html, HtmlFlow ensures HTML safety, restricting the available
-HTML builders and attributes after the dot (\texttt{.}) operator.
+HtmlFlow provides two APIs: one in idiomatic Kotlin, similar to the
+\texttt{KotlinX.html} API, and another that employs the \textit{method
+  chaining} idiom, as illustrated in the snippet of
+Figure~\ref{fig:dsl-idioms}.c). In this approach, the receiver object is
+implicitly passed as an argument to each method call, enabling subsequent
+methods to be invoked on the result of the preceding one. This facilitates the
+composition of methods, with each call building upon the other. Similar to
+KotlinX.html, HtmlFlow ensures HTML safety, restricting the available HTML
+builders and attributes after the dot (\texttt{.}) operator.
 
@@
 Pursuing the optimal user experience for web users has been a consistent goal
-since the inception of the World Wide Web.
-Consequently, various approaches have been explored to deliver the initial
-meaningful content to the end-user as promptly as possible, while the remaining
-content \textit{progressively} (or incrementally) loads as the server streams
-the HTML content.
+since the inception of the World Wide Web. Consequently, various approaches
+have been explored to deliver the initial meaningful content to the end-user as
+promptly as possible, while the remaining content \textit{progressively} (or
+incrementally) loads as the server streams the HTML content.
 
@@
 % fragments or partials to enable PSSR.
-An example of this limitation in SSR web templates
-was highlighted by Jeff Atwood in 2005, who criticized Microsoft ASP.NET for loading the
-entire web page into memory before sending any data to the browser~\cite{pssr2005}.
+An example of this limitation in SSR web templates was highlighted by Jeff
+Atwood in 2005, who criticized Microsoft ASP.NET for loading the entire web
+page into memory before sending any data to the browser~\cite{pssr2005}.
 % This practice delays the display of content to the user until the server
@@
 % HTML content from their inception.
-Despite historical critiques and HTML's inherent capabilities, most Web application
-frameworks, such as ASP.Net, Express.js, Spring, and others, 
-persistently lack support for progressive rendering, leading to the
-appearance of alternative techniques leveraging client-side JavaScript.
-Since 2007, various patents have addressed the PSSR issue, with Microsoft's
-patent enabling the infinite scrolling technique by displaying a single page of
+Despite historical critiques and HTML's inherent capabilities, most Web
+application frameworks, such as ASP.Net, Express.js, Spring, and others,
+persistently lack support for progressive rendering, leading to the appearance
+of alternative techniques leveraging client-side JavaScript. Since 2007,
+various patents have addressed the PSSR issue, with Microsoft's patent enabling
+the infinite scrolling technique by displaying a single page of
 results~\cite{scroll2007}, and Yahoo's patent focusing on differentiating
@@
 multiple data models and asynchronous APIs while ensuring well-formed HTML,
-PSSR, and non-blocking template resolution.
-That solution builds upon HtmlFlow, a Java DSL for HTML, with an internal
-processing mechanism that streams HTML in chunks as data items become available
-from data sources, without requiring client-side JavaScript.
-Like Thymeleaf, their proposal avoids the use of client-side JavaScript;
-however, unlike Thymeleaf, it supports a wide range of asynchronous APIs and
-multiple data sources, not limited to the \texttt{Publisher} API.
-However, the main counter-argument is the non-trivial management of the
-\emph{resume} callback in continuations~\cite{von2003events,callbackhell}, which
-is used to linearize the execution flow between asynchronous calls.
+PSSR, and non-blocking template resolution. That solution builds upon HtmlFlow,
+a Java DSL for HTML, with an internal processing mechanism that streams HTML in
+chunks as data items become available from data sources, without requiring
+client-side JavaScript. Like Thymeleaf, their proposal avoids the use of
+client-side JavaScript; however, unlike Thymeleaf, it supports a wide range of
+asynchronous APIs and multiple data sources, not limited to the
+\texttt{Publisher} API. However, the main counter-argument is the non-trivial
+management of the \emph{resume} callback in
+continuations~\cite{von2003events,callbackhell}, which is used to linearize the
+execution flow between asynchronous calls.

### 2025-05-23 docs/article/Sections/state-of-the-art.tex
Chapter 2: add Spring Webflux subsection

@@
 \subsubsection{Spring WebFlux}
+
+Spring WebFlux is a reactive web framework built on a non-blocking,
+event-driven architecture that enables efficient handling of a large number of
+concurrent connections with minimal resource usage. Unlike traditional
+servlet-based stacks that rely on a thread-per-request model, WebFlux operates
+on reactive runtimes like Netty, where a small, fixed-size event loop manages
+all I/O events. Incoming HTTP requests are routed to handler functions or
+annotated controllers, which return reactive types such as \texttt{Mono<T>} and
+\texttt{Flux<T>}—Publisher implementations defined by the Reactive Streams
+specification. A \texttt{Mono} represents a single asynchronous value (or
+none), while a \texttt{Flux} represents a stream of zero or more values. These
+abstractions allow the entire request lifecycle—from controller invocation to
+response rendering—to be orchestrated as a chain of non-blocking, asynchronous
+operations. I/O-bound tasks such as database queries or external API calls do
+not occupy threads during execution, enabling the framework to scale gracefully
+under high load and maintain responsiveness even in resource-constrained
+environments.
+
+Progressive Server-Side Rendering (PSSR) is natively supported in WebFlux via
+\texttt{Flux}-based controllers. When a controller returns a
+\texttt{Flux<String>} or another stream of HTML fragments, WebFlux can begin
+streaming content to the client as soon as the first elements are emitted. This
+allows browsers to incrementally render parts of the response without waiting
+for the entire payload, significantly reducing Time to First Byte (TTFB) and
+improving user-perceived performance. By integrating PSSR directly into the
+reactive processing pipeline, WebFlux facilitates responsive and resilient web
+applications well-suited for modern microservices architectures.
+
+An illustrative example of PSSR in WebFlux is shown below:
+
+\begin{lstlisting}[language=Java]
+@GetMapping(value = "/pssr", produces = MediaType.TEXT_HTML_VALUE)
+public Flux<String> renderChunks() {
+    return Flux.range(1, 10)
+               .delayElements(Duration.ofMillis(500))
+               .map(i -> "<p>Chunk " + i + "</p>\n");
+}
+\end{lstlisting}
+
+In this example, the method emits one HTML chunk every 500 milliseconds. As
+each chunk is generated, it is immediately written to the HTTP response,
+allowing the client to begin rendering content even as more data is still being
+produced. Unlike \texttt{StreamingResponseBody} in Spring MVC, which is
+constrained by servlet buffer thresholds and blocking semantics, WebFlux
+ensures that each emitted item is pushed to the client as soon as the network
+is ready, without occupying a dedicated thread. Moreover, WebFlux integrates
+seamlessly with reactive data sources such as R2DBC (Reactive Relational
+Database Connectivity) and reactive NoSQL drivers, making it possible to build
+end-to-end non-blocking applications. HTML builders and DSLs used in
+conjunction with WebFlux must be designed to accommodate reactive streams,
+ensuring that view generation does not violate the non-blocking contract by
+performing blocking I/O or awaiting asynchronous operations imperatively.
 

### 2025-05-23 docs/article/Sections/state-of-the-art.tex
Chapter 2: add Quarkus subsection
TODO: references for chapter 2 MVC, Webflux and Quarkus

@@
 \subsubsection{Quarkus}
+
+Quarkus is a modern, cloud-native Java framework that supports both imperative
+and reactive programming models. It is built on top of Vert.x,
+which provides a non-blocking, event-driven runtime similar to Netty. Incoming
+HTTP requests are processed asynchronously on a small, fixed-size event loop
+thread pool, allowing the framework to handle many concurrent connections
+efficiently without blocking threads. Quarkus supports Kotlin coroutines and
+reactive APIs such as Mutiny for asynchronous programming, enabling developers
+to write sequential-style code that runs in a non-blocking manner.
+
+For traditional imperative-style request handling, Quarkus provides integration
+with JAX-RS, where resource methods can return synchronous types or
+asynchronous constructs such as \texttt{CompletionStage}. To enable Progressive
+Server-Side Rendering (PSSR) in imperative endpoints, Quarkus leverages the
+JAX-RS \texttt{StreamingOutput} interface. By returning a
+\texttt{StreamingOutput} instance, developers can write parts of the HTTP
+response body incrementally as data becomes available. This approach allows the
+server to flush HTML fragments progressively to the client, improving the
+perceived responsiveness especially for long-running or large responses.
+However, like other streaming mechanisms based on servlet buffers, the initial
+data flush may be delayed until internal buffers (commonly around 8KB) are
+filled. However, Quarkus allows configuring the buffer size to a smaller value, which
+can help mitigate this issue.
+
+\begin{lstlisting}[language=java,caption={Quarkus example using StreamingOutput for progressive HTML streaming}]
+@GET
+@Path("/stream")
+@Produces(MediaType.TEXT_HTML)
+public StreamingOutput streamHtml() {
+    return output -> {
+        PrintWriter writer = new PrintWriter(output);
+        writer.println("<html><body>");
+        writer.flush(); // Flush initial HTML header
+
+        for (int i = 1; i <= 5; i++) {
+            writer.println("<p>Chunk " + i + "</p>");
+            writer.flush(); // Flush each chunk incrementally
+            Thread.sleep(500); // Simulate processing delay
+        }
+
+        writer.println("</body></html>");
+        writer.flush();
+    };
+}
+\end{lstlisting}
+
+This example demonstrates how \texttt{StreamingOutput} can be used to send
+partial HTML content in chunks, allowing browsers to progressively render the
+response while the server continues processing. Quarkus’ architecture ensures
+that the underlying event loop threads are not blocked by these writes, as the
+framework offloads blocking I/O operations to worker threads. This combination
+of non-blocking event-driven runtime and incremental response streaming enables
+efficient PSSR even within imperative programming styles.
 

### 2025-05-23 docs/article/Sections/problem-statement.tex
section 3 - Add missing biblio refs. Formatting listings. Clarify the onCompletion use in await of HtmlFlow and its behavior similar to resume in coroutines.

@@
 
-Reactive streams, such as those based on \texttt{Observable} (from RxJava) and
-\texttt{Flow} (from the Java 9+ standard library), are abstractions that enable
-progressive server-side rendering (PSSR) in environments utilizing
-\textit{internal} DSLs. These abstractions provide a non-blocking, asynchronous
-way to handle data streams, making them ideal for generating and sending HTML
-content incrementally as data becomes available. In PSSR, the server does not
+Reactive streams~\cite{ReactiveStreams}, such as those based on
+\texttt{Observable} (from RxJava)~\cite{rx-observable} and \texttt{Flow} (from
+the Java 9+ standard library~\footnote{~\url{https://openjdk.org/jeps/266}}),
+are abstractions that provide a non-blocking, asynchronous
+way to handle data streams.
+These abstractions support the incremental generation and transmission of HTML
+content as data becomes available, enabling \textit{progressive server-side
+rendering} (PSSR)~\cite{pssr2005}.
+In PSSR, the server does not
 wait for the entire data model to be ready before beginning to render HTML.
 Instead, it processes and streams each piece of data to the client as soon as
-it arrives. Reactive types like \texttt{Observable<T>} or \texttt{Flow<T>}
+it arrives.
+
+Reactive types like \texttt{Observable<T>} or \texttt{Flow<T>}
 facilitate this by representing data as a sequence of asynchronous events.
-These streams emit values over time and support backpressure, allowing the
+These streams emit values over time, allowing the
 rendering engine to consume data at a manageable rate without overwhelming
-system resources. A reactive stream might emit a sequence of
-\texttt{Presentation} objects, each representing a talk in a conference
-schedule. As each \texttt{Presentation} is emitted by the \texttt{Observable}
+system resources.
+For example, a reactive stream might emit a sequence of \texttt{Presentation}
+objects, each representing a talk in a conference schedule.
+As each \texttt{Presentation} is emitted by the \texttt{Observable}
 or \texttt{Flow}, the internal DSL-based engine—such as HtmlFlow—can render the
@@
 presentation is rendered asynchronously as it is emitted by an
-\texttt{Observable}, and in \autoref{lst:presentation-flow}, which shows an
-equivalent coroutine-based implementation using Kotlin's \texttt{Flow}. Both
-examples highlight how internal DSLs can natively integrate with reactive types
+\texttt{Observable}.
+Note that the \texttt{await} builder receives an additional parameter, the
+\texttt{onCompletion} callback, which is used to signal HtmlFlow that it can
+proceed to render the next HTML element in the web
+template~\cite{carvalho2023async}.
+HtmlFlow pauses the rendering process until \texttt{onCompletion} is called,
+similar to how the \emph{resume} function works in continuations and
+coroutines~\cite{coroutines_continuations}.
+In \autoref{lst:presentation-flow}, we show an
+equivalent suspend-based implementation using Kotlin's \texttt{Flow}~\cite{PSSR-WISE2024}.
+Both examples highlight how internal DSLs can natively integrate with reactive types
 to enable non-blocking, progressive rendering on the server side.
@@
 \lstset{style=listingstyle}
+\begin{center}
+\begin{minipage}{0.50\textwidth}
 \begin{lstlisting}[
@@
 ]
-await { div, model: Observable<Presentation>, onCompletion ->
+await { div, model, onCompletion ->
   model
-        .doOnNext { 
+    .doOnNext { presentation ->
       presentationFragmentAsync
-                .renderAsync(it)
+        .renderAsync(presentation)
         .thenApply { frag -> div.raw(frag) }
@@
 \end{lstlisting}
-\hfil
+\end{minipage}
+\hfill
+\begin{minipage}{0.46\textwidth}
 \begin{lstlisting}[
@@
 ]
-suspending { model: Flow<Presentation> ->
-    model.collect {
+suspending { model ->
+  model
+    .toFlowable()
+    .asFlow()
+    .collect { presentation ->
       presentationFragmentAsync
-            .renderAsync(it)
+        .renderAsync(presentation)
         .thenApply { frag -> raw(frag) }
@@
 \end{lstlisting}
+\end{minipage}
+\end{center}
 
@@
 \texttt{Iterable} or \texttt{List}. These interfaces require the entire data
-model to be materialized before rendering can begin, which blocks server
+model, to be materialized in memory before rendering can begin, which blocks server
 threads during template expansion and significantly limits scalability under
-high concurrency. These interfaces require the entire data model to be
-materialized before rendering can begin, which blocks server threads during
-template expansion and significantly limits scalability under high concurrency.
+high concurrency.
 Some reactive libraries, such as RxJava, provide bridging mechanisms like
 \texttt{Observable.blockingIterable()}, which allows asynchronous data sources
-to be exposed as Iterable by blocking the thread until new items are available.
+to be exposed as \texttt{Iterable} by blocking the thread until new items are available.
 While useful for compatibility with traditional APIs, this approach
@@
 
-\vspace{1cm}
-
 \lstset{style=listingstyle}
+\begin{center}
+\begin{minipage}{0.70\textwidth}
 \begin{lstlisting}[
@@
 blocking, synchronous rendering. However, this tradeoff becomes critical under
-high concurrency, where blocking threads severely degrades throughput. Emerging
-features in the Java ecosystem, particularly \textit{virtual threads}
-introduced in Java 21 as part of Project Loom, offer a promising solution to
+high concurrency, where blocking threads severely degrades throughput~\cite{PSSR-WISE2024}.
+Emerging features in the Java ecosystem, particularly \textit{virtual threads}
+introduced in Java 21 as part of Project Loom~\cite{Veen2024}, offer a promising solution to
 this challenge. Virtual threads drastically reduce the overhead of blocking
@@
 
-If successful, this approach would enable developers to reap the ergonomic
-benefits of external DSLs while maintaining scalable performance, without
-adopting complex asynchronous paradigms. It would also simplify the
-implementation of PSSR across a broader range of frameworks and engines, making
-the technique more accessible to mainstream web development.
+% If successful, this approach would enable developers to reap the ergonomic
+% benefits of external DSLs while maintaining scalable performance, without
+% adopting complex asynchronous paradigms. It would also simplify the
+% implementation of PSSR across a broader range of frameworks and engines, making
+% the technique more accessible to mainstream web development.

### 2025-05-23 docs/article/Sections/state-of-the-art.tex
Section 2 - Minor fixes. Add some biblio refs.

@@
 synchronous code while benefiting from the scalability of non-blocking I/O.
-
-\subsubsection{Spring MVC}
 
@@
 
-Explicar Observable e modelos reactivos
-
-Distinguish between template engines that support and do not support
-asynchronous data models.
-
-1. Nem se quer Existe API para iterar sobre um Async data model.
-E.g JStachio, .... apenas suportam interface Iterable. => Compilation Error.
-
-2. Consegue usar qq API e.g Kotlinx.Html and Groovy  => usa mal => produzir HTML out of order
-
-3. Sim suporta => well-formed HTML => Thymleaf e o HtmlFlow
-
 One of the reasons for legacy web templates not supporting asynchronous APIs is
@@
 page depending on client-side JavaScript.
-
-Introduce Thymeleaf
-
-Spring resolver does ot natively
-
+The Java Thymeleaf, the default SSR template engine for Spring web servers,
+added support for PSSR in 2018, through the use of a specific non-blocking
+Spring \texttt{ViewResolver} driver and without requiring client-side
+JavaScript.
 Carvalho~\cite{carvalho2023async} introduced an SSR solution that manages

### 2025-05-26 docs/article/Sections/benchmark.tex
Chapter 1: improve clarity, add references
Sections 2,4,5: formatting

@@
 \textit{view} and \textit{model} layers from the \textit{controller} layer
-\cite{Bucanek2009}, which allows for easy extension and integration of new
+~\cite{Bucanek2009}, which allows for easy extension and integration of new
 template engines and frameworks. It also includes a set of tests to ensure the
@@
 
-\vspace{0.5cm}
-
-\lstset{style=listingstyle}
-\begin{minipage}{0.5\textwidth}
+\begin{center}
+\begin{minipage}{0.48\textwidth}
   \begin{lstlisting}[
-  language=Kotlin,
+  language=kotlin,
   numbers=none,
@@
 \hfill
-\begin{minipage}{0.4\textwidth}
+\begin{minipage}{0.48\textwidth}
   \begin{lstlisting}[
-    language=Kotlin,
+    language=kotlin,
     basicstyle=\scriptsize\ttfamily,
@@
 \end{minipage}
+\end{center}
 
@@
   OutputStream without holding up the Servlet container thread}
-\cite{Spring-StreamingResponseBody}.
 

### 2025-05-26 docs/article/Sections/problem-statement.tex
Sections 2,4,5: formatting

@@
 
-\lstset{style=listingstyle}
 \begin{center}
@@
 reintroduces blocking behavior and undermines the benefits of non-blocking
-I/O—especially under high concurrency. \autoref{lst:presentation-jstachio}
+I/O—especially under high concurrency.\autoref{lst:presentation-jstachio}
 illustrates this model using a JStachio template, where the engine performs a

### 2025-05-26 docs/article/Sections/state-of-the-art.tex
Section 3: formatting, completed template engine table

@@
 
-Spring MVC is a classical, synchronous web framework built on the
+Spring MVC is a synchronous web framework built on the
 thread-per-request model. When an HTTP request is received, it is handled by a
@@
 data-intensive views. However, Spring MVC provides a mechanism for response
-streaming via the \texttt{StreamingResponseBody} interface, introduced in
+streaming via the \texttt{StreamingResponseBody}\footnote{https://docs.spring.io/spring-framework/docs/current/javadoc-api/org/springframework/web/servlet/mvc/method/annotation/StreamingResponseBody.html} interface, introduced in
 Spring 4.2. When a controller method returns a \texttt{StreamingResponseBody},
 Spring writes directly to the \texttt{HttpServletResponse} output stream,
-allowing developers to send parts of the response incrementally as they become
+allowing the server to send parts of the response incrementally as they become
 available. This is particularly useful for:
@@
 
-A regular usage might look like the following:
-\begin{lstlisting}[language=Java]
+\begin{lstlisting}[
+    language=Java,
+    basicstyle=\scriptsize\ttfamily,
+    numbers=none,
+    caption={StreamingResponseBody handler in Spring MVC},
+    label={lst:streaming-response-body}
+]
   @GetMapping("/stream")
@@
 
-In this example, HTML content is written and flushed in discrete chunks,
+In the example shown in \autoref{lst:streaming-response-body}, HTML content is written and flushed in discrete chunks,
 allowing the client to progressively render the response as it arrives, while
@@
 
-An illustrative example of PSSR in WebFlux is shown below:
-
-\begin{lstlisting}[language=Java]
+\begin{lstlisting}[
+    language=Java,
+    basicstyle=\scriptsize\ttfamily,
+    numbers=none,
+    caption={Progressive Server-Side Rendering in Spring WebFlux},
+    label={lst:pssr-webflux}
+]
 @GetMapping(value = "/pssr", produces = MediaType.TEXT_HTML_VALUE)
@@
 
-In this example, the method emits one HTML chunk every 500 milliseconds. As
+\autoref{lst:pssr-webflux} shows a method that emits one HTML chunk every 500 milliseconds. As
 each chunk is generated, it is immediately written to the HTTP response,
@@
 
-\begin{lstlisting}[language=java,caption={Quarkus example using StreamingOutput for progressive HTML streaming}]
+\begin{lstlisting}[
+  language=java,
+  basicstyle=\scriptsize\ttfamily,
+  numbers=none,
+  caption={Progressive Server-Side Rendering in Quarkus using StreamingOutput},
+  label={lst:pssr-quarkus}
+]
 @GET
@@
         writer.flush(); // Flush initial HTML header
-
         for (int i = 1; i <= 5; i++) {
@@
         }
-
         writer.println("</body></html>");
@@
 
-This example demonstrates how \texttt{StreamingOutput} can be used to send
+\autoref{lst:pssr-quarkus} shows how \texttt{StreamingOutput} can be used to send
 partial HTML content in chunks, allowing browsers to progressively render the
@@
      & External DSL
-     & \shortstack{\texttt{Iterable} \\\texttt{Publisher}}
+     & \shortstack{\texttt{Iterable} \\\texttt{Stream} \\\texttt{Publisher}}
      & \texttt{Publisher}$^{(1)}$
@@
      & External DSL
-     & \texttt{Iterable}
+     & \shortstack{\texttt{Iterable} \\\texttt{Sequence}}
      & \large{$\textcolor{red}{\times}$}
@@
 page depending on client-side JavaScript.
-The Java Thymeleaf, the default SSR template engine for Spring web servers,
+Java Thymeleaf, the default SSR template engine for Spring web servers,
 added support for PSSR in 2018, through the use of a specific non-blocking

### 2025-05-26 docs/article/Sections/benchmark.tex
article: formatting

@@
 content to the client. The implementation includes three main approaches to
-PSSR:
+PSSR\@:
 \begin{itemize}
@@
   \item \textbf{Suspendable}: The template engine is used in a suspendable
-        context, where the HTML content is rendered within the context of a suspending function
+        context, where the HTML content is rendered within the context of a suspending function.
         An example of this approach is the HtmlFlow template engine, which

### 2025-05-27 docs/article/Sections/state-of-the-art.tex
Clarify distinguish between Reactor and Mutiny and Reactive streams specification. Add biblio refs.

@@
 data-intensive views. However, Spring MVC provides a mechanism for response
-streaming via the \texttt{StreamingResponseBody}\footnote{https://docs.spring.io/spring-framework/docs/current/javadoc-api/org/springframework/web/servlet/mvc/method/annotation/StreamingResponseBody.html} interface, introduced in
+streaming via the \texttt{StreamingResponseBody}\footnote{\url{https://docs.spring.io/spring-framework/docs/current/javadoc-api/org/springframework/web/servlet/mvc/method/annotation/StreamingResponseBody.html}} interface, introduced in
 Spring 4.2. When a controller method returns a \texttt{StreamingResponseBody},
@@
 allowing the client to progressively render the response as it arrives, while
-the server continues processing additional data. Although
-\texttt{StreamingResponseBody} does not eliminate the blocking nature inherent
-in Spring MVC—since the handling thread remains active during the streaming
-process—it does enable a basic form of PSSR by facilitating incremental content
-delivery. This approach is particularly useful for long-running operations or
-large response bodies, where providing early partial output can enhance
-perceived performance.
-
-However, a key limitation we encountered with this method involves the servlet
-response buffer. By default, most servlet containers (e.g., Tomcat) use a
+the server continues processing additional data.
+However, this technique has two limitations. First, it does not eliminate the
+blocking nature inherent in Spring MVC, as the handling thread remains active
+during the streaming process.
+Second, it is constrained by the servlet response buffer.
+By default, most servlet containers (e.g., Tomcat) use a
 response buffer size of approximately 8KB. Data written to the response output
@@
 on reactive runtimes like Netty, where a small, fixed-size event loop manages
-all I/O events. Incoming HTTP requests are routed to handler functions or
-annotated controllers, which return reactive types such as \texttt{Mono<T>} and
-\texttt{Flux<T>}—Publisher implementations defined by the Reactive Streams
-specification. A \texttt{Mono} represents a single asynchronous value (or
-none), while a \texttt{Flux} represents a stream of zero or more values. These
+all I/O events.
+Incoming HTTP requests are routed to handler functions or annotated controllers,
+which return reactive types such as \texttt{Mono<T>} and
+\texttt{Flux<T>}—Publisher implementations provided by the Project Reactor
+library~\cite{projectreactor}. These implementations adhere to the Reactive
+Streams specification~\cite{ReactiveStreams}, enabling interoperability with
+other compliant reactive stream libraries.
+A \texttt{Mono} represents a single asynchronous value (or
+none)~\cite{promise}, while a \texttt{Flux} represents a stream of zero or more values~\cite{rx-observable}. These
 abstractions allow the entire request lifecycle—from controller invocation to
@@
 \texttt{Flux<String>} or another stream of HTML fragments, WebFlux can begin
-streaming content to the client as soon as the first elements are emitted. This
-allows browsers to incrementally render parts of the response without waiting
-for the entire payload, significantly reducing Time to First Byte (TTFB) and
-improving user-perceived performance. By integrating PSSR directly into the
-reactive processing pipeline, WebFlux facilitates responsive and resilient web
-applications well-suited for modern microservices architectures.
+streaming content to the client as soon as the first elements are emitted.
+% This allows browsers to incrementally render parts of the response without waiting
+% for the entire payload, significantly reducing Time to First Byte (TTFB) and
+% improving user-perceived performance. By integrating PSSR directly into the
+% reactive processing pipeline, WebFlux facilitates responsive and resilient web
+% applications well-suited for modern microservices architectures.
 
@@
 thread pool, allowing the framework to handle many concurrent connections
-efficiently without blocking threads. Quarkus supports Kotlin coroutines and
-reactive APIs such as Mutiny for asynchronous programming, enabling developers
-to write sequential-style code that runs in a non-blocking manner.
-
+efficiently without blocking threads. 
+Quarkus provides support for reactive APIs like Mutiny~\cite{mutiny2021}, a
+modern alternative to Spring Reactor~\cite{projectreactor}. Like Reactor,
+Mutiny complies with the Reactive Streams specification~\cite{ReactiveStreams}.
 For traditional imperative-style request handling, Quarkus provides integration

### 2025-05-27 docs/article/Sections/state-of-the-art-web-servers.tex
Refactor Section 2 in two files to enable easy reordering

--- 2025-05-27 docs/article/Sections/state-of-the-art.tex
@@
-\section{Background and Related Work}
-
-In this section, we first examine the different design choices adopted by web
-servers in their internal architectures, and how these choices impact the
-behavior of the web template engines. Then, in Subsection~2.2, we present the
-main properties that characterize each web template technology approach, along
-with the advantages and drawbacks that result from these characteristics.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%-------------------------------------------------------------------
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \subsection{Web Framework Architectures and Approaches to PSSR}
@@
 efficient PSSR even within imperative programming styles.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%-------------------------------------------------------------------
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{Web Templates}
-
-\textit{Web templates} have been the most widely adopted approach for
-constructing dynamic HTML pages.
-Web templates or \textit{web views}~\cite{Fowler02,Alur01} (e.g. JSP, Handlebars,
-or Thymeleaf), are based on HTML documents augmented with template-specific
-markers (e.g., \texttt{<\%>}, \texttt{\{\{\}\}}, or \texttt{\$\{\}}), which
-represent \textit{dynamic} information to be replaced at runtime with the
-results of corresponding computations, producing the final HTML page.
-The process of parsing and replacing these markers---i.e.,
-\textit{resolution}---is the primary responsibility of the \textit{template
-  engine}~\cite{Parr04}.
-One key characteristic of \textit{web templates} is their ability to receive a
-\textit{context object}---equivalent to the \textit{model} in the model-view
-design pattern~\cite{mvc88,Parr04}---which provides the data used to fill
-template placeholders at runtime.
-Web templates can be distinguished by several properties, namely:
-\begin{enumerate}
-  \item \textit{Domain-specific language} idiom
-  \item Supported data model APIs
-  \item Asynchronous support
-  \item Type safety and HTML safety
-  \item Progressive rendering
-\end{enumerate}
-
-Although some of the aforementioned characteristics apply to both
-\textit{server-side} and \textit{client-side} approaches, we focus solely on
-web template technologies for \textit{server-side rendering}, as our work is
-centered on that approach. Before getting deep on each of the aforementioned
-characteristics, Table~\ref{table:cmplibs} presents a breakdown of mainstream
-template engines, classified according to the identified properties.
-
-\begin{table}[h]
-  \small
-  \tabcolsep=0.1cm
-  \def\arraystretch{1.2}
-  \begin{tabular}{|c|c|c|c|c|c|c|}
-    \hline
-    \textbf{Library}
-     & \textbf{DSL idiom}
-     & \textbf{Data Model APIs}
-     & \shortstack{\textbf{Asynchronous}           \\\textbf{Support}}
-     & \shortstack{\textbf{Type}                   \\\textbf{Safety}}
-     & \shortstack{\textbf{HTML}                   \\\textbf{Safety}}
-     & \shortstack{\textbf{Progressive}            \\\textbf{Rendering}}
-    \\
-    \hline
-    \textbf{Freemarker}
-     & External DSL
-     & \texttt{Iterable}
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{PineGreen}{\checkmark}$}
-    \\
-    \hline
-    \textbf{JSP}
-     & External DSL
-     & \texttt{Iterable}
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{red}{\times}$}
-    \\
-    \hline
-    \textbf{JStachio}
-     & External DSL
-     & \texttt{Iterable}
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{PineGreen}{\checkmark}$}
-    \\\hline
-    \textbf{Pebble}
-     & External DSL
-     & \texttt{Iterable}
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{PineGreen}{\checkmark}$}
-    \\
-    \hline
-    \textbf{Rocker}
-     & External DSL
-     & \texttt{Iterable}
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{PineGreen}{\checkmark}$}
-    \\
-    \hline
-    \textbf{Thymeleaf}
-     & External DSL
-     & \shortstack{\texttt{Iterable} \\\texttt{Stream} \\\texttt{Publisher}}
-     & \texttt{Publisher}$^{(1)}$
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{PineGreen}{\checkmark}$}
-    \\
-    \hline
-    \textbf{Trimou}
-     & External DSL
-     & \texttt{Iterable}
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{PineGreen}{\checkmark}$}
-    \\
-    \hline
-    \textbf{Velocity}
-     & External DSL
-     & \shortstack{\texttt{Iterable} \\\texttt{Sequence}}
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{PineGreen}{\checkmark}$}
-    \\
-    \hline
-    \textbf{Clojure Hiccup}
-     & Nested Eager
-     & All
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{PineGreen}{\checkmark}$}
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{red}{\times}$}
-    \\
-    \hline
-    \shortstack{\textbf{Groovy } \\\textbf{MarkupBuilder}}
-     & Nested Lazy
-     & All
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{PineGreen}{\checkmark}$}
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{PineGreen}{\checkmark}$}
-    \\
-    \hline
-    \textbf{HtmlFlow}
-     & Chaining
-     & All
-     & \large{$\textcolor{PineGreen}{\checkmark}$}
-     & \large{$\textcolor{PineGreen}{\checkmark}$}
-     & \large{$\textcolor{PineGreen}{\checkmark}$}
-     & \large{$\textcolor{PineGreen}{\checkmark}$}
-    \\
-    \hline
-    \textbf{j2html}
-     & Nested Eager
-     & All
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{PineGreen}{\checkmark}$}
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{red}{\times}$}
-    \\
-    \hline
-    \textbf{KotlinX}
-     & Nested Lazy
-     & All
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{PineGreen}{\checkmark}$}
-     & \textcolor{PineGreen}{\checkmark}$^{(2)}$
-     & \large{$\textcolor{PineGreen}{\checkmark}$}
-    \\
-    \hline
-    \textbf{ScalaTags}
-     & Nested Eager
-     & All
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{PineGreen}{\checkmark}$}
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{red}{\times}$}
-    \\
-    \hline
-  \end{tabular}
-  \caption{
-    Comparison of web template technologies in the Java ecosystem.
-    \\$^{(1)}$ Limited to a single model per web template and compatible with
-    the reactive stream \texttt{Publisher}.
-    \\$^{(2)}$ Non-safety for HTML attributes.
-  }
-  \label{table:cmplibs}
-\end{table}
-
-The first half of Table~\ref{table:cmplibs} lists template engines that use
-their own templating dialects, referred to as \textit{external DSLs}. The
-second half lists Java libraries that provide an \textit{internal DSL},
-typically using a \textit{nested} or \textit{chaining} style to build HTML.
-Note that, by leveraging the host language (e.g., Clojure, Groovy, Java,
-Kotlin, or Scala) as the templating idiom, the latter impose no restrictions on
-the data model and fully support all styles of data access APIs. They also
-benefit from static type checking, which helps ensure type safety.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%-------------------------------------------------------------------
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsubsection{Templating idiom}
-
-Web templates are based on a \textit{domain-specific language}
-(DSL)~\cite{landin1966next}, which defines a language tailored to a specific
-\textit{domain}~\cite{evans2004domain}—in this case, HTML for expressing web
-documents. The DSL constrains the template's syntax and semantics to match the
-structure and purpose of HTML.
-
-DSLs can be divided in two types: \textit{external} or
-\textit{internal}\cite{dslbook}. \textit{External} DSLs are languages created
-without any affiliation to a concrete programming language. An example of an
-\textit{external} DSL is the regular expressions search
-pattern\cite{thompson1968}, since it defines its own syntax without any
-dependency of programming languages. On the other hand an \textit{internal} DSL
-is defined within a host programming language as a library and tends to be
-limited to the syntax of the host language, such as Java.
-JQuery\cite{resig2007pro} is one of the most well-known examples of an internal
-DSL in Javascript, designed to simplify HTML DOM\cite{dom} tree traversal and
-manipulation.
-
-Traditionally, web template technologies use an \textit{external} DSL to define
-control flow constructs and data binding primitives. Early web template engines
-such as JSP, ASP, Velocity, PHP, and others adopted this \textit{external} DSL
-approach as their templating dialect. For example, a \texttt{foreach} loop can
-be expressed in each technology using its own DSL: \\\texttt{<\% for(String
-  item : items) \%>} in JSP, \texttt{<\% For Each item In items \%>} in legacy
-ASP with VBScript, \texttt{\#foreach(\$item in \$items)} in Velocity, or
-\texttt{<?php foreach (\$items as \$item):?>} in PHP.
-
-On the other hand, \textit{internal} DSLs for HTML allow templates to be
-defined directly within the \emph{host} language (such as Java, Kotlin, Groovy,
-Scala, or other general-purpose programming languages), rather than using
-text-based template files~\cite{carvalho2019hot}. In this case, a web template
-is not limited to templating constructs but may instead leverage any available
-primitive of the host language or any external API.
-
-Using an internal DSL can have several benefits over using textual templates:
-\begin{enumerate}
-  \item \emph{Type safety}: Because the templates are defined with the host
-        programming language, the compiler can check the syntax and types of the
-        templates at compile time, which can help catch errors earlier in the
-        development process.
-
-  \item \emph{IDE support}: Many modern IDEs provide code completion, syntax
-        highlighting, and other features, which can make it easier to write and
-        maintain templates.
-
-  \item \emph{Flexibility}: Use all the features of the host programming language
-        to generate HTML, can make it easier to write complex templates and reuse code.
-
-  \item \emph{Integration}: Because the templates are defined in Java code, for
-        example, you can easily integrate them with other Java code in your
-        application, such as controllers, services, repositories and models.
-
-\end{enumerate}
-
-DSLs for HTML provide an API where methods or functions correspond to the names
-of available HTML elements. These methods, also known as \textit{builders}, can
-be combined in a chain of calls to mimic the construction of an HTML document
-in a fluent manner. Martin Fowler\cite{dslbook} identifies three different
-patterns for combining functions to create a DSL: 1) \textit{function
-  sequence}; 2) \textit{nested function}, and 3) \textit{method chaining}, which
-are illustrated in the snippets of Figure~\ref{fig:dsl-idioms}.
-\begin{figure}[h]
-  \centering
-
-  % Subfloat 1
-\begin{minipage}[c]{0.33\linewidth}
-  \centering
-  \begin{lstlisting}[
-  language=java,
-  basicstyle=\scriptsize\ttfamily,
-  numbers=none
-]
-html();
- head();
-  title();text("JT");end();
- end();
- body();
-  p();
-   text("Hi JATL");
-  end();
- end();
-end();
-  \end{lstlisting}
-    \caption*{(a) Function sequence}
-  \end{minipage}
-  \hfil
-  % Subfloat 2
-  \begin{minipage}[c]{0.23\linewidth}
-    \centering
-    \begin{lstlisting}[
-    language=java,
-    basicstyle=\scriptsize\ttfamily,
-    numbers=none
-  ]
-html(
- head(
-  title("ST")
- ),
- body(
-  p("Hi ScalaTags")
- )
-);
-  \end{lstlisting}
-    \caption*{(b) Nested function}
-  \end{minipage}
-  \hfil
-  % Subfloat 3
-  \begin{minipage}[c]{0.32\linewidth}
-    \centering
-    \begin{lstlisting}[
-    language=java,
-    basicstyle=\scriptsize\ttfamily,
-    numbers=none
-  ]
-html()
- .head()
-  .title().text("HF").__()
- .__()
- .body()
-  .p()
-   .text("Hi HtmlFlow")
-  .__()
- .__()
-.__();
-  \end{lstlisting}
-    \caption*{(c) Method chaining}
-  \end{minipage}
-
-  \caption{Utilizing DSL for HTML libraries with JATL, ScalaTags, and HtmlFlow.}
-  \label{fig:dsl-idioms}
-\end{figure}
-
-From the three examples depicted in Figure~\ref{fig:dsl-idioms}, the
-\textit{nested function} idiom used by \textbf{ScalaTags}, as shown in the
-snippet of Figure~\ref{fig:dsl-idioms}.b), is the least verbose, requiring
-fewer statements than the other two cases. This approach of combining function
-calls is also utilized by \textbf{Hiccup} Clojure Library and \textbf{j2html}
-Java library. One reason for verbosity in \textit{function sequence} and
-\textit{method chaining} approaches is the requirement of a dedicated HTML
-builder to emit the closing tag, exemplified by \texttt{end()} in \textbf{JATL}
-(Figure~\ref{fig:dsl-idioms}.a) and \texttt{\_\_()} in \textbf{HtmlFlow}
-(Figure~\ref{fig:dsl-idioms}.c). But the \textit{nested function} approach
-comes with a notable drawback: \textbf{they do not support PSSR} because the
-sequence of nested functions is evaluated backward to the order in which they
-are written. In other words, arguments are evaluated before the functions are
-invoked. Taking the example in Figure~\ref{fig:dsl-idioms}.b), the
-\texttt{title()} function is first evaluated, and its resulting paragraph
-becomes the argument for the \texttt{head()} call, which, in turn, becomes the
-argument for \texttt{html()}, and so on. If HTML is emitted as functions are
-called, it will print tags in reverse order. The aforementioned DSLs must
-collect resulting nodes into an internal data structure, which is later
-traversed to produce the HTML output. Therefore, they cannot progressively emit
-the output as builders are called.
-
-Two other JVM libraries, \textbf{Groovy MarkupBuilder} and
-\textbf{KotlinX.html}, also adopt a \textit{nested function} approach, but they
-address the backward evaluation issue by implementing \textit{lazy evaluation}
-of arguments~\cite{Landin65}, expressed in lambda expressions (also known as
-function literals). Due to the concise form of expressing lambdas with brackets
-(i.e., \texttt{\{\}}) in both Groovy and Kotlin, translating the template shown
-in Figure~\ref{fig:dsl-idioms}.b) to use Groovy MarkupBuilder or KotlinX.html
-only requires replacing parentheses with brackets for parent elements.
-\textbf{HtmlFlow} also adopts this approach and provides a Kotlin-idiomatic API
-as an alternative to its original Java API.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%-------------------------------------------------------------------
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsubsection{Supported data model APIs}
-
-Template engines with dedicated templating dialects typically rely on specific
-control flow constructs—such as a \texttt{foreach}-like statement—to iterate
-over elements in a data source. In the Java ecosystem, \texttt{Iterable} is the
-common supertype implemented by most data structures and serves as the standard
-API for iteration. However, other standard interfaces, such as
-\texttt{java.util.Stream}, also represent sequences of elements but are not
-compatible with \texttt{Iterable}. This fragmentation becomes more evident when
-considering other JVM languages, such as Kotlin, which introduces additional
-abstractions like the \texttt{Sequence} interface—also incompatible with
-\texttt{Iterable}. As a result, template engines based on \textit{external}
-DSLs must explicitly support each of these interfaces to accommodate the
-various iteration protocols across languages and libraries.
-
-On the other hand, template engines that employ an \textit{internal} DSL do not
-face such limitations, as they can leverage any available API within the host
-programming language. This flexibility extends beyond the standard library to
-include third-party APIs as well. For example, in Java, developers can iterate
-over data using a traditional \texttt{for} loop, the \texttt{java.util.stream}
-API, or external libraries such as Guava, Vavr, StreamEx, and others—depending
-on their specific needs or preferences.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%-------------------------------------------------------------------
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsubsection{Asynchronous support}
-\label{sec:async-support}
-
-One of the reasons for legacy web templates not supporting asynchronous APIs is
-the absence of a unified standard calling convention for asynchronous calls.
-While there is a single, straightforward way to use a synchronous API with a
-direct style, where the result of a method call corresponds to its returned
-value, there is no equivalent standard in the asynchronous approach. Instead,
-we may encounter various asynchronous conventions depending on the programming
-language and runtime environment. Some of these approaches include
-\emph{continuation-passing style} (CPS)~\cite{scheme},
-\textit{promises}~\cite{promise} async/await idiom~\cite{async_await}, reactive
-streams \cite{ReactiveStreams}, Kotlin Flow~\cite{kotlinlang}, and others.
-
-Many \textit{general-purpose languages} (GPLs) have embraced the
-\texttt{async}/\texttt{await} feature~\cite{async_await} enabling non-blocking
-routines to mimic the structure of synchronous ones, allowing developers to
-reason about instruction flow sequentially. The simplicity and broad adoption
-of this programming model have led to its incorporation into mainstream
-languages like C\#, JavaScript, Python, Perl, Swift, Kotlin, and others,
-excluding Java. However, implementing \texttt{async}/\texttt{a}wait requires
-compiler support to translate \textit{suspension points} (i.e., \texttt{await}
-statements) into state machines. Most template engines operate using an
-external DSL with their own templating dialect (e.g., Thymeleaf, JSP, Jade,
-Handlebars, and others), which do not inherently leverage asynchronous
-capabilities from their host GPLs.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%-------------------------------------------------------------------
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsubsection{Type safety and HTML safety}
-
-Another key characteristic of such DSLs is \textbf{HTML safety}, which refers
-to whether they produce only valid HTML conforming to a well-formed document.
-To ensure HTML safety, a DSL API should only allow combining calls to builders
-that result in valid HTML. However, most HTML DSLs using the \textit{function
-  sequence} or \textit{nested function} approach cannot ensure HTML safety at
-compile time.
-
-The \textit{function sequence} approach, as illustrated in the snippet of
-Figure~\ref{fig:dsl-idioms}.a), involves combining function calls as a sequence
-of statements, making it challenging to restrict the order of statements. In
-the \textit{nested function} approach shown in Figure~\ref{fig:dsl-idioms}.b),
-variable-length arguments are often used to allow an undefined number of child
-elements, which cannot be strongly typed in every programming language.
-
-KotlinX.html, which utilizes the \textit{nested function} idiom with lazy
-evaluation, mitigates this issue through function types with a receiver. In
-this approach, the receiver (i.e., \texttt{this} within the lambda) is strongly
-typed and provides a set of methods corresponding to legal child elements. This
-enables KotlinX.html to enforce HTML safety by restricting the available
-methods during compile time.
-
-To achieve this KotlinX.html provides HTML builders using \textit{function
-  literals with receiver}~\cite{kotlinlang}. In Kotlin, a block of code enclosed
-in curly braces \texttt{\{...\}} is known as a \emph{lambda}, and can be used
-as an argument to a function that expects a \emph{function literal}. When we
-write, for example, \texttt{body \{ div \{ hr() \} \}}, we are invoking the
-\texttt{body} function with a lambda as its argument. This lambda, in turn,
-calls the \texttt{div} function with another lambda as an argument that creates
-a horizontal row (i.e. \texttt{hr}). Each call to an HTML builder (e.g.,
-\texttt{body}, \texttt{div}, \texttt{hr}) creates the child element within the
-element generated by the outer function call.
-
-HtmlFlow provides two APIs: one in idiomatic Kotlin, similar to the
-\texttt{KotlinX.html} API, and another that employs the \textit{method
-  chaining} idiom, as illustrated in the snippet of
-Figure~\ref{fig:dsl-idioms}.c). In this approach, the receiver object is
-implicitly passed as an argument to each method call, enabling subsequent
-methods to be invoked on the result of the preceding one. This facilitates the
-composition of methods, with each call building upon the other. Similar to
-KotlinX.html, HtmlFlow ensures HTML safety, restricting the available HTML
-builders and attributes after the dot (\texttt{.}) operator.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%-------------------------------------------------------------------
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsubsection{Progressive rendering}
-
-Pursuing the optimal user experience for web users has been a consistent goal
-since the inception of the World Wide Web. Consequently, various approaches
-have been explored to deliver the initial meaningful content to the end-user as
-promptly as possible, while the remaining content \textit{progressively} (or
-incrementally) loads as the server streams the HTML content.
-
-\textit{\textbf{Progressive rendering}} and \textit{\textbf{progressive loading}}
-encompass different concepts. The former pertains to the \textit{dynamic} content of a
-\textit{dynamic web page}, encompassing elements with logic and placeholders
-that are fulfilled by data from an object \textit{model} constructed at runtime.
-On the other hand, the latter is associated with \textit{render-blocking
-  resources} such as scripts, stylesheets, and HTML imports, which may hinder the
-browser from rendering page content to the screen~\cite{progressive-2022}.
-The notion of \textit{progressive rendering} has typically been aligned with
-\textit{client-side rendering} (CSR), where a single HTML page with static
-content is delivered upfront, while the dynamic content is fetched as the data
-becomes available to complete the web page.
-However, despite being overlooked, HTTP and browsers were designed from their
-inception to also support this feature in the context of SSR approaches,
-which offers the advantage of not being dependent on
-\textit{rendering-blocking resources}, such as the required JavaScript for CSR.
-% \textit{Progressive server-side rendering} (PSSR) is the
-% ability of streaming HTML content to the client in chunks or data frames, as it
-% gets resolved on the server at runtime.
-% This lets the user-interface to be rendered incrementally (i.e.
-% \textit{progressively}) by the browser in line with the availability of the data.
-% However, most SSR template engines do not inherently support this feature at the
-% template level, requiring developers to manually break down web pages into
-% fragments or partials to enable PSSR.
-An example of this limitation in SSR web templates was highlighted by Jeff
-Atwood in 2005, who criticized Microsoft ASP.NET for loading the entire web
-page into memory before sending any data to the browser~\cite{pssr2005}.
-% This practice delays the display of content to the user until the server
-% completes its processing, contradicting the original design intent of HTML to
-% render progressively as content is received.
-% Notably, both Netscape and Internet Explorer were capable of rendering partial
-% HTML content from their inception.
-Despite historical critiques and HTML's inherent capabilities, most Web
-application frameworks, such as ASP.Net, Express.js, Spring, and others,
-persistently lack support for progressive rendering, leading to the appearance
-of alternative techniques leveraging client-side JavaScript. Since 2007,
-various patents have addressed the PSSR issue, with Microsoft's patent enabling
-the infinite scrolling technique by displaying a single page of
-results~\cite{scroll2007}, and Yahoo's patent focusing on differentiating
-elements based on their position relative to the visible
-portion~\cite{schiller2007progressive}.
-
-Former techniques all describe methods of progressively adding content to a web
-page depending on client-side JavaScript.
-Java Thymeleaf, the default SSR template engine for Spring web servers,
-added support for PSSR in 2018, through the use of a specific non-blocking
-Spring \texttt{ViewResolver} driver and without requiring client-side
-JavaScript.
-Carvalho~\cite{carvalho2023async} introduced an SSR solution that manages
-multiple data models and asynchronous APIs while ensuring well-formed HTML,
-PSSR, and non-blocking template resolution. That solution builds upon HtmlFlow,
-a Java DSL for HTML, with an internal processing mechanism that streams HTML in
-chunks as data items become available from data sources, without requiring
-client-side JavaScript. Like Thymeleaf, their proposal avoids the use of
-client-side JavaScript; however, unlike Thymeleaf, it supports a wide range of
-asynchronous APIs and multiple data sources, not limited to the
-\texttt{Publisher} API. However, the main counter-argument is the non-trivial
-management of the \emph{resume} callback in
-continuations~\cite{von2003events,callbackhell}, which is used to linearize the
-execution flow between asynchronous calls.

### 2025-05-27 docs/article/Sections/state-of-the-art-web-templates.tex
Refactor Section 2 in two files to enable easy reordering

--- 2025-05-26 docs/article/Sections/state-of-the-art.tex
@@
-\section{Background and Related Work}
-
-In this section, we first examine the different design choices adopted by web
-servers in their internal architectures, and how these choices impact the
-behavior of the web template engines. Then, in Subsection~2.2, we present the
-main properties that characterize each web template technology approach, along
-with the advantages and drawbacks that result from these characteristics.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%-------------------------------------------------------------------
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{Web Framework Architectures and Approaches to PSSR}
-
-In traditional thread-per-request architectures, each incoming request is
-handled by a dedicated thread. The web server maintains a thread pool from
-which a thread is assigned to handle each request. As the load increases, the
-number of active threads can grow rapidly, potentially exhausting the pool.
-This can lead to performance and scalability issues, as the system may become
-bogged down by context switching and thread management
-overhead~\cite{kant2000scalable}. In the Java ecosystem, Spring Boot is one of
-the most widely used frameworks that follow this model, according to several
-reports such as the Stack Overflow 2024 Developer
-Survey~\footnote{~\url{https://survey.stackoverflow.co/2024/technology\#1-web-frameworks-and-technologies}}.
-
-On the other hand, in modern \textit{low-thread} architectures, the server uses
-a small number of threads to handle a large number of requests.
-\textit{Low-thread} servers, also known as
-\textit{event-driven}~\cite{event-driven-servers}, offer a significant
-advantage in efficiently managing a high number of concurrent I/O operations
-with minimal resource usage.
-
-The \textit{non-blocking} I/O model employed in low-thread servers is
-well-suited for handling large volumes of data asynchronously~\cite{Meijer12}.
-This combination of low-thread servers and asynchronous data models has
-facilitated the development of highly scalable, responsive, and resilient web
-applications capable of managing substantial data loads~\cite{Jin15}. The
-prominence of this concept increased with the advent of Node.js in 2009, and
-subsequently, various technologies adopted this approach in the Java ecosystem,
-including Netty, Akka HTTP, Vert.X, and Spring WebFlux.
-% Among these, Spring WebFlux stands out as the most widely used web framework in
-% Java, according to surveys like JetBrains' State of Java report (2021) and
-% community metrics from platforms like Github and StackOverflow.
-The \textit{non-blocking} I/O model in low-thread servers functions optimally
-only when HTTP handlers avoid blocking. Therefore, HTML templates need to be
-proficient in dealing with the asynchronous APIs provided by data models. While
-most legacy web templates struggle with asynchronous models, DSLs for HTML face
-no such limitations, leveraging all constructions available in the host
-programming language. However, the unexpected intertwining of asynchronous
-handlers' completion and HTML builders' execution may potentially lead to
-malformed HTML with an unexpected layout, as demonstrated in the subsequent
-subsection~\ref{sec:async-support}.
-
-An alternative approach involves utilizing user-level threads, while
-maintaining a blocking I/O and a synchronous programming paradigm. However,
-this approach still requires a user-level I/O subsystem capable of mitigating
-system-level blocking, which is crucial for the performance of I/O-intensive
-applications. This technique offers a lightweight solution for efficiently
-managing a larger number of concurrent sessions by minimizing per-thread
-overhead. In 2020, Karsten~\cite{karsten2020} demonstrated how this strategy
-supports a synchronous programming style, effectively abstracting away the
-complexities associated with managing continuations in asynchronous
-programming.
-
-Among mainstream technologies, the Kotlin programming language introduces a new
-abstraction for managing coroutines and provides structured concurrency, which
-ensures that coroutines are \textit{scoped}, \textit{cancellable}, and
-\textit{coordinated} within a well-defined
-lifecycle~\cite{elizarov2021coroutines}. Although this model embraces the
-\texttt{async}/\texttt{await} feature~\cite{async_await}, enabling non-blocking
-routines to mimic the structure of synchronous ones, it still lacks a
-user-level I/O subsystem and relies on an explicit I/O dispatcher tied to a
-dedicated thread pool that frees worker threads for handling processing tasks.
-Moreover, web templates using their own templating dialects, such as JSP,
-Thymeleaf, or Handlebars, are unable to take advantage of such constructs, as
-they typically support only a limited subset of the host library's API—most
-commonly just the \texttt{Iterable} interface.
-
-Only Java virtual threads, introduced in
-JEP~444~\footnote{\url{https://openjdk.org/jeps/444}}, adopt a similar approach
-to Karsten's proposal, using user-mode threads to preserve a synchronous
-programming model while still allowing blocking I/O operations without tying up
-platform threads. When a virtual thread performs a blocking I/O operation, the
-JVM intercepts the call and transparently parks the virtual thread, freeing the
-underlying platform thread to perform other tasks. Once the I/O operation
-completes, the virtual thread is rescheduled on a platform thread and resumes
-execution. This mechanism is enabled by the JVM's integration with non-blocking
-system calls at the OS level, allowing developers to write traditional
-synchronous code while benefiting from the scalability of non-blocking I/O.
-
-\subsubsection{Spring MVC}
-
-Spring MVC is a synchronous web framework built on the
-thread-per-request model. When an HTTP request is received, it is handled by a
-dedicated thread from the server’s thread pool (e.g., Tomcat, Jetty, or
-Undertow). This thread is responsible for executing the full request lifecycle
-synchronously, including invoking controllers, executing business logic, and
-rendering the response view. Because the model is blocking, the thread remains
-occupied throughout the request—even during I/O operations such as database
-queries or external API calls—which can lead to performance bottlenecks under
-high concurrency.
-
-Controllers in Spring MVC typically return Java objects or models that are
-processed by view resolvers in conjunction with template engines like JSP,
-Thymeleaf, or FreeMarker. These templates generate the full HTML response in
-one pass, which is then sent to the client only after all necessary data has
-been gathered and rendered. By default, this architecture limits the ability to
-perform \textit{Progressive Server-Side Rendering (PSSR)}, as traditional
-rendering waits for all data before sending any output. This results in higher
-latency for clients and less responsive page loads, particularly for
-data-intensive views. However, Spring MVC provides a mechanism for response
-streaming via the \texttt{StreamingResponseBody}\footnote{https://docs.spring.io/spring-framework/docs/current/javadoc-api/org/springframework/web/servlet/mvc/method/annotation/StreamingResponseBody.html} interface, introduced in
-Spring 4.2. When a controller method returns a \texttt{StreamingResponseBody},
-Spring writes directly to the \texttt{HttpServletResponse} output stream,
-allowing the server to send parts of the response incrementally as they become
-available. This is particularly useful for:
-\begin{itemize}
-  \item Sending large responses without buffering the entire output in memory.
-  \item Writing dynamic HTML in fragments as data is fetched or computed.
-  \item Reducing Time to First Byte (TTFB) and improving perceived performance.
-\end{itemize}
-
-\begin{lstlisting}[
-    language=Java,
-    basicstyle=\scriptsize\ttfamily,
-    numbers=none,
-    caption={StreamingResponseBody handler in Spring MVC},
-    label={lst:streaming-response-body}
-]
-  @GetMapping("/stream")
-  fun handleStream(): StreamingResponseBody? {
-      return StreamingResponseBody { outputStream ->
-          for (i in 0..9) {
-              val htmlFragment = "<p>Chunk " + i + "</p>\n"
-              outputStream.write(htmlFragment.toByteArray())
-              outputStream.flush() // ensure partial response is sent
-              Thread.sleep(500) // simulate delay
-          }
-      }
-  }
-\end{lstlisting}
-
-In the example shown in \autoref{lst:streaming-response-body}, HTML content is written and flushed in discrete chunks,
-allowing the client to progressively render the response as it arrives, while
-the server continues processing additional data. Although
-\texttt{StreamingResponseBody} does not eliminate the blocking nature inherent
-in Spring MVC—since the handling thread remains active during the streaming
-process—it does enable a basic form of PSSR by facilitating incremental content
-delivery. This approach is particularly useful for long-running operations or
-large response bodies, where providing early partial output can enhance
-perceived performance.
-
-However, a key limitation we encountered with this method involves the servlet
-response buffer. By default, most servlet containers (e.g., Tomcat) use a
-response buffer size of approximately 8KB. Data written to the response output
-stream is held in this buffer until it reaches capacity. Consequently, if the
-total size of the initial HTML content is less than the buffer threshold, the
-server will not transmit any data to the client until the buffer is filled.
-This buffering behavior introduces a delay in sending the first chunk of
-content, thereby negating the benefits of progressive rendering. In scenarios
-where the rendered HTML template does not produce sufficient data to exceed the
-buffer threshold early, PSSR becomes ineffective, and the client experiences a
-delayed initial response.
-
-\subsubsection{Spring WebFlux}
-
-Spring WebFlux is a reactive web framework built on a non-blocking,
-event-driven architecture that enables efficient handling of a large number of
-concurrent connections with minimal resource usage. Unlike traditional
-servlet-based stacks that rely on a thread-per-request model, WebFlux operates
-on reactive runtimes like Netty, where a small, fixed-size event loop manages
-all I/O events. Incoming HTTP requests are routed to handler functions or
-annotated controllers, which return reactive types such as \texttt{Mono<T>} and
-\texttt{Flux<T>}—Publisher implementations defined by the Reactive Streams
-specification. A \texttt{Mono} represents a single asynchronous value (or
-none), while a \texttt{Flux} represents a stream of zero or more values. These
-abstractions allow the entire request lifecycle—from controller invocation to
-response rendering—to be orchestrated as a chain of non-blocking, asynchronous
-operations. I/O-bound tasks such as database queries or external API calls do
-not occupy threads during execution, enabling the framework to scale gracefully
-under high load and maintain responsiveness even in resource-constrained
-environments.
-
-Progressive Server-Side Rendering (PSSR) is natively supported in WebFlux via
-\texttt{Flux}-based controllers. When a controller returns a
-\texttt{Flux<String>} or another stream of HTML fragments, WebFlux can begin
-streaming content to the client as soon as the first elements are emitted. This
-allows browsers to incrementally render parts of the response without waiting
-for the entire payload, significantly reducing Time to First Byte (TTFB) and
-improving user-perceived performance. By integrating PSSR directly into the
-reactive processing pipeline, WebFlux facilitates responsive and resilient web
-applications well-suited for modern microservices architectures.
-
-\begin{lstlisting}[
-    language=Java,
-    basicstyle=\scriptsize\ttfamily,
-    numbers=none,
-    caption={Progressive Server-Side Rendering in Spring WebFlux},
-    label={lst:pssr-webflux}
-]
-@GetMapping(value = "/pssr", produces = MediaType.TEXT_HTML_VALUE)
-public Flux<String> renderChunks() {
-    return Flux.range(1, 10)
-               .delayElements(Duration.ofMillis(500))
-               .map(i -> "<p>Chunk " + i + "</p>\n");
-}
-\end{lstlisting}
-
-\autoref{lst:pssr-webflux} shows a method that emits one HTML chunk every 500 milliseconds. As
-each chunk is generated, it is immediately written to the HTTP response,
-allowing the client to begin rendering content even as more data is still being
-produced. Unlike \texttt{StreamingResponseBody} in Spring MVC, which is
-constrained by servlet buffer thresholds and blocking semantics, WebFlux
-ensures that each emitted item is pushed to the client as soon as the network
-is ready, without occupying a dedicated thread. Moreover, WebFlux integrates
-seamlessly with reactive data sources such as R2DBC (Reactive Relational
-Database Connectivity) and reactive NoSQL drivers, making it possible to build
-end-to-end non-blocking applications. HTML builders and DSLs used in
-conjunction with WebFlux must be designed to accommodate reactive streams,
-ensuring that view generation does not violate the non-blocking contract by
-performing blocking I/O or awaiting asynchronous operations imperatively.
-
-\subsubsection{Quarkus}
-
-Quarkus is a modern, cloud-native Java framework that supports both imperative
-and reactive programming models. It is built on top of Vert.x,
-which provides a non-blocking, event-driven runtime similar to Netty. Incoming
-HTTP requests are processed asynchronously on a small, fixed-size event loop
-thread pool, allowing the framework to handle many concurrent connections
-efficiently without blocking threads. Quarkus supports Kotlin coroutines and
-reactive APIs such as Mutiny for asynchronous programming, enabling developers
-to write sequential-style code that runs in a non-blocking manner.
-
-For traditional imperative-style request handling, Quarkus provides integration
-with JAX-RS, where resource methods can return synchronous types or
-asynchronous constructs such as \texttt{CompletionStage}. To enable Progressive
-Server-Side Rendering (PSSR) in imperative endpoints, Quarkus leverages the
-JAX-RS \texttt{StreamingOutput} interface. By returning a
-\texttt{StreamingOutput} instance, developers can write parts of the HTTP
-response body incrementally as data becomes available. This approach allows the
-server to flush HTML fragments progressively to the client, improving the
-perceived responsiveness especially for long-running or large responses.
-However, like other streaming mechanisms based on servlet buffers, the initial
-data flush may be delayed until internal buffers (commonly around 8KB) are
-filled. However, Quarkus allows configuring the buffer size to a smaller value, which
-can help mitigate this issue.
-
-\begin{lstlisting}[
-  language=java,
-  basicstyle=\scriptsize\ttfamily,
-  numbers=none,
-  caption={Progressive Server-Side Rendering in Quarkus using StreamingOutput},
-  label={lst:pssr-quarkus}
-]
-@GET
-@Path("/stream")
-@Produces(MediaType.TEXT_HTML)
-public StreamingOutput streamHtml() {
-    return output -> {
-        PrintWriter writer = new PrintWriter(output);
-        writer.println("<html><body>");
-        writer.flush(); // Flush initial HTML header
-        for (int i = 1; i <= 5; i++) {
-            writer.println("<p>Chunk " + i + "</p>");
-            writer.flush(); // Flush each chunk incrementally
-            Thread.sleep(500); // Simulate processing delay
-        }
-        writer.println("</body></html>");
-        writer.flush();
-    };
-}
-\end{lstlisting}
-
-\autoref{lst:pssr-quarkus} shows how \texttt{StreamingOutput} can be used to send
-partial HTML content in chunks, allowing browsers to progressively render the
-response while the server continues processing. Quarkus’ architecture ensures
-that the underlying event loop threads are not blocked by these writes, as the
-framework offloads blocking I/O operations to worker threads. This combination
-of non-blocking event-driven runtime and incremental response streaming enables
-efficient PSSR even within imperative programming styles.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%-------------------------------------------------------------------
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \subsection{Web Templates}

### 2025-05-27 docs/article/Sections/state-of-the-art.tex
Refactor Section 2 in two files to enable easy reordering

--- 2025-05-23 docs/article/Sections/state-of-the-art.tex
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%-------------------------------------------------------------------
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{Web Framework Architectures and Approaches to PSSR}
-
-In traditional thread-per-request architectures, each incoming request is
-handled by a dedicated thread. The web server maintains a thread pool from
-which a thread is assigned to handle each request. As the load increases, the
-number of active threads can grow rapidly, potentially exhausting the pool.
-This can lead to performance and scalability issues, as the system may become
-bogged down by context switching and thread management
-overhead~\cite{kant2000scalable}. In the Java ecosystem, Spring Boot is one of
-the most widely used frameworks that follow this model, according to several
-reports such as the Stack Overflow 2024 Developer
-Survey~\footnote{~\url{https://survey.stackoverflow.co/2024/technology\#1-web-frameworks-and-technologies}}.
-
-On the other hand, in modern \textit{low-thread} architectures, the server uses
-a small number of threads to handle a large number of requests.
-\textit{Low-thread} servers, also known as
-\textit{event-driven}~\cite{event-driven-servers}, offer a significant
-advantage in efficiently managing a high number of concurrent I/O operations
-with minimal resource usage.
-
-The \textit{non-blocking} I/O model employed in low-thread servers is
-well-suited for handling large volumes of data asynchronously~\cite{Meijer12}.
-This combination of low-thread servers and asynchronous data models has
-facilitated the development of highly scalable, responsive, and resilient web
-applications capable of managing substantial data loads~\cite{Jin15}. The
-prominence of this concept increased with the advent of Node.js in 2009, and
-subsequently, various technologies adopted this approach in the Java ecosystem,
-including Netty, Akka HTTP, Vert.X, and Spring WebFlux.
-% Among these, Spring WebFlux stands out as the most widely used web framework in
-% Java, according to surveys like JetBrains' State of Java report (2021) and
-% community metrics from platforms like Github and StackOverflow.
-The \textit{non-blocking} I/O model in low-thread servers functions optimally
-only when HTTP handlers avoid blocking. Therefore, HTML templates need to be
-proficient in dealing with the asynchronous APIs provided by data models. While
-most legacy web templates struggle with asynchronous models, DSLs for HTML face
-no such limitations, leveraging all constructions available in the host
-programming language. However, the unexpected intertwining of asynchronous
-handlers' completion and HTML builders' execution may potentially lead to
-malformed HTML with an unexpected layout, as demonstrated in the subsequent
-subsection~\ref{sec:async-support}.
-
-An alternative approach involves utilizing user-level threads, while
-maintaining a blocking I/O and a synchronous programming paradigm. However,
-this approach still requires a user-level I/O subsystem capable of mitigating
-system-level blocking, which is crucial for the performance of I/O-intensive
-applications. This technique offers a lightweight solution for efficiently
-managing a larger number of concurrent sessions by minimizing per-thread
-overhead. In 2020, Karsten~\cite{karsten2020} demonstrated how this strategy
-supports a synchronous programming style, effectively abstracting away the
-complexities associated with managing continuations in asynchronous
-programming.
-
-Among mainstream technologies, the Kotlin programming language introduces a new
-abstraction for managing coroutines and provides structured concurrency, which
-ensures that coroutines are \textit{scoped}, \textit{cancellable}, and
-\textit{coordinated} within a well-defined
-lifecycle~\cite{elizarov2021coroutines}. Although this model embraces the
-\texttt{async}/\texttt{await} feature~\cite{async_await}, enabling non-blocking
-routines to mimic the structure of synchronous ones, it still lacks a
-user-level I/O subsystem and relies on an explicit I/O dispatcher tied to a
-dedicated thread pool that frees worker threads for handling processing tasks.
-Moreover, web templates using their own templating dialects, such as JSP,
-Thymeleaf, or Handlebars, are unable to take advantage of such constructs, as
-they typically support only a limited subset of the host library's API—most
-commonly just the \texttt{Iterable} interface.
-
-Only Java virtual threads, introduced in
-JEP~444~\footnote{\url{https://openjdk.org/jeps/444}}, adopt a similar approach
-to Karsten's proposal, using user-mode threads to preserve a synchronous
-programming model while still allowing blocking I/O operations without tying up
-platform threads. When a virtual thread performs a blocking I/O operation, the
-JVM intercepts the call and transparently parks the virtual thread, freeing the
-underlying platform thread to perform other tasks. Once the I/O operation
-completes, the virtual thread is rescheduled on a platform thread and resumes
-execution. This mechanism is enabled by the JVM's integration with non-blocking
-system calls at the OS level, allowing developers to write traditional
-synchronous code while benefiting from the scalability of non-blocking I/O.
-
-\subsubsection{Spring MVC}
-
-Spring MVC is a classical, synchronous web framework built on the
-thread-per-request model. When an HTTP request is received, it is handled by a
-dedicated thread from the server’s thread pool (e.g., Tomcat, Jetty, or
-Undertow). This thread is responsible for executing the full request lifecycle
-synchronously, including invoking controllers, executing business logic, and
-rendering the response view. Because the model is blocking, the thread remains
-occupied throughout the request—even during I/O operations such as database
-queries or external API calls—which can lead to performance bottlenecks under
-high concurrency.
-
-Controllers in Spring MVC typically return Java objects or models that are
-processed by view resolvers in conjunction with template engines like JSP,
-Thymeleaf, or FreeMarker. These templates generate the full HTML response in
-one pass, which is then sent to the client only after all necessary data has
-been gathered and rendered. By default, this architecture limits the ability to
-perform \textit{Progressive Server-Side Rendering (PSSR)}, as traditional
-rendering waits for all data before sending any output. This results in higher
-latency for clients and less responsive page loads, particularly for
-data-intensive views. However, Spring MVC provides a mechanism for response
-streaming via the \texttt{StreamingResponseBody} interface, introduced in
-Spring 4.2. When a controller method returns a \texttt{StreamingResponseBody},
-Spring writes directly to the \texttt{HttpServletResponse} output stream,
-allowing developers to send parts of the response incrementally as they become
-available. This is particularly useful for:
-\begin{itemize}
-  \item Sending large responses without buffering the entire output in memory.
-  \item Writing dynamic HTML in fragments as data is fetched or computed.
-  \item Reducing Time to First Byte (TTFB) and improving perceived performance.
-\end{itemize}
-
-A regular usage might look like the following:
-\begin{lstlisting}[language=Java]
-  @GetMapping("/stream")
-  fun handleStream(): StreamingResponseBody? {
-      return StreamingResponseBody { outputStream ->
-          for (i in 0..9) {
-              val htmlFragment = "<p>Chunk " + i + "</p>\n"
-              outputStream.write(htmlFragment.toByteArray())
-              outputStream.flush() // ensure partial response is sent
-              Thread.sleep(500) // simulate delay
-          }
-      }
-  }
-\end{lstlisting}
-
-In this example, HTML content is written and flushed in discrete chunks,
-allowing the client to progressively render the response as it arrives, while
-the server continues processing additional data. Although
-\texttt{StreamingResponseBody} does not eliminate the blocking nature inherent
-in Spring MVC—since the handling thread remains active during the streaming
-process—it does enable a basic form of PSSR by facilitating incremental content
-delivery. This approach is particularly useful for long-running operations or
-large response bodies, where providing early partial output can enhance
-perceived performance.
-
-However, a key limitation we encountered with this method involves the servlet
-response buffer. By default, most servlet containers (e.g., Tomcat) use a
-response buffer size of approximately 8KB. Data written to the response output
-stream is held in this buffer until it reaches capacity. Consequently, if the
-total size of the initial HTML content is less than the buffer threshold, the
-server will not transmit any data to the client until the buffer is filled.
-This buffering behavior introduces a delay in sending the first chunk of
-content, thereby negating the benefits of progressive rendering. In scenarios
-where the rendered HTML template does not produce sufficient data to exceed the
-buffer threshold early, PSSR becomes ineffective, and the client experiences a
-delayed initial response.
-
-\subsubsection{Spring WebFlux}
-
-Spring WebFlux is a reactive web framework built on a non-blocking,
-event-driven architecture that enables efficient handling of a large number of
-concurrent connections with minimal resource usage. Unlike traditional
-servlet-based stacks that rely on a thread-per-request model, WebFlux operates
-on reactive runtimes like Netty, where a small, fixed-size event loop manages
-all I/O events. Incoming HTTP requests are routed to handler functions or
-annotated controllers, which return reactive types such as \texttt{Mono<T>} and
-\texttt{Flux<T>}—Publisher implementations defined by the Reactive Streams
-specification. A \texttt{Mono} represents a single asynchronous value (or
-none), while a \texttt{Flux} represents a stream of zero or more values. These
-abstractions allow the entire request lifecycle—from controller invocation to
-response rendering—to be orchestrated as a chain of non-blocking, asynchronous
-operations. I/O-bound tasks such as database queries or external API calls do
-not occupy threads during execution, enabling the framework to scale gracefully
-under high load and maintain responsiveness even in resource-constrained
-environments.
-
-Progressive Server-Side Rendering (PSSR) is natively supported in WebFlux via
-\texttt{Flux}-based controllers. When a controller returns a
-\texttt{Flux<String>} or another stream of HTML fragments, WebFlux can begin
-streaming content to the client as soon as the first elements are emitted. This
-allows browsers to incrementally render parts of the response without waiting
-for the entire payload, significantly reducing Time to First Byte (TTFB) and
-improving user-perceived performance. By integrating PSSR directly into the
-reactive processing pipeline, WebFlux facilitates responsive and resilient web
-applications well-suited for modern microservices architectures.
-
-An illustrative example of PSSR in WebFlux is shown below:
-
-\begin{lstlisting}[language=Java]
-@GetMapping(value = "/pssr", produces = MediaType.TEXT_HTML_VALUE)
-public Flux<String> renderChunks() {
-    return Flux.range(1, 10)
-               .delayElements(Duration.ofMillis(500))
-               .map(i -> "<p>Chunk " + i + "</p>\n");
-}
-\end{lstlisting}
-
-In this example, the method emits one HTML chunk every 500 milliseconds. As
-each chunk is generated, it is immediately written to the HTTP response,
-allowing the client to begin rendering content even as more data is still being
-produced. Unlike \texttt{StreamingResponseBody} in Spring MVC, which is
-constrained by servlet buffer thresholds and blocking semantics, WebFlux
-ensures that each emitted item is pushed to the client as soon as the network
-is ready, without occupying a dedicated thread. Moreover, WebFlux integrates
-seamlessly with reactive data sources such as R2DBC (Reactive Relational
-Database Connectivity) and reactive NoSQL drivers, making it possible to build
-end-to-end non-blocking applications. HTML builders and DSLs used in
-conjunction with WebFlux must be designed to accommodate reactive streams,
-ensuring that view generation does not violate the non-blocking contract by
-performing blocking I/O or awaiting asynchronous operations imperatively.
-
-\subsubsection{Quarkus}
-
-Quarkus is a modern, cloud-native Java framework that supports both imperative
-and reactive programming models. It is built on top of Vert.x,
-which provides a non-blocking, event-driven runtime similar to Netty. Incoming
-HTTP requests are processed asynchronously on a small, fixed-size event loop
-thread pool, allowing the framework to handle many concurrent connections
-efficiently without blocking threads. Quarkus supports Kotlin coroutines and
-reactive APIs such as Mutiny for asynchronous programming, enabling developers
-to write sequential-style code that runs in a non-blocking manner.
-
-For traditional imperative-style request handling, Quarkus provides integration
-with JAX-RS, where resource methods can return synchronous types or
-asynchronous constructs such as \texttt{CompletionStage}. To enable Progressive
-Server-Side Rendering (PSSR) in imperative endpoints, Quarkus leverages the
-JAX-RS \texttt{StreamingOutput} interface. By returning a
-\texttt{StreamingOutput} instance, developers can write parts of the HTTP
-response body incrementally as data becomes available. This approach allows the
-server to flush HTML fragments progressively to the client, improving the
-perceived responsiveness especially for long-running or large responses.
-However, like other streaming mechanisms based on servlet buffers, the initial
-data flush may be delayed until internal buffers (commonly around 8KB) are
-filled. However, Quarkus allows configuring the buffer size to a smaller value, which
-can help mitigate this issue.
-
-\begin{lstlisting}[language=java,caption={Quarkus example using StreamingOutput for progressive HTML streaming}]
-@GET
-@Path("/stream")
-@Produces(MediaType.TEXT_HTML)
-public StreamingOutput streamHtml() {
-    return output -> {
-        PrintWriter writer = new PrintWriter(output);
-        writer.println("<html><body>");
-        writer.flush(); // Flush initial HTML header
-
-        for (int i = 1; i <= 5; i++) {
-            writer.println("<p>Chunk " + i + "</p>");
-            writer.flush(); // Flush each chunk incrementally
-            Thread.sleep(500); // Simulate processing delay
-        }
-
-        writer.println("</body></html>");
-        writer.flush();
-    };
-}
-\end{lstlisting}
-
-This example demonstrates how \texttt{StreamingOutput} can be used to send
-partial HTML content in chunks, allowing browsers to progressively render the
-response while the server continues processing. Quarkus’ architecture ensures
-that the underlying event loop threads are not blocked by these writes, as the
-framework offloads blocking I/O operations to worker threads. This combination
-of non-blocking event-driven runtime and incremental response streaming enables
-efficient PSSR even within imperative programming styles.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%-------------------------------------------------------------------
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsection{Web Templates}
-
-\textit{Web templates} have been the most widely adopted approach for
-constructing dynamic HTML pages.
-Web templates or \textit{web views}~\cite{Fowler02,Alur01} (e.g. JSP, Handlebars,
-or Thymeleaf), are based on HTML documents augmented with template-specific
-markers (e.g., \texttt{<\%>}, \texttt{\{\{\}\}}, or \texttt{\$\{\}}), which
-represent \textit{dynamic} information to be replaced at runtime with the
-results of corresponding computations, producing the final HTML page.
-The process of parsing and replacing these markers---i.e.,
-\textit{resolution}---is the primary responsibility of the \textit{template
-  engine}~\cite{Parr04}.
-One key characteristic of \textit{web templates} is their ability to receive a
-\textit{context object}---equivalent to the \textit{model} in the model-view
-design pattern~\cite{mvc88,Parr04}---which provides the data used to fill
-template placeholders at runtime.
-Web templates can be distinguished by several properties, namely:
-\begin{enumerate}
-  \item \textit{Domain-specific language} idiom
-  \item Supported data model APIs
-  \item Asynchronous support
-  \item Type safety and HTML safety
-  \item Progressive rendering
-\end{enumerate}
-
-Although some of the aforementioned characteristics apply to both
-\textit{server-side} and \textit{client-side} approaches, we focus solely on
-web template technologies for \textit{server-side rendering}, as our work is
-centered on that approach. Before getting deep on each of the aforementioned
-characteristics, Table~\ref{table:cmplibs} presents a breakdown of mainstream
-template engines, classified according to the identified properties.
-
-\begin{table}[h]
-  \small
-  \tabcolsep=0.1cm
-  \def\arraystretch{1.2}
-  \begin{tabular}{|c|c|c|c|c|c|c|}
-    \hline
-    \textbf{Library}
-     & \textbf{DSL idiom}
-     & \textbf{Data Model APIs}
-     & \shortstack{\textbf{Asynchronous}           \\\textbf{Support}}
-     & \shortstack{\textbf{Type}                   \\\textbf{Safety}}
-     & \shortstack{\textbf{HTML}                   \\\textbf{Safety}}
-     & \shortstack{\textbf{Progressive}            \\\textbf{Rendering}}
-    \\
-    \hline
-    \textbf{Freemarker}
-     & External DSL
-     & \texttt{Iterable}
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{PineGreen}{\checkmark}$}
-    \\
-    \hline
-    \textbf{JSP}
-     & External DSL
-     & \texttt{Iterable}
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{red}{\times}$}
-    \\
-    \hline
-    \textbf{JStachio}
-     & External DSL
-     & \texttt{Iterable}
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{PineGreen}{\checkmark}$}
-    \\\hline
-    \textbf{Pebble}
-     & External DSL
-     & \texttt{Iterable}
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{PineGreen}{\checkmark}$}
-    \\
-    \hline
-    \textbf{Rocker}
-     & External DSL
-     & \texttt{Iterable}
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{PineGreen}{\checkmark}$}
-    \\
-    \hline
-    \textbf{Thymeleaf}
-     & External DSL
-     & \shortstack{\texttt{Iterable} \\\texttt{Publisher}}
-     & \texttt{Publisher}$^{(1)}$
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{PineGreen}{\checkmark}$}
-    \\
-    \hline
-    \textbf{Trimou}
-     & External DSL
-     & \texttt{Iterable}
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{PineGreen}{\checkmark}$}
-    \\
-    \hline
-    \textbf{Velocity}
-     & External DSL
-     & \texttt{Iterable}
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{PineGreen}{\checkmark}$}
-    \\
-    \hline
-    \textbf{Clojure Hiccup}
-     & Nested Eager
-     & All
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{PineGreen}{\checkmark}$}
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{red}{\times}$}
-    \\
-    \hline
-    \shortstack{\textbf{Groovy }                   \\\textbf{MarkupBuilder}}
-     & Nested Lazy
-     & All
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{PineGreen}{\checkmark}$}
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{PineGreen}{\checkmark}$}
-    \\
-    \hline
-    \textbf{HtmlFlow}
-     & Chaining
-     & All
-     & \large{$\textcolor{PineGreen}{\checkmark}$}
-     & \large{$\textcolor{PineGreen}{\checkmark}$}
-     & \large{$\textcolor{PineGreen}{\checkmark}$}
-     & \large{$\textcolor{PineGreen}{\checkmark}$}
-    \\
-    \hline
-    \textbf{j2html}
-     & Nested Eager
-     & All
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{PineGreen}{\checkmark}$}
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{red}{\times}$}
-    \\
-    \hline
-    \textbf{KotlinX}
-     & Nested Lazy
-     & All
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{PineGreen}{\checkmark}$}
-     & \textcolor{PineGreen}{\checkmark}$^{(2)}$
-     & \large{$\textcolor{PineGreen}{\checkmark}$}
-    \\
-    \hline
-    \textbf{ScalaTags}
-     & Nested Eager
-     & All
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{PineGreen}{\checkmark}$}
-     & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{red}{\times}$}
-    \\
-    \hline
-  \end{tabular}
-  \caption{
-    Comparison of web template technologies in the Java ecosystem.
-    \\$^{(1)}$ Limited to a single model per web template and compatible with
-    the reactive stream \texttt{Publisher}.
-    \\$^{(2)}$ Non-safety for HTML attributes.
-  }
-  \label{table:cmplibs}
-\end{table}
-
-The first half of Table~\ref{table:cmplibs} lists template engines that use
-their own templating dialects, referred to as \textit{external DSLs}. The
-second half lists Java libraries that provide an \textit{internal DSL},
-typically using a \textit{nested} or \textit{chaining} style to build HTML.
-Note that, by leveraging the host language (e.g., Clojure, Groovy, Java,
-Kotlin, or Scala) as the templating idiom, the latter impose no restrictions on
-the data model and fully support all styles of data access APIs. They also
-benefit from static type checking, which helps ensure type safety.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%-------------------------------------------------------------------
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsubsection{Templating idiom}
-
-Web templates are based on a \textit{domain-specific language}
-(DSL)~\cite{landin1966next}, which defines a language tailored to a specific
-\textit{domain}~\cite{evans2004domain}—in this case, HTML for expressing web
-documents. The DSL constrains the template's syntax and semantics to match the
-structure and purpose of HTML.
-
-DSLs can be divided in two types: \textit{external} or
-\textit{internal}\cite{dslbook}. \textit{External} DSLs are languages created
-without any affiliation to a concrete programming language. An example of an
-\textit{external} DSL is the regular expressions search
-pattern\cite{thompson1968}, since it defines its own syntax without any
-dependency of programming languages. On the other hand an \textit{internal} DSL
-is defined within a host programming language as a library and tends to be
-limited to the syntax of the host language, such as Java.
-JQuery\cite{resig2007pro} is one of the most well-known examples of an internal
-DSL in Javascript, designed to simplify HTML DOM\cite{dom} tree traversal and
-manipulation.
-
-Traditionally, web template technologies use an \textit{external} DSL to define
-control flow constructs and data binding primitives. Early web template engines
-such as JSP, ASP, Velocity, PHP, and others adopted this \textit{external} DSL
-approach as their templating dialect. For example, a \texttt{foreach} loop can
-be expressed in each technology using its own DSL: \\\texttt{<\% for(String
-  item : items) \%>} in JSP, \texttt{<\% For Each item In items \%>} in legacy
-ASP with VBScript, \texttt{\#foreach(\$item in \$items)} in Velocity, or
-\texttt{<?php foreach (\$items as \$item):?>} in PHP.
-
-On the other hand, \textit{internal} DSLs for HTML allow templates to be
-defined directly within the \emph{host} language (such as Java, Kotlin, Groovy,
-Scala, or other general-purpose programming languages), rather than using
-text-based template files~\cite{carvalho2019hot}. In this case, a web template
-is not limited to templating constructs but may instead leverage any available
-primitive of the host language or any external API.
-
-Using an internal DSL can have several benefits over using textual templates:
-\begin{enumerate}
-  \item \emph{Type safety}: Because the templates are defined with the host
-        programming language, the compiler can check the syntax and types of the
-        templates at compile time, which can help catch errors earlier in the
-        development process.
-
-  \item \emph{IDE support}: Many modern IDEs provide code completion, syntax
-        highlighting, and other features, which can make it easier to write and
-        maintain templates.
-
-  \item \emph{Flexibility}: Use all the features of the host programming language
-        to generate HTML, can make it easier to write complex templates and reuse code.
-
-  \item \emph{Integration}: Because the templates are defined in Java code, for
-        example, you can easily integrate them with other Java code in your
-        application, such as controllers, services, repositories and models.
-
-\end{enumerate}
-
-DSLs for HTML provide an API where methods or functions correspond to the names
-of available HTML elements. These methods, also known as \textit{builders}, can
-be combined in a chain of calls to mimic the construction of an HTML document
-in a fluent manner. Martin Fowler\cite{dslbook} identifies three different
-patterns for combining functions to create a DSL: 1) \textit{function
-  sequence}; 2) \textit{nested function}, and 3) \textit{method chaining}, which
-are illustrated in the snippets of Figure~\ref{fig:dsl-idioms}.
-\begin{figure}[h]
-  \centering
-
-  % Subfloat 1
-  \begin{minipage}[c]{0.33\linewidth}
-    \centering
-    \begin{lstlisting}[
-    language=java,
-    basicstyle=\scriptsize\ttfamily,
-    numbers=none
-  ]
-html();
- head();
-  title();text("JT");end();
- end();
- body();
-  p();
-   text("Hi JATL");
-  end();
- end();
-end();
-  \end{lstlisting}
-    \caption*{(a) Function sequence}
-  \end{minipage}
-  \hfil
-  % Subfloat 2
-  \begin{minipage}[c]{0.23\linewidth}
-    \centering
-    \begin{lstlisting}[
-    language=java,
-    basicstyle=\scriptsize\ttfamily,
-    numbers=none
-  ]
-html(
- head(
-  title("ST")
- ),
- body(
-  p("Hi ScalaTags")
- )
-);
-  \end{lstlisting}
-    \caption*{(b) Nested function}
-  \end{minipage}
-  \hfil
-  % Subfloat 3
-  \begin{minipage}[c]{0.32\linewidth}
-    \centering
-    \begin{lstlisting}[
-    language=java,
-    basicstyle=\scriptsize\ttfamily,
-    numbers=none
-  ]
-html()
- .head()
-  .title().text("HF").__()
- .__()
- .body()
-  .p()
-   .text("Hi HtmlFlow")
-  .__()
- .__()
-.__();
-  \end{lstlisting}
-    \caption*{(c) Method chaining}
-  \end{minipage}
-
-  \caption{Utilizing DSL for HTML libraries with JATL, ScalaTags, and HtmlFlow.}
-  \label{fig:dsl-idioms}
-\end{figure}
-
-From the three examples depicted in Figure~\ref{fig:dsl-idioms}, the
-\textit{nested function} idiom used by \textbf{ScalaTags}, as shown in the
-snippet of Figure~\ref{fig:dsl-idioms}.b), is the least verbose, requiring
-fewer statements than the other two cases. This approach of combining function
-calls is also utilized by \textbf{Hiccup} Clojure Library and \textbf{j2html}
-Java library. One reason for verbosity in \textit{function sequence} and
-\textit{method chaining} approaches is the requirement of a dedicated HTML
-builder to emit the closing tag, exemplified by \texttt{end()} in \textbf{JATL}
-(Figure~\ref{fig:dsl-idioms}.a) and \texttt{\_\_()} in \textbf{HtmlFlow}
-(Figure~\ref{fig:dsl-idioms}.c). But the \textit{nested function} approach
-comes with a notable drawback: \textbf{they do not support PSSR} because the
-sequence of nested functions is evaluated backward to the order in which they
-are written. In other words, arguments are evaluated before the functions are
-invoked. Taking the example in Figure~\ref{fig:dsl-idioms}.b), the
-\texttt{title()} function is first evaluated, and its resulting paragraph
-becomes the argument for the \texttt{head()} call, which, in turn, becomes the
-argument for \texttt{html()}, and so on. If HTML is emitted as functions are
-called, it will print tags in reverse order. The aforementioned DSLs must
-collect resulting nodes into an internal data structure, which is later
-traversed to produce the HTML output. Therefore, they cannot progressively emit
-the output as builders are called.
-
-Two other JVM libraries, \textbf{Groovy MarkupBuilder} and
-\textbf{KotlinX.html}, also adopt a \textit{nested function} approach, but they
-address the backward evaluation issue by implementing \textit{lazy evaluation}
-of arguments~\cite{Landin65}, expressed in lambda expressions (also known as
-function literals). Due to the concise form of expressing lambdas with brackets
-(i.e., \texttt{\{\}}) in both Groovy and Kotlin, translating the template shown
-in Figure~\ref{fig:dsl-idioms}.b) to use Groovy MarkupBuilder or KotlinX.html
-only requires replacing parentheses with brackets for parent elements.
-\textbf{HtmlFlow} also adopts this approach and provides a Kotlin-idiomatic API
-as an alternative to its original Java API.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%-------------------------------------------------------------------
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsubsection{Supported data model APIs}
-
-Template engines with dedicated templating dialects typically rely on specific
-control flow constructs—such as a \texttt{foreach}-like statement—to iterate
-over elements in a data source. In the Java ecosystem, \texttt{Iterable} is the
-common supertype implemented by most data structures and serves as the standard
-API for iteration. However, other standard interfaces, such as
-\texttt{java.util.Stream}, also represent sequences of elements but are not
-compatible with \texttt{Iterable}. This fragmentation becomes more evident when
-considering other JVM languages, such as Kotlin, which introduces additional
-abstractions like the \texttt{Sequence} interface—also incompatible with
-\texttt{Iterable}. As a result, template engines based on \textit{external}
-DSLs must explicitly support each of these interfaces to accommodate the
-various iteration protocols across languages and libraries.
-
-On the other hand, template engines that employ an \textit{internal} DSL do not
-face such limitations, as they can leverage any available API within the host
-programming language. This flexibility extends beyond the standard library to
-include third-party APIs as well. For example, in Java, developers can iterate
-over data using a traditional \texttt{for} loop, the \texttt{java.util.stream}
-API, or external libraries such as Guava, Vavr, StreamEx, and others—depending
-on their specific needs or preferences.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%-------------------------------------------------------------------
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsubsection{Asynchronous support}
-\label{sec:async-support}
-
-One of the reasons for legacy web templates not supporting asynchronous APIs is
-the absence of a unified standard calling convention for asynchronous calls.
-While there is a single, straightforward way to use a synchronous API with a
-direct style, where the result of a method call corresponds to its returned
-value, there is no equivalent standard in the asynchronous approach. Instead,
-we may encounter various asynchronous conventions depending on the programming
-language and runtime environment. Some of these approaches include
-\emph{continuation-passing style} (CPS)~\cite{scheme},
-\textit{promises}~\cite{promise} async/await idiom~\cite{async_await}, reactive
-streams \cite{ReactiveStreams}, Kotlin Flow~\cite{kotlinlang}, and others.
-
-Many \textit{general-purpose languages} (GPLs) have embraced the
-\texttt{async}/\texttt{await} feature~\cite{async_await} enabling non-blocking
-routines to mimic the structure of synchronous ones, allowing developers to
-reason about instruction flow sequentially. The simplicity and broad adoption
-of this programming model have led to its incorporation into mainstream
-languages like C\#, JavaScript, Python, Perl, Swift, Kotlin, and others,
-excluding Java. However, implementing \texttt{async}/\texttt{a}wait requires
-compiler support to translate \textit{suspension points} (i.e., \texttt{await}
-statements) into state machines. Most template engines operate using an
-external DSL with their own templating dialect (e.g., Thymeleaf, JSP, Jade,
-Handlebars, and others), which do not inherently leverage asynchronous
-capabilities from their host GPLs.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%-------------------------------------------------------------------
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsubsection{Type safety and HTML safety}
-
-Another key characteristic of such DSLs is \textbf{HTML safety}, which refers
-to whether they produce only valid HTML conforming to a well-formed document.
-To ensure HTML safety, a DSL API should only allow combining calls to builders
-that result in valid HTML. However, most HTML DSLs using the \textit{function
-  sequence} or \textit{nested function} approach cannot ensure HTML safety at
-compile time.
-
-The \textit{function sequence} approach, as illustrated in the snippet of
-Figure~\ref{fig:dsl-idioms}.a), involves combining function calls as a sequence
-of statements, making it challenging to restrict the order of statements. In
-the \textit{nested function} approach shown in Figure~\ref{fig:dsl-idioms}.b),
-variable-length arguments are often used to allow an undefined number of child
-elements, which cannot be strongly typed in every programming language.
-
-KotlinX.html, which utilizes the \textit{nested function} idiom with lazy
-evaluation, mitigates this issue through function types with a receiver. In
-this approach, the receiver (i.e., \texttt{this} within the lambda) is strongly
-typed and provides a set of methods corresponding to legal child elements. This
-enables KotlinX.html to enforce HTML safety by restricting the available
-methods during compile time.
-
-To achieve this KotlinX.html provides HTML builders using \textit{function
-  literals with receiver}~\cite{kotlinlang}. In Kotlin, a block of code enclosed
-in curly braces \texttt{\{...\}} is known as a \emph{lambda}, and can be used
-as an argument to a function that expects a \emph{function literal}. When we
-write, for example, \texttt{body \{ div \{ hr() \} \}}, we are invoking the
-\texttt{body} function with a lambda as its argument. This lambda, in turn,
-calls the \texttt{div} function with another lambda as an argument that creates
-a horizontal row (i.e. \texttt{hr}). Each call to an HTML builder (e.g.,
-\texttt{body}, \texttt{div}, \texttt{hr}) creates the child element within the
-element generated by the outer function call.
-
-HtmlFlow provides two APIs: one in idiomatic Kotlin, similar to the
-\texttt{KotlinX.html} API, and another that employs the \textit{method
-  chaining} idiom, as illustrated in the snippet of
-Figure~\ref{fig:dsl-idioms}.c). In this approach, the receiver object is
-implicitly passed as an argument to each method call, enabling subsequent
-methods to be invoked on the result of the preceding one. This facilitates the
-composition of methods, with each call building upon the other. Similar to
-KotlinX.html, HtmlFlow ensures HTML safety, restricting the available HTML
-builders and attributes after the dot (\texttt{.}) operator.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%-------------------------------------------------------------------
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\subsubsection{Progressive rendering}
-
-Pursuing the optimal user experience for web users has been a consistent goal
-since the inception of the World Wide Web. Consequently, various approaches
-have been explored to deliver the initial meaningful content to the end-user as
-promptly as possible, while the remaining content \textit{progressively} (or
-incrementally) loads as the server streams the HTML content.
-
-\textit{\textbf{Progressive rendering}} and \textit{\textbf{progressive loading}}
-encompass different concepts. The former pertains to the \textit{dynamic} content of a
-\textit{dynamic web page}, encompassing elements with logic and placeholders
-that are fulfilled by data from an object \textit{model} constructed at runtime.
-On the other hand, the latter is associated with \textit{render-blocking
-  resources} such as scripts, stylesheets, and HTML imports, which may hinder the
-browser from rendering page content to the screen~\cite{progressive-2022}.
-The notion of \textit{progressive rendering} has typically been aligned with
-\textit{client-side rendering} (CSR), where a single HTML page with static
-content is delivered upfront, while the dynamic content is fetched as the data
-becomes available to complete the web page.
-However, despite being overlooked, HTTP and browsers were designed from their
-inception to also support this feature in the context of SSR approaches,
-which offers the advantage of not being dependent on
-\textit{rendering-blocking resources}, such as the required JavaScript for CSR.
-% \textit{Progressive server-side rendering} (PSSR) is the
-% ability of streaming HTML content to the client in chunks or data frames, as it
-% gets resolved on the server at runtime.
-% This lets the user-interface to be rendered incrementally (i.e.
-% \textit{progressively}) by the browser in line with the availability of the data.
-% However, most SSR template engines do not inherently support this feature at the
-% template level, requiring developers to manually break down web pages into
-% fragments or partials to enable PSSR.
-An example of this limitation in SSR web templates was highlighted by Jeff
-Atwood in 2005, who criticized Microsoft ASP.NET for loading the entire web
-page into memory before sending any data to the browser~\cite{pssr2005}.
-% This practice delays the display of content to the user until the server
-% completes its processing, contradicting the original design intent of HTML to
-% render progressively as content is received.
-% Notably, both Netscape and Internet Explorer were capable of rendering partial
-% HTML content from their inception.
-Despite historical critiques and HTML's inherent capabilities, most Web
-application frameworks, such as ASP.Net, Express.js, Spring, and others,
-persistently lack support for progressive rendering, leading to the appearance
-of alternative techniques leveraging client-side JavaScript. Since 2007,
-various patents have addressed the PSSR issue, with Microsoft's patent enabling
-the infinite scrolling technique by displaying a single page of
-results~\cite{scroll2007}, and Yahoo's patent focusing on differentiating
-elements based on their position relative to the visible
-portion~\cite{schiller2007progressive}.
-
-Former techniques all describe methods of progressively adding content to a web
-page depending on client-side JavaScript.
-The Java Thymeleaf, the default SSR template engine for Spring web servers,
-added support for PSSR in 2018, through the use of a specific non-blocking
-Spring \texttt{ViewResolver} driver and without requiring client-side
-JavaScript.
-Carvalho~\cite{carvalho2023async} introduced an SSR solution that manages
-multiple data models and asynchronous APIs while ensuring well-formed HTML,
-PSSR, and non-blocking template resolution. That solution builds upon HtmlFlow,
-a Java DSL for HTML, with an internal processing mechanism that streams HTML in
-chunks as data items become available from data sources, without requiring
-client-side JavaScript. Like Thymeleaf, their proposal avoids the use of
-client-side JavaScript; however, unlike Thymeleaf, it supports a wide range of
-asynchronous APIs and multiple data sources, not limited to the
-\texttt{Publisher} API. However, the main counter-argument is the non-trivial
-management of the \emph{resume} callback in
-continuations~\cite{von2003events,callbackhell}, which is used to linearize the
-execution flow between asynchronous calls.
+\input{Sections/state-of-the-art-web-servers.tex}
+\input{Sections/state-of-the-art-web-templates.tex}

### 2025-05-27 docs/article/Sections/state-of-the-art-web-servers.tex
Section 2 - Web Servers - minor fixes

@@
 bogged down by context switching and thread management
-overhead~\cite{kant2000scalable}. In the Java ecosystem, Spring Boot is one of
+overhead~\cite{kant2000scalable}. In the Java ecosystem, Spring MVC is one of
 the most widely used frameworks that follow this model, according to several
@@
 handlers' completion and HTML builders' execution may potentially lead to
-malformed HTML with an unexpected layout, as demonstrated in the subsequent
-subsection~\ref{sec:async-support}.
+malformed HTML with an unexpected layout~\cite{wise2024pssr}.
 
@@
 Only Java virtual threads, introduced in
-JEP~444~\footnote{\url{https://openjdk.org/jeps/444}}, adopt a similar approach
+JEP~444\footnote{\url{https://openjdk.org/jeps/444}}, adopt a similar approach
 to Karsten's proposal, using user-mode threads to preserve a synchronous
@@
 been gathered and rendered. By default, this architecture limits the ability to
-perform \textit{Progressive Server-Side Rendering (PSSR)}, as traditional
+perform PSSR, as traditional
 rendering waits for all data before sending any output. This results in higher
@@
 which return reactive types such as \texttt{Mono<T>} and
-\texttt{Flux<T>}—Publisher implementations provided by the Project Reactor
+\texttt{Flux<T>}—\texttt{Publisher} implementations provided by the Project Reactor
 library~\cite{projectreactor}. These implementations adhere to the Reactive
@@
 
-Progressive Server-Side Rendering (PSSR) is natively supported in WebFlux via
+PSSR is natively supported in WebFlux via
 \texttt{Flux}-based controllers. When a controller returns a
@@
 with JAX-RS, where resource methods can return synchronous types or
-asynchronous constructs such as \texttt{CompletionStage}. To enable Progressive
-Server-Side Rendering (PSSR) in imperative endpoints, Quarkus leverages the
+asynchronous constructs such as \texttt{CompletionStage}. To enable 
+PSSR in imperative endpoints, Quarkus leverages the
 JAX-RS \texttt{StreamingOutput} interface. By returning a
@@
         PrintWriter writer = new PrintWriter(output);
-        writer.println("<html><body>");
-        writer.flush(); // Flush initial HTML header
         for (int i = 1; i <= 5; i++) {
@@
         }
-        writer.println("</body></html>");
         writer.flush();

### 2025-05-27 docs/article/Sections/introduction.tex

@@
 
-Efficient thread utilization is crucial for low-thread servers. These servers
-must handle many concurrent requests without being overwhelmed, which
-necessitates asynchronous, non-blocking request handling
-~\cite{carvalho2023async}. By leveraging asynchronous I/O operations—such as
+\textit{Low-thread} servers, also known as
+\textit{event-driven}~\cite{event-driven-servers}, have gained prominence in
+contemporary web applications due to their ability to efficiently manage a large
+number of concurrent I/O operations with minimal resources, thus promoting
+better scalability.
+By leveraging asynchronous I/O operations—such as
 database queries and API calls, servers can avoid blocking threads while
@@
 this non-blocking architecture, PSSR implementations require template engines
-that are compatible with asynchronous data models and I/O. Some modern template
+that are compatible with asynchronous data models~\cite{carvalho2023async}.
+Some modern template
 engines, such as HtmlFlow~\cite{htmlflow} and Thymeleaf~\cite{thymeleaf}, have
@@
 template engines—particularly those using external domain-specific languages
-~\cite{Fowler03} (DSLs)—still depend on blocking interfaces like
+(DSLs)~\cite{Fowler03}—still depend on blocking interfaces like
 \texttt{Iterable} for data processing. This blocking behavior forces server
@@
 high-concurrency, non-blocking environments without requiring complex
-asynchronous programming models. As a result, Progressive Server-Side Rendering
-(PSSR) can now be implemented using familiar HTML templates, 
+asynchronous programming models. As a result, PSSR
+can now be implemented using familiar HTML templates, 
 simplifying development and improving maintainability.
@@
 primary paradigms: \textit{reactive programming} and \textit{coroutines}, both
-of which have traditionally been used to achieve asynchronous I/O. As an
+of which have been used to achieve asynchronous I/O in the Java ecosystem. As an
 alternative, we investigate whether Java’s Virtual Threads can offer comparable

### 2025-05-27 docs/article/Sections/problem-statement.tex

--- 2025-05-23 docs/article/Sections/problem-statement.tex
@@
 
-In this section, we examine the challenges of implementing Progressive
-Server-Side Rendering (PSSR) in modern web applications, with a focus on the
+In this section, we examine the challenges of implementing \textit{Progressive
+Server-Side Rendering} (PSSR) in modern web applications, with a focus on the
 limitations of current template engine designs. Our goal is to broaden the
 range of options available for PSSR, particularly within JVM-based frameworks.
-
-Reactive streams~\cite{ReactiveStreams}, such as those based on
-\texttt{Observable} (from RxJava)~\cite{rx-observable} and \texttt{Flow} (from
-the Java 9+ standard library~\footnote{~\url{https://openjdk.org/jeps/266}}),
-are abstractions that provide a non-blocking, asynchronous
-way to handle data streams.
-These abstractions support the incremental generation and transmission of HTML
-content as data becomes available, enabling \textit{progressive server-side
-rendering} (PSSR)~\cite{pssr2005}.
+% Reactive streams~\cite{ReactiveStreams}, such as those based on
+% \texttt{Observable} (from RxJava)~\cite{rx-observable} and \texttt{Flow} (from
+% the Java 9+ standard library~\footnote{~\url{https://openjdk.org/jeps/266}}),
+% are abstractions that provide a non-blocking, asynchronous
+% way to handle data streams.
+% These abstractions support the incremental generation and transmission of HTML
+% content as data becomes available, enabling \textit{progressive server-side
+% rendering} (PSSR)~\cite{pssr2005}.
 In PSSR, the server does not
@@
 
-Reactive types like \texttt{Observable<T>} or \texttt{Flow<T>}
+Reactive types like \texttt{Observable<T>} of ReactiveX~\cite{rxjava} or Kotlin \texttt{Flow<T>}~\cite{kotlinlang}
 facilitate this by representing data as a sequence of asynchronous events.
-These streams emit values over time, allowing the
-rendering engine to consume data at a manageable rate without overwhelming
-system resources.
 For example, a reactive stream might emit a sequence of \texttt{Presentation}
@@
 In \autoref{lst:presentation-flow}, we show an
-equivalent suspend-based implementation using Kotlin's \texttt{Flow}~\cite{PSSR-WISE2024}.
+equivalent suspend-based implementation using Kotlin's \texttt{Flow}~\cite{wise2024pssr}.
 Both examples highlight how internal DSLs can natively integrate with reactive types
@@
     numbers=none,
-    caption={\textit{HtmlFlow} presentation template with \texttt{Observable}},
+    caption={\textit{HtmlFlow} presentation template in Koltin with an \texttt{Observable} model},
     label={lst:presentation-observable}
@@
     label={lst:presentation-flow},
-    caption={\textit{HtmlFlow} presentation template with \texttt{Flow}},
+    caption={\textit{HtmlFlow} presentation template in Kotlin with a \texttt{Flow} model},
 ]
@@
 documents using custom markers, and rely on blocking interfaces like
-\texttt{Iterable} or \texttt{List}. These interfaces require the entire data
+\texttt{java.util.Iterable} or \texttt{java.util.stream.Stream}.
+These interfaces require the entire data
 model, to be materialized in memory before rendering can begin, which blocks server
@@
 \texttt{Observable.blockingIterable()}, which allows asynchronous data sources
-to be exposed as \texttt{Iterable} by blocking the thread until new items are available.
+to be exposed as \texttt{Iterable} by blocking the thread until all items are available.
 While useful for compatibility with traditional APIs, this approach
@@
 blocking, synchronous rendering. However, this tradeoff becomes critical under
-high concurrency, where blocking threads severely degrades throughput~\cite{PSSR-WISE2024}.
+high concurrency, where blocking threads severely degrades throughput~\cite{wise2024pssr}.
 Emerging features in the Java ecosystem, particularly \textit{virtual threads}

### 2025-05-27 docs/article/Sections/state-of-the-art-web-templates.tex

@@
      & External DSL
-     & \shortstack{\texttt{Iterable} \\\texttt{Stream} \\\texttt{Publisher}}
+     & \shortstack{\texttt{Iterable} \\\texttt{Stream} \\\texttt{Publisher}$^{(1)}$}
      & \texttt{Publisher}$^{(1)}$
@@
      & External DSL
-     & \shortstack{\texttt{Iterable} \\\texttt{Sequence}}
+     & \shortstack{\texttt{Iterable} \\\texttt{Sequence}$^{(2)}$}
      & \large{$\textcolor{red}{\times}$}
@@
      & \large{$\textcolor{PineGreen}{\checkmark}$}
-     & \textcolor{PineGreen}{\checkmark}$^{(2)}$
+     & \textcolor{PineGreen}{\checkmark}$^{(3)}$
      & \large{$\textcolor{PineGreen}{\checkmark}$}
@@
     Comparison of web template technologies in the Java ecosystem.
-    \\$^{(1)}$ Limited to a single model per web template and compatible with
-    the reactive stream \texttt{Publisher}.
-    \\$^{(2)}$ Non-safety for HTML attributes.
+    \\$^{(1)}$ \texttt{Publisher} of Reactive Streams~\cite{ReactiveStreams}.
+    Limited to a single model per web template.
+    \\$^{(2)}$ \texttt{kotlin.sequences.Sequence}
+    \\$^{(3)}$ Non-safety for HTML attributes.
   }
@@
 the data model and fully support all styles of data access APIs. They also
-benefit from static type checking, which helps ensure type safety.
+benefit from static type checking, which helps ensure \textit{type safety}.
 
@@
 
-\subsubsection{Templating idiom}
+\subsubsection{Domain-specific language idiom}
 
@@
 structure and purpose of HTML.
-
 DSLs can be divided in two types: \textit{external} or
@@
 approach as their templating dialect. For example, a \texttt{foreach} loop can
-be expressed in each technology using its own DSL: \\\texttt{<\% for(String
-  item : items) \%>} in JSP, \texttt{<\% For Each item In items \%>} in legacy
-ASP with VBScript, \texttt{\#foreach(\$item in \$items)} in Velocity, or
-\texttt{<?php foreach (\$items as \$item):?>} in PHP.
+be expressed in each technology using its own DSL:
+\begin{itemize}
+  \item \texttt{<\% for(String item : items) \%>} in JSP
+  \item \texttt{<\% For Each item In items \%>} in legacy ASP with VBScript
+  \item \texttt{\#foreach(\$item in \$items)} in Velocity
+  \item \texttt{<?php foreach (\$items as \$item):?>} in PHP
+\end{itemize}
 
@@
 Scala, or other general-purpose programming languages), rather than using
-text-based template files~\cite{carvalho2019hot}. In this case, a web template
+text-based template files~\cite{carvalho2020}. In this case, a web template
 is not limited to templating constructs but may instead leverage any available
@@
 
+For example, Thymeleaf supports both Java \texttt{Iterable} and \texttt{Stream}
+iteration protocols but is incompatible with Kotlin's \texttt{Sequence}
+interface.
+Velocity, on the other hand, uses internal reflection-based template processing
+that allows it to iterate over any type defining an \texttt{iterator()} method
+returning an \texttt{Iterator}. This enables support for both Java
+\texttt{Iterable} and Kotlin \texttt{Sequence}, but it cannot handle Java
+\texttt{Stream} since \texttt{Stream} does not expose a compatible
+\texttt{iterator()} method.
+The remaining analyzed template engines that use external DSLs support only Java
+\texttt{Iterable} as their sole iteration protocol.
+
 On the other hand, template engines that employ an \textit{internal} DSL do not
@@
 \emph{continuation-passing style} (CPS)~\cite{scheme},
-\textit{promises}~\cite{promise} async/await idiom~\cite{async_await}, reactive
-streams \cite{ReactiveStreams}, Kotlin Flow~\cite{kotlinlang}, and others.
+\textit{promises}~\cite{promise}, async/await idiom~\cite{async_await},
+or \textit{suspend functions}~\cite{elizarov2021coroutines}.
 
@@
 Handlebars, and others), which do not inherently leverage asynchronous
-capabilities from their host GPLs.
+capabilities from their host programming languages.
+
+An alternative approach to asynchronous programming involves handling data as
+a sequence of asynchronous events, commonly known as a \textit{reactive stream}.
+Instead of materializing all items in memory, reactive streams emit values over
+time, allowing clients to subscribe and consume data incrementally as it becomes
+available.
+This idea was first proposed by Meijer~\cite{rx-observable} in the .NET
+framework, leading to the development of \textit{Reactive Extensions} (Rx).
+Subsequently, several alternative implementations emerged in the Java ecosystem,
+such as Rx Java~\cite{rxjava}, Project Reactor~\cite{projectreactor}, or the
+Akka Streams~\cite{akka} library.
+Later, due to the proliferation of alternative implementations, the need arose to
+create a common interface to ensure interoperability between different reactive
+stream libraries. This led to the formation of a working group that included
+engineers from notable companies such as Lightbend, Pivotal, and Netflix.
+The group began defining the \textit{Reactive Streams}
+specification~\cite{ReactiveStreams}, which established a standard for
+asynchronous stream processing using non-blocking, backpressure-aware data
+pipelines.
+The more recent Mutiny~\cite{mutiny2021}, while conforming to the
+\textit{Reactive Streams} standard, proposes a different approach that avoids
+complex optimizations, resulting in more straightforward and easier-to-maintain
+code.
+
+Thymeleaf was one of the first web template engines—and remains the only one
+among external DSLs for HTML—to support reactive streams based on the
+\texttt{Publisher} model. However, it is limited to using a single
+\texttt{Publisher} within a given template.
 
@@
 
-\subsubsection{Type safety and HTML safety}
+\subsubsection{HTML safety}
 
-Another key characteristic of such DSLs is \textbf{HTML safety}, which refers
+Another key characteristic of such DSLs is \textit{HTML safety}, which refers
 to whether they produce only valid HTML conforming to a well-formed document.

### 2025-05-27 docs/article/Sections/state-of-the-art.tex

@@
 
-In this section, we first examine the different design choices adopted by web
-servers in their internal architectures, and how these choices impact the
-behavior of the web template engines. Then, in Subsection~2.2, we present the
-main properties that characterize each web template technology approach, along
-with the advantages and drawbacks that result from these characteristics.
+In this section, we first present the main properties that characterize each web
+template technology approach, along with the advantages and drawbacks that result
+from these characteristics. Then, in Subsection~2.2, we examine the different
+design choices adopted by web servers in their internal architectures, and how
+these choices impact the behavior of the web template engines.
 
+\input{Sections/state-of-the-art-web-templates.tex}
 \input{Sections/state-of-the-art-web-servers.tex}
-\input{Sections/state-of-the-art-web-templates.tex}

### 2025-05-28 docs/article/Sections/benchmark.tex
Section 4: Include approaches from Spring MVC and Quarkus

@@
 
-The benchmark is designed with a modular architecture, separating the
-\textit{view} and \textit{model} layers from the \textit{controller} layer
-~\cite{Bucanek2009}, which allows for easy extension and integration of new
-template engines and frameworks. It also includes a set of tests to ensure the
-correctness of implementations and to validate the \textit{HTML} output. It
-includes two different data models, defined as \texttt{Presentation} and
-\texttt{Stock}, shown in \autoref{lst:stock-class} and
-\autoref{lst:presentation-class}, respectively. The \texttt{Presentation} class
-represents a presentation with a title, speaker name, and summary, while the
-\texttt{Stock} class represents a stock with a name, URL, symbol, price,
-change, and ratio.
+The benchmark\footnote{\url{https://github.com/fmcarvalho/comparing-non-blocking-progressive-ssr}}
+is designed with a modular architecture, separating the \textit{view} and
+\textit{model} layers from the \textit{controller} layer ~\cite{Bucanek2009},
+which allows for easy extension and integration of new template engines and
+frameworks. It also includes a set of tests to ensure the correctness of
+implementations and to validate the \textit{HTML} output. It includes two
+different data models, defined as \texttt{Presentation} and \texttt{Stock}, as
+shown in \autoref{lst:stock-class} and \autoref{lst:presentation-class},
+following the same model as
+\textit{spring-webflux-comparing-template-engines}~\footnote{\url{https://github.com/xmlet/spring-webflux-comparing-template-engines}}.
+The \texttt{Presentation} class represents a presentation with a title, speaker
+name, and summary, while the \texttt{Stock} class represents a stock with a
+name, URL, symbol, price, change, and ratio.
 
@@
 response body, which acts as a publisher that progressively streams the HTML
-content to the client. The implementation includes three main approaches to
+content to the client. The implementation includes four main approaches to
 PSSR\@:
@@
 request processing where the application can write directly to the response
-  OutputStream without holding up the Servlet container thread}
+OutputStream without holding up the Servlet container thread.}
 
@@
 
+This implementation includes two main approaches to PSSR\@:
+\begin{itemize}
+      \item \textbf{Blocking}: The template engine is used in a blocking context,
+            where the HTML content is rendered using the blocking interface of the
+            \texttt{Observable} class.
+      \item \textbf{Virtual}: The template engine is used in a non-blocking context,
+            where the HTML content is rendered within the context of Virtual Threads.
+\end{itemize}
+
 The Quarkus implementation also uses handlers based on the blocking interface
@@
 to be sent to the client progressively as the HTML content is rendered.
+
+This implementation includes three main approaches to PSSR\@:
+\begin{itemize}
+      \item \textbf{Blocking}: The template engine is used in a blocking context,
+            where the HTML content is rendered using the blocking interface of the
+            \texttt{Observable} class.
+      \item \textbf{Virtual}: The template engine is used in a non-blocking context,
+            where the HTML content is rendered within the context of Virtual Threads.
+      \item \textbf{Reactive}: The template engine is used in a reactive context,
+            where the HTML content is rendered using the reactive programming model.
+            In this case, we use the HtmlFlow template engine, which
+            supports asynchronous rendering through the \texttt{writeAsync} method.
+\end{itemize}

### 2025-05-28 docs/article/Sections/results.tex
Sections 2,4,5: formatting
Section 3 & 5: update template engines table; Add HtmlFlow-Rx to Quarkus Results and change Sync to Blocking

@@
 and Thymeleaf using the reactive View Resolver driver (Thymeleaf-Rx).
-\textit{Sync} and \textit{Virtual} represent the average throughput of the
+\textit{Blocking} and \textit{Virtual} represent the average throughput of the
 blocking approaches (i.e., KotlinX, Rocker, Jstachio, Pebble, Freemarker,
@@
 separately to observe the performance of the non-blocking engines when using
-the Suspending, Virtual Thread, and Reactive approaches. The \textit{Sync} and
-\textit{Virtual} are aggregated due to the similar performance of different
+the Suspending, Virtual Thread, and Reactive approaches. The \textit{Blocking}
+and \textit{Virtual} are aggregated due to the similar performance of different
 engines when using those approaches.
@@
 \begin{figure}[h]
-     \isPreprints{\centering}{}
+     \centering
      \includegraphics[width=0.8\textwidth]{./Graphs/presentations-webflux-jmeter.png}
@@
 \begin{figure}[h]
-     \isPreprints{\centering}{}
+     \centering
      \includegraphics[width=0.8\textwidth]{./Graphs/presentations-springmvc-jmeter.png}
@@
 
+% The results for the Spring MVC implementation, shown in
+% Figure~\ref{fig:presentations-springmvc-jmeter}, compare two approaches:
+% \textit{Blocking}, which uses platform threads with \texttt{StreamingResponseBody},
+% and \textit{Virtual}, which uses Virtual Threads. Both approaches scale
+% effectively up to 128 concurrent users, with the Virtual Thread approach
+% achieving a slightly higher throughput of 9,000 requests per second. However,
+% these values are slightly lower than those observed in the Spring WebFlux
+% implementation.
+
 The results for the Spring MVC implementation, shown in
-Figure~\ref{fig:presentations-springmvc-jmeter}, compare two approaches:
-\textit{Sync}, which uses platform threads with \texttt{StreamingResponseBody},
-and \textit{Virtual}, which uses Virtual Threads. Both approaches scale
-effectively up to 128 concurrent users, with the Virtual Thread approach
-achieving a slightly higher throughput of 9,000 requests per second. However,
-these values are slightly lower than those observed in the Spring WebFlux
-implementation.
+Figure~\ref{fig:presentations-springmvc-jmeter}, compare two synchronous
+approaches: \textit{Blocking}, which uses platform threads with
+\texttt{StreamingResponseBody}, and \textit{Virtual}, which uses Virtual
+Threads. Note that for Spring MVC, which follows a thread-per-request
+architecture, neither of the asynchronous approaches—\textbf{Reactive} nor
+\textbf{Suspendable}—described in Section~5 are applicable. Regarding the
+synchronous strategy, both \textit{Blocking} and \textit{Virtual} scale
+effectively up to 128 concurrent users, with the Virtual Threads approach
+achieving a slightly higher throughput of approximately 9,000 requests per
+second. However, these values are slightly lower than those observed in the
+Spring WebFlux implementation.
 
 \begin{figure}[h]
-     \isPreprints{\centering}{}
+     \centering
      \includegraphics[width=0.8\textwidth]{./Graphs/presentations-quarkus-jmeter.png}
@@
 The results for the Quarkus implementation, shown in
-Figure~\ref{fig:presentations-quarkus-jmeter} demonstrate that Quarkus handles
-blocking approaches more effectively than Spring WebFlux, with the blocking
-engines scaling up to 128 concurrent users and achieving 10,000 requests per
+Figure~\ref{fig:presentations-quarkus-jmeter}, indicate that Quarkus handles
+synchronous approaches more efficiently than Spring WebFlux. The blocking
+engines scale up to 128 concurrent users, achieving up to 10,000 requests per
 second. While the use of Virtual Threads results in a slightly higher
-throughput, the difference is not significant, as both approaches deliver
-similar performance.
+throughput, the difference is negligible, with both approaches delivering
+comparable performance.
+
+Additionally, \textit{HtmlFlow-Rx}, a reactive implementation of the HtmlFlow
+template engine that utilizes Quarkus's reactive programming model, achieved
+throughput similar to the \textit{Blocking} and \textit{Virtual}
+approaches—approximately 10,000 requests per second. This demonstrates that
+Quarkus's reactive programming model is effective for enabling PSSR\@. However,
+there is no significant difference in performance between the reactive and
+synchronous approaches, indicating that the reactive model does not provide a
+substantial advantage over synchronous approaches in this context.
 
@@
 \begin{figure}[h]
-     \isPreprints{\centering}{}
+     \centering
      \includegraphics[width=0.8\textwidth]{./Graphs/stocks-webflux-jmeter.png}
@@
 \begin{figure}[h]
-     \isPreprints{\centering}{}
+     \centering
      \includegraphics[width=0.8\textwidth]{./Graphs/stocks-springmvc-jmeter.png}
@@
 \begin{figure}[h]
-     \isPreprints{\centering}{}
+     \centering
      \includegraphics[width=0.8\textwidth]{./Graphs/stocks-quarkus-jmeter.png}
@@
 performance comparable to the Spring WebFlux implementation. The blocking
-engines reach 6,000 requests per second. Again, there is no significant
-difference between the Virtual Threads and platform threads approaches, with
-both achieving similar results.
+engines reach approximately 5,500 requests per second, and once again, there is
+no significant difference between the \textit{Blocking} and \textit{Virtual}
+approaches, with both achieving similar results.
+
+In addition to the blocking engines, the \textit{HtmlFlow-Rx} approach also
+achieves a throughput of approximately 5,500 requests per second, matching the
+performance of the \textit{Blocking} and \textit{Virtual} approaches. This
+further demonstrates that there is no significant difference in performance
+between the reactive and synchronous approaches in this context.
 
@@
 to scale as effectively as those using reactive programming or Kotlin
-coroutines, allowing \textit{external} DSLs to be used in a non-blocking context, 
-effectively enabling PSSR\@.
+coroutines, allowing \textit{external} DSLs to be used in a non-blocking
+context, effectively enabling PSSR\@.

### 2025-05-29 docs/article/Sections/conclusion.tex
article 50%~ completed
Completed Article disclaimers

--- 2025-04-30 docs/progress-report/sections/conclusion.tex
@@
-\documentclass[../apG48.tex]{subfiles}
+\section{Conclusion}
 
-\begin{document}
+In recent decades, non-blocking I/O has become the standard approach for building
+highly responsive and scalable web servers. However, traditional synchronous
+programming models are not compatible with non-blocking APIs, which typically
+rely on callback-based conventions such as
+\textit{continuation-passing style} (CPS)~\cite{scheme} or
+\textit{promises}~\cite{promise}. These approaches not only hinder
+sequential readability but also increase code verbosity, making them more
+error-prone.
+Alternatives like the \texttt{async}/\texttt{await} idiom~\cite{async_await}
+or \textit{suspend functions}~\cite{elizarov2021coroutines} simplify
+asynchronous programming by mimicking a sequential style without blocking
+threads.
+Recent proposals~\cite{carvalho2023async,wise2024pssr} have applied contemporary
+asynchronous idioms to SSR web templates, demonstrating how they can overcome
+the scalability bottlenecks present in traditional web template engines.
 
-Through benchmarking across Spring WebFlux, Spring MVC, and Quarkus, we evaluated eight different template engines and found that non-blocking implementations—especially those using virtual threads—consistently deliver performance on par with traditional blocking approaches under high concurrency.
-These results highlight virtual threads as a promising alternative to complex asynchronous programming models, offering a simpler development experience without sacrificing scalability or responsiveness.
-
-The next steps include a deeper analysis of the benchmark results in the final report, as well as expanded coverage of the current state of non-blocking PSSR and asynchronous programming techniques.
-Additionally, the project may be extended to support other frameworks such as Vert.x.
-
-\end{document}
+As an alternative, Java Virtual Threads can be applied to any blocking I/O call,
+leveraging the Java runtime to transparently intercept and convert it into non-blocking I/O,
+without requiring any changes to the calling code or exposing its internal complexities.
+In this work, we explored how this technique can be applied to traditional
+web template engines and whether it can achieve performance competitive with
+reactive approaches provided by frameworks such as Thymeleaf~\cite{webflux} and HtmlFlow~\cite{htmlflow}.
+Our benchmarks across Spring WebFlux, Spring MVC, and Quarkus show that synchronous
+non-blocking execution using virtual threads consistently delivers performance
+comparable to asynchronous non-blocking approaches under high concurrency.
+These findings highlight virtual threads as a promising alternative to complex
+asynchronous programming models, offering a simpler development experience
+without compromising scalability or responsiveness.

### 2025-05-29 docs/article/Sections/introduction.tex
Completed Article disclaimers

@@
 engines, such as HtmlFlow~\cite{htmlflow} and Thymeleaf~\cite{thymeleaf}, have
-been designed with these capabilities in mind. However, many traditional
+been designed with these capabilities in mind. However, many legacy
 template engines—particularly those using external domain-specific languages
@@
 in Java 21, it is now possible to execute blocking I/O operations in a
-scalable, lightweight manner. This capability allows traditional template
+scalable, lightweight manner. This capability allows legacy template
 engines—often reliant on blocking interfaces—to operate efficiently in

### 2025-05-29 docs/article/Sections/problem-statement.tex
Highlight Reactive versus Suspend for HtmlFlow templates and cross ref thos Listings in Results analysis.

@@
     numbers=none,
-    caption={\textit{HtmlFlow} presentation template in Koltin with an \texttt{Observable} model},
+    caption={\textit{HtmlFlow reactive} presentation template in Koltin with an \texttt{Observable} model},
     label={lst:presentation-observable}
@@
     label={lst:presentation-flow},
-    caption={\textit{HtmlFlow} presentation template in Kotlin with a \texttt{Flow} model},
+    caption={\textit{HtmlFlow suspend} presentation template in Kotlin with a \texttt{Flow} model},
 ]

### 2025-05-29 docs/article/Sections/results.tex
Highlight Reactive versus Suspend for HtmlFlow templates and cross ref thos Listings in Results analysis.

@@
 concurrent users ranging from 1 to 128, from left to right. The benchmarks
-include HtmlFlow using suspendable web templates (HtmlFlow-Susp), Jstachio
-using Virtual Threads with the \texttt{Iterable} interface (Jstachio-Virtual),
-and Thymeleaf using the reactive View Resolver driver (Thymeleaf-Rx).
+include HtmlFlow using suspendable web templates (\textit{HtmlFlow-Susp},
+equivalent to the approach shown in Listing~\ref{lst:presentation-flow}),
+Jstachio using Virtual Threads with the \texttt{Iterable} interface
+(\textit{Jstachio-Virtual}), and Thymeleaf using the reactive View Resolver
+driver (Thymeleaf-Rx).
 \textit{Blocking} and \textit{Virtual} represent the average throughput of the
@@
 
-We show the HtmlFlow-Susp, Jstachio-Virtual, and Thymeleaf-Rx engines
+We show the \textit{HtmlFlow-Susp}, \textit{Jstachio-Virtual}, and \textit{Thymeleaf-Rx} engines
 separately to observe the performance of the non-blocking engines when using
@@
 architecture, neither of the asynchronous approaches—\textbf{Reactive} nor
-\textbf{Suspendable}—described in Section~5 are applicable. Regarding the
+\textbf{Suspendable}—described in Section~\ref{sec:bench} are applicable. Regarding the
 synchronous strategy, both \textit{Blocking} and \textit{Virtual} scale
@@
 Additionally, \textit{HtmlFlow-Rx}, a reactive implementation of the HtmlFlow
-template engine that utilizes Quarkus's reactive programming model, achieved
-throughput similar to the \textit{Blocking} and \textit{Virtual}
+template engine (equivalent to the approach shown in
+Listing~\ref{lst:presentation-observable}) that utilizes Quarkus's reactive
+programming model, achieved throughput similar to the \textit{Blocking} and
+\textit{Virtual}
 approaches—approximately 10,000 requests per second. This demonstrates that

### 2025-05-29 benches/ab/bench-mvc.sh
Added benchmark workflows; Changed graphs x axis to Approaches

@@ -34,6 +34,7 @@ ROUTES=(
 #   stocks/velocity
    stocks/thymeleaf
    stocks/htmlFlow
+   stocks/kotlinx
 )
 
 #
@@ -55,14 +56,12 @@ echo "##########################################"
 ./run-ab.sh "${ROUTES[@]}" | tee spring-mvc-results.log
 
 
-# Gracefully terminate the Spring Boot application when running on local machine.
+# Gracefully terminate the Spring Boot application.
 # It will send a SIGTERM corresponding to Exit code 143.
-if [ "$GH" != "true" ]; then
-  kill $PID_GRADLE
+kill $PID_GRADLE
 
-  # Wait for the process to exit
-  wait $PID_GRADLE
-fi
+# Wait for the process to exit
+wait $PID_GRADLE
 
 echo ":::::::::::::::::::::::::::::::     Sync Bench Done"
 

### 2025-05-29 benches/ab/bench-quarkus.sh

@@ -19,23 +19,26 @@ echo ":::::::::::::::::::::::::::::::     Quarkus running PID = $PID_QUARKUS"
 # Define routes for benchmark
 #
 ROUTES=(
-  presentations/rocker
-  presentations/jstachio
-  presentations/pebble
-  presentations/freemarker
-  presentations/trimou
-#  presentations/velocity
-  presentations/thymeleaf
-  presentations/htmlFlow
-  presentations/kotlinx
-  stocks/rocker
-  stocks/jstachio
-  stocks/pebble
-  stocks/freemarker
-  stocks/trimou
-#  stocks/velocity
-  stocks/thymeleaf
-  stocks/htmlFlow
+   presentations/rocker
+   presentations/jstachio
+   presentations/pebble
+   presentations/freemarker
+   presentations/trimou
+#   presentations/velocity
+   presentations/thymeleaf
+   presentations/htmlFlow
+   presentations/kotlinx
+   presentations/reactive/htmlFlow
+   stocks/rocker
+   stocks/jstachio
+   stocks/pebble
+   stocks/freemarker
+   stocks/trimou
+#   stocks/velocity
+   stocks/thymeleaf
+   stocks/htmlFlow
+   stocks/kotlinx
+   stocks/reactive/htmlFlow
 )
 
 #
@@ -57,15 +60,14 @@ echo "##########################################"
 ./run-ab.sh "${ROUTES[@]}" | tee quarkus-results.log
 
 
-# Gracefully terminate the Spring Boot application when running on local machine.
+# Gracefully terminate the Spring Boot application.
 # It will send a SIGTERM corresponding to Exit code 143.
-if [ "$GH" != "true" ]; then
-  kill $PID_GRADLE
-  kill $PID_QUARKUS
+kill $PID_GRADLE
+kill $PID_QUARKUS
 
-  # Wait for the process to exit
-  wait $PID_GRADLE
-fi
+# Wait for the process to exit
+wait $PID_GRADLE
+
 
 echo ":::::::::::::::::::::::::::::::     Sync Bench Done"
 

### 2025-05-29 benches/ab/bench-webflux.sh

@@ -48,14 +48,14 @@ ROUTES=(
 #   presentations/velocity/sync
 #   presentations/velocity/virtualSync
    presentations/thymeleaf
-    stocks/thymeleaf
+   stocks/thymeleaf
    stocks/thymeleaf/sync
    stocks/thymeleaf/virtualSync
-#   stocks/htmlFlow
+   stocks/htmlFlow
    stocks/htmlFlow/suspending
    stocks/htmlFlow/sync
    stocks/htmlFlow/virtualSync
-#   stocks/kotlinx
+   stocks/kotlinx
    stocks/kotlinx/sync
    stocks/kotlinx/virtualSync
    stocks/rocker/sync

### 2025-05-29 benches/ab/run-ab.sh

@@ -9,7 +9,7 @@ for ths in "${THREADS[@]}"; do   # For each number of threads
     for ((n=0;n<$ITERATIONS;n++)); do
       total_requests=$((REQUESTS * ths))
       result=$(ab -q -s 240 -n $total_requests -c $ths http://localhost:8080/$path | grep "Requests per second")
-      echo ":::::::::::::::::::::::::::::::     $path:$ths:$total_requests:$result"
+      echo "$path::$ths:$result"
     done
   done
 done

### 2025-05-29 benches/jmeter/bench-mvc.sh

@@ -34,12 +34,45 @@ ROUTES=(
 #   stocks/velocity
    stocks/thymeleaf
    stocks/htmlFlow
+   stocks/kotlinx
 )
 
 echo "##########################################"
 echo "############# RUN BENCH ##################"
 echo "##########################################"
 ./run-jmeter.sh "${ROUTES[@]}" | tee spring-mvc-results.log
+
+# Gracefully terminate the Spring Boot application.
+# It will send a SIGTERM corresponding to Exit code 143.
+kill $PID_GRADLE
+kill "$PID_MVC"
+
+# Wait for the process to exit
+wait $PID_GRADLE
+
+
+cd ../../code || exit
+
+./gradlew runMVCVirtual -DbenchTimeout=1 -DXms=1024m -DXmx=16G > ../benches/jmeter/spring-mvc.log &
+
+PID_GRADLE=$!
+
+cd ../benches/jmeter || exit
+
+sleep 1
+while ! grep -m1 'Tomcat started on port 8080' < spring-mvc.log; do
+    sleep 1
+done
+
+PID_MVC=$(grep -oP 'with PID \K[0-9]+' spring-mvc.log)
+
+echo ":::::::::::::::::::::::::::::::     Spring running PID = $PID_MVC"
+echo ":::::::::::::::::::::::::::::::     Gradle running PID = $PID_GRADLE"
+
+echo "##########################################"
+echo "############# RUN BENCH ##################"
+echo "##########################################"
+./run-jmeter.sh "${ROUTES[@]}" | tee spring-mvc-virtual-results.log
 
 # Gracefully terminate the Spring Boot application when running on local machine.
 # It will send a SIGTERM corresponding to Exit code 143.
@@ -50,37 +83,4 @@ if [ "$GH" != "true" ]; then
   # Wait for the process to exit
   wait $PID_GRADLE
 fi
-
-#cd ../../code || exit
-#
-#./gradlew runMVCVirtual -DbenchTimeout=1 -DXms=1024m -DXmx=16G > ../benches/jmeter/spring-mvc.log &
-#
-#PID_GRADLE=$!
-#
-#cd ../benches/jmeter || exit
-#
-#sleep 1
-#while ! grep -m1 'Tomcat started on port 8080' < spring-mvc.log; do
-#    sleep 1
-#done
-#
-#PID_MVC=$(grep -oP 'with PID \K[0-9]+' spring-mvc.log)
-#
-#echo ":::::::::::::::::::::::::::::::     Spring running PID = $PID_MVC"
[…]

### 2025-05-29 benches/jmeter/bench-quarkus.sh

@@ -25,6 +25,7 @@ ROUTES=(
   presentations/thymeleaf
   presentations/htmlFlow
   presentations/kotlinx
+  presentations/reactive/htmlFlow
   stocks/rocker
   stocks/jstachio
   stocks/pebble
@@ -33,6 +34,8 @@ ROUTES=(
 #  stocks/velocity
   stocks/thymeleaf
   stocks/htmlFlow
+  stocks/kotlinx
+  stocks/reactive/htmlFlow
 )
 
 #
@@ -44,15 +47,14 @@ echo "##########################################"
 ./run-jmeter.sh "${ROUTES[@]}" | tee quarkus-results.log
 
 
-# Gracefully terminate the Spring Boot application when running on local machine.
+# Gracefully terminate the Quarkus application.
 # It will send a SIGTERM corresponding to Exit code 143.
-if [ "$GH" != "true" ]; then
-  kill $PID_GRADLE
-  kill $PID_QUARKUS
+kill $PID_GRADLE
+kill $PID_QUARKUS
 
-  # Wait for the process to exit
-  wait $PID_GRADLE
-fi
+# Wait for the process to exit
+wait $PID_GRADLE
+
 
 echo ":::::::::::::::::::::::::::::::     Sync Bench Done"
 

### 2025-05-29 benches/jmeter/bench-webflux.sh

@@ -37,11 +37,11 @@ ROUTES=(
   stocks/thymeleaf
   stocks/thymeleaf/sync
   stocks/thymeleaf/virtualSync
-#  stocks/htmlFlow
+  stocks/htmlFlow
   stocks/htmlFlow/suspending
   stocks/htmlFlow/sync
   stocks/htmlFlow/virtualSync
-#  stocks/kotlinx
+  stocks/kotlinx
   stocks/kotlinx/sync
   stocks/kotlinx/virtualSync
   stocks/rocker/sync

### 2025-05-29 results/plot-quarkus.sh

@@ -1,0 +1,53 @@
+#!/bin/bash
+
+INPUT_BLOCKING="quarkus-results.log"
+INPUT_VIRTUAL="quarkus-virtual-results.log"
+
+PRESENTATIONS_CSV="presentations-quarkus.csv"
+STOCKS_CSV="stocks-quarkus.csv"
+
+TMP_PRESENTATIONS=$(mktemp)
+TMP_STOCKS=$(mktemp)
+
+extract_data() {
+  local file="$1"
+  local label="$2"
+
+  while IFS= read -r line; do
+    [[ -z "$line" ]] && continue
+
+    route=$(echo "$line" | cut -d':' -f1)
+    concurrency=$(echo "$line" | cut -d':' -f3 | awk '{print $1}')
+    req_per_sec=$(echo "$line" | grep -oP '[0-9.]+(?= req/s)')
+
+    if [[ "$route" == *"/reactive/htmlFlow"* ]]; then
+      label_out="HtmlFlow-Rx"
+    else
+      label_out="$label"
+      engine=$(echo "$route" | cut -d'/' -f2)
+    fi
+
+    if [[ "$route" == presentations/* ]]; then
+      echo "$engine,$concurrency,$req_per_sec,$label_out" >> "$TMP_PRESENTATIONS"
+    elif [[ "$route" == stocks/* ]]; then
+      echo "$engine,$concurrency,$req_per_sec,$label_out" >> "$TMP_STOCKS"
+    fi
+  done < "$file"
+}
+
+extract_data "$INPUT_BLOCKING" "Blocking"
+extract_data "$INPUT_VIRTUAL" "Virtual"
+
+./aggregate.sh "$TMP_PRESENTATIONS" "$PRESENTATIONS_CSV"
+./aggregate.sh "$TMP_STOCKS" "$STOCKS_CSV"
+
+# Clean up
+rm "$TMP_PRESENTATIONS" "$TMP_STOCKS"
+
+echo "Generated wide format CSV files:"
+echo "- $PRESENTATIONS_CSV"
+echo "- $STOCKS_CSV"
+
+gnuplot plots/quarkus.plot
+
+echo "Generated plots using gnuplot."

### 2025-05-29 benches/ab/bench-webflux.sh
Remove kotlinX reactive approach from webflux bench

@@ -55,7 +55,7 @@ ROUTES=(
    stocks/htmlFlow/suspending
    stocks/htmlFlow/sync
    stocks/htmlFlow/virtualSync
-   stocks/kotlinx
+#   stocks/kotlinx
    stocks/kotlinx/sync
    stocks/kotlinx/virtualSync
    stocks/rocker/sync

### 2025-05-29 benches/jmeter/bench-webflux.sh

@@ -41,7 +41,7 @@ ROUTES=(
   stocks/htmlFlow/suspending
   stocks/htmlFlow/sync
   stocks/htmlFlow/virtualSync
-  stocks/kotlinx
+#  stocks/kotlinx
   stocks/kotlinx/sync
   stocks/kotlinx/virtualSync
   stocks/rocker/sync

### 2025-05-29 benches/ab/bench-mvc.sh
Fix quarkus workflow gradle build

@@ -1,6 +1,6 @@
 cd ../../code || exit
 
-./gradlew runMVC -DbenchTimeout=1 -DXms=1024m -DXmx=16G > ../benches/ab/spring-mvc.log &
+./gradlew runMVC -DbenchTimeout=1 -Dorg.gradle.jvmargs="-Xms512M -Xmx16g" > ../benches/ab/spring-mvc.log &
 PID_GRADLE=$!
 
 cd ../benches/ab || exit
@@ -67,7 +67,7 @@ echo ":::::::::::::::::::::::::::::::     Sync Bench Done"
 
 cd ../../code || exit
 
-./gradlew runMVCVirtual -DbenchTimeout=1 -DXms=1024m -DXmx=16G > ../benches/ab/spring-mvc.log &
+./gradlew runMVCVirtual -DbenchTimeout=1 -Dorg.gradle.jvmargs="-Xms512M -Xmx16g" > ../benches/ab/spring-mvc.log &
 PID_GRADLE=$!
 
 cd ../benches/ab || exit

### 2025-05-29 benches/ab/bench-quarkus.sh

@@ -1,6 +1,6 @@
 cd ../../code || exit
 
-./gradlew runQuarkus -DbenchTimeout=1 -DXms=1024m -DXmx=16G -Djdk.tracePinnedThreads > ../benches/ab/quarkus.log &
+./gradlew runQuarkus -DbenchTimeout=1 -Dorg.gradle.jvmargs="-Xms512M -Xmx16g" -Djdk.tracePinnedThreads > ../benches/ab/quarkus.log &
 PID_GRADLE=$!
 
 cd ../benches/ab || exit
@@ -73,7 +73,7 @@ echo ":::::::::::::::::::::::::::::::     Sync Bench Done"
 
 cd ../../code || exit
 
-./gradlew runQuarkusVirtual -DbenchTimeout=1 -DXms=1024m -DXmx=16G > ../benches/ab/quarkus.log &
+./gradlew runQuarkusVirtual -Dorg.gradle.jvmargs="-Xms512M -Xmx16g" -DbenchTimeout=1 > ../benches/ab/quarkus.log &
 PID_GRADLE=$!
 
 cd ../benches/ab || exit

### 2025-05-29 benches/ab/bench-webflux.sh

@@ -6,7 +6,7 @@
 
 cd ../../code || exit
 
-./gradlew runWebflux -DbenchTimeout=1 -DXms=1024m -DXmx=16G > ../benches/ab/spring-webflux.log &
+./gradlew runWebflux -DbenchTimeout=1 -Dorg.gradle.jvmargs="-Xms512M -Xmx16g" > ../benches/ab/spring-webflux.log &
 PID_GRADLE=$!
 
 cd ../benches/ab || exit

### 2025-05-29 benches/jmeter/bench-mvc.sh

@@ -1,6 +1,6 @@
 cd ../../code || exit
 
-./gradlew runMVC -DbenchTimeout=1 -DXms=1024m -DXmx=16G > ../benches/jmeter/spring-mvc.log &
+./gradlew runMVC -DbenchTimeout=1 -Dorg.gradle.jvmargs="-Xms512M -Xmx16g" > ../benches/jmeter/spring-mvc.log &
 
 PID_GRADLE=$!
 
@@ -53,7 +53,7 @@ wait $PID_GRADLE
 
 cd ../../code || exit
 
-./gradlew runMVCVirtual -DbenchTimeout=1 -DXms=1024m -DXmx=16G > ../benches/jmeter/spring-mvc.log &
+./gradlew runMVCVirtual -DbenchTimeout=1 -Dorg.gradle.jvmargs="-Xms512M -Xmx16g" > ../benches/jmeter/spring-mvc.log &
 
 PID_GRADLE=$!
 

### 2025-05-29 benches/jmeter/bench-quarkus.sh

@@ -1,6 +1,6 @@
 cd ../../code || exit
 
-./gradlew runQuarkus -DbenchTimeout=1 -DXms=1024m -DXmx=16G -Djdk.tracePinnedThreads > ../benches/jmeter/quarkus.log &
+./gradlew runQuarkus -DbenchTimeout=1 -Dorg.gradle.jvmargs="-Xms512M -Xmx16g" -Djdk.tracePinnedThreads > ../benches/jmeter/quarkus.log &
 PID_GRADLE=$!
 
 cd ../benches/jmeter || exit
@@ -60,7 +60,7 @@ echo ":::::::::::::::::::::::::::::::     Sync Bench Done"
 
 cd ../../code || exit
 
-./gradlew runQuarkusVirtual -DbenchTimeout=1 -DXms=1024m -DXmx=16G > ../benches/jmeter/quarkus.log &
+./gradlew runQuarkusVirtual -DbenchTimeout=1 -Dorg.gradle.jvmargs="-Xms512M -Xmx16g" > ../benches/jmeter/quarkus.log &
 PID_GRADLE=$!
 
 cd ../benches/jmeter || exit

### 2025-05-29 benches/jmeter/bench-webflux.sh

@@ -1,6 +1,6 @@
 cd ../../code || exit
 
-./gradlew runWebflux -DbenchTimeout=1 -DXms=1024m -DXmx=16G > ../benches/jmeter/spring-webflux.log &
+./gradlew runWebflux -DbenchTimeout=1 -Dorg.gradle.jvmargs="-Xms512M -Xmx16g" > ../benches/jmeter/spring-webflux.log &
 
 PID_GRADLE=$!
 

### 2025-05-29 benches/ab/bench-webflux.sh
Remove webflux kotlinx reactive presentations bench

@@ -32,7 +32,7 @@ ROUTES=(
    presentations/htmlFlow/suspending
    presentations/htmlFlow/sync
    presentations/htmlFlow/virtualSync
-   presentations/kotlinx
+#   presentations/kotlinx
    presentations/kotlinx/sync
    presentations/kotlinx/virtualSync
    presentations/rocker/sync

### 2025-05-29 benches/jmeter/bench-webflux.sh

@@ -19,7 +19,7 @@ ROUTES=(
   presentations/htmlFlow/suspending
   presentations/htmlFlow/sync
   presentations/htmlFlow/virtualSync
-  presentations/kotlinx
+#  presentations/kotlinx
   presentations/kotlinx/sync
   presentations/kotlinx/virtualSync
   presentations/rocker/sync

### 2025-05-29 benches/jmeter/bench-quarkus.sh
fix quarkus bench workflow artifact names

@@ -81,7 +81,7 @@ echo ":::::::::::::::::::::::::::::::     Quarkus running PID = $PID_QUARKUS"
 echo "##########################################"
 echo "############# RUN BENCH ##################"
 echo "##########################################"
-./run-jmeter.sh "${ROUTES[@]}" | tee quarkus-virtual-results.log
+./run-jmeter.sh "${ROUTES[@]}" | tee quarkus-results-virtual.log
 
 if [ "$GH" != "true" ]; then
   kill $PID_GRADLE

### 2025-05-29 benches/ab/bench-webflux.sh
reduce number of blocking routes in webflux bench

@@ -26,47 +26,47 @@ echo ":::::::::::::::::::::::::::::::     Gradle running PID = $PID_GRADLE"
 #
 ROUTES=(
    presentations/thymeleaf
-   presentations/thymeleaf/sync
+#   presentations/thymeleaf/sync
    presentations/thymeleaf/virtualSync
    presentations/htmlFlow
    presentations/htmlFlow/suspending
    presentations/htmlFlow/sync
    presentations/htmlFlow/virtualSync
 #   presentations/kotlinx
-   presentations/kotlinx/sync
+#   presentations/kotlinx/sync
    presentations/kotlinx/virtualSync
    presentations/rocker/sync
    presentations/rocker/virtualSync
    presentations/jstachio/sync
    presentations/jstachio/virtualSync
-   presentations/pebble/sync
+#   presentations/pebble/sync
    presentations/pebble/virtualSync
-   presentations/freemarker/sync
+#   presentations/freemarker/sync
    presentations/freemarker/virtualSync
-   presentations/trimou/sync
+#   presentations/trimou/sync
    presentations/trimou/virtualSync
 #   presentations/velocity/sync
 #   presentations/velocity/virtualSync
    presentations/thymeleaf
    stocks/thymeleaf
-   stocks/thymeleaf/sync
+#   stocks/thymeleaf/sync
    stocks/thymeleaf/virtualSync
    stocks/htmlFlow
    stocks/htmlFlow/suspending
    stocks/htmlFlow/sync
    stocks/htmlFlow/virtualSync
 #   stocks/kotlinx
-   stocks/kotlinx/sync
+#   stocks/kotlinx/sync
    stocks/kotlinx/virtualSync
    stocks/rocker/sync
    stocks/rocker/virtualSync
    stocks/jstachio/sync
    stocks/jstachio/virtualSync
-   stocks/pebble/sync
+#   stocks/pebble/sync
    stocks/pebble/virtualSync
-   stocks/freemarker/sync
+#   stocks/freemarker/sync
    stocks/freemarker/virtualSync
-   stocks/trimou/sync
+#   stocks/trimou/sync
    stocks/trimou/virtualSync
 #   stocks/velocity/sync
 #   stocks/velocity/virtualSync

### 2025-05-29 benches/jmeter/bench-webflux.sh

@@ -13,49 +13,50 @@ done
 
 ROUTES=(
   presentations/thymeleaf
-  presentations/thymeleaf/sync
-  presentations/thymeleaf/virtualSync
-  presentations/htmlFlow
-  presentations/htmlFlow/suspending
-  presentations/htmlFlow/sync
-  presentations/htmlFlow/virtualSync
-#  presentations/kotlinx
-  presentations/kotlinx/sync
-  presentations/kotlinx/virtualSync
-  presentations/rocker/sync
-  presentations/rocker/virtualSync
-  presentations/jstachio/sync
-  presentations/jstachio/virtualSync
-  presentations/pebble/sync
-  presentations/pebble/virtualSync
-  presentations/freemarker/sync
-  presentations/freemarker/virtualSync
-  presentations/trimou/sync
-  presentations/trimou/virtualSync
-#  presentations/velocity/sync
-#  presentations/velocity/virtualSync
-  stocks/thymeleaf
-  stocks/thymeleaf/sync
-  stocks/thymeleaf/virtualSync
-  stocks/htmlFlow
-  stocks/htmlFlow/suspending
-  stocks/htmlFlow/sync
-  stocks/htmlFlow/virtualSync
-#  stocks/kotlinx
-  stocks/kotlinx/sync
-  stocks/kotlinx/virtualSync
-  stocks/rocker/sync
-  stocks/rocker/virtualSync
-  stocks/jstachio/sync
-  stocks/jstachio/virtualSync
-  stocks/pebble/sync
-  stocks/pebble/virtualSync
-  stocks/freemarker/sync
-  stocks/freemarker/virtualSync
-  stocks/trimou/sync
-  stocks/trimou/virtualSync
-#  stocks/velocity/sync
-#  stocks/velocity/virtualSync
+  #   presentations/thymeleaf/sync
+     presentations/thymeleaf/virtualSync
+     presentations/htmlFlow
+     presentations/htmlFlow/suspending
+     presentations/htmlFlow/sync
+     presentations/htmlFlow/virtualSync
+  #   presentations/kotlinx
+  #   presentations/kotlinx/sync
+     presentations/kotlinx/virtualSync
+     presentations/rocker/sync
+     presentations/rocker/virtualSync
+     presentations/jstachio/sync
+     presentations/jstachio/virtualSync
+  #   presentations/pebble/sync
+     presentations/pebble/virtualSync
[…]

### 2025-05-30 benches/ab/bench-webflux.sh
remove freemarker from webflux bench

@@ -42,7 +42,7 @@ ROUTES=(
 #   presentations/pebble/sync
    presentations/pebble/virtualSync
 #   presentations/freemarker/sync
-   presentations/freemarker/virtualSync
+#   presentations/freemarker/virtualSync
 #   presentations/trimou/sync
    presentations/trimou/virtualSync
 #   presentations/velocity/sync
@@ -65,7 +65,7 @@ ROUTES=(
 #   stocks/pebble/sync
    stocks/pebble/virtualSync
 #   stocks/freemarker/sync
-   stocks/freemarker/virtualSync
+#   stocks/freemarker/virtualSync
 #   stocks/trimou/sync
    stocks/trimou/virtualSync
 #   stocks/velocity/sync

### 2025-05-30 benches/jmeter/bench-webflux.sh

@@ -29,7 +29,7 @@ ROUTES=(
   #   presentations/pebble/sync
      presentations/pebble/virtualSync
   #   presentations/freemarker/sync
-     presentations/freemarker/virtualSync
+#     presentations/freemarker/virtualSync
   #   presentations/trimou/sync
      presentations/trimou/virtualSync
   #   presentations/velocity/sync
@@ -52,7 +52,7 @@ ROUTES=(
   #   stocks/pebble/sync
      stocks/pebble/virtualSync
   #   stocks/freemarker/sync
-     stocks/freemarker/virtualSync
+#     stocks/freemarker/virtualSync
   #   stocks/trimou/sync
      stocks/trimou/virtualSync
   #   stocks/velocity/sync

### 2025-05-30 docs/article/Sections/results.tex
Update article results to Github workflow results

@@
 
-All the following benchmarks were conducted on a local machine running Ubuntu
-22.04 LTS with a 6-core, 12-thread CPU and 32GB of RAM\@. All tests were
-conducted on the OpenJDK VM Corretto 21. The JVM was configured with a minimum
-heap size of 1024MB and a maximum heap size of 16GB\@.
+All the following benchmarks were conducted on a Github-hosted virtual machine
+under GitHub Actions, running Ubuntu 22.04 LTS with a 4-core, 8-thread CPU and
+16GB of RAM\@. All tests were conducted on the OpenJDK VM Corretto 21. The JVM
+was configured with a minimum heap size of 512MB and a maximum heap size of
+16GB\@.
 
@@
 (\textit{Jstachio-Virtual}), and Thymeleaf using the reactive View Resolver
-driver (Thymeleaf-Rx).
-\textit{Blocking} and \textit{Virtual} represent the average throughput of the
-blocking approaches (i.e., KotlinX, Rocker, Jstachio, Pebble, Freemarker,
-Trimou, HtmlFlow, and Thymeleaf) when run in the context of a separate
-coroutine dispatcher or Virtual Threads, respectively.
+driver (Thymeleaf-Rx). \textit{Blocking} and \textit{Virtual} represent the
+average throughput of the blocking approaches (i.e., KotlinX, Rocker, Jstachio,
+Pebble, Freemarker, Trimou, HtmlFlow, and Thymeleaf) when run in the context of
+a separate coroutine dispatcher or Virtual Threads, respectively.
 
-We show the \textit{HtmlFlow-Susp}, \textit{Jstachio-Virtual}, and \textit{Thymeleaf-Rx} engines
-separately to observe the performance of the non-blocking engines when using
-the Suspending, Virtual Thread, and Reactive approaches. The \textit{Blocking}
-and \textit{Virtual} are aggregated due to the similar performance of different
-engines when using those approaches.
+We show the \textit{HtmlFlow-Susp}, \textit{Jstachio-Virtual}, and
+\textit{Thymeleaf-Rx} engines separately to observe the performance of the
+non-blocking engines when using the Suspending, Virtual Thread, and Reactive
+approaches. The \textit{Blocking} and \textit{Virtual} are aggregated due to
+the similar performance of different engines when using those approaches.
 
@@
 
-The results show that when using the blocking template engines with a separate
-coroutine dispatcher, the engines are unable to scale effectively beyond 16
-concurrent users. In contrast, the non-blocking engines scale effectively up to
-128 concurrent users, with HtmlFlow achieving approximately 11,000 requests per
-second. When using the blocking approaches in the context of Virtual Threads
-(achieving non-blocking I/O), the engines scale effectively up to 128
-concurrent users, with Jstachio matching HtmlFlow`s performance at
-approximately 11,000 requests per second. Thymeleaf, using the reactive View
-Resolver driver, also scales to 128 users, albeit less effectively, achieving
-around 8,500 requests per second.
+The results in Figure~\ref{fig:presentations-webflux-jmeter} show that when
+using blocking template engines with a separate coroutine dispatcher, the
+engines are unable to scale effectively beyond 4 concurrent users. In contrast,
+the non-blocking engines scale efficiently up to 128 concurrent users, with
+HtmlFlow achieving 4,487 requests per second. When blocking approaches are
+executed in the context of Virtual Threads—thus enabling non-blocking I/O—the
+engines also scale up to 128 concurrent users, with Jstachio using Virtual
+Threads reaching 3,514 requests per second. The Thymeleaf implementation using
+the reactive View Resolver driver also scales up to 128 concurrent users, but
+achieves a lower throughput of approximately 2,559 requests per second.
+
+It is important to note that the differences in scalability and throughput
+between the \textit{HtmlFlow-Susp}, \textit{Jstachio-Virtual}, and
+\textit{Thymeleaf-Rx} approaches may be influenced by the specific template
+engines used, rather than the approach itself. When comparing the
+\textit{Reactive}, \textit{Suspendable}, and \textit{Virtual} approaches
+specifically with HtmlFlow, we found that all three achieve similar
+performance: HtmlFlow using a blocking approach with Virtual Threads reaches
+4,691 requests per second, while HtmlFlow using a reactive approach achieves
+4,792 requests per second.
 
@@
 % implementation.
-
 The results for the Spring MVC implementation, shown in
@@
 \texttt{StreamingResponseBody}, and \textit{Virtual}, which uses Virtual
-Threads. Note that for Spring MVC, which follows a thread-per-request
-architecture, neither of the asynchronous approaches—\textbf{Reactive} nor
-\textbf{Suspendable}—described in Section~\ref{sec:bench} are applicable. Regarding the
-synchronous strategy, both \textit{Blocking} and \textit{Virtual} scale
-effectively up to 128 concurrent users, with the Virtual Threads approach
-achieving a slightly higher throughput of approximately 9,000 requests per
-second. However, these values are slightly lower than those observed in the
-Spring WebFlux implementation.
+Threads. Since Spring MVC follows a thread-per-request architecture, the
+asynchronous approaches—\textbf{Reactive} and \textbf{Suspendable}—described in
+Section~\ref{sec:bench} are not applicable. Both the \textit{Blocking} and
+\textit{Virtual} strategies scale effectively up to 128 concurrent users, with
+the Virtual Threads approach achieving a slightly higher maximum throughput of
+2,797 requests per second, compared to 2,498 requests per second for the
+blocking approach. These results indicate that while Spring MVC can handle a
+moderate level of concurrency, it does not reach the scalability of the
+reactive or suspendable approaches available in Spring WebFlux. Furthermore,
+Spring MVC does not enable Progressive Server-Side Rendering (PSSR) for the
+tested templates, as previously discussed in Section~\ref{sec:bench}.
 
@@
 synchronous approaches more efficiently than Spring WebFlux. The blocking
-engines scale up to 128 concurrent users, achieving up to 10,000 requests per
-second. While the use of Virtual Threads results in a slightly higher
-throughput, the difference is negligible, with both approaches delivering
-comparable performance.
+engines scale up to 128 concurrent users, achieving up to 3744 requests per
+second. When using Virtual Threads, the throughput increases even further,
+reaching 4856 requests per second. This demonstrates that Quarkus's
+implementation of Virtual Threads is effective for enabling PSSR\@, and
+comparable to the \textit{Suspendable} and \textit{Reactive} used approaches in
+Spring WebFlux in terms of scalability and throughput.
 
@@
 Listing~\ref{lst:presentation-observable}) that utilizes Quarkus's reactive
-programming model, achieved throughput similar to the \textit{Blocking} and
-\textit{Virtual}
-approaches—approximately 10,000 requests per second. This demonstrates that
-Quarkus's reactive programming model is effective for enabling PSSR\@. However,
-there is no significant difference in performance between the reactive and
-synchronous approaches, indicating that the reactive model does not provide a
-substantial advantage over synchronous approaches in this context.
+programming model, achieved a lower throughput than the the \textit{Blocking}
+and \textit{Virtual} approaches—3088 requests per second. This demonstrates
+that Quarkus's reactive programming model is effective for enabling PSSR\@,
+although it does not achieve the same level of performance as the same approach
+in Spring WebFlux.
 
@@
 engines and approaches as the previous benchmark, but replace the data model
-with the more complex Stock class, including 20 instances. Despite the
-increased complexity and number of instances, the scalability of the engines
-remains largely unaffected. However, throughput is reduced by approximately 50
-percent across all engines, with HtmlFlow achieving 6,000 requests per second.
-Since the \texttt{Stock} class\ contains more data properties than the
-\texttt{Presentation} class, the data binding to the template engine adds
-overhead, which explains the reduced performance. The Thymeleaf implementation
-using the reactive View Resolver driver reaches 5,000 requests per second,
-suggesting that the performance drop compared to the Presentation benchmark is
-not substantial.
+with the more complex Stock class, including 20 instances. When using this data
+model, we observe that the scalability of the engines is not significantly
+affected, however the throughput is reduced across all engines. When compared
+to the Presentation benchmark, we observe that particularly the approaches
+using Virtual Threads took a greater performance hit when compared to the
+\textit{Reactive} and \textit{Suspending} approaches, with Jstachio using
+Virtual Threads now achieving 1509 requests per second, compared to the 1,750
+requests per achieved by the Thymeleaf implementation using the reactive View
+Resolver driver.
+
+It is again important to note that the differences in scalability and
+throughput between the \textit{HtmlFlow-Susp}, \textit{Jstachio-Virtual}, and
+\textit{Thymeleaf-Rx} approaches may be influenced by the specific template
+engines used, rather than the approach itself. When comparing the
+\textit{Reactive}, \textit{Suspendable}, and \textit{Virtual} approaches
+specifically with HtmlFlow, we found that all three achieve similar
+performance: HtmlFlow using a blocking approach with Virtual Threads reaches
+3,090 requests per second, while HtmlFlow using a reactive approach achieves
+3,026 requests per second.
+
+The overall throughput reduction is expected, as the Stock class contains more
+data properties than the \texttt{Presentation} class, adding overhead related
+to the data binding process of each template engine.
 
@@
 
-The results observed in Figure~\ref{fig:stocks-springmvc-jmeter} show that the
+The results shown in Figure~\ref{fig:stocks-springmvc-jmeter} indicate that the
 Spring MVC implementation using the blocking approach with
-\texttt{StreamingResponseBody} achieves a throughput of up to 5500 requests per
-second, while no significant change is observed when using Virtual Threads. As
-such, both approaches scale effectively up to 128 concurrent users.
+\texttt{StreamingResponseBody} achieves a throughput of up to 1,916 requests
+per second, with no significant improvement observed when using Virtual
+Threads. Both approaches scale effectively up to 128 concurrent users. Although
+these approaches achieve higher throughput in Spring MVC than in Spring
+WebFlux, their overall performance remains lower than that of the reactive and
+suspendable approaches.
 
@@
 performance comparable to the Spring WebFlux implementation. The blocking
-engines reach approximately 5,500 requests per second, and once again, there is
-no significant difference between the \textit{Blocking} and \textit{Virtual}
-approaches, with both achieving similar results.
+approach reaches a throughput of 3,019 requests per second, while the Virtual
+Threads approach achieves a throughput of 3,357 requests per second.
 
 In addition to the blocking engines, the \textit{HtmlFlow-Rx} approach also
-achieves a throughput of approximately 5,500 requests per second, matching the
-performance of the \textit{Blocking} and \textit{Virtual} approaches. This
-further demonstrates that there is no significant difference in performance
-between the reactive and synchronous approaches in this context.
+achieves a throughput of 1760 requests per second, indicating that this
+approach achieves lower performance in Quarkus than in Spring WebFlux, where it
+reached 3,026 requests per second, as previously mentioned.
 

### 2025-05-30 docs/article/Sections/benchmark.tex
Highlight Reactive versus Suspend for HtmlFlow templates and cross ref thos Listings in Results analysis.
review sections 4 and 5

@@
 \section{Benchmark Implementation}
+\label{sec:bench}
 
-The benchmark\footnote{\url{https://github.com/fmcarvalho/comparing-non-blocking-progressive-ssr}}
+This benchmark\footnote{\url{https://github.com/xmlet/comparing-non-blocking-progressive-ssr}}
 is designed with a modular architecture, separating the \textit{view} and
@@
 shown in \autoref{lst:stock-class} and \autoref{lst:presentation-class}, 
-following the same model as
-\textit{spring-webflux-comparing-template-engines}~\footnote{\url{https://github.com/xmlet/spring-webflux-comparing-template-engines}}.
+following the proposal of the benchmarks~\cite{templatesbenchmark,springbenchmark}.
 The \texttt{Presentation} class represents a presentation with a title, speaker

### 2025-05-30 docs/article/Sections/results.tex
review sections 4 and 5

@@
 engines are unable to scale effectively beyond 4 concurrent users. In contrast,
-the non-blocking engines scale efficiently up to 128 concurrent users, with
-HtmlFlow achieving 4,487 requests per second. When blocking approaches are
-executed in the context of Virtual Threads—thus enabling non-blocking I/O—the
-engines also scale up to 128 concurrent users, with Jstachio using Virtual
-Threads reaching 3,514 requests per second. The Thymeleaf implementation using
-the reactive View Resolver driver also scales up to 128 concurrent users, but
-achieves a lower throughput of approximately 2,559 requests per second.
+\textit{HtmlFlow-Susp} scales up to 128 concurrent users, achieving 4,487
+requests per second. When blocking approaches are executed in the context of
+Virtual Threads—thus enabling non-blocking I/O—the engines scale up to 64
+concurrent users, with Jstachio using Virtual Threads reaching 3,514 requests
+per second. The Thymeleaf implementation using the reactive View Resolver
+driver scales up to 32 concurrent users, achieving a lower maximum throughput
+of 2,559 requests per second.
 
@@
 Section~\ref{sec:bench} are not applicable. Both the \textit{Blocking} and
-\textit{Virtual} strategies scale effectively up to 128 concurrent users, with
+\textit{Virtual} strategies scale effectively up to 32 concurrent users, with
 the Virtual Threads approach achieving a slightly higher maximum throughput of
@@
 synchronous approaches more efficiently than Spring WebFlux. The blocking
-engines scale up to 128 concurrent users, achieving up to 3744 requests per
+engines scale up to 64 concurrent users, achieving up to 3,744 requests per
 second. When using Virtual Threads, the throughput increases even further,
-reaching 4856 requests per second. This demonstrates that Quarkus's
-implementation of Virtual Threads is effective for enabling PSSR\@, and
-comparable to the \textit{Suspendable} and \textit{Reactive} used approaches in
-Spring WebFlux in terms of scalability and throughput.
+reaching 4,856 requests per second, allowing scalability up to 128 users. This
+demonstrates that Quarkus's implementation of Virtual Threads is effective for
+enabling PSSR\@, and comparable to the \textit{Suspendable} and
+\textit{Reactive} approaches used in Spring WebFlux in terms of scalability and
+throughput.
 
@@
 programming model, achieved a lower throughput than the the \textit{Blocking}
-and \textit{Virtual} approaches—3088 requests per second. This demonstrates
+and \textit{Virtual} approaches—3,088 requests per second. This demonstrates
 that Quarkus's reactive programming model is effective for enabling PSSR\@,
-although it does not achieve the same level of performance as the same approach
-in Spring WebFlux.
+although it does not achieve the same level of performance or scalability as
+the same approach in Spring WebFlux, stagnating after 32 concurrent users.
 
@@
 The results in Figure~\ref{fig:stocks-webflux-jmeter} use the same template
-engines and approaches as the previous benchmark, but replace the data model
-with the more complex Stock class, including 20 instances. When using this data
-model, we observe that the scalability of the engines is not significantly
-affected, however the throughput is reduced across all engines. When compared
-to the Presentation benchmark, we observe that particularly the approaches
-using Virtual Threads took a greater performance hit when compared to the
+engines and approaches as the previous benchmark, but with a more complex data
+model: the Stock class, which includes 20 instances and approximately two
+times as many data bindings. With this data model, the scalability of the
+engines remains largely unchanged; however, throughput is reduced across all
+engines. Compared to the Presentation benchmark, the approaches using Virtual
+Threads experience a more pronounced decrease in performance relative to the
 \textit{Reactive} and \textit{Suspending} approaches, with Jstachio using
-Virtual Threads now achieving 1509 requests per second, compared to the 1,750
-requests per achieved by the Thymeleaf implementation using the reactive View
-Resolver driver.
+Virtual Threads now achieving 1,509 requests per second, compared to 1,750
+requests per second achieved by the Thymeleaf implementation using the reactive
+View Resolver driver.
 
@@
 per second, with no significant improvement observed when using Virtual
-Threads. Both approaches scale effectively up to 128 concurrent users. Although
+Threads. Both approaches scale effectively up to 64 concurrent users. Although
 these approaches achieve higher throughput in Spring MVC than in Spring
@@
 The results depicted in Figure~\ref{fig:stocks-quarkus-jmeter} show that the
-Quarkus implementation scales effectively up to 128 concurrent users, achieving
-performance comparable to the Spring WebFlux implementation. The blocking
-approach reaches a throughput of 3,019 requests per second, while the Virtual
-Threads approach achieves a throughput of 3,357 requests per second.
-
-In addition to the blocking engines, the \textit{HtmlFlow-Rx} approach also
+Quarkus synchronous approaches scale effectively up to 128 concurrent users,
+achieving performance comparable to the Spring WebFlux implementation. The
+blocking approach reaches a throughput of 3,019 requests per second, while the
+Virtual Threads approach achieves a throughput of 3,357 requests per second. In
+addition to the synchronous engines, the \textit{HtmlFlow-Rx} approach also
 achieves a throughput of 1760 requests per second, indicating that this
@@
 
-The results of the benchmarks show that non-blocking engines, through the use
-of reactive programming, Kotlin coroutines, or Java virtual threads, are able
-to scale effectively up to 128 concurrent users. Out of all the tested
-frameworks, Spring Webflux showed itself the most effective at enabling PSSR,
-mostly due to its native support for Publish and Subscriber interfaces, which
-allow for HTML content to be progressively streamed to the client. Quarkus also
-enabled PSSR effectively, but it required additional configuration of the
-\texttt{OutputBuffer} size to achieve the same results as Spring Webflux. The
-Spring MVC implementation, on the other hand, did not enable PSSR for the
-tested templates.
+The results of the benchmarks show that non-blocking engines—whether using
+reactive programming, Kotlin coroutines, or Java Virtual Threads—are able to
+scale effectively, supporting between 32 and 128 concurrent users depending on
+the approach and framework. Out of all the tested frameworks, Spring Webflux
+showed itself the most effective at enabling PSSR, mostly due to its native
+support for Publish and Subscriber interfaces, which allow for HTML content to
+be progressively streamed to the client. Quarkus also enabled PSSR effectively,
+but it required additional configuration of the \texttt{OutputBuffer} size to
+achieve the same results as Spring Webflux. The Spring MVC implementation, on
+the other hand, did not enable PSSR for the tested templates.
 
-Additionally, the results show that approaches using Virtual Threads are able
-to scale as effectively as those using reactive programming or Kotlin
-coroutines, allowing \textit{external} DSLs to be used in a non-blocking
-context, effectively enabling PSSR\@.
+% Additionally, the results show that approaches using Virtual Threads are able
+% to scale as effectively as those using reactive programming or Kotlin
+% coroutines, allowing \textit{external} DSLs to be used in a non-blocking
+% context, effectively enabling PSSR\@.

### 2025-05-30 docs/article/Sections/state-of-the-art-web-servers.tex
First release for submission to MDPI Software Journal

@@
 For traditional imperative-style request handling, Quarkus provides integration
-with JAX-RS, where resource methods can return synchronous types or
+with JAX-RS~\cite{burke2013restful}, where resource methods can return synchronous types or
 asynchronous constructs such as \texttt{CompletionStage}. To enable 

### 2025-07-13 docs/article/Sections/benchmark.tex
Post MDPI review changes.

@@
 shown in \autoref{lst:stock-class} and \autoref{lst:presentation-class},
-following the proposal of the benchmarks~\cite{templatesbenchmark,springbenchmark}.
-The \texttt{Presentation} class represents a presentation with a title, speaker
-name, and summary, while the \texttt{Stock} class represents a stock with a
-name, URL, symbol, price, change, and ratio.
+following the proposal of the
+benchmarks~\cite{templatesbenchmark,springbenchmark}. The \texttt{Presentation}
+class represents a presentation with a title, speaker name, and summary, while
+the \texttt{Stock} class represents a stock with a name, URL, symbol, price,
+change, and ratio.
 
@@
             non-blocking Spring \texttt{ViewResolver}.
+      \item \textbf{Suspendable}: The template engine is used in a suspendable
+            context, where the HTML content is rendered within the context of a suspending function.
+            An example of this approach is the HtmlFlow template engine, which
+            supports suspendable templates with the use of the \texttt{Flow} class from the
+            Kotlin standard library.
       \item \textbf{Virtual}: The template engine is used in a non-blocking
@@
             because they do not support the asynchronous rendering of HTML content.
-  \item \textbf{Suspendable}: The template engine is used in a suspendable
-        context, where the HTML content is rendered within the context of a suspending function.
-        An example of this approach is the HtmlFlow template engine, which
-        supports suspendable templates with the use of the \texttt{Flow} class from the
-        Kotlin standard library.
+      \item \textbf{Blocking}: The template engine is used in a blocking
+            context, where the HTML content is rendered using the blocking interface of the
+            \texttt{Observable} class, within the context of a OS-thread.
 \end{itemize}
@@
 application to write directly to the response \texttt{OutputStream} without
-blocking the servlet container thread. According to the Spring documentation,
+blocking the servlet container thread.
+According to the Spring documentation\footnote{\url{https://docs.spring.io/spring-framework/docs/current/javadoc-api/org/springframework/web/servlet/mvc/method/annotation/StreamingResponseBody.html}},
 this class is \textit{a controller method return value type for asynchronous
@@
 these templates, as HTML content is not streamed progressively to the client.
-This is because the response is only sent once the content written to the
-\texttt{OutputStream} exceeds the output buffer size, which defaults to 8KB\@.
-As a result, the client receives the response only after the entire HTML
-content is rendered, defeating the purpose of PSSR in this context.
+This limitation occurs because the response is only flushed to the client once
+the content written to the \texttt{OutputStream} exceeds the configured output
+buffer size, as explained in Section~\ref{sec:web-frameworks}. Consequently,
+the client receives the complete response only after the entire HTML content
+has been rendered and buffered, which negates the primary benefits of
+PSSR—namely, reduced time-to-first-byte and progressive content delivery.
+Furthermore, Spring MVC does not provide configuration options for the response
+buffer size, preventing developers from reducing it to smaller values that
+would enable more frequent flushing and achieve true progressive streaming of
+HTML content. This architectural constraint makes Spring MVC unsuitable for
+implementing effective PSSR compared to reactive frameworks like Spring
+WebFlux.
 
@@
 to be streamed to the client. While \texttt{StreamingOutput} also uses blocking
-I/O, it operates on Vert.x worker threads, which prevents blocking of the event
+I/O, it operates on Vert.X worker threads, which prevents blocking of the event
 loop. When Virtual Threads are used, the I/O operations are handled

### 2025-07-13 docs/article/Sections/introduction.tex
Post MDPI review changes.

@@
 by streaming HTML content progressively. This technique enhances user-perceived performance by
-allowing progressive rendering as data becomes available. In this respect,
+allowing progressive rendering as data becomes available, significantly reducing the
+\textit{Time to First Byte (TTFB)} and improving perceived load times compared to traditional
+SSR approaches~\cite{wise2024pssr}. In this respect,
 PSSR is similar to CSR in that the server initially sends a minimal HTML
@@
 not need to execute JavaScript or make additional requests to retrieve page content.
+The streaming nature of PSSR allows users to see content progressively as it becomes
+available, rather than waiting for the complete page to be rendered server-side,
+thus providing a more responsive user experience with measurably lower TTFB values.
 
@@
 
-With the introduction of \textit{Virtual Threads}~\footnote{\url{https://openjdk.org/jeps/444}}
-in Java 21, it is now possible to execute blocking I/O operations in a
-scalable, lightweight manner. This capability allows legacy template
-engines—often reliant on blocking interfaces—to operate efficiently in
-high-concurrency, non-blocking environments without requiring complex
-asynchronous programming models. As a result, PSSR
-can now be implemented using familiar HTML templates, 
-simplifying development and improving maintainability.
+With the introduction of \textit{Virtual
+    Threads}~\footnote{\url{https://openjdk.org/jeps/444}} in Java 21, it is now
+possible to execute blocking I/O operations in a scalable, lightweight manner.
+This capability allows legacy template engines—often reliant on blocking
+interfaces—to operate efficiently in high-concurrency, non-blocking
+environments without requiring complex asynchronous programming models. As a
+result, PSSR can now be implemented using familiar HTML templates, simplifying
+development and improving maintainability.
 
@@
 primary paradigms: \textit{reactive programming} and \textit{coroutines}, both
-of which have been used to achieve asynchronous I/O in the Java ecosystem. As an
-alternative, we investigate whether Java’s Virtual Threads can offer comparable
-performance while preserving the simplicity of synchronous code. Section 2
-reviews the state-of-the-art in PSSR and template engine design. Section 3
-outlines the limitations of conventional engines in asynchronous settings and
-presents our proposed approach. Section 4 details the benchmark methodology,
-followed by the results and analysis in Section 5. Conclusions are presented in Section 6.
+of which have been used to achieve asynchronous I/O in the Java ecosystem. As
+an alternative, we investigate whether Java’s Virtual Threads can offer
+comparable performance while preserving the simplicity of synchronous code.
+Section 2 reviews the state-of-the-art in PSSR and template engine design.
+Section 3 outlines the limitations of conventional engines in asynchronous
+settings and presents our proposed approach. Section 4 details the benchmark
+methodology, followed by the results and analysis in Section 5. Conclusions are
+presented in Section 6.

### 2025-07-13 docs/article/Sections/results.tex
Post MDPI review changes.

@@
 
-All the following benchmarks were conducted on a Github-hosted virtual machine
-under GitHub Actions, running Ubuntu 22.04 LTS with a 4-core, 8-thread CPU and
-16GB of RAM\@. All tests were conducted on the OpenJDK VM Corretto 21. The JVM
-was configured with a minimum heap size of 512MB and a maximum heap size of
-16GB\@.
+All the following benchmarks were conducted on a GitHub-hosted virtual machine
+under GitHub Actions, running Ubuntu 24.04.2 LTS (Noble Numbat) with an AMD
+EPYC 7763 64-Core Processor (x86\_64 architecture) operating at 3.24 GHz,
+configured with 2 CPU cores and 2 threads per core, and 7.8GB of available RAM.
+The system utilizes a 75GB SSD with ext4 filesystem, achieving 1.5 GB/s write
+throughput in basic I/O tests. Network connectivity is provided through a 1500
+MTU Ethernet interface. All tests were conducted on OpenJDK 21 (Correto build)
+with the G1 garbage collector enabled by default. The following . JVM arguments
+were used: \texttt{-Xms512m -Xmx16g}. Each test was executed 5 times per
+configuration to ensure the reliability of the results.
+
+All template engines were configured to use UTF-8 encoding and disable
+automatic HTML escaping to ensure fair comparison and consistent output across
+all template engines. Caching behaviour was left as the default for each
+engine, allowing them to optimize template loading and rendering as per their
+design. These configurations ensure that all engines operate under equivalent
+conditions, with template loading, encoding, and escaping behavior normalized
+across implementations.
 
@@
 
-Both the Quarkus and Spring MVC implementations were configured with an output
-buffer size of 8KB, despite the Quarkus implementation not enabling PSSR at
-this size. This configuration was chosen to ensure consistency across Spring
-MVC and Quarkus, as the Spring MVC implementation does not support PSSR for the
-tested templates.
+Both Quarkus and Spring MVC implementations were configured with an 8KB output
+buffer size to ensure consistency, despite Quarkus not enabling PSSR at this
+size and Spring MVC not supporting PSSR for the tested templates. Testing with
+a reduced 512B buffer size showed only a 0.046\% performance difference with
+Rocker, indicating negligible impact. Both frameworks use unlimited platform
+thread pools, enabling on-demand thread creation up to system limits for
+maximum throughput and scalability observation under high concurrent load.
 
 Since the obtained results for JMeter and Apache Bench show no significant
-differences, only the JMeter results will be presented.
+differences, only the JMeter results will be presented. Statistical analysis of
+test configurations revealed an average absolute percentage difference of
+2.83\% between the two load testing tools. While individual approach
+differences ranged from -16.53\% to +14.66\% at the extremes, the consistent
+directional bias and small magnitude of differences indicate that both tools
+provide comparable performance measurements. Given this statistical equivalence
+and the need for brevity, only JMeter results are presented, as they provide
+representative performance characteristics across all tested configurations.
 
-\subsection{Presentations Results}
+\subsection{Scalability results for the \texttt{Presentations} class}
 
@@
      \includegraphics[width=0.8\textwidth]{./Graphs/presentations-webflux-jmeter.png}
-     \caption{Presentation Benchmark Results in Spring WebFlux with JMeter}\label{fig:presentations-webflux-jmeter}
+     \caption{Throughput (requests per second) scalability results for Spring Webflux with \texttt{Presentations} class}\label{fig:presentations-webflux-jmeter}
 \end{figure}
@@
      \includegraphics[width=0.8\textwidth]{./Graphs/presentations-springmvc-jmeter.png}
-     \caption{Presentation Benchmark Results in Spring MVC with JMeter}\label{fig:presentations-springmvc-jmeter}
+     \caption{Throughput (requests per second) scalability results for Spring MVC with \texttt{Presentations} class}\label{fig:presentations-springmvc-jmeter}
 \end{figure}
@@
      \includegraphics[width=0.8\textwidth]{./Graphs/presentations-quarkus-jmeter.png}
-     \caption{Presentation Benchmark Results in Quarkus with JMeter}\label{fig:presentations-quarkus-jmeter}
+     \caption{Throughput (requests per second) scalability results for Quarkus with \texttt{Presentations} class}\label{fig:presentations-quarkus-jmeter}
 \end{figure}
@@
 
-\subsection{Stocks Results}
+\subsection{Scalability results for the \texttt{Stocks} class}
 
@@
      \includegraphics[width=0.8\textwidth]{./Graphs/stocks-webflux-jmeter.png}
-     \caption{Stocks Benchmark Results in Spring WebFlux with JMeter}\label{fig:stocks-webflux-jmeter}
+     \caption{Throughput (requests per second) scalability results for Spring Webflux with \texttt{Stocks} class}\label{fig:stocks-webflux-jmeter}
 \end{figure}
@@
 engines and approaches as the previous benchmark, but with a more complex data
-model: the Stock class, which includes 20 instances and approximately two
-times as many data bindings. With this data model, the scalability of the
-engines remains largely unchanged; however, throughput is reduced across all
-engines. Compared to the Presentation benchmark, the approaches using Virtual
-Threads experience a more pronounced decrease in performance relative to the
+model: the Stock class, which includes 20 instances and approximately two times
+as many data bindings. With this data model, the scalability of the engines
+remains largely unchanged; however, throughput is reduced across all engines.
+Compared to the Presentation benchmark, Jstachio using virtual threads
+experienced a more pronounced decrease in performance relative to the
 \textit{Reactive} and \textit{Suspending} approaches, with Jstachio using
@@
 3,090 requests per second, while HtmlFlow using a reactive approach achieves
-3,026 requests per second.
+3,026 requests per second. This indicates that the more pronounced decrease in
+performance for Jstachio using Virtual Threads is likely due to the specific
+implementation of the template engine, rather than the use of Virtual
+Threads itself.
 
-The overall throughput reduction is expected, as the Stock class contains more
-data properties than the \texttt{Presentation} class, adding overhead related
-to the data binding process of each template engine.
+The overall throughput reduction across all engines is expected, as the Stock
+class contains more data properties than the \texttt{Presentation} class,
+adding overhead related to the data binding process of each template engine.
 
@@
      \includegraphics[width=0.8\textwidth]{./Graphs/stocks-springmvc-jmeter.png}
-     \caption{Stocks Benchmark Results in Spring MVC with JMeter}\label{fig:stocks-springmvc-jmeter}
+     \caption{Throughput (requests per second) scalability results for Spring MVC with \texttt{Stocks} class}\label{fig:stocks-springmvc-jmeter}
 \end{figure}
@@
      \includegraphics[width=0.8\textwidth]{./Graphs/stocks-quarkus-jmeter.png}
-     \caption{Stocks Benchmark Results in Quarkus with JMeter}\label{fig:stocks-quarkus-jmeter}
+     \caption{Throughput (requests per second) scalability results for Quarkus with \texttt{Stocks} class}\label{fig:stocks-quarkus-jmeter}
 \end{figure}
@@
 showed itself the most effective at enabling PSSR, mostly due to its native
-support for Publish and Subscriber interfaces, which allow for HTML content to
-be progressively streamed to the client. Quarkus also enabled PSSR effectively,
-but it required additional configuration of the \texttt{OutputBuffer} size to
-achieve the same results as Spring Webflux. The Spring MVC implementation, on
-the other hand, did not enable PSSR for the tested templates.
+support for publish-subscribe interfaces, allowing for content to be streamed
+as data becomes available, instead of when the response buffer is flushed.
+Quarkus also enabled PSSR effectively, but it required additional configuration
+of the \texttt{OutputBuffer} size to achieve the same results as Spring
+Webflux. The Spring MVC implementation, on the other hand, did not enable PSSR
+for the tested templates.
 
-% Additionally, the results show that approaches using Virtual Threads are able
-% to scale as effectively as those using reactive programming or Kotlin
-% coroutines, allowing \textit{external} DSLs to be used in a non-blocking
-% context, effectively enabling PSSR\@.
+However, it is important to acknowledge the limitations of our chosen data
+models for generalizability. The tested templates used relatively simple data
+structures with limited nesting and straightforward property bindings.
+Real-world applications often involve deeply nested data structures, complex
+conditional logic, and iterative rendering over thousands of items. Under such
+conditions, we anticipate significantly different performance characteristics:
+increased memory consumption due to object traversal overhead, higher CPU
+utilization for complex evaluations, and altered scalability patterns where
+non-blocking approaches may become more advantageous. The performance
+degradation observed with our Stock class benchmark—which merely doubled the
+number of properties—suggests that enterprise-level template complexity would
+result in substantially higher performance degradation. Future work should
+investigate these scenarios with more realistic data models to better
+understand performance boundaries and optimization strategies for complex PSSR
+implementations.
+
+\subsection{Memory Consumption and Resource Utilization Analysis}
+
+This section evaluates the memory and CPU resource characteristics of three
+representative concurrency models: Virtual Threads, suspendable coroutines, and
+reactive streams. While previous benchmarks demonstrated that Virtual Threads
+can offer comparable throughput to non-blocking paradigms, we now assess
+whether this simplicity in code comes with higher runtime resource costs.
+
+\textbf{Virtual Threads vs Suspendables}
+
+\begin{figure}[h]
+     \centering
+     \includegraphics[width=0.8\textwidth]{./Graphs/cpu-susp.png}
+     \caption{Benchmark CPU utilization profiling with HtmlFlow-Susp in Spring WebFlux}\label{fig:cpu-susp}
+\end{figure}
+
+\begin{figure}[h]
+     \centering
+     \includegraphics[width=0.8\textwidth]{./Graphs/cpu-virt.png}
+     \caption{Benchmark CPU utilization profiling with HtmlFlow running on Virtual Threads in Spring WebFlux}\label{fig:cpu-virt}
+\end{figure}
+
+Figures~\ref{fig:cpu-virt} and~\ref{fig:cpu-susp} show CPU utilization profiles
+for the \texttt{HtmlFlow-Virtual} and \texttt{HtmlFlow-Susp} implementations,
+respectively. \texttt{HtmlFlow-Virtual} reaches peak CPU usage of around 90\%
+during load, with sharp transitions between idle and active phases. In
+contrast, \texttt{HtmlFlow-Susp} peaks at approximately 75\% and exhibits
+smoother transitions, suggesting more efficient CPU scheduling and lower
+contention.
+
+\begin{figure}[h]
+     \centering
+     \includegraphics[width=0.8\textwidth]{./Graphs/gc-susp.png}
+     \caption{Benchmark memory utilization profiling with HtmlFlow-Susp in Spring WebFlux}\label{fig:gc-susp}
+\end{figure}
+
+\begin{figure}[h]
+     \centering
+     \includegraphics[width=0.8\textwidth]{./Graphs/gc-virt.png}
+     \caption{Benchmark memory utilization profiling with HtmlFlow running on Virtual Threads in Spring WebFlux}\label{fig:gc-virt}
+\end{figure}
+
+The memory usage disparity is even more pronounced. As shown in
+Figure~\ref{fig:gc-virt}, the Virtual Threads implementation gradually climbs
+to a peak heap usage of 900MB, with large but infrequent garbage collection
+events. Figure~\ref{fig:gc-susp} shows the suspendable version peaks at only
+245MB, following a regular sawtooth pattern of small, frequent garbage
+collections. Overall, \texttt{HtmlFlow-Virtual} allocates a total of 939MB heap
+compared to just 348MB in \texttt{HtmlFlow-Susp}—a 2.7× difference.
+
+\textbf{RxJava Blocking vs Structured Concurrency}
+
+To further assess the efficiency of blocking versus structured concurrency, we
+compared the memory behavior of RxJava's \texttt{BlockingObservableIterable}
+with \texttt{PublisherAsFlow}. The blocking iterator internally relies on a
+lock-based condition variable and an unbounded queue
+(\texttt{SpscLinkedArrayQueue}), blocking the thread when no elements are
+available. Under high concurrency, this model tends to retain large numbers of
+elements in memory, especially when producers outpace consumers. This was
+reflected in our profiling, where heap usage for blocking consumption showed
+linear growth and irregular GC behavior.
+
+In contrast, \texttt{PublisherAsFlow} employs coroutine-based suspension with
+bounded \texttt{Channel} capacity and configurable \texttt{BufferOverflow}
+strategies. It requests upstream data based on demand, leveraging structured
+concurrency and cooperative cancellation.
+
+It is important to note that our benchmark keeps data items in memory as static
+objects, which represents a best-case scenario for memory allocation patterns.
+In real-world applications, data typically comes from external sources such as
+databases, web services, or file systems, requiring dynamic object creation and
+serialization for each request. This would result in significantly higher
+memory allocation rates and more frequent garbage collection cycles across all
+concurrency models, with the observed differences between Virtual Threads and
+suspendable coroutines likely becoming even more pronounced under such
+realistic I/O-intensive workloads. While our benchmark demonstrates that
+Virtual Threads can achieve comparable throughput to suspendable coroutines, or
+other non-blocking paradigms, future work should further explore the impact of
+dynamic data sources on memory allocation patterns and garbage collection
+behavior to better understand the trade-offs between these concurrency models
+in real-world scenarios.
+
+\textbf{Summary of Trade-offs}
+
+While the approaches can achieve similar throughput under load, they differ
+significantly in runtime efficiency. Virtual Threads, despite offering a
+simplified synchronous style, incur the highest memory and CPU overhead.
+structured concurrency delivers superior resource efficiency but
+requires familiarity with asynchronous programming constructs. The RxJava
+blocking iterator, while easier to adopt in traditional reactive settings,
+performs worse in memory-constrained environments due to its unbounded internal
+queuing model.
+
+These results suggest that system resource constraints and predictability
+requirements should inform the choice of concurrency model. In environments
+where memory and CPU resources are limited, structured concurrency via
+suspendable coroutines is the most efficient and robust option.

### 2025-07-13 docs/article/Sections/state-of-the-art-web-servers.tex
Post MDPI review changes.

@@
 \subsection{Web Framework Architectures and Approaches to PSSR}
+\label{sec:web-frameworks}
 
@@
 
+The primary advantage of virtual threads lies in their ability to eliminate the
+need for complex asynchronous programming models. Unlike reactive or structured
+concurrency approaches that require developers to adopt complex paradigms, virtual
+threads enable non-blocking execution of existing synchronous APIs without code
+modification. This significantly reduces the learning curve and maintenance
+burden associated with asynchronous programming while maintaining comparable
+performance characteristics.
+
 \subsubsection{Spring MVC}
 
-Spring MVC is a synchronous web framework built on the
-thread-per-request model. When an HTTP request is received, it is handled by a
-dedicated thread from the server’s thread pool (e.g., Tomcat, Jetty, or
-Undertow). This thread is responsible for executing the full request lifecycle
-synchronously, including invoking controllers, executing business logic, and
-rendering the response view. Because the model is blocking, the thread remains
-occupied throughout the request—even during I/O operations such as database
-queries or external API calls—which can lead to performance bottlenecks under
-high concurrency.
+Spring MVC is a synchronous web framework built on the thread-per-request
+model. When an HTTP request is received, it is handled by a dedicated thread
+from the server’s thread pool (e.g., Tomcat, Jetty, or Undertow). This thread
+is responsible for executing the full request lifecycle synchronously,
+including invoking controllers, executing business logic, and rendering the
+response view. Because the model is blocking, the thread remains occupied
+throughout the request—even during I/O operations such as database queries or
+external API calls—which can lead to performance bottlenecks under high
+concurrency.
 
@@
 been gathered and rendered. By default, this architecture limits the ability to
-perform PSSR, as traditional
-rendering waits for all data before sending any output. This results in higher
-latency for clients and less responsive page loads, particularly for
-data-intensive views. However, Spring MVC provides a mechanism for response
-streaming via the \texttt{StreamingResponseBody}\footnote{\url{https://docs.spring.io/spring-framework/docs/current/javadoc-api/org/springframework/web/servlet/mvc/method/annotation/StreamingResponseBody.html}} interface, introduced in
-Spring 4.2. When a controller method returns a \texttt{StreamingResponseBody},
-Spring writes directly to the \texttt{HttpServletResponse} output stream,
-allowing the server to send parts of the response incrementally as they become
-available. This is particularly useful for:
+perform PSSR, as traditional rendering waits for all data before sending any
+output. This results in higher latency for clients and less responsive page
+loads, particularly for data-intensive views. However, Spring MVC provides a
+mechanism for response streaming via the
+\texttt{StreamingResponseBody}\footnote{\url{https://docs.spring.io/spring-framework/docs/current/javadoc-api/org/springframework/web/servlet/mvc/method/annotation/StreamingResponseBody.html}}
+interface, introduced in Spring 4.2. When a controller method returns a
+\texttt{StreamingResponseBody}, Spring writes directly to the
+\texttt{HttpServletResponse} output stream, allowing the server to send parts
+of the response incrementally as they become available. This is particularly
+useful for:
 \begin{itemize}
@@
 
-In the example shown in \autoref{lst:streaming-response-body}, HTML content is written and flushed in discrete chunks,
-allowing the client to progressively render the response as it arrives, while
-the server continues processing additional data.
-However, this technique has two limitations. First, it does not eliminate the
-blocking nature inherent in Spring MVC, as the handling thread remains active
-during the streaming process.
-Second, it is constrained by the servlet response buffer.
-By default, most servlet containers (e.g., Tomcat) use a
-response buffer size of approximately 8KB. Data written to the response output
-stream is held in this buffer until it reaches capacity. Consequently, if the
-total size of the initial HTML content is less than the buffer threshold, the
-server will not transmit any data to the client until the buffer is filled.
-This buffering behavior introduces a delay in sending the first chunk of
-content, thereby negating the benefits of progressive rendering. In scenarios
-where the rendered HTML template does not produce sufficient data to exceed the
-buffer threshold early, PSSR becomes ineffective, and the client experiences a
-delayed initial response.
+In the example shown in \autoref{lst:streaming-response-body}, HTML content is
+written and flushed in discrete chunks, allowing the client to progressively
+render the response as it arrives, while the server continues processing
+additional data. However, this technique has two limitations. First, it does
+not eliminate the blocking nature inherent in Spring MVC, as the handling
+thread remains active during the streaming process. Second, it is constrained
+by the servlet response buffer. By default, most servlet containers (e.g.,
+Tomcat) use a response buffer size of approximately 8KB. Data written to the
+response output stream is held in this buffer until it reaches capacity.
+Consequently, if the total size of the initial HTML content is less than the
+buffer threshold, the server will not transmit any data to the client until the
+buffer is filled. This buffering behavior introduces a delay in sending the
+first chunk of content, thereby negating the benefits of progressive rendering.
+In scenarios where the rendered HTML template does not produce sufficient data
+to exceed the buffer threshold early, PSSR becomes ineffective, and the client
+experiences a delayed initial response.
 
@@
 on reactive runtimes like Netty, where a small, fixed-size event loop manages
-all I/O events.
-Incoming HTTP requests are routed to handler functions or annotated controllers,
-which return reactive types such as \texttt{Mono<T>} and
-\texttt{Flux<T>}—\texttt{Publisher} implementations provided by the Project Reactor
-library~\cite{projectreactor}. These implementations adhere to the Reactive
-Streams specification~\cite{ReactiveStreams}, enabling interoperability with
-other compliant reactive stream libraries.
-A \texttt{Mono} represents a single asynchronous value (or
-none)~\cite{promise}, while a \texttt{Flux} represents a stream of zero or more values~\cite{rx-observable}. These
-abstractions allow the entire request lifecycle—from controller invocation to
-response rendering—to be orchestrated as a chain of non-blocking, asynchronous
-operations. I/O-bound tasks such as database queries or external API calls do
-not occupy threads during execution, enabling the framework to scale gracefully
-under high load and maintain responsiveness even in resource-constrained
-environments.
+all I/O events. Incoming HTTP requests are routed to handler functions or
+annotated controllers, which return reactive types such as \texttt{Mono<T>} and
+\texttt{Flux<T>}—\texttt{Publisher} implementations provided by the Project
+Reactor library~\cite{projectreactor}. These implementations adhere to the
+Reactive Streams specification~\cite{ReactiveStreams}, enabling
+interoperability with other compliant reactive stream libraries. A
+\texttt{Mono} represents a single asynchronous value (or none)~\cite{promise},
+while a \texttt{Flux} represents a stream of zero or more
+values~\cite{rx-observable}. These abstractions allow the entire request
+lifecycle—from controller invocation to response rendering—to be orchestrated
+as a chain of non-blocking, asynchronous operations. I/O-bound tasks such as
+database queries or external API calls do not occupy threads during execution,
+enabling the framework to scale gracefully under high load and maintain
+responsiveness even in resource-constrained environments.
 
-PSSR is natively supported in WebFlux via
-\texttt{Flux}-based controllers. When a controller returns a
-\texttt{Flux<String>} or another stream of HTML fragments, WebFlux can begin
-streaming content to the client as soon as the first elements are emitted.
+PSSR is natively supported in WebFlux via \texttt{Flux}-based controllers. When
+a controller returns a \texttt{Flux<String>} or another stream of HTML
+fragments, WebFlux can begin streaming content to the client as soon as the
+first elements are emitted.
 % This allows browsers to incrementally render parts of the response without waiting
@@
 Quarkus is a modern, cloud-native Java framework that supports both imperative
-and reactive programming models. It is built on top of Vert.x,
-which provides a non-blocking, event-driven runtime similar to Netty. Incoming
-HTTP requests are processed asynchronously on a small, fixed-size event loop
-thread pool, allowing the framework to handle many concurrent connections
-efficiently without blocking threads. 
-Quarkus provides support for reactive APIs like Mutiny~\cite{mutiny2021}, a
-modern alternative to Spring Reactor~\cite{projectreactor}. Like Reactor,
-Mutiny complies with the Reactive Streams specification~\cite{ReactiveStreams}.
-For traditional imperative-style request handling, Quarkus provides integration
-with JAX-RS~\cite{burke2013restful}, where resource methods can return synchronous types or
-asynchronous constructs such as \texttt{CompletionStage}. To enable 
-PSSR in imperative endpoints, Quarkus leverages the
-JAX-RS \texttt{StreamingOutput} interface. By returning a
-\texttt{StreamingOutput} instance, developers can write parts of the HTTP
-response body incrementally as data becomes available. This approach allows the
-server to flush HTML fragments progressively to the client, improving the
-perceived responsiveness especially for long-running or large responses.
-However, like other streaming mechanisms based on servlet buffers, the initial
-data flush may be delayed until internal buffers (commonly around 8KB) are
-filled. However, Quarkus allows configuring the buffer size to a smaller value, which
-can help mitigate this issue.
+and reactive programming models. It is built on top of Vert.X, which provides a
+non-blocking, event-driven runtime similar to Netty. Incoming HTTP requests are
+processed asynchronously on a small, fixed-size event loop thread pool,
+allowing the framework to handle many concurrent connections efficiently
+without blocking threads. Quarkus provides support for reactive APIs like
+Mutiny~\cite{mutiny2021}, a modern alternative to Spring
+Reactor~\cite{projectreactor}. Like Reactor, Mutiny complies with the Reactive
+Streams specification~\cite{ReactiveStreams}. For traditional imperative-style
+request handling, Quarkus provides integration with
+JAX-RS~\cite{burke2013restful}, where resource methods can return synchronous
+types or asynchronous constructs such as \texttt{CompletionStage}. To enable
+PSSR in imperative endpoints, Quarkus leverages the JAX-RS
+\texttt{StreamingOutput} interface. By returning a \texttt{StreamingOutput}
+instance, developers can write parts of the HTTP response body incrementally as
+data becomes available. This approach allows the server to flush HTML fragments
+progressively to the client, improving the perceived responsiveness especially
+for long-running or large responses. Like other streaming mechanisms
+based on servlet buffers, the initial data flush may be delayed until internal
+buffers (commonly around 8KB) are filled. However, Quarkus allows configuring
+the buffer size to a smaller value, which can help mitigate this issue.
 

### 2025-07-13 docs/article/Sections/state-of-the-art-web-templates.tex
Article section 2: Update template engines table type-safety section; added Qute.
Section 3 & 5: update template engines table; Add HtmlFlow-Rx to Quarkus Results and change Sync to Blocking
Post MDPI review changes.

@@
      & \large{$\textcolor{red}{\times}$}
-     & \large{$\textcolor{red}{\times}$}
+     & \textcolor{PineGreen}{\checkmark}$^{(1)}$
      & \large{$\textcolor{red}{\times}$}
@@
      & \large{$\textcolor{red}{\times}$}
+     & \large{$\textcolor{red}{\times}$}
+     & \large{$\textcolor{PineGreen}{\checkmark}$}
+    \\
+    \hline
+    \textbf{Qute}
+     & External DSL
+     & \texttt{Iterable}
+     & \large{$\textcolor{red}{\times}$}
+     & \textcolor{PineGreen}{\checkmark}$^{(2)}$
      & \large{$\textcolor{red}{\times}$}
@@
      & External DSL
-     & \shortstack{\texttt{Iterable} \\\texttt{Stream} \\\texttt{Publisher}$^{(1)}$}
-     & \texttt{Publisher}$^{(1)}$
+     & \shortstack{\texttt{Iterable} \\\texttt{Stream} \\\texttt{Publisher}$^{(3)}$}
+     & \texttt{Publisher}$^{(3)}$
      & \large{$\textcolor{red}{\times}$}
@@
      & External DSL
-     & \shortstack{\texttt{Iterable} \\\texttt{Sequence}$^{(2)}$}
+     & \shortstack{\texttt{Iterable} \\\texttt{Sequence}$^{(4)}$}
      & \large{$\textcolor{red}{\times}$}
@@
      & \large{$\textcolor{PineGreen}{\checkmark}$}
-     & \textcolor{PineGreen}{\checkmark}$^{(3)}$
+     & \textcolor{PineGreen}{\checkmark}$^{(5)}$
      & \large{$\textcolor{PineGreen}{\checkmark}$}
@@
     Comparison of web template technologies in the Java ecosystem.
-    \\$^{(1)}$ \texttt{Publisher} of Reactive Streams~\cite{ReactiveStreams}.
-    Limited to a single model per web template.
-    \\$^{(2)}$ \texttt{kotlin.sequences.Sequence}
-    \\$^{(3)}$ Non-safety for HTML attributes.
+    \\$^{(1)}$ Enforced via annotation processor that checks templates against typed interfaces at compile time.
+    \\$^{(2)}$ Compile-time expression validation available via \texttt{@CheckedTemplate} and build-time metadata in Quarkus.
+    \\$^{(3)}$ \texttt{Publisher} of Reactive Streams~\cite{ReactiveStreams}. Limited to a single model per web template.
+    \\$^{(4)}$ \texttt{kotlin.sequences.Sequence}
+    \\$^{(5)}$ Non-safety for HTML attributes.
   }
@@
 pattern\cite{thompson1968}, since it defines its own syntax without any
-dependency of programming languages. On the other hand an \textit{internal} DSL
+dependency of programming languages. On the other hand, an \textit{internal} DSL
 is defined within a host programming language as a library and tends to be
@@
 (Figure~\ref{fig:dsl-idioms}.c). But the \textit{nested function} approach
-comes with a notable drawback: \textbf{they do not support PSSR} because the
+comes with a notable drawback: \textbf{it does not support PSSR} because the
 sequence of nested functions is evaluated backward to the order in which they
@@
 
-To achieve this KotlinX.html provides HTML builders using \textit{function
+To achieve this, KotlinX.html provides HTML builders using \textit{function
   literals with receiver}~\cite{kotlinlang}. In Kotlin, a block of code enclosed
@@
 which offers the advantage of not being dependent on
-\textit{rendering-blocking resources}, such as the required JavaScript for CSR.
+\textit{render-blocking resources}, such as the required JavaScript for CSR.
 % \textit{Progressive server-side rendering} (PSSR) is the

### 2025-07-13 results/plot-quarkus.sh

@@ -1,7 +1,7 @@
 #!/bin/bash
 
 INPUT_BLOCKING="quarkus-results.log"
-INPUT_VIRTUAL="quarkus-virtual-results.log"
+INPUT_VIRTUAL="quarkus-results-virtual.log"
 
 PRESENTATIONS_CSV="presentations-quarkus.csv"
 STOCKS_CSV="stocks-quarkus.csv"

### 2025-07-13 docs/article/Sections/introduction.tex
Small changes to article.

@@
 the client, resulting in a faster \textit{First Contentful Paint (FCP)}
-~\cite{Edgar2024-FCP} and better Search Engine Optimization (SEO). However, SSR
+~\cite{Edgar2024-FCP} and better Search Engine Optimization (SEO). Nevertheless, SSR
 can increase server load and reduce \textit{throughput}\footnote{The number of
@@
 document to the client and subsequently streams additional HTML fragments.
-However, unlike CSR, PSSR retains all rendering responsibilities on the server side,
-thereby reducing the load on the client. As a result, the client does
+Even so, unlike CSR, PSSR retains all rendering responsibilities on the server side,
+thereby reducing the load on the client. Consequently, the client does
 not need to execute JavaScript or make additional requests to retrieve page content.
@@
 interfaces—to operate efficiently in high-concurrency, non-blocking
-environments without requiring complex asynchronous programming models. As a
-result, PSSR can now be implemented using familiar HTML templates, simplifying
+environments without requiring complex asynchronous programming models. Due to this, PSSR can now be implemented using familiar HTML templates, simplifying
 development and improving maintainability.
 
-This work explores the current landscape of non-blocking PSSR, focusing on two
+We investigate the current landscape of non-blocking PSSR, focusing on two
 primary paradigms: \textit{reactive programming} and \textit{coroutines}, both

### 2025-07-13 docs/article/Sections/problem-statement.tex
Small changes to article.

@@
 this challenge. Virtual threads drastically reduce the overhead of blocking
-operations by decoupling thread execution from OS-level threads. As a result,
+operations by decoupling thread execution from OS-level threads. For this reason,
 engines that rely on blocking interfaces—like those used in

### 2025-07-13 docs/article/Sections/state-of-the-art-web-servers.tex
Small changes to article.

@@
 subsequently, various technologies adopted this approach in the Java ecosystem,
-including Netty, Akka HTTP, Vert.X, and Spring WebFlux.
+including Netty, Akka HTTP, Vert.x, and Spring WebFlux.
 % Among these, Spring WebFlux stands out as the most widely used web framework in
@@
 An alternative approach involves utilizing user-level threads, while
-maintaining a blocking I/O and a synchronous programming paradigm. However,
+maintaining a blocking I/O and a synchronous programming paradigm. On the other hand,
 this approach still requires a user-level I/O subsystem capable of mitigating
@@
 output. This results in higher latency for clients and less responsive page
-loads, particularly for data-intensive views. However, Spring MVC provides a
+loads, particularly for data-intensive views. In any case, Spring MVC provides a
 mechanism for response streaming via the
@@
 render the response as it arrives, while the server continues processing
-additional data. However, this technique has two limitations. First, it does
+additional data. This technique has two limitations. First, it does
 not eliminate the blocking nature inherent in Spring MVC, as the handling
@@
 Quarkus is a modern, cloud-native Java framework that supports both imperative
-and reactive programming models. It is built on top of Vert.X, which provides a
+and reactive programming models. It is built on top of Vert.x, which provides a
 non-blocking, event-driven runtime similar to Netty. Incoming HTTP requests are

### 2025-07-13 docs/article/Sections/state-of-the-art-web-templates.tex
Small changes to article.

@@
 common supertype implemented by most data structures and serves as the standard
-API for iteration. However, other standard interfaces, such as
+API for iteration. Nevertheless, other standard interfaces, such as
 \texttt{java.util.Stream}, also represent sequences of elements but are not
@@
 abstractions like the \texttt{Sequence} interface—also incompatible with
-\texttt{Iterable}. As a result, template engines based on \textit{external}
+\texttt{Iterable}. In consequence, template engines based on \textit{external}
 DSLs must explicitly support each of these interfaces to accommodate the
@@
 among external DSLs for HTML—to support reactive streams based on the
-\texttt{Publisher} model. However, it is limited to using a single
+\texttt{Publisher} model. Yet, it is limited to using a single
 \texttt{Publisher} within a given template.
@@
 To ensure HTML safety, a DSL API should only allow combining calls to builders
-that result in valid HTML. However, most HTML DSLs using the \textit{function
+that result in valid HTML. Nonetheless, most HTML DSLs using the \textit{function
   sequence} or \textit{nested function} approach cannot ensure HTML safety at
@@
 becomes available to complete the web page.
-However, despite being overlooked, HTTP and browsers were designed from their
+Nevertheless, despite being overlooked, HTTP and browsers were designed from their
 inception to also support this feature in the context of SSR approaches,
@@
 client-side JavaScript. Like Thymeleaf, their proposal avoids the use of
-client-side JavaScript; however, unlike Thymeleaf, it supports a wide range of
+client-side JavaScript; yet, unlike Thymeleaf, it supports a wide range of
 asynchronous APIs and multiple data sources, not limited to the
-\texttt{Publisher} API. However, the main counter-argument is the non-trivial
+\texttt{Publisher} API. Nonetheless, the main counter-argument is the non-trivial
 management of the \emph{resume} callback in

### 2025-07-19 docs/article/Sections/introduction.tex
Changes to introduction, results. Removed redundancy in 2.2.2.

@@
 available, rather than waiting for the complete page to be rendered server-side,
-thus providing a more responsive user experience with measurably lower TTFB values.
+thus providing a more responsive user experience with measurably lower TTFB values~\cite{wise2024pssr}.
 
@@
 that are compatible with asynchronous data models~\cite{carvalho2023async}.
-Some modern template
-engines, such as HtmlFlow~\cite{htmlflow} and Thymeleaf~\cite{thymeleaf}, have
+Some modern template engines, such as HtmlFlow~\cite{htmlflow} and Thymeleaf~\cite{thymeleaf}, have
 been designed with these capabilities in mind. However, many legacy

### 2025-07-19 docs/article/Sections/results.tex
Changes to introduction, results. Removed redundancy in 2.2.2.

@@
 
-All the following benchmarks were conducted on a GitHub-hosted virtual machine
-under GitHub Actions, running Ubuntu 24.04.2 LTS (Noble Numbat) with an AMD
-EPYC 7763 64-Core Processor (x86\_64 architecture) operating at 3.24 GHz,
-configured with 2 CPU cores and 2 threads per core, and 7.8GB of available RAM.
-The system utilizes a 75GB SSD with ext4 filesystem, achieving 1.5 GB/s write
-throughput in basic I/O tests. Network connectivity is provided through a 1500
-MTU Ethernet interface. All tests were conducted on OpenJDK 21 (Correto build)
-with the G1 garbage collector enabled by default. The following . JVM arguments
-were used: \texttt{-Xms512m -Xmx16g}. Each test was executed 5 times per
-configuration to ensure the reliability of the results.
+This section presents the evaluation results organized into four main parts:
+Section~\ref{sec:environment} details the testing environment and hardware
+specifications; Section~\ref{sec:presentations-results} presents scalability
+results for the \texttt{Presentations} data model;
+Section~\ref{sec:stocks-results} analyzes performance with the more complex
+\texttt{Stocks} data model; and Section~\ref{sec:memory-analysis} provides
+detailed memory consumption and resource utilization analysis.
+
+\subsection{Environment Specifications}
+\label{sec:environment}
+
+All benchmarks were conducted on a GitHub-hosted virtual machine under GitHub
+Actions, running Ubuntu 24.04.2 LTS with an AMD EPYC 7763 64-Core Processor
+(x86\_64 architecture) operating at 3.24 GHz, configured with 2 CPU cores and 2
+threads per core, and 7.8 GB of available RAM. The system utilizes a 75 GB SSD
+with ext4 filesystem, achieving 1.5 GB/s write throughput in basic I/O tests.
+Network connectivity is provided through a 1500 MTU Ethernet interface. All
+tests were conducted on OpenJDK 21 (Corretto build) with the G1 garbage
+collector enabled by default. The JVM arguments used were: \texttt{-Xms512m
+     -Xmx16g}. Each test was executed 5 times per configuration to ensure result
+reliability.
 
@@
 \subsection{Scalability results for the \texttt{Presentations} class}
+\label{sec:presentations-results}
 
@@
 \subsection{Scalability results for the \texttt{Stocks} class}
+\label{sec:stocks-results}
 
@@
 \subsection{Memory Consumption and Resource Utilization Analysis}
+\label{sec:memory-analysis}
 
@@
 
-\textbf{Virtual Threads vs Suspendables}
+Resource utilization data was collected using VisualVM during 40-second
+benchmark runs, capturing detailed CPU and memory usage patterns under
+sustained load conditions. Each profiling session monitored system performance
+during the 1 to 128 concurrent user load test, providing comprehensive insights
+into runtime behavior.
+
+\subsubsection{Virtual Threads vs Structured Concurrency}
+
+\begin{figure}[h]
+     \centering
+     \includegraphics[width=0.8\textwidth]{./Graphs/cpu-virt.png}
+     \caption{CPU utilization profiling with HtmlFlow running on Virtual Threads in Spring WebFlux during 40-second load test}\label{fig:cpu-virt}
+\end{figure}
 
@@
      \includegraphics[width=0.8\textwidth]{./Graphs/cpu-susp.png}
-     \caption{Benchmark CPU utilization profiling with HtmlFlow-Susp in Spring WebFlux}\label{fig:cpu-susp}
+     \caption{CPU utilization profiling with HtmlFlow-Susp in Spring WebFlux during 40-second load test}\label{fig:cpu-susp}
+\end{figure}
+
+Figures~\ref{fig:cpu-virt} and~\ref{fig:cpu-susp} show CPU utilization profiles
+for the \texttt{HtmlFlow-Virtual} and \texttt{HtmlFlow-Susp} implementations,
+respectively. In both cases, the initial spike in CPU usage corresponds to the
+system's bootstrap and initial load phase. The subsequent gradual increase in
+CPU utilization reflects the ramp-up of concurrent users during the benchmark.
+\texttt{HtmlFlow-Virtual} exhibits a distinctive bell-shaped curve, rapidly
+climbing from near-zero to peak CPU usage of around 90\% during the initial
+load phase, maintaining high utilization during sustained load, and then
+dropping sharply back to baseline. The profile shows more aggressive CPU
+consumption with steep ramp-up and ramp-down phases. In contrast,
+\texttt{HtmlFlow-Susp} displays a more gradual and controlled resource
+utilization pattern, with a smoother ascent to a lower peak of approximately
+75\%, followed by a more gentle descent. This smoother profile suggests more
+efficient CPU scheduling and better resource management under load. The GC
+activity indicators in both profiles remain relatively low throughout the test
+duration, confirming that garbage collection overhead does not significantly
+impact CPU utilization in either approach.
+
+\begin{figure}[h]
+     \centering
+     \includegraphics[width=0.8\textwidth]{./Graphs/gc-virt.png}
+     \caption{Memory utilization and garbage collection behavior with HtmlFlow running on Virtual Threads in Spring WebFlux during 40-second load test}\label{fig:gc-virt}
 \end{figure}
@@
      \centering
-     \includegraphics[width=0.8\textwidth]{./Graphs/cpu-virt.png}
-     \caption{Benchmark CPU utilization profiling with HtmlFlow running on Virtual Threads in Spring WebFlux}\label{fig:cpu-virt}
+     \includegraphics[width=0.8\textwidth]{./Graphs/gc-susp.png}
+     \caption{Memory utilization and garbage collection behavior with HtmlFlow-Susp in Spring WebFlux during 40-second load test}\label{fig:gc-susp}
 \end{figure}
 
-Figures~\ref{fig:cpu-virt} and~\ref{fig:cpu-susp} show CPU utilization profiles
-for the \texttt{HtmlFlow-Virtual} and \texttt{HtmlFlow-Susp} implementations,
-respectively. \texttt{HtmlFlow-Virtual} reaches peak CPU usage of around 90\%
-during load, with sharp transitions between idle and active phases. In
-contrast, \texttt{HtmlFlow-Susp} peaks at approximately 75\% and exhibits
-smoother transitions, suggesting more efficient CPU scheduling and lower
-contention.
-
-\begin{figure}[h]
-     \centering
-     \includegraphics[width=0.8\textwidth]{./Graphs/gc-susp.png}
-     \caption{Benchmark memory utilization profiling with HtmlFlow-Susp in Spring WebFlux}\label{fig:gc-susp}
-\end{figure}
-
-\begin{figure}[h]
-     \centering
-     \includegraphics[width=0.8\textwidth]{./Graphs/gc-virt.png}
-     \caption{Benchmark memory utilization profiling with HtmlFlow running on Virtual Threads in Spring WebFlux}\label{fig:gc-virt}
-\end{figure}
-
-The memory usage disparity is even more pronounced. As shown in
-Figure~\ref{fig:gc-virt}, the Virtual Threads implementation gradually climbs
-to a peak heap usage of 900MB, with large but infrequent garbage collection
-events. Figure~\ref{fig:gc-susp} shows the suspendable version peaks at only
-245MB, following a regular sawtooth pattern of small, frequent garbage
+The memory usage disparity is more pronounced. Memory utilization data was
+collected using VisualVM profiling during the same 40-second benchmark runs. As
+shown in Figure~\ref{fig:gc-virt}, the Virtual Threads implementation gradually
+climbs to a peak heap usage of 900MB, with large but infrequent garbage
+collection events. Figure~\ref{fig:gc-susp} shows the suspendable version peaks
+at only 245MB, following a regular sawtooth pattern of small, frequent garbage
 collections. Overall, \texttt{HtmlFlow-Virtual} allocates a total of 939MB heap
@@
 
-\textbf{RxJava Blocking vs Structured Concurrency}
-
-To further assess the efficiency of blocking versus structured concurrency, we
-compared the memory behavior of RxJava's \texttt{BlockingObservableIterable}
-with \texttt{PublisherAsFlow}. The blocking iterator internally relies on a
-lock-based condition variable and an unbounded queue
-(\texttt{SpscLinkedArrayQueue}), blocking the thread when no elements are
-available. Under high concurrency, this model tends to retain large numbers of
-elements in memory, especially when producers outpace consumers. This was
-reflected in our profiling, where heap usage for blocking consumption showed
-linear growth and irregular GC behavior.
-
-In contrast, \texttt{PublisherAsFlow} employs coroutine-based suspension with
-bounded \texttt{Channel} capacity and configurable \texttt{BufferOverflow}
-strategies. It requests upstream data based on demand, leveraging structured
-concurrency and cooperative cancellation.
+The higher observed heap usage with Virtual Threads stems from how they manage
+concurrency. Virtual Threads allocate a full stack per thread, and when they
+are suspended (e.g., during blocking I/O), their entire call stack is unmounted
+and copied to the heap, allowing the physical carrier thread to be reused.
+Kotlin coroutines, by contrast, also store their execution state on the heap
+when suspended — but the compiler transforms coroutine code into a state
+machine, so only the minimal necessary state is saved rather than a full stack.
+This results in significantly lower memory overhead per suspended task.
 
@@
 suspendable coroutines likely becoming even more pronounced under such
-realistic I/O-intensive workloads. While our benchmark demonstrates that
-Virtual Threads can achieve comparable throughput to suspendable coroutines, or
-other non-blocking paradigms, future work should further explore the impact of
-dynamic data sources on memory allocation patterns and garbage collection
-behavior to better understand the trade-offs between these concurrency models
-in real-world scenarios.
+realistic I/O-intensive workloads.
 
-\textbf{Summary of Trade-offs}
+\subsubsection{Summary of Trade-offs}
 

### 2025-07-19 docs/article/Sections/results.tex
Updated results section

@@
 
-\subsection{Scalability results for the \texttt{Presentations} class}
-\label{sec:presentations-results}
+\subsection{Scalability results for the \texttt{Presentations} class} \label{sec:presentations-results}
 
@@
 
-\subsection{Scalability results for the \texttt{Stocks} class}
-\label{sec:stocks-results}
+\subsection{Scalability results for the \texttt{Stocks} class} \label{sec:stocks-results}
 
@@
 performance for Jstachio using Virtual Threads is likely due to the specific
-implementation of the template engine, rather than the use of Virtual
-Threads itself.
+implementation of the template engine, rather than the use of Virtual Threads
+itself.
 
@@
 degradation observed with our Stock class benchmark—which merely doubled the
-number of properties—suggests that enterprise-level template complexity would
-result in substantially higher performance degradation. Future work should
-investigate these scenarios with more realistic data models to better
-understand performance boundaries and optimization strategies for complex PSSR
+number of properties—suggests that higher template complexity would result in
+substantially higher performance degradation. Future work should investigate
+these scenarios with more realistic data models to better understand
+performance boundaries and optimization strategies for complex PSSR
 implementations.
@@
 
-This section evaluates the memory and CPU resource characteristics of three
-representative concurrency models: Virtual Threads, suspendable coroutines, and
-reactive streams. While previous benchmarks demonstrated that Virtual Threads
-can offer comparable throughput to non-blocking paradigms, we now assess
-whether this simplicity in code comes with higher runtime resource costs.
+This section evaluates the memory and CPU resource usage characteristics of
+structured concurrency approaches, specifically comparing Virtual Threads and
+suspendable coroutines.
 
-Resource utilization data was collected using VisualVM during 40-second
+Resource utilization data was collected using VisualVM during 30-second
 benchmark runs, capturing detailed CPU and memory usage patterns under
@@
 
-\begin{figure}[h]
+\begin{center}
+     \begin{minipage}[t]{0.48\textwidth}
           \centering
-     \includegraphics[width=0.8\textwidth]{./Graphs/cpu-virt.png}
-     \caption{CPU utilization profiling with HtmlFlow running on Virtual Threads in Spring WebFlux during 40-second load test}\label{fig:cpu-virt}
-\end{figure}
+          \includegraphics[width=1.0\textwidth,height=1.0\textwidth]{./Graphs/cpu-susp.png}
+          \captionof{figure}{CPU utilization profiling with HtmlFlow-Susp in Spring WebFlux during 30-second load test}
+          \label{fig:cpu-susp}
+     \end{minipage}
+     \hfill
+     \begin{minipage}[t]{0.48\textwidth}
+          \centering
+          \includegraphics[width=1.0\textwidth,height=1.0\textwidth]{./Graphs/cpu-virt.png}
+          \captionof{figure}{CPU utilization profiling with HtmlFlow-Virtual in Spring WebFlux during 30-second load test}
+          \label{fig:cpu-virt}
+     \end{minipage}
+\end{center}
 
-\begin{figure}[h]
+\begin{center}
+     \begin{minipage}[t]{0.48\textwidth}
           \centering
-     \includegraphics[width=0.8\textwidth]{./Graphs/cpu-susp.png}
-     \caption{CPU utilization profiling with HtmlFlow-Susp in Spring WebFlux during 40-second load test}\label{fig:cpu-susp}
-\end{figure}
+          \includegraphics[width=1.0\textwidth,height=1.0\textwidth]{./Graphs/mem-susp.png}
+          \captionof{figure}{Memory utilization and garbage collection behavior with HtmlFlow-Susp in Spring WebFlux during 30-second load test}
+          \label{fig:gc-susp}
+     \end{minipage}
+     \hfill
+     \begin{minipage}[t]{0.48\textwidth}
+          \centering
+          \includegraphics[width=1.0\textwidth,height=1.0\textwidth]{./Graphs/mem-virt.png}
+          \captionof{figure}{Memory utilization and garbage collection behavior with HtmlFlow-Virtual in Spring WebFlux during 30-second load test}
+          \label{fig:gc-virt}
+     \end{minipage}
+\end{center}
 
@@
 for the \texttt{HtmlFlow-Virtual} and \texttt{HtmlFlow-Susp} implementations,
-respectively. In both cases, the initial spike in CPU usage corresponds to the
-system's bootstrap and initial load phase. The subsequent gradual increase in
-CPU utilization reflects the ramp-up of concurrent users during the benchmark.
-\texttt{HtmlFlow-Virtual} exhibits a distinctive bell-shaped curve, rapidly
-climbing from near-zero to peak CPU usage of around 90\% during the initial
-load phase, maintaining high utilization during sustained load, and then
-dropping sharply back to baseline. The profile shows more aggressive CPU
-consumption with steep ramp-up and ramp-down phases. In contrast,
-\texttt{HtmlFlow-Susp} displays a more gradual and controlled resource
-utilization pattern, with a smoother ascent to a lower peak of approximately
-75\%, followed by a more gentle descent. This smoother profile suggests more
-efficient CPU scheduling and better resource management under load. The GC
-activity indicators in both profiles remain relatively low throughout the test
-duration, confirming that garbage collection overhead does not significantly
-impact CPU utilization in either approach.
+respectively. The subsequent gradual increase in CPU utilization reflects the
+ramp-up of concurrent users during the benchmark. \texttt{HtmlFlow-Virtual}
+exhibits a distinctive bell-shaped curve, rapidly climbing from near-zero to
+peak CPU usage of around 50\% during the sustained load phase, maintaining
+moderate utilization during the test period, and then dropping sharply back to
+baseline. \texttt{HtmlFlow-Susp} displays a similar pattern but with a slightly
+lower peak CPU utilization of approximately 42\%, showing marginally lower
+intensive CPU usage during the sustained load phase. Both profiles demonstrate
+comparable resource consumption patterns, with the virtual thread
+implementation showing only slightly higher CPU demands. The GC activity
+indicators in both profiles remain relatively low throughout the test duration,
+confirming that garbage collection overhead does not significantly impact CPU
+utilization in either approach.
 
-\begin{figure}[h]
-     \centering
-     \includegraphics[width=0.8\textwidth]{./Graphs/gc-virt.png}
-     \caption{Memory utilization and garbage collection behavior with HtmlFlow running on Virtual Threads in Spring WebFlux during 40-second load test}\label{fig:gc-virt}
-\end{figure}
+Figures~\ref{fig:gc-susp} and~\ref{fig:gc-virt} show similar memory usage for
+both approaches, with peaks around 300~MB during load. The initial spike to
+450~MB in both graphs corresponds to the applications' bootstrapping phase.
+Overall, these results indicate that under the tested conditions, Virtual
+Threads and suspendable coroutines exhibit comparable memory consumption.
 
-\begin{figure}[h]
-     \centering
-     \includegraphics[width=0.8\textwidth]{./Graphs/gc-susp.png}
-     \caption{Memory utilization and garbage collection behavior with HtmlFlow-Susp in Spring WebFlux during 40-second load test}\label{fig:gc-susp}
-\end{figure}
+Virtual Threads allocate a full call stack per thread, which is unmounted and
+stored on the heap when suspended (e.g., during blocking I/O). In contrast,
+Kotlin coroutines compile into state machines that capture only the minimal
+execution state needed to resume computation. This typically allows coroutines
+to scale more efficiently in terms of memory usage, especially under high
+concurrency. These findings contrasts with prior research, such as Beronić et
+al.~\cite{9803765}, which reported lower heap usage for Kotlin coroutines
+(52--99~MB) compared to Java Virtual Threads (16--64~MB), albeit under
+different workloads.
 
-The memory usage disparity is more pronounced. Memory utilization data was
-collected using VisualVM profiling during the same 40-second benchmark runs. As
-shown in Figure~\ref{fig:gc-virt}, the Virtual Threads implementation gradually
-climbs to a peak heap usage of 900MB, with large but infrequent garbage
-collection events. Figure~\ref{fig:gc-susp} shows the suspendable version peaks
-at only 245MB, following a regular sawtooth pattern of small, frequent garbage
-collections. Overall, \texttt{HtmlFlow-Virtual} allocates a total of 939MB heap
-compared to just 348MB in \texttt{HtmlFlow-Susp}—a 2.7× difference.
-
-The higher observed heap usage with Virtual Threads stems from how they manage
-concurrency. Virtual Threads allocate a full stack per thread, and when they
-are suspended (e.g., during blocking I/O), their entire call stack is unmounted
-and copied to the heap, allowing the physical carrier thread to be reused.
-Kotlin coroutines, by contrast, also store their execution state on the heap
-when suspended — but the compiler transforms coroutine code into a state
-machine, so only the minimal necessary state is saved rather than a full stack.
-This results in significantly lower memory overhead per suspended task.
-
-It is important to note that our benchmark keeps data items in memory as static
-objects, which represents a best-case scenario for memory allocation patterns.
-In real-world applications, data typically comes from external sources such as
-databases, web services, or file systems, requiring dynamic object creation and
-serialization for each request. This would result in significantly higher
-memory allocation rates and more frequent garbage collection cycles across all
-concurrency models, with the observed differences between Virtual Threads and
-suspendable coroutines likely becoming even more pronounced under such
-realistic I/O-intensive workloads.
-
-\subsubsection{Summary of Trade-offs}
-
-While the approaches can achieve similar throughput under load, they differ
-significantly in runtime efficiency. Virtual Threads, despite offering a
-simplified synchronous style, incur the highest memory and CPU overhead.
-structured concurrency delivers superior resource efficiency but
-requires familiarity with asynchronous programming constructs. The RxJava
-blocking iterator, while easier to adopt in traditional reactive settings,
-performs worse in memory-constrained environments due to its unbounded internal
-queuing model.
-
-These results suggest that system resource constraints and predictability
-requirements should inform the choice of concurrency model. In environments
-where memory and CPU resources are limited, structured concurrency via
-suspendable coroutines is the most efficient and robust option.
+These observations suggest that the choice between Virtual Threads and
+suspendable coroutines should be guided by both runtime behavior and
+development preferences. Virtual Threads offer the simplicity of synchronous
+programming with competitive performance in moderate-load scenarios. In
+contrast, suspendable coroutines, while requiring an asynchronous programming
+model, may offer better efficiency in applications with higher concurrency,
+deeper stacks, or more dynamic workloads.

### 2025-07-19 docs/article/Sections/results.tex
fix typo in results section, re-add graphs

@@
 to scale more efficiently in terms of memory usage, especially under high
-concurrency. These findings contrasts with prior research, such as Beronić et
-al.~\cite{9803765}, which reported lower heap usage for Kotlin coroutines
-(52--99~MB) compared to Java Virtual Threads (16--64~MB), albeit under
-different workloads.
+concurrency. These findings contrast with prior research, such as Beronić et
+al.~\cite{9803765}, which reported lower heap usage for Java Virtual Threads (16--64~MB)
+compared to Kotlin coroutines (52--99~MB), albeit under different workloads.
 

### 2025-07-28 docs/article/Sections/results.tex
New Section 6 - Discussion

@@
 to scale more efficiently in terms of memory usage, especially under high
-concurrency. These findings contrast with prior research, such as Beronić et
-al.~\cite{9803765}, which reported lower heap usage for Java Virtual Threads (16--64~MB)
-compared to Kotlin coroutines (52--99~MB), albeit under different workloads.
+concurrency.
 
-These observations suggest that the choice between Virtual Threads and
+\section{Discussion}
+
+Beronić et al.~\cite{9803765} compared different structured concurrency constructs 
+in Java and Kotlin in the context of a multithreaded HTTP server. 
+Their benchmark scenario differs from ours: their server awaited incoming requests 
+and, upon receiving one, executed a task involving object creation, writing the 
+object's information to a file, and returning the data as a response.
+Consistent with our findings, they concluded that both Kotlin's coroutines and 
+Java's virtual threads offer significant performance improvements over traditional 
+JVM-based threads in concurrent applications.
+
+However, our results diverge from theirs regarding memory usage. 
+They reported lower heap usage for Java virtual threads (16–64~MB) compared to 
+Kotlin coroutines (52–99~MB), which contrasts with our measurements. 
+This discrepancy may be attributed to the different workloads: while their 
+benchmark includes a single I/O read-write operation per request, our experiments 
+focus exclusively on read-only I/O tasks.
+
+The work of Navarro et al.~\cite{navarro2023considerations} focused on the
+performance of Quarkus, which relies on Eclipse Vert.x~\cite{vertx}, itself
+built on top of Netty~\cite{netty}. 
+This represents a similar environment to the one evaluated in our experiments
+with Quarkus. 
+Their study also emphasized template rendering, using a blocking HTML template
+engine (Qute). 
+The benchmark they employed is the Fortunes
+test\footnote{\url{https://www.techempower.com/benchmarks}}, which involves
+rendering a simple HTML table with only two data bindings per row.
+
+Their results show that Quarkus with Virtual Threads outperformed the
+traditional thread-per-request model under increased concurrency. Moreover,
+compared to the reactive model, virtual threads demonstrated competitive
+performance, particularly when executing blocking operations within template
+engines.
+In contrast, our experiments revealed that the reactive approach underperforms 
+relative to virtual threads, likely due to the significantly larger number of 
+data bindings in our use cases—namely, the \texttt{Presentation} and \texttt{Stock} 
+models.
+
+The work of Šimatović et al.\cite{vsimatovic2025evaluating} presents an
+evaluation of Java Virtual Threads under different garbage collectors. They
+explored three types of workloads, only one of which was based on a web
+application scenario. This scenario replicated large-scale web scraping by
+executing 1,000 parallel tasks, each consisting of an I/O-bound operation
+followed by CPU-bound string processing.
+
+In this workload, garbage collection activity remained low across all 
+collectors, indicating reduced memory pressure. These findings suggest 
+that Virtual Threads improve memory allocation efficiency and reduce the 
+frequency of garbage collection cycles—particularly when used with 
+concurrent garbage collectors.
+
+Our observations suggest that the choice between Virtual Threads and
 suspendable coroutines should be guided by both runtime behavior and

### 2025-07-29 docs/report/Sections/benchmark.tex
Add report

--- 2025-05-30 docs/article/Sections/benchmark.tex
@@
-\section{Benchmark Implementation}
+\chapter{Benchmark Implementation}
 \label{sec:bench}
@@
 which allows for easy extension and integration of new template engines and
-frameworks. It also includes a set of tests to ensure the correctness of
-implementations and to validate the \textit{HTML} output. It includes two
-different data models, defined as \texttt{Presentation} and \texttt{Stock}, as
-shown in \autoref{lst:stock-class} and \autoref{lst:presentation-class}, 
+frameworks. It also includes a set of tests to ensure the correctness of the
+implementations and to validate the HTML output.
+
+\section{Data Model}
+
+The benchmark includes two different data models, defined as \texttt{Presentation} and \texttt{Stock}, as
+shown in Listing \ref{lst:stock-class} and Listing \ref{lst:presentation-class}, 
 following the proposal of the benchmarks~\cite{templatesbenchmark,springbenchmark}.
@@
 
+\section{Spring WebFlux}
+
 The Spring WebFlux core implementation uses Project Reactor to support a
@@
 % using \textit{continuations}.
+
+\section{Spring MVC}
+
 The Spring MVC implementation uses handlers based solely on the blocking
@@
 
+\section{Quarkus}
+
 The Quarkus implementation also uses handlers based on the blocking interface

### 2025-07-29 docs/report/Sections/introduction.tex
Add report

--- 2025-07-13 docs/article/Sections/introduction.tex
@@
-\section{Introduction}
+\chapter{Introduction}
 
 % Falta adicionar referências
+
+\section{Motivation}
 
@@
 by streaming HTML content progressively. This technique enhances user-perceived performance by
-allowing progressive rendering as data becomes available, significantly reducing the
-\textit{Time to First Byte (TTFB)} and improving perceived load times compared to traditional
-SSR approaches~\cite{wise2024pssr}. In this respect,
+allowing progressive rendering as data becomes available. In this respect,
 PSSR is similar to CSR in that the server initially sends a minimal HTML
@@
 not need to execute JavaScript or make additional requests to retrieve page content.
-The streaming nature of PSSR allows users to see content progressively as it becomes
-available, rather than waiting for the complete page to be rendered server-side,
-thus providing a more responsive user experience with measurably lower TTFB values.
 
@@
 
-With the introduction of \textit{Virtual
-    Threads}~\footnote{\url{https://openjdk.org/jeps/444}} in Java 21, it is now
-possible to execute blocking I/O operations in a scalable, lightweight manner.
-This capability allows legacy template engines—often reliant on blocking
-interfaces—to operate efficiently in high-concurrency, non-blocking
-environments without requiring complex asynchronous programming models. As a
-result, PSSR can now be implemented using familiar HTML templates, simplifying
-development and improving maintainability.
+With the introduction of \textit{Virtual Threads}~\footnote{\url{https://openjdk.org/jeps/444}}
+in Java 21, it is now possible to execute blocking I/O operations in a
+scalable, lightweight manner. This capability allows legacy template
+engines—often reliant on blocking interfaces—to operate efficiently in
+high-concurrency, non-blocking environments without requiring complex
+asynchronous programming models. As a result, PSSR
+can now be implemented using familiar HTML templates, 
+simplifying development and improving maintainability.
+
+\section{Goals}
 
@@
 primary paradigms: \textit{reactive programming} and \textit{coroutines}, both
-of which have been used to achieve asynchronous I/O in the Java ecosystem. As
-an alternative, we investigate whether Java’s Virtual Threads can offer
-comparable performance while preserving the simplicity of synchronous code.
-Section 2 reviews the state-of-the-art in PSSR and template engine design.
-Section 3 outlines the limitations of conventional engines in asynchronous
-settings and presents our proposed approach. Section 4 details the benchmark
-methodology, followed by the results and analysis in Section 5. Conclusions are
-presented in Section 6.
+of which have been used to achieve asynchronous I/O in the Java ecosystem. As an
+alternative, we investigate whether Java’s Virtual Threads can offer comparable
+performance while preserving the simplicity of synchronous code. Chapter 2
+reviews the state-of-the-art in PSSR and template engine design. Chapter 3
+outlines the limitations of conventional engines in asynchronous settings and
+presents our proposed approach.
+
+\section{Document Structure}
+
+Chapter 4 details the benchmark methodology,
+followed by the results and analysis in Chapter 5. Conclusions are presented in Chapter 6.

### 2025-07-29 docs/report/Sections/problem-statement.tex
Add report

--- 2025-05-29 docs/article/Sections/problem-statement.tex
@@
-\section{Problem Statement}
+\chapter{Problem Statement}
 
@@
 % rendering} (PSSR)~\cite{pssr2005}.
+
+\section{Reactive and Suspendable Templates}
+
 In PSSR, the server does not
@@
 Instead, it processes and streams each piece of data to the client as soon as
-it arrives.
-
-Reactive types like \texttt{Observable<T>} of ReactiveX~\cite{rxjava} or Kotlin \texttt{Flow<T>}~\cite{kotlinlang}
+it arrives. Reactive types like \texttt{Observable<T>} of ReactiveX~\cite{rxjava} or Kotlin \texttt{Flow<T>}~\cite{kotlinlang}
 facilitate this by representing data as a sequence of asynchronous events.
@@
 corresponding HTML fragment and immediately flush it to the client. This
-approach is demonstrated in \autoref{lst:presentation-observable}, where each
+approach is demonstrated in Listing \ref{lst:presentation-observable}, where each
 presentation is rendered asynchronously as it is emitted by an
@@
 coroutines~\cite{coroutines_continuations}.
-In \autoref{lst:presentation-flow}, we show an
+In Listing \ref{lst:presentation-flow}, we show an
 equivalent suspend-based implementation using Kotlin's \texttt{Flow}~\cite{wise2024pssr}.
@@
 \begin{center}
-\begin{minipage}{0.50\textwidth}
+\begin{minipage}{0.51\textwidth}
 \begin{lstlisting}[
@@
 \hfill
-\begin{minipage}{0.46\textwidth}
+\begin{minipage}{0.47\textwidth}
 \begin{lstlisting}[
@@
 
+\section{External DSLs and Blocking Interfaces}
+
 By contrast, template engines that use \textit{external} DSLs—such as JStachio,
@@
 These interfaces require the entire data
-model, to be materialized in memory before rendering can begin, which blocks server
+model to be materialized in memory before rendering can begin, which blocks server
 threads during template expansion and significantly limits scalability under
@@
 reintroduces blocking behavior and undermines the benefits of non-blocking
-I/O—especially under high concurrency. \autoref{lst:presentation-jstachio}
+I/O—especially under high concurrency. Listing \ref{lst:presentation-jstachio}
 illustrates this model using a JStachio template, where the engine performs a
@@
 external DSLs—can potentially achieve scalability levels that approach those of
-non-blocking, asynchronous engines.
+non-blocking, asynchronous engines, without requiring developers to adopt complex
+asynchronous programming paradigms.
 

### 2025-07-29 docs/report/Sections/results.tex
Add report

--- 2025-07-19 docs/article/Sections/results.tex
@@
-\section{Results}
+\chapter{Results}
 
-This section presents the evaluation results organized into four main parts:
-Section~\ref{sec:environment} details the testing environment and hardware
-specifications; Section~\ref{sec:presentations-results} presents scalability
-results for the \texttt{Presentations} data model;
-Section~\ref{sec:stocks-results} analyzes performance with the more complex
-\texttt{Stocks} data model; and Section~\ref{sec:memory-analysis} provides
-detailed memory consumption and resource utilization analysis.
-
-\subsection{Environment Specifications}
-\label{sec:environment}
-
-All benchmarks were conducted on a GitHub-hosted virtual machine under GitHub
-Actions, running Ubuntu 24.04.2 LTS with an AMD EPYC 7763 64-Core Processor
-(x86\_64 architecture) operating at 3.24 GHz, configured with 2 CPU cores and 2
-threads per core, and 7.8 GB of available RAM. The system utilizes a 75 GB SSD
-with ext4 filesystem, achieving 1.5 GB/s write throughput in basic I/O tests.
-Network connectivity is provided through a 1500 MTU Ethernet interface. All
-tests were conducted on OpenJDK 21 (Corretto build) with the G1 garbage
-collector enabled by default. The JVM arguments used were: \texttt{-Xms512m
-     -Xmx16g}. Each test was executed 5 times per configuration to ensure result
-reliability.
-
-All template engines were configured to use UTF-8 encoding and disable
-automatic HTML escaping to ensure fair comparison and consistent output across
-all template engines. Caching behaviour was left as the default for each
-engine, allowing them to optimize template loading and rendering as per their
-design. These configurations ensure that all engines operate under equivalent
-conditions, with template loading, encoding, and escaping behavior normalized
-across implementations.
+All the following benchmarks were conducted on a Github-hosted virtual machine
+under GitHub Actions, running Ubuntu 22.04 LTS with a 4-core, 8-thread CPU and
+16GB of RAM\@. All tests were conducted on the OpenJDK VM Corretto 21. The JVM
+was configured with a minimum heap size of 512MB and a maximum heap size of
+16GB\@.
 
@@
 
-Both Quarkus and Spring MVC implementations were configured with an 8KB output
-buffer size to ensure consistency, despite Quarkus not enabling PSSR at this
-size and Spring MVC not supporting PSSR for the tested templates. Testing with
-a reduced 512B buffer size showed only a 0.046\% performance difference with
-Rocker, indicating negligible impact. Both frameworks use unlimited platform
-thread pools, enabling on-demand thread creation up to system limits for
-maximum throughput and scalability observation under high concurrent load.
+Both the Quarkus and Spring MVC implementations were configured with an output
+buffer size of 8KB, despite the Quarkus implementation not enabling PSSR at
+this size. This configuration was chosen to ensure consistency across Spring
+MVC and Quarkus, as the Spring MVC implementation does not support PSSR for the
+tested templates.
 
 Since the obtained results for JMeter and Apache Bench show no significant
-differences, only the JMeter results will be presented. Statistical analysis of
-test configurations revealed an average absolute percentage difference of
-2.83\% between the two load testing tools. While individual approach
-differences ranged from -16.53\% to +14.66\% at the extremes, the consistent
-directional bias and small magnitude of differences indicate that both tools
-provide comparable performance measurements. Given this statistical equivalence
-and the need for brevity, only JMeter results are presented, as they provide
-representative performance characteristics across all tested configurations.
+differences, only the JMeter results will be presented.
 
-\subsection{Scalability results for the \texttt{Presentations} class} \label{sec:presentations-results}
+\section{Presentations Results}
 
@@
      \includegraphics[width=0.8\textwidth]{./Graphs/presentations-webflux-jmeter.png}
-     \caption{Throughput (requests per second) scalability results for Spring Webflux with \texttt{Presentations} class}\label{fig:presentations-webflux-jmeter}
+     \caption{Presentation Benchmark Results in Spring WebFlux with JMeter}\label{fig:presentations-webflux-jmeter}
 \end{figure}
@@
      \includegraphics[width=0.8\textwidth]{./Graphs/presentations-springmvc-jmeter.png}
-     \caption{Throughput (requests per second) scalability results for Spring MVC with \texttt{Presentations} class}\label{fig:presentations-springmvc-jmeter}
+     \caption{Presentation Benchmark Results in Spring MVC with JMeter}\label{fig:presentations-springmvc-jmeter}
 \end{figure}
@@
      \includegraphics[width=0.8\textwidth]{./Graphs/presentations-quarkus-jmeter.png}
-     \caption{Throughput (requests per second) scalability results for Quarkus with \texttt{Presentations} class}\label{fig:presentations-quarkus-jmeter}
+     \caption{Presentation Benchmark Results in Quarkus with JMeter}\label{fig:presentations-quarkus-jmeter}
 \end{figure}
@@
 
-\subsection{Scalability results for the \texttt{Stocks} class} \label{sec:stocks-results}
+\section{Stocks Results}
 
@@
      \includegraphics[width=0.8\textwidth]{./Graphs/stocks-webflux-jmeter.png}
-     \caption{Throughput (requests per second) scalability results for Spring Webflux with \texttt{Stocks} class}\label{fig:stocks-webflux-jmeter}
+     \caption{Stocks Benchmark Results in Spring WebFlux with JMeter}\label{fig:stocks-webflux-jmeter}
 \end{figure}
@@
 engines and approaches as the previous benchmark, but with a more complex data
-model: the Stock class, which includes 20 instances and approximately two times
-as many data bindings. With this data model, the scalability of the engines
-remains largely unchanged; however, throughput is reduced across all engines.
-Compared to the Presentation benchmark, Jstachio using virtual threads
-experienced a more pronounced decrease in performance relative to the
+model: the Stock class, which includes 20 instances and approximately two
+times as many data bindings. With this data model, the scalability of the
+engines remains largely unchanged; however, throughput is reduced across all
+engines. Compared to the Presentation benchmark, the approaches using Virtual
+Threads experience a more pronounced decrease in performance relative to the
 \textit{Reactive} and \textit{Suspending} approaches, with Jstachio using
@@
 3,090 requests per second, while HtmlFlow using a reactive approach achieves
-3,026 requests per second. This indicates that the more pronounced decrease in
-performance for Jstachio using Virtual Threads is likely due to the specific
-implementation of the template engine, rather than the use of Virtual Threads
-itself.
+3,026 requests per second.
 
-The overall throughput reduction across all engines is expected, as the Stock
-class contains more data properties than the \texttt{Presentation} class,
-adding overhead related to the data binding process of each template engine.
+The overall throughput reduction is expected, as the Stock class contains more
+data properties than the \texttt{Presentation} class, adding overhead related
+to the data binding process of each template engine.
 
@@
      \includegraphics[width=0.8\textwidth]{./Graphs/stocks-springmvc-jmeter.png}
-     \caption{Throughput (requests per second) scalability results for Spring MVC with \texttt{Stocks} class}\label{fig:stocks-springmvc-jmeter}
+     \caption{Stocks Benchmark Results in Spring MVC with JMeter}\label{fig:stocks-springmvc-jmeter}
 \end{figure}
@@
      \includegraphics[width=0.8\textwidth]{./Graphs/stocks-quarkus-jmeter.png}
-     \caption{Throughput (requests per second) scalability results for Quarkus with \texttt{Stocks} class}\label{fig:stocks-quarkus-jmeter}
+     \caption{Stocks Benchmark Results in Quarkus with JMeter}\label{fig:stocks-quarkus-jmeter}
 \end{figure}
@@
 showed itself the most effective at enabling PSSR, mostly due to its native
-support for publish-subscribe interfaces, allowing for content to be streamed
-as data becomes available, instead of when the response buffer is flushed.
-Quarkus also enabled PSSR effectively, but it required additional configuration
-of the \texttt{OutputBuffer} size to achieve the same results as Spring
-Webflux. The Spring MVC implementation, on the other hand, did not enable PSSR
-for the tested templates.
+support for Publish and Subscriber interfaces, which allow for HTML content to
+be progressively streamed to the client. Quarkus also enabled PSSR effectively,
+but it required additional configuration of the \texttt{OutputBuffer} size to
+achieve the same results as Spring Webflux. The Spring MVC implementation, on
+the other hand, did not enable PSSR for the tested templates.
 
-However, it is important to acknowledge the limitations of our chosen data
-models for generalizability. The tested templates used relatively simple data
-structures with limited nesting and straightforward property bindings.
-Real-world applications often involve deeply nested data structures, complex
-conditional logic, and iterative rendering over thousands of items. Under such
-conditions, we anticipate significantly different performance characteristics:
-increased memory consumption due to object traversal overhead, higher CPU
-utilization for complex evaluations, and altered scalability patterns where
-non-blocking approaches may become more advantageous. The performance
-degradation observed with our Stock class benchmark—which merely doubled the
-number of properties—suggests that higher template complexity would result in
-substantially higher performance degradation. Future work should investigate
-these scenarios with more realistic data models to better understand
-performance boundaries and optimization strategies for complex PSSR
-implementations.
-
-\subsection{Memory Consumption and Resource Utilization Analysis}
-\label{sec:memory-analysis}
-
-This section evaluates the memory and CPU resource usage characteristics of
-structured concurrency approaches, specifically comparing Virtual Threads and
-suspendable coroutines.
-
-Resource utilization data was collected using VisualVM during 30-second
-benchmark runs, capturing detailed CPU and memory usage patterns under
-sustained load conditions. Each profiling session monitored system performance
-during the 1 to 128 concurrent user load test, providing comprehensive insights
-into runtime behavior.
-
-\subsubsection{Virtual Threads vs Structured Concurrency}
-
-\begin{center}
-     \begin{minipage}[t]{0.48\textwidth}
-          \centering
-          \includegraphics[width=1.0\textwidth,height=1.0\textwidth]{./Graphs/cpu-susp.png}
-          \captionof{figure}{CPU utilization profiling with HtmlFlow-Susp in Spring WebFlux during 30-second load test}
-          \label{fig:cpu-susp}
-     \end{minipage}
-     \hfill
-     \begin{minipage}[t]{0.48\textwidth}
-          \centering
-          \includegraphics[width=1.0\textwidth,height=1.0\textwidth]{./Graphs/cpu-virt.png}
-          \captionof{figure}{CPU utilization profiling with HtmlFlow-Virtual in Spring WebFlux during 30-second load test}
-          \label{fig:cpu-virt}
-     \end{minipage}
-\end{center}
-
-\begin{center}
-     \begin{minipage}[t]{0.48\textwidth}
-          \centering
-          \includegraphics[width=1.0\textwidth,height=1.0\textwidth]{./Graphs/mem-susp.png}
-          \captionof{figure}{Memory utilization and garbage collection behavior with HtmlFlow-Susp in Spring WebFlux during 30-second load test}
-          \label{fig:gc-susp}
-     \end{minipage}
-     \hfill
-     \begin{minipage}[t]{0.48\textwidth}
-          \centering
-          \includegraphics[width=1.0\textwidth,height=1.0\textwidth]{./Graphs/mem-virt.png}
-          \captionof{figure}{Memory utilization and garbage collection behavior with HtmlFlow-Virtual in Spring WebFlux during 30-second load test}
-          \label{fig:gc-virt}
-     \end{minipage}
-\end{center}
-
-Figures~\ref{fig:cpu-virt} and~\ref{fig:cpu-susp} show CPU utilization profiles
-for the \texttt{HtmlFlow-Virtual} and \texttt{HtmlFlow-Susp} implementations,
-respectively. The subsequent gradual increase in CPU utilization reflects the
-ramp-up of concurrent users during the benchmark. \texttt{HtmlFlow-Virtual}
-exhibits a distinctive bell-shaped curve, rapidly climbing from near-zero to
-peak CPU usage of around 50\% during the sustained load phase, maintaining
-moderate utilization during the test period, and then dropping sharply back to
-baseline. \texttt{HtmlFlow-Susp} displays a similar pattern but with a slightly
-lower peak CPU utilization of approximately 42\%, showing marginally lower
-intensive CPU usage during the sustained load phase. Both profiles demonstrate
-comparable resource consumption patterns, with the virtual thread
-implementation showing only slightly higher CPU demands. The GC activity
-indicators in both profiles remain relatively low throughout the test duration,
-confirming that garbage collection overhead does not significantly impact CPU
-utilization in either approach.
-
-Figures~\ref{fig:gc-susp} and~\ref{fig:gc-virt} show similar memory usage for
-both approaches, with peaks around 300~MB during load. The initial spike to
-450~MB in both graphs corresponds to the applications' bootstrapping phase.
-Overall, these results indicate that under the tested conditions, Virtual
-Threads and suspendable coroutines exhibit comparable memory consumption.
-
-Virtual Threads allocate a full call stack per thread, which is unmounted and
-stored on the heap when suspended (e.g., during blocking I/O). In contrast,
-Kotlin coroutines compile into state machines that capture only the minimal
-execution state needed to resume computation. This typically allows coroutines
-to scale more efficiently in terms of memory usage, especially under high
-concurrency. These findings contrast with prior research, such as Beronić et
-al.~\cite{9803765}, which reported lower heap usage for Java Virtual Threads (16--64~MB)
-compared to Kotlin coroutines (52--99~MB), albeit under different workloads.
-
-These observations suggest that the choice between Virtual Threads and
-suspendable coroutines should be guided by both runtime behavior and
-development preferences. Virtual Threads offer the simplicity of synchronous
-programming with competitive performance in moderate-load scenarios. In
-contrast, suspendable coroutines, while requiring an asynchronous programming
-model, may offer better efficiency in applications with higher concurrency,
-deeper stacks, or more dynamic workloads.
+% Additionally, the results show that approaches using Virtual Threads are able
+% to scale as effectively as those using reactive programming or Kotlin
+% coroutines, allowing \textit{external} DSLs to be used in a non-blocking
+% context, effectively enabling PSSR\@.

### 2025-07-29 docs/report/Sections/state-of-the-art-web-servers.tex
Add report

--- 2025-05-30 docs/article/Sections/state-of-the-art-web-servers.tex
@@
-\subsection{Web Framework Architectures and Approaches to PSSR}
+\section{Web Framework Architectures and Approaches to PSSR}
 
@@
 
-\subsubsection{Spring MVC}
+\subsection{Spring MVC}
 
@@
 
-In the example shown in \autoref{lst:streaming-response-body}, HTML content is written and flushed in discrete chunks,
+In the example shown in Listing \ref{lst:streaming-response-body}, HTML content is written and flushed in discrete chunks,
 allowing the client to progressively render the response as it arrives, while
@@
 
-\subsubsection{Spring WebFlux}
+\subsection{Spring WebFlux}
 
@@
 
-\autoref{lst:pssr-webflux} shows a method that emits one HTML chunk every 500 milliseconds. As
+Listing \ref{lst:pssr-webflux} shows a method that emits one HTML chunk every 500 milliseconds. As
 each chunk is generated, it is immediately written to the HTTP response,
@@
 
-\subsubsection{Quarkus}
+\subsection{Quarkus}
 
@@
 For traditional imperative-style request handling, Quarkus provides integration
-with JAX-RS~\cite{burke2013restful}, where resource methods can return synchronous types or
+with JAX-RS, where resource methods can return synchronous types or
 asynchronous constructs such as \texttt{CompletionStage}. To enable 
@@
 
-\autoref{lst:pssr-quarkus} shows how \texttt{StreamingOutput} can be used to send
+Listing \ref{lst:pssr-quarkus} shows how \texttt{StreamingOutput} can be used to send
 partial HTML content in chunks, allowing browsers to progressively render the

### 2025-07-29 docs/report/rfG48.tex
Add report

% Classe do documento e parâmetros gerais.
\documentclass[a4paper,openright,twoside,11pt]{report}

% Packages utilizadas e respetivos parâmetros.
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\addto{\captionsenglish}{\renewcommand{\bibname}{References}}
\addto{\captionsenglish}{\renewcommand{\contentsname}{Contents}}
\usepackage{lipsum} % gerador de texto
\usepackage{graphicx}
\usepackage{url}
\usepackage[Algoritmo]{algorithm}
\usepackage{algorithmicx}
\usepackage{algpseudocode}
\usepackage[dvipsnames]{xcolor}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{listingstyles}
\usepackage{amssymb}
\usepackage{bookmark}

\renewcommand{\algorithmicrequire}{\textbf{Dados: }}
\renewcommand{\algorithmicensure}{\textbf{Resultado: }}

% Definições das dimensões das páginas
\setlength{\textheight}{24.00cm}
\setlength{\textwidth}{15.50cm}
\setlength{\topmargin}{0.35cm}
\setlength{\headheight}{0cm}
\setlength{\headsep}{0cm}
\setlength{\oddsidemargin}{0.25cm}
\setlength{\evensidemargin}{0.25cm}

%\renewcommand{\baselinestretch}{1}

% Página inicial (capa)
\title{
   \vspace{-50mm}
   \begin{minipage}[l]{\textwidth}
      \hspace{-20mm}\resizebox{75mm}{!}{\includegraphics{./figures/logoISELnew2.png}}\\
   \end{minipage}\\[20mm]
   {\fontsize{22pt}{26pt}\selectfont\textbf{Non-Blocking Progressive Server-Side Rendering Benchmark}}\\[10mm]
   {\fontsize{14pt}{18pt}\selectfont\textbf{A Performance Comparison of Progressive SSR Techniques}}\\[5mm]
}

% Nome dos autores (um por linha)
\author{
\begin{tabular}{ll}
             & Bernardo Pereira \\[50mm]
\end{tabular}}

\date{
\begin{tabular}{ll}
  {Supervisor:} & Miguel Carvalho, ISEL \\
\end{tabular}\\[10mm]
% Deixar o indicador respetivo em função da versão do relatório.
Final Report written for Project and Seminar\\
BSc in Computer Science and Engineering\\[20mm]
June 2025}

\begin{document}
\pagenumbering{roman}
\thispagestyle{empty}
\maketitle

\baselineskip 18pt % line spacing: 12pt for single, 18pt for 1 1/2, and 24pt for double spacing

\newpage
\thispagestyle{empty}
% Fim da contracapa

% Página com identificação completa (número e nome) e assinaturas do(s) estudante(s) e do(s) orientador(es)
\cleardoublepage
\setcounter{page}{1}
\begin{center}
  \textsc{\LARGE Instituto Superior de Engenharia de Lisboa}\\[30mm]

  {\fontsize{22pt}{26pt}\selectfont\textbf{Non-Blocking Progressive Server-Side Rendering Benchmark}}\\[10mm]
  {\fontsize{14pt}{18pt}\selectfont\textbf{A Performance Comparison of Progressive SSR Techniques}}\\[30mm]

  \begin{tabular}{rl}
    50493 & Bernardo André Batista Correia Murta Pereira \\[10mm]
          & \rule{75mm}{0.5pt}                           \\[5mm]
  \end{tabular}\\[10mm]

  \begin{tabular}{rl}
    Supervisor: & Miguel Carvalho, ISEL \\[10mm]
                & \rule{75mm}{0.5pt}    \\[5mm]
  \end{tabular}\\[10mm]

  Final report written for Project and Seminar\\ BSc in Computer Science and Engineering\\[20mm]
  June 2025\\
\end{center}

% Página de resumo em Português
\cleardoublepage
\chapter*{Abstract}

Modern web applications increasingly demand rendering techniques that optimize performance, responsiveness, and scalability. Progressive Server-Side Rendering (PSSR) bridges the gap between Server-Side Rendering and Client-Side Rendering by progressively streaming HTML content, improving perceived load times.

However, traditional HTML template engines often rely on blocking interfaces
that hinder their use in asynchronous, non-blocking contexts required for PSSR.

This work explores how Java Virtual Threads, introduced in Java 21, enable
non-blocking execution of blocking I/O operations, allowing the reuse of
traditional template engines for PSSR without complex asynchronous programming
models. We benchmark multiple engines across Spring WebFlux, Spring MVC, and
Quarkus using reactive, suspendable, and virtual-thread-based approaches.
Results show that virtual threads allow blocking engines to scale comparably to
those designed for non-blocking I/O, achieving high throughput and
responsiveness under load. This demonstrates that Virtual Threads provide a
compelling path to simplify the implementation of PSSR with familiar HTML
templates, significantly lowering the barrier to entry while maintaining
performance.

\vspace{5mm}

{\bf Keywords:} Web Templates; Server-Side Rendering; Non-Blocking; Java Virtual Threads; Asynchronous.

%% Página de agradecimentos
%\cleardoublepage
%\chapter*{Agradecimentos}
%Texto dos agradecimentos. É opcional.\\

% Geração do índice de conteúdos
\cleardoublepage
\tableofcontents \cleardoublepage

% Geração do índice de figuras e de tabelas
%\listoffigures \cleardoublepage
%\listoftables \cleardoublepage

% Reiniciar a numeração de páginas
\setcounter{page}{1}
\pagenumbering{arabic}

% Capitulo 1
\include{Sections/introduction}

% Capitulo 2
\include{Sections/state-of-the-art}

% Capitulo 3
\include{Sections/problem-statement}

% Capitulo 4
\include{Sections/benchmark}

% Capitulo 5
\include{Sections/results}

% Capitulo 6
\include{Sections/conclusion}

% Referências
\bibliographystyle{unsrt}
\bibliography{bibliography}
\addcontentsline{toc}{chapter}{References}

\end{document}

### 2025-07-29 docs/article/Sections/results.tex
section 6 improvements

@@
 
-Beronić et al.~\cite{9803765} compared different structured concurrency constructs 
-in Java and Kotlin in the context of a multithreaded HTTP server. 
-Their benchmark scenario differs from ours: their server awaited incoming requests 
-and, upon receiving one, executed a task involving object creation, writing the 
-object's information to a file, and returning the data as a response.
-Consistent with our findings, they concluded that both Kotlin's coroutines and 
-Java's virtual threads offer significant performance improvements over traditional 
-JVM-based threads in concurrent applications.
+Beronić et al.~\cite{9803765} compared different structured concurrency
+constructs in Java and Kotlin in the context of a multithreaded HTTP server.
+Their benchmark scenario differs from ours: their server awaited incoming
+requests and, upon receiving one, executed a task involving object creation,
+writing the object's information to a file, and returning the data as a
+response. Consistent with our findings, they concluded that both Kotlin's
+coroutines and Java's virtual threads offer significant performance
+improvements over traditional JVM-based threads in concurrent applications.
 
-However, our results diverge from theirs regarding memory usage. 
-They reported lower heap usage for Java virtual threads (16–64~MB) compared to 
-Kotlin coroutines (52–99~MB), which contrasts with our measurements. 
-This discrepancy may be attributed to the different workloads: while their 
-benchmark includes a single I/O read-write operation per request, our experiments 
-focus exclusively on read-only I/O tasks.
+Our results diverge from theirs regarding memory usage. They reported lower
+heap usage for Java virtual threads (16–64~MB) compared to Kotlin coroutines
+(52–99~MB), which contrasts with our measurements. This discrepancy may be
+attributed to the different workloads: while their benchmark includes a single
+I/O read-write operation per request, our experiments focus exclusively on
+read-only I/O tasks.
 
@@
 performance of Quarkus, which relies on Eclipse Vert.x~\cite{vertx}, itself
-built on top of Netty~\cite{netty}. 
-This represents a similar environment to the one evaluated in our experiments
-with Quarkus. 
-Their study also emphasized template rendering, using a blocking HTML template
-engine (Qute). 
-The benchmark they employed is the Fortunes
+built on top of Netty~\cite{netty}. This represents a similar environment to
+the one evaluated in our experiments with Quarkus. Their study also emphasized
+template rendering, using a blocking HTML template engine (Qute). The benchmark
+they employed is the Fortunes
 test\footnote{\url{https://www.techempower.com/benchmarks}}, which involves
@@
 performance, particularly when executing blocking operations within template
-engines.
-In contrast, our experiments revealed that the reactive approach underperforms 
-relative to virtual threads, likely due to the significantly larger number of 
-data bindings in our use cases—namely, the \texttt{Presentation} and \texttt{Stock} 
-models.
+engines. In contrast, our experiments revealed that the reactive approach
+underperforms relative to virtual threads, likely due to the significantly
+larger number of data bindings in our use cases—namely, the
+\texttt{Presentation} and \texttt{Stock} models.
 
@@
 In this workload, garbage collection activity remained low across all
-collectors, indicating reduced memory pressure. These findings suggest 
-that Virtual Threads improve memory allocation efficiency and reduce the 
-frequency of garbage collection cycles—particularly when used with 
-concurrent garbage collectors.
+collectors, indicating reduced memory pressure. These findings suggest that
+Virtual Threads improve memory allocation efficiency and reduce the frequency
+of garbage collection cycles—particularly when used with concurrent garbage
+collectors.
 
@@
 suspendable coroutines should be guided by both runtime behavior and
-development preferences. Virtual Threads offer the simplicity of synchronous
-programming with competitive performance in moderate-load scenarios. In
-contrast, suspendable coroutines, while requiring an asynchronous programming
-model, may offer better efficiency in applications with higher concurrency,
-deeper stacks, or more dynamic workloads.
+development constraints. Virtual Threads offer the compelling advantage of
+preserving synchronous programming semantics while achieving competitive
+performance across moderate to high-load scenarios—a significant benefit for
+teams migrating existing codebases or working with legacy template engines.
+Suspendable coroutines, while requiring developers to adopt asynchronous
+programming paradigms, may provide greater efficiency in scenarios involving
+high concurrency, deeper call stacks, or highly dynamic workloads.
+
+The framework-specific variations observed in our results—particularly the
+performance differences between Spring WebFlux and Quarkus
+implementations—suggest that the underlying reactive runtime architecture
+significantly influences the effectiveness of each approach. This finding
+underscores the importance of considering not only the concurrency model but
+also the specific framework ecosystem when making architectural decisions for
+PSSR implementations.

### 2025-07-30 docs/article/Sections/results.tex
Fix paragraph.

@@
 performance, particularly when executing blocking operations within template
-engines. In contrast, our experiments revealed that the reactive approach
-underperforms relative to virtual threads, likely due to the significantly
-larger number of data bindings in our use cases—namely, the
-\texttt{Presentation} and \texttt{Stock} models.
+engines.
+In contrast, our experiments revealed that the reactive approach in Spring
+WebFlux performs better than virtual threads in the same experiment conducted
+with Quarkus, suggesting that WebFlux provides a more efficient infrastructure
+for managing reactive processing flows.
 

### 2025-08-06 docs/article/software-3702556.tex
Merge Sections in single File from MDPI

%  LaTeX support: latex@mdpi.com 
%  For support, please attach all files needed for compiling as well as the log file, and specify your operating system, LaTeX version, and LaTeX editor.

%=================================================================
\documentclass[software,article,accept,pdftex,moreauthors]{Definitions/mdpi}
%\usepackage[dvipsnames]{xcolor}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{listingstyles}
\definecolor{PineGreen}{RGB}{1, 121, 111} 
%\documentclass[preprints,article,submit,pdftex,moreauthors]{Definitions/mdpi} 
% For posting an early version of this manuscript as a preprint, you may use "preprints" as the journal. Changing "submit" to "accept" before posting will remove line numbers.

% Below journals will use APA reference format:
% admsci, behavsci, businesses, econometrics, economies, education, ejihpe, famsci, games, humans, ijcs, ijfs, journalmedia, jrfm, languages, psycholint, publications, tourismhosp, youth

% Below journals will use Chicago reference format:
% arts, genealogy, histories, humanities, jintelligence, laws, literature, religions, risks, socsci

%--------------------
% Class Options:
%--------------------
%----------r

% journal
%----------
% Choose between the following MDPI journals:
% accountaudit, acoustics, actuators, addictions, adhesives, admsci, adolescents, aerobiology, aerospace, agriculture, agriengineering, agrochemicals, agronomy, ai, air, algorithms, allergies, alloys, amh, analytica, analytics, anatomia, anesthres, animals, antibiotics, antibodies, antioxidants, applbiosci, appliedchem, appliedmath, appliedphys, applmech, applmicrobiol, applnano, applsci, aquacj, architecture, arm, arthropoda, arts, asc, asi, astronomy, atmosphere, atoms, audiolres, automation, axioms, bacteria, batteries, bdcc, behavsci, beverages, biochem, bioengineering, biologics, biology, biomass, biomechanics, biomed, biomedicines, biomedinformatics, biomimetics, biomolecules, biophysica, biosensors, biosphere, biotech, birds, blockchains, bloods, blsf, brainsci, breath, buildings, businesses, cancers, carbon, cardiogenetics, catalysts, cells, ceramics, challenges, chemengineering, chemistry, chemosensors, chemproc, children, chips, cimb, civileng, cleantechnol, climate, clinbioenerg, clinpract, clockssleep, cmd, cmtr, coasts, coatings, colloids, colorants, commodities, complications, compounds, computation, computers, condensedmatter, conservation, constrmater, cosmetics, covid, crops, cryo, cryptography, crystals, csmf, ctn, curroncol, cyber, dairy, data, ddc, dentistry, dermato, dermatopathology, designs, devices, diabetology, diagnostics, dietetics, digital, disabilities, diseases, diversity, dna, drones, dynamics, earth, ebj, ecm, ecologies, econometrics, economies, education, eesp, ejihpe, electricity, electrochem, electronicmat, electronics, encyclopedia, endocrines, energies, eng, engproc, ent, entomology, entropy, environments, epidemiologia, epigenomes, esa, est, famsci, fermentation, fibers, fintech, fire, fishes, fluids, foods, forecasting, forensicsci, forests, fossstud, foundations, fractalfract, fuels, future, futureinternet, futureparasites, futurepharmacol, futurephys, futuretransp, galaxies, games, gases, gastroent, gastrointestdisord, gastronomy, gels, genealogy, genes, geographies, geohazards, geomatics, geometry, geosciences, geotechnics, geriatrics, glacies, grasses, greenhealth, gucdd, hardware, hazardousmatters, healthcare, hearts, hemato, hematolrep, heritage, higheredu, highthroughput, histories, horticulturae, hospitals, humanities, humans, hydrobiology, hydrogen, hydrology, hygiene, idr, iic, ijerph, ijfs, ijgi, ijmd, ijms, ijns, ijpb, ijt, ijtm, ijtpp, ime, immuno, informatics, information, infrastructures, inorganics, insects, instruments, inventions, iot, j, jal, jcdd, jcm, jcp, jcs, jcto, jdad, jdb, jeta, jfb, jfmk, jimaging, jintelligence, jlpea, jmahp, jmmp, jmms, jmp, jmse, jne, jnt, jof, joitmc, joma, jop, jor, journalmedia, jox, jpbi, jpm, jrfm, jsan, jtaer, jvd, jzbg, kidney, kidneydial, kinasesphosphatases, knowledge, labmed, laboratories, land, languages, laws, life, lights, limnolrev, lipidology, liquids, literature, livers, logics, logistics, lubricants, lymphatics, machines, macromol, magnetism, magnetochemistry, make, marinedrugs, materials, materproc, mathematics, mca, measurements, medicina, medicines, medsci, membranes, merits, metabolites, metals, meteorology, methane, metrics, metrology, micro, microarrays, microbiolres, microelectronics, micromachines, microorganisms, microplastics, microwave, minerals, mining, mmphys, modelling, molbank, molecules, mps, msf, mti, multimedia, muscles, nanoenergyadv, nanomanufacturing, nanomaterials, ncrna, ndt, network, neuroglia, neurolint, neurosci, nitrogen, notspecified, nursrep, nutraceuticals, nutrients, obesities, oceans, ohbm, onco, oncopathology, optics, oral, organics, organoids, osteology, oxygen, parasites, parasitologia, particles, pathogens, pathophysiology, pediatrrep, pets, pharmaceuticals, pharmaceutics, pharmacoepidemiology, pharmacy, philosophies, photochem, photonics, phycology, physchem, physics, physiologia, plants, plasma, platforms, pollutants, polymers, polysaccharides, populations, poultry, powders, preprints, proceedings, processes, prosthesis, proteomes, psf, psych, psychiatryint, psychoactives, psycholint, publications, purification, quantumrep, quaternary, qubs, radiation, reactions, realestate, receptors, recycling, regeneration, religions, remotesensing, reports, reprodmed, resources, rheumato, risks, robotics, rsee, ruminants, safety, sci, scipharm, sclerosis, seeds, sensors, separations, sexes, signals, sinusitis, siuj, skins, smartcities, sna, societies, socsci, software, soilsystems, solar, solids, spectroscj, sports, standards, stats, std, stresses, surfaces, surgeries, suschem, sustainability, symmetry, synbio, systems, tae, targets, taxonomy, technologies, telecom, test, textiles, thalassrep, therapeutics, thermo, timespace, tomography, tourismhosp, toxics, toxins, transplantology, transportation, traumacare, traumas, tropicalmed, universe, urbansci, uro, vaccines, vehicles, venereology, vetsci, vibration, virtualworlds, viruses, vision, waste, water, wem, wevj, wild, wind, women, world, youth, zoonoticdis

%---------
% article
%---------
% The default type of manuscript is "article", but can be replaced by: 
% abstract, addendum, article, benchmark, book, bookreview, briefcommunication, briefreport, casereport, changes, clinicopathologicalchallenge, comment, commentary, communication, conceptpaper, conferenceproceedings, correction, conferencereport, creative, datadescriptor, discussion, entry, expressionofconcern, extendedabstract, editorial, essay, erratum, fieldguide, hypothesis, interestingimages, letter, meetingreport, monograph, newbookreceived, obituary, opinion, proceedingpaper, projectreport, reply, retraction, review, perspective, protocol, shortnote, studyprotocol, supfile, systematicreview, technicalnote, viewpoint, guidelines, registeredreport, tutorial,  giantsinurology, urologyaroundtheworld
% supfile = supplementary materials

%----------
% submit
%----------
% The class option "submit" will be changed to "accept" by the Editorial Office when the paper is accepted. This will only make changes to the frontpage (e.g., the logo of the journal will get visible), the headings, and the copyright information. Also, line numbering will be removed. Journal info and pagination for accepted papers will also be assigned by the Editorial Office.

%------------------
% moreauthors
%------------------
% If there is only one author the class option oneauthor should be used. Otherwise use the class option moreauthors.

%---------
% pdftex
%---------
% The option pdftex is for use with pdfLaTeX. Remove "pdftex" for (1) compiling with LaTeX & dvi2pdf (if eps figures are used) or for (2) compiling with XeLaTeX.

%=================================================================
% MDPI internal commands - do not modify
\firstpage{1} 
\makeatletter 
\setcounter{page}{\@firstpage} 
\makeatother
\pubvolume{1}
\issuenum{1}
\articlenumber{0}
\pubyear{2025}
\copyrightyear{2025}
\externaleditor{Firstname Lastname} % More than 1 editor, please add `` and '' before the last editor name
\datereceived{30 May 2025} 
\daterevised{19 July 2025} % Comment out if no revised date
\dateaccepted{ } 
\datepublished{ } 
%\datecorrected{} % For corrected papers: "Corrected: XXX" date in the original paper.
%\dateretracted{} % For retracted papers: "Retracted: XXX" date in the original paper.
\hreflink{https://doi.org/} % If needed use \linebreak
%\doinum{}
%\pdfoutput=1 % Uncommented for upload to arXiv.org
%\CorrStatement{yes}  % For updates
%\longauthorlist{yes} % For many authors that exceed the left citation part

%=================================================================
% Add packages and commands here. The following packages are loaded in our class file: fontenc, inputenc, calc, indentfirst, fancyhdr, graphicx, epstopdf, lastpage, ifthen, float, amsmath, amssymb, lineno, setspace, enumitem, mathpazo, booktabs, titlesec, etoolbox, tabto, xcolor, colortbl, soul, multirow, microtype, tikz, totcount, changepage, attrib, upgreek, array, tabularx, pbox, ragged2e, tocloft, marginnote, marginfix, enotez, amsthm, natbib, hyperref, cleveref, scrextend, url, geometry, newfloat, caption, draftwatermark, seqsplit
% cleveref: load \crefname definitions after \begin{document}

%=================================================================
% Please use the following mathematics environments: Theorem, Lemma, Corollary, Proposition, Characterization, Property, Problem, Example, ExamplesandDefinitions, Hypothesis, Remark, Definition, Notation, Assumption
%% For proofs, please use the proof environment (the amsthm package is loaded by the MDPI class).

%=================================================================
% Full title of the paper (Capitalized)
\Title{Enabling Progressive Server-Side Rendering for Traditional Web Template Engines with Java Virtual Threads}

% MDPI internal command: Title for citation in the left column
\TitleCitation{Enabling Progressive Server-Side Rendering for Traditional Template Engines with Java Virtual Threads}

% Author Orchid ID: enter ID or remove command
\newcommand{\orcidauthorA}{0009-0009-1863-3782} % Add \orcidA{} behind the author's name
\newcommand{\orcidauthorB}{0000-0002-4281-3195} % Add \orcidB{} behind the author's name

% Authors, for the paper (add full first names)
\Author{\hl{Bernardo Pereira} %MDPI: 1. Please carefully check the accuracy of names and affiliations. 
 \orcidA{} and \hl{Fernando Miguel Carvalho} %MDPI: The names highlighted are different from the ones submitted online at susy.mdpi.com. Please confirm which are correct. 
 *\orcidB{}}

%\longauthorlist{yes}

% MDPI internal command: Authors, for metadata in PDF
\AuthorNames{Bernardo Pereira, Miguel Carvalho}

% MDPI internal command: Authors, for citation in the left column, only choose below one of them according to the journal style
% If this is a Chicago style journal 
% (arts, genealogy, histories, humanities, jintelligence, laws, literature, religions, risks, socsci): 
% Lastname, Firstname, Firstname Lastname, and Firstname Lastname.

% If this is a APA style journal 
% (admsci, behavsci, businesses, econometrics, economies, education, ejihpe, games, humans, ijfs, journalmedia, jrfm, languages, psycholint, publications, tourismhosp, youth): 
% Lastname, F., Lastname, F., \& Lastname, F.

% If this is a ACS style journal (Except for the above Chicago and APA journals, all others are in the ACS format): 
% Lastname, F.; Lastname, F.; Lastname, F.
 \isAPAStyle{%
        \AuthorCitation{Lastname, F., Lastname, F., \& Lastname, F.}
          }{%
         \isChicagoStyle{%
         \AuthorCitation{Lastname, Firstname, Firstname Lastname, and Firstname Lastname.}
         }{
         \AuthorCitation{\hl{Pereira}%MDPI: Please carefully check the accuracy of names
, B.; Carvalho, F.M.}
         }
 }

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address[1]{Polytechnical Institute of Lisbon, \hl{ISEL}%MDPI: Please add full name if possible
, \hl{Lisbon}%MDPI: please add post code
, Portugal; \hl{a50493@alunos.isel.pt} %MDPI: We added these email addresses here according to those submitted online at susy.mdpi.com. Please confirm.
}

% Contact information of the corresponding author
\corres{\hangafter=1 \hangindent=1.0em \hspace{-1em}Correspondence: miguel.gamboa@isel.pt}

% Current address and/or shared authorship
% \firstnote{Current address: Affiliation.}  % Current address should not be the same as any items in the Affiliation section.

% The commands \thirdnote{} till \eighthnote{} are available for further notes

%\simplesumm{} % Simple summary

%\conference{} % An extended version of a conference paper

% Abstract (Do not insert blank lines, i.e., \\) 
\abstract{Modern web applications increasingly demand rendering techniques that optimize performance, responsiveness, and scalability. Progressive Server-Side Rendering (PSSR) bridges the gap between Server-Side Rendering and Client-Side Rendering by progressively streaming HTML content, improving perceived load times. Still, traditional HTML template engines often rely on blocking interfaces that hinder their use in asynchronous, non-blocking contexts required for PSSR. This paper analyzes how Java Virtual Threads, introduced in Java 21, enable non-blocking execution of blocking I/O operations, allowing the reuse of traditional template engines for PSSR without complex asynchronous programming models. We benchmark multiple engines across Spring WebFlux, Spring MVC, and Quarkus using reactive, suspendable, and virtual-thread-based approaches.~Results show that virtual threads allow blocking engines to scale comparably to those designed for non-blocking I/O, achieving high throughput and responsiveness under load. This demonstrates that Virtual Threads provide a compelling path to simplify the implementation of PSSR with familiar HTML templates, significantly lowering the barrier to entry while maintaining~performance.}

% Keywords
\keyword{web templates; Server-Side Rendering; non-blocking; Java virtual threads; asynchronous}

% The fields PACS, MSC, and JEL may be left empty or commented out if not applicable
%\PACS{J0101}
%\MSC{}
%\JEL{}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Diversity
%\LSID{\url{http://}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Applied Sciences
%\featuredapplication{Authors are encouraged to provide a concise description of the specific application or a potential application of the work. This section is not mandatory.}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Data
%\dataset{DOI number or link to the deposited data set if the data set is published separately. If the data set shall be published as a supplement to this paper, this field will be filled by the journal editors. In this case, please submit the data set as a supplement.}
%\datasetlicense{License under which the data set is made available (CC0, CC-BY, CC-BY-SA, CC-BY-NC, etc.)}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Toxins
%\keycontribution{The breakthroughs or highlights of the manuscript. Authors can write one or two sentences to describe the most important part of the paper.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Encyclopedia
%\encyclopediadef{For entry manuscripts only: please provide a brief overview of the entry title instead of an abstract.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Advances in Respiratory Medicine, Smart Cities and Sensors
%\addhighlights{yes}
%\renewcommand{\addhighlights}{%
%
%\noindent This is an obligatory section in “Advances in Respiratory Medicine'' and ``Smart Cities”, whose goal is to increase the discoverability and readability of the article via search engines and other scholars. Highlights should not be a copy of the abstract, but a simple text allowing the reader to quickly and simplified find out what the article is about and what can be cited from it. Each of these parts should be devoted up to 2~bullet points.\vspace{3pt}\\
%\textbf{What are the main findings?}
% \begin{itemize}[labelsep=2.5mm,topsep=-3pt]
% \item First bullet.
% \item Second bullet.
% \end{itemize}\vspace{3pt}
%\textbf{What is the implication of the main finding?}
% \begin{itemize}[labelsep=2.5mm,topsep=-3pt]
% \item First bullet.
% \item Second bullet.
% \end{itemize}
%}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}

% Falta adicionar referências

Modern web applications rely on different rendering strategies to optimize
performance, user experience, and \textit{\hl{scalability}%MDPI: Please confirm if the italics are necessary; if not, please remove them. The following highlights are the same.
}. The two most dominant
approaches are \textit{\hl{Server-Side Rendering (SSR)}} and \textit{\hl{Client-Side Rendering (CSR)}}. SSR generates HTML content on the server before sending it to the client, resulting in a faster \textit{\hl{First Contentful Paint (FCP)}}~\cite{Edgar2024-FCP} and better Search Engine Optimization (SEO). Nevertheless, SSR
can increase server load and reduce \textit{\hl{throughput}} ({\hl{The} %MDPI: footnote is not allowed in this journal, so we moved it to here, please confirm it
 number of
    requests the server can handle per second (RPS)}) since each request requires
additional processing before responding. In contrast, CSR shifts the rendering
workload to the browser---the server initially sends a minimal HTML document with
JavaScript, which dynamically loads the page content. While CSR reduces the
server’s burden, it can lead to a slower \textit{\hl{FCP}}, as users must wait for
JavaScript execution before meaningful content appears.

\textit{\hl{Progressive Server-Side Rendering (PSSR)}} combines benefits from both SSR and CSR
by streaming HTML content progressively. This technique enhances user-perceived performance by
allowing progressive rendering as data becomes available, significantly reducing the
\textit{\hl{Time to First Byte (TTFB)}} and improving perceived load times compared to traditional
SSR approaches~\cite{wise2024pssr}. In this respect,
PSSR is similar to CSR in that the server initially sends a minimal HTML
document to the client and subsequently streams additional HTML fragments.
Even so, unlike CSR, PSSR retains all rendering responsibilities on the server side,
thereby reducing the load on the client. Consequently, the client does
not need to execute JavaScript or make additional requests to retrieve page content.
The streaming nature of PSSR allows users to see content progressively as it becomes
available, rather than waiting for the complete page to be rendered server-side,
thus providing a more responsive user experience with measurably lower TTFB values~\cite{wise2024pssr}.

\textit{\hl{Low-thread}} servers, also known as
\textit{\hl{event-driven}}~\cite{event-driven-servers}, have gained prominence in
contemporary web applications due to their ability to efficiently manage a large
number of concurrent I/O operations with minimal resources, thus promoting
better scalability.
By leveraging asynchronous I/O operations---such as
database queries and API calls, servers can avoid blocking threads while
waiting for data, thereby maximizing throughput and responsiveness. To support
this non-blocking architecture, PSSR implementations require template engines
that are compatible with asynchronous data models~\cite{carvalho2023async}.
Some modern template engines, such as HtmlFlow~\cite{htmlflow} and Thymeleaf~\cite{thymeleaf}, have
been designed with these capabilities in mind. However, many legacy
template engines---particularly those using external domain-specific languages
(DSLs)~\cite{Fowler03}---still depend on blocking interfaces like
\texttt{Iterable} for data processing. This blocking behavior forces server
threads to remain idle until the entire HTML output is ready, undermining the
performance benefits of non-blocking I/O and limiting scalability in
high-concurrency environments.

With the introduction of \textit{\hl{Virtual
    Threads}}~({\url{https://openjdk.org/jeps/444}}, \hl{accessed on}%MDPI: footnote is not allowed in this journal, so we moved it to here, please confirm it, and please add the access date (format: Date Month Year), e.g., accessed on 1 January 2020.
) in Java 21, it is now
possible to execute blocking I/O operations in a scalable, lightweight manner.
This capability allows legacy template engines---often reliant on blocking
interfaces---to operate efficiently in high-concurrency, non-blocking
environments without requiring complex asynchronous programming models. Due to this, PSSR can now be implemented using familiar HTML templates, simplifying
development and improving maintainability.

We investigate the current landscape of non-blocking PSSR, focusing on two
primary paradigms: \textit{\hl{reactive programming}} and \textit{\hl{coroutines}}, both
of which have been used to achieve asynchronous I/O in the Java ecosystem. As
an alternative, we investigate whether Java’s Virtual Threads can offer
comparable performance while preserving the simplicity of synchronous code.
Section \ref{s2} reviews the state-of-the-art in PSSR and template engine design.
Section \ref{s3} outlines the limitations of conventional engines in asynchronous
settings and presents our proposed approach. Section \ref{sec:bench} details the benchmark
methodology, followed by the results and analysis in Section \ref{s5}.
Section~\ref{s6} compares our results with those from other studies.
Conclusions are presented in Section \ref{s7}.

% The introduction should briefly place the study in a broad context and
% highlight why it is important. It should define the purpose of the work and its
% significance. The current state of the research field should be reviewed
% carefully and key publications cited. Please highlight controversial and
% diverging hypotheses when necessary. Finally, briefly mention the main aim of
% the work and highlight the principal conclusions. As far as possible, please
% keep the introduction comprehensible to scientists outside your particular
% field of research. Citing a journal paper \citep{ref-journal}. Now citing a
% book reference \citep{ref-book1,ref-book2} or other reference types
% \citep{ref-unpublish,ref-url}. Please use the command
% \citep{ref-proceeding,ref-thesis} for the following MDPI journals, which use
% author--date citation: Administrative Sciences, Arts, Behavioral Sciences,
% Businesses, Econometrics, Economies, Education Sciences, European Journal of
% Investigation in Health, Psychology and Education, Games, Genealogy, Histories,
% Humanities, Humans, IJFS, Journal of Intelligence, Journalism and Media, JRFM,
% Languages, Laws, Literature, Psychology International, Publications, Religions,
% Risks, Social Sciences, Tourism and Hospitality, Youth.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Background and Related Work}\label{s2}

In this section, we first present the main properties that characterize each web
template technology approach, along with the advantages and drawbacks that result
from these characteristics. Then, in Section~\ref{sec:web-frameworks}, we examine the different
design choices adopted by web servers in their internal architectures, and how
these choices impact the behavior of the web template engines.

\subsection{Web Templates}\label{s2.1}

\textit{\hl{Web templates}} have been the most widely adopted approach for
constructing dynamic HTML pages.
Web templates or \textit{\hl{web views}}~\cite{Fowler02,Alur01} (e.g., JSP, Handlebars,
or Thymeleaf), are based on HTML documents augmented with template-specific
markers (e.g., \texttt{<\%>}, \texttt{\{\{\}\}}, or \texttt{\$\{\}}), which
represent \textit{\hl{dynamic}} information to be replaced at runtime with the
results of corresponding computations, producing the final HTML page.
The process of parsing and replacing these markers---i.e.,
\textit{\hl{resolution}}---is the primary responsibility of the \textit{\hl{template
  engine}}~\cite{Parr04}.
One key characteristic of \textit{\hl{web templates}} is their ability to receive a
\textit{context object}---equivalent to the \textit{model} in the model-view
design pattern~\cite{mvc88,Parr04}---which provides the data used to fill
template placeholders at runtime.
Web templates can be distinguished by several properties, namely:
\begin{enumerate}
  \item \textit{\hl{Domain-specific language}} idiom
  \item Supported data model APIs
  \item Asynchronous support
  \item Type safety and HTML safety
  \item Progressive rendering
\end{enumerate}

Although some of the aforementioned characteristics apply to both
\textit{\hl{server-side}} and \textit{\hl{client-side}} approaches, we focus solely on
web template technologies for \textit{\hl{server-side rendering}}, as our work is
centered on that approach. Before getting deep on each of the aforementioned
characteristics, Table~\ref{table:cmplibs} presents a breakdown of mainstream
template engines, classified according to the identified properties.

\begin{table}[H]
  \small
    \caption{
    \hl{Comparison} %MDPI: 1. Please confirm whether the different font colors are retained. If so, please confirm whether additional explanation is needed. 2. Please confirm whether √ and × require additional explanation. 3. Please confirm whether the bold text in the table can be deleted
 of web template technologies in the Java ecosystem.
$^{(1)}$ Enforced via annotation processor that checks templates against typed interfaces at compile time.
$^{(2)}$ Compile-time expression validation available via \texttt{@CheckedTemplate} and build-time metadata in Quarkus.
$^{(3)}$ \texttt{Publisher} of Reactive Streams~\cite{ReactiveStreams}. Limited to a single model per web template.
$^{(4)}$ \texttt{kotlin.sequences.Sequence}
$^{(5)}$ {Non-safety} for HTML attributes.}

\begin{adjustwidth}{-\extralength}{0cm}
%\centering %% If there is a figure in wide page, please release command \centering, for Table, ``\textwidth" should be ``\fulllength"
\begin{tabularx}{\fulllength}{CCCCCCC}
    \toprule
    \textbf{Library}
     & \textbf{DSL Idiom}
     & \textbf{Data Model APIs}
     & \shortstack{\textbf{Asynchronous} \\\textbf{Support}}
     & \shortstack{\textbf{Type}         \\\textbf{Safety}}
     & \shortstack{\textbf{HTML}         \\\textbf{Safety}}
     & \shortstack{\textbf{Progressive}  \\\textbf{Rendering}}
    \\
    \midrule
    \textbf{\hl{Freemarker}}
     & External DSL
     & \texttt{Iterable}
     & \large{$\textcolor{red}{\times}$}
     & \large{$\textcolor{red}{\times}$}
     & \large{$\textcolor{red}{\times}$}
     & \large{$\textcolor{PineGreen}{\checkmark}$}
    \\
    \midrule
    \textbf{JSP}
     & External DSL
     & \texttt{Iterable}
     & \large{$\textcolor{red}{\times}$}
     & \large{$\textcolor{red}{\times}$}
     & \large{$\textcolor{red}{\times}$}
     & \large{$\textcolor{red}{\times}$}
    \\
    \hline
    \textbf{JStachio}
     & External DSL
     & \texttt{Iterable}
     & \large{$\textcolor{red}{\times}$}
     & \textcolor{PineGreen}{\checkmark} $^{(1)}$
     & \large{$\textcolor{red}{\times}$}
     & \large{$\textcolor{PineGreen}{\checkmark}$}
    \\\midrule
    \textbf{Pebble}
     & External DSL
     & \texttt{Iterable}
     & \large{$\textcolor{red}{\times}$}
     & \large{$\textcolor{red}{\times}$}
     & \large{$\textcolor{red}{\times}$}
     & \large{$\textcolor{PineGreen}{\checkmark}$}
    \\
    \midrule
    \textbf{Qute}
     & External DSL
     & \texttt{Iterable}
     & \large{$\textcolor{red}{\times}$}
     & \textcolor{PineGreen}{\checkmark} $^{(2)}$
     & \large{$\textcolor{red}{\times}$}
     & \large{$\textcolor{PineGreen}{\checkmark}$}
    \\
    \midrule
    \textbf{Rocker}
     & External DSL
     & \texttt{Iterable}
     & \large{$\textcolor{red}{\times}$}
     & \large{$\textcolor{red}{\times}$}
     & \large{$\textcolor{red}{\times}$}
     & \large{$\textcolor{PineGreen}{\checkmark}$}
    \\
    \midrule
    \textbf{Thymeleaf}
     & External DSL
     & \shortstack{\texttt{Iterable} \\\texttt{Stream} \\\texttt{Publisher} $^{(3)}$}
     & \texttt{Publisher} $^{(3)}$
     & \large{$\textcolor{red}{\times}$}
     & \large{$\textcolor{red}{\times}$}
     & \large{$\textcolor{PineGreen}{\checkmark}$}
    \\
    \midrule
    \textbf{Trimou}
     & External DSL
     & \texttt{Iterable}
     & \large{$\textcolor{red}{\times}$}
     & \large{$\textcolor{red}{\times}$}
     & \large{$\textcolor{red}{\times}$}
     & \large{$\textcolor{PineGreen}{\checkmark}$}
    \\
    \midrule
    \textbf{Velocity}
     & External DSL
     & \shortstack{\texttt{Iterable} \\\texttt{Sequence} $^{(4)}$}
     & \large{$\textcolor{red}{\times}$}
     & \large{$\textcolor{red}{\times}$}
     & \large{$\textcolor{red}{\times}$}
     & \large{$\textcolor{PineGreen}{\checkmark}$}
    \\
    \midrule
    \textbf{Clojure Hiccup}
     & Nested Eager
     & All
     & \large{$\textcolor{red}{\times}$}
     & \large{$\textcolor{PineGreen}{\checkmark}$}
     & \large{$\textcolor{red}{\times}$}
     & \large{$\textcolor{red}{\times}$}
    \\
    \midrule
    \shortstack{\textbf{Groovy } \\\textbf{MarkupBuilder}}
     & Nested Lazy
     & All
     & \large{$\textcolor{red}{\times}$}
     & \large{$\textcolor{PineGreen}{\checkmark}$}
     & \large{$\textcolor{red}{\times}$}
     & \large{$\textcolor{PineGreen}{\checkmark}$}
    \\
    \midrule
    \textbf{HtmlFlow}
     & Chaining
     & All
     & \large{$\textcolor{PineGreen}{\checkmark}$}
     & \large{$\textcolor{PineGreen}{\checkmark}$}
     & \large{$\textcolor{PineGreen}{\checkmark}$}
     & \large{$\textcolor{PineGreen}{\checkmark}$}
    \\
    \midrule
    \textbf{j2html}
     & Nested Eager
     & All
     & \large{$\textcolor{red}{\times}$}
     & \large{$\textcolor{PineGreen}{\checkmark}$}
     & \large{$\textcolor{red}{\times}$}
     & \large{$\textcolor{red}{\times}$}
    \\
    \midrule
    \textbf{KotlinX}
     & Nested Lazy
     & All
     & \large{$\textcolor{red}{\times}$}
     & \large{$\textcolor{PineGreen}{\checkmark}$}
     & \textcolor{PineGreen}{\checkmark} $^{(5)}$
     & \large{$\textcolor{PineGreen}{\checkmark}$}
    \\
    \midrule
    \textbf{ScalaTags}
     & Nested Eager
     & All
     & \large{$\textcolor{red}{\times}$}
     & \large{$\textcolor{PineGreen}{\checkmark}$}
     & \large{$\textcolor{red}{\times}$}
     & \large{$\textcolor{red}{\times}$}
    \\
    \bottomrule
  \end{tabularx}

\end{adjustwidth}
  \label{table:cmplibs}
\end{table}

The first half of Table~\ref{table:cmplibs} lists template engines that use
their own templating dialects, referred to as \textit{\hl{external DSLs}}. The
second half lists Java libraries that provide an \textit{\hl{internal DSL}},
typically using a \textit{\hl{nested}} or \textit{\hl{chaining}} style to build HTML.
Note that, by leveraging the host language (e.g., Clojure, Groovy, Java,
Kotlin, or Scala) as the templating idiom, the latter impose no restrictions on
the data model and fully support all styles of data access APIs. They also
benefit from static type checking, which helps ensure \textit{\hl{type safety}}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%-------------------------------------------------------------------
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsubsection{Domain-Specific Language Idiom}\label{s2.1.1}

Web templates are based on a \textit{\hl{domain-specific language}}
(DSL)~\cite{landin1966next}, which defines a language tailored to a specific
\textit{\hl{domain}}~\cite{evans2004domain}---in this case, HTML for expressing web
documents. The DSL constrains the template's syntax and semantics to match the
structure and purpose of HTML.
DSLs can be divided in two types: \textit{\hl{external}} or
\textit{\hl{internal}}~\cite{dslbook}. \textit{\hl{External}} DSLs are languages created
without any affiliation to a concrete programming language. An example of an
\textit{\hl{external}} DSL is the regular expressions search
pattern~\cite{thompson1968}, since it defines its own syntax without any
dependency of programming languages. On the other hand, an \textit{\hl{internal}} DSL
is defined within a host programming language as a library and tends to be
limited to the syntax of the host language, such as Java.
JQuery~\cite{resig2007pro} is one of the most well-known examples of an internal
DSL in Javascript, designed to simplify HTML DOM~\cite{dom} tree traversal and
manipulation.

Traditionally, web template technologies use an \textit{external} DSL to define
control flow constructs and data binding primitives. Early web template engines
such as JSP, ASP, Velocity, PHP, and others adopted this \textit{external} DSL
approach as their templating dialect. For example, a \texttt{foreach} loop can
be expressed in each technology using its own DSL:
\begin{itemize}
  \item \texttt{<\% for(String item : items) \%>} in JSP
  \item \texttt{<\% For Each item In items \%>} in legacy ASP with VBScript
  \item \texttt{\#foreach(\$item in \$items)} in Velocity
  \item \texttt{<?php foreach (\$items as \$item):?>} in PHP
\end{itemize}

On the other hand, \textit{\hl{internal}} DSLs for HTML allow templates to be
defined directly within the \emph{\hl{host}} language (such as Java, Kotlin, Groovy,
Scala, or other general-purpose programming languages), rather than using
text-based template files~\cite{carvalho2020}. In this case, a web template
is not limited to templating constructs but may instead leverage any available
primitive of the host language or any external API.

Using an internal DSL can have several benefits over using textual templates:
\begin{enumerate}
  \item \emph{\hl{Type safety}}: Because the templates are defined with the host
        programming language, the compiler can check the syntax and types of the
        templates at compile time, which can help catch errors earlier in the
        development process.

  \item \emph{\hl{IDE support}}: Many modern IDEs provide code completion, syntax
        highlighting, and other features, which can make it easier to write and
        maintain templates.

  \item \emph{\hl{Flexibility}}: Use all the features of the host programming language
        to generate HTML, can make it easier to write complex templates and reuse code.

  \item \emph{\hl{Integration}}: Because the templates are defined in Java code, for
        example, you can easily integrate them with other Java code in your
        application, such as controllers, services, repositories and models.

\end{enumerate}

DSLs for HTML provide an API where methods or functions correspond to the names
of available HTML elements. These methods, also known as \textit{builders}, can
be combined in a chain of calls to mimic the construction of an HTML document
in a fluent manner. Martin Fowler~\cite{dslbook} identifies three different
patterns for combining functions to create a DSL: (1) \textit{\hl{function
  sequence}}; (2) \textit{\hl{nested function}}, and (3) \textit{\hl{method chaining}}, which
are illustrated in the snippets of Figure~\ref{fig:dsl-idioms}.

From the three examples depicted in Figure~\ref{fig:dsl-idioms}, the
\textit{\hl{nested function}} idiom used by \textbf{\hl{ScalaTags}%MDPI: Please confirm if the bold formatting is necessary; if not, please remove it. The following highlights are the same.
}, as shown in the
snippet of Figure~\ref{fig:dsl-idioms}b, is the least verbose, requiring
fewer statements than the other two cases. This approach of combining function
calls is also utilized by \textbf{\hl{Hiccup}} Clojure Library and \textbf{\hl{j2html}}
Java library. One reason for verbosity in \textit{\hl{function sequence}} and
\textit{\hl{method chaining}} approaches is the requirement of a dedicated HTML
builder to emit the closing tag, exemplified by \texttt{end()} in \textbf{JATL}
(Figure~\ref{fig:dsl-idioms}a) and \texttt{\_\_()} in \textbf{\hl{HtmlFlow}}
(Figure~\ref{fig:dsl-idioms}c). But the \textit{\hl{nested function}} approach
comes with a notable drawback: \textbf{\hl{it does not support PSSR}} because the
sequence of nested functions is evaluated backward to the order in which they
are written. In other words, arguments are evaluated before the functions are
invoked. Taking the example in Figure~\ref{fig:dsl-idioms}b, the
\texttt{title()} function is first evaluated, and its resulting paragraph
becomes the argument for the \texttt{head()} call, which, in turn, becomes the
argument for \texttt{html()}, and so on. If HTML is emitted as functions are
called, it will print tags in reverse order. The aforementioned DSLs must
collect resulting nodes into an internal data structure, which is later
traversed to produce the HTML output. Therefore, they cannot progressively emit
the output as builders are called.

\begin{figure}[H]
  \centering

  % Subfloat 1
\begin{minipage}[c]{0.33\linewidth}
  \centering
  \begin{lstlisting}[
  language=java,
  basicstyle=\scriptsize\ttfamily,
  numbers=none
]
html();
 head();
  title();text("JT");end();
 end();
 body();
  p();
   text("Hi JATL");
  end();
 end();
end();
  \end{lstlisting}
    \caption*{\centering(\textbf{a}) Function sequence}
  \end{minipage}
  \hfil
  % Subfloat 2
  \begin{minipage}[c]{0.23\linewidth}
    \centering
    \begin{lstlisting}[
    language=java,
    basicstyle=\scriptsize\ttfamily,
    numbers=none
  ]
html(
 head(
  title("ST")
 ),
 body(
  p("Hi ScalaTags")
 )
);
  \end{lstlisting}
    \caption*{\centering(\textbf{b}) Nested function}
  \end{minipage}
  \hfil
  % Subfloat 3
  \begin{minipage}[c]{0.32\linewidth}
    \centering
    \begin{lstlisting}[
    language=java,
    basicstyle=\scriptsize\ttfamily,
    numbers=none
  ]
html()
 .head()
  .title().text("HF").__()
 .__()
 .body()
  .p()
   .text("Hi HtmlFlow")
  .__()
 .__()
.__();
  \end{lstlisting}
    \caption*{\centering(\textbf{c}) Method chaining}
  \end{minipage}
\vspace*{6pt}
  \caption{\hl{Utilizing} %MDPI: Please note that changes to the position/size of figures or tables may occur during the production stage.
 DSL for HTML libraries with JATL, ScalaTags, and HtmlFlow.}
  \label{fig:dsl-idioms}
\end{figure}

Two other JVM libraries, \textbf{\hl{Groovy MarkupBuilder}} and
\textbf{\hl{KotlinX.html}}, also adopt a \textit{\hl{nested function}} approach, but they
address the backward evaluation issue by implementing \textit{\hl{lazy evaluation}}
of arguments~\cite{Landin65}, expressed in lambda expressions (also known as
function literals). Due to the concise form of expressing lambdas with brackets
(i.e., \texttt{\{\}}) in both Groovy and Kotlin, translating the template shown
in Figure~\ref{fig:dsl-idioms}b to use Groovy MarkupBuilder or KotlinX.html
only requires replacing parentheses with brackets for parent elements.
\textbf{\hl{HtmlFlow}} also adopts this approach and provides a Kotlin-idiomatic API
as an alternative to its original Java API.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%-------------------------------------------------------------------
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsubsection{Supported Data Model APIs}\label{s2.1.2}

Template engines with dedicated templating dialects typically rely on specific
control flow constructs---such as a \texttt{foreach}-like statement---to iterate
over elements in a data source. In the Java ecosystem, \texttt{Iterable} is the
common supertype implemented by most data structures and serves as the standard
API for iteration. Nevertheless, other standard interfaces, such as
\texttt{java.util.Stream}, also represent sequences of elements but are not
compatible with \texttt{Iterable}. This fragmentation becomes more evident when
considering other JVM languages, such as Kotlin, which introduces additional
abstractions like the \texttt{Sequence} interface---also incompatible with
\texttt{Iterable}. In consequence, template engines based on \textit{\hl{external}}
DSLs must explicitly support each of these interfaces to accommodate the
various iteration protocols across languages and libraries.

For example, Thymeleaf supports both Java \texttt{Iterable} and \texttt{Stream}
iteration protocols but is incompatible with Kotlin's \texttt{Sequence}
interface.
Velocity, on the other hand, uses internal reflection-based template processing
that allows it to iterate over any type defining an \texttt{iterator()} method
returning an \texttt{Iterator}. This enables support for both Java
\texttt{Iterable} and Kotlin \texttt{Sequence}, but it cannot handle Java
\texttt{Stream} since \texttt{Stream} does not expose a compatible
\texttt{iterator()} method.
The remaining analyzed template engines that use external DSLs support only Java
\texttt{Iterable} as their sole iteration protocol.

On the other hand, template engines that employ an \textit{\hl{internal}} DSL do not
face such limitations, as they can leverage any available API within the host
programming language. This flexibility extends beyond the standard library to
include third-party APIs as well. For example, in Java, developers can iterate
over data using a traditional \texttt{for} loop, the \texttt{java.util.stream}
API, or external libraries such as Guava, Vavr, StreamEx, and \mbox{others---depending}
on their specific needs or preferences.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%-------------------------------------------------------------------
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsubsection{Asynchronous Support}
\label{sec:async-support}

One of the reasons for legacy web templates not supporting asynchronous APIs is
the absence of a unified standard calling convention for asynchronous calls.
While there is a single, straightforward way to use a synchronous API with a
direct style, where the result of a method call corresponds to its returned
value, there is no equivalent standard in the asynchronous approach. Instead,
we may encounter various asynchronous conventions depending on the programming
language and runtime environment. Some of these approaches include
\emph{\hl{continuation-passing style}} (CPS)~\cite{scheme},
\textit{\hl{promises}}~\cite{promise}, async/await idiom~\cite{async_await},
or \textit{\hl{suspend functions}}~\cite{elizarov2021coroutines}.

Many \textit{\hl{general-purpose languages}} (GPLs) have embraced the
\texttt{async}/\texttt{await} feature~\cite{async_await} enabling non-blocking
routines to mimic the structure of synchronous ones, allowing developers to
reason about instruction flow sequentially. The simplicity and broad adoption
of this programming model have led to its incorporation into mainstream
languages like \hl{C\#, JavaScript, Python, Perl, Swift, Kotlin} %MDPI: Please confirm whether this is the software used in this article. If so, please provide the version number.
, and others,
excluding Java. However, implementing \texttt{async}/\texttt{a}wait requires
compiler support to translate \textit{\hl{suspension points}} (i.e., \texttt{await}
statements) into state machines. Most template engines operate using an
external DSL with their own templating dialect (e.g., Thymeleaf, JSP, Jade,
Handlebars, and others), which do not inherently leverage asynchronous
capabilities from their host programming languages.

An alternative approach to asynchronous programming involves handling data as
a sequence of asynchronous events, commonly known as a \textit{\hl{reactive stream}}.
Instead of materializing all items in memory, reactive streams emit values over
time, allowing clients to subscribe and consume data incrementally as it becomes
available.
This idea was first proposed by Meijer~\cite{rx-observable} in the .NET
framework, leading to the development of \textit{Reactive Extensions} (Rx).
Subsequently, several alternative implementations emerged in the Java ecosystem,
such as Rx Java~\cite{rxjava}, Project Reactor~\cite{projectreactor}, or the
Akka Streams~\cite{akka} library.
Later, due to the proliferation of alternative implementations, the need arose to
create a common interface to ensure interoperability between different reactive
stream libraries. This led to the formation of a working group that included
engineers from notable companies such as Lightbend, Pivotal, and Netflix.
The group began defining the \textit{\hl{Reactive Streams}}
specification~\cite{ReactiveStreams}, which established a standard for
asynchronous stream processing using non-blocking, backpressure-aware data
pipelines.
The more recent Mutiny~\cite{mutiny2021}, while conforming to the
\textit{\hl{Reactive Streams}} standard, proposes a different approach that avoids
complex optimizations, resulting in more straightforward and easier-to-maintain
code.

Thymeleaf was one of the first web template engines---and remains the only one
among external DSLs for HTML---to support reactive streams based on the
\texttt{Publisher} model. Yet, it is limited to using a single
\texttt{Publisher} within a given template.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%-------------------------------------------------------------------
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsubsection{HTML safety}\label{s2.1.4}

Another key characteristic of such DSLs is \textit{\hl{HTML safety}}, which refers
to whether they produce only valid HTML conforming to a well-formed document.
To ensure HTML safety, a DSL API should only allow combining calls to builders
that result in valid HTML. Nonetheless, most HTML DSLs using the \textit{\hl{function
  sequence}} or \textit{\hl{nested function}} approach cannot ensure HTML safety at
compile time.

The \textit{\hl{function sequence}} approach, as illustrated in the snippet of
Figure~\ref{fig:dsl-idioms}a, involves combining function calls as a sequence
of statements, making it challenging to restrict the order of statements. In
the \textit{\hl{nested function}} approach shown in Figure~\ref{fig:dsl-idioms}b,
variable-length arguments are often used to allow an undefined number of child
elements, which cannot be strongly typed in every programming language.

KotlinX.html, which utilizes the \textit{\hl{nested function}} idiom with lazy
evaluation, mitigates this issue through function types with a receiver. In
this approach, the receiver (i.e., \texttt{this} within the lambda) is strongly
typed and provides a set of methods corresponding to legal child elements. This
enables KotlinX.html to enforce HTML safety by restricting the available
methods during compile time.

To achieve this, KotlinX.html provides HTML builders using \textit{\hl{function
  literals with receiver}}~\cite{kotlinlang}. In Kotlin, a block of code enclosed
in curly braces \texttt{\{...\}} is known as a \emph{\hl{lambda}}, and can be used
as an argument to a function that expects a \emph{\hl{function literal}}. When we
write, for example, \texttt{body \{ div \{ hr() \} \}}, we are invoking the
\texttt{body} function with a lambda as its argument. This lambda, in turn,
calls the \texttt{div} function with another lambda as an argument that creates
a horizontal row (i.e., \texttt{hr}). Each call to an HTML builder (e.g.,
\texttt{body}, \texttt{div}, \texttt{hr}) creates the child element within the
element generated by the outer function call.

HtmlFlow provides two APIs: one in idiomatic Kotlin, similar to the
\texttt{KotlinX.html} API, and another that employs the \textit{\hl{method
  chaining}} idiom, as illustrated in the snippet of
Figure~\ref{fig:dsl-idioms}c. In this approach, the receiver object is
implicitly passed as an argument to each method call, enabling subsequent
methods to be invoked on the result of the preceding one. This facilitates the
composition of methods, with each call building upon the other. Similar to
KotlinX.html, HtmlFlow ensures HTML safety, restricting the available HTML
builders and attributes after the dot (\texttt{.}) operator.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%-------------------------------------------------------------------
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsubsection{Progressive Rendering}\label{s2.1.5}

Pursuing the optimal user experience for web users has been a consistent goal
since the inception of the World Wide Web. Consequently, various approaches
have been explored to deliver the initial meaningful content to the end-user as
promptly as possible, while the remaining content \textit{\hl{progressively}} (or
incrementally) loads as the server streams the HTML~content.

\textit{\textbf{\hl{Progressive rendering}}} and \textit{\textbf{\hl{progressive loading}}}
encompass different concepts. The former pertains to the \textit{\hl{dynamic}} content of a
\textit{\hl{dynamic web page}}, encompassing elements with logic and placeholders
that are fulfilled by data from an object \textit{\hl{model}} constructed at runtime.
On the other hand, the latter is associated with \textit{\hl{render-blocking
  resources}} such as scripts, stylesheets, and HTML imports, which may hinder the
browser from rendering page content to the screen~\cite{progressive-2022}.
The notion of \textit{\hl{progressive rendering}} has typically been aligned with
\textit{\hl{client-side rendering}} (CSR), where a single HTML page with static
content is delivered upfront, while the dynamic content is fetched as the data
becomes available to complete the web page.
Nevertheless, despite being overlooked, HTTP and browsers were designed from their
inception to also support this feature in the context of SSR approaches,
which offers the advantage of not being dependent on
\textit{\hl{render-blocking resources}}, such as the required JavaScript for CSR.
% \textit{Progressive server-side rendering} (PSSR) is the
% ability of streaming HTML content to the client in chunks or data frames, as it
% gets resolved on the server at runtime.
% This lets the user-interface to be rendered incrementally (i.e.,
% \textit{progressively}) by the browser in line with the availability of the data.
% However, most SSR template engines do not inherently support this feature at the
% template level, requiring developers to manually break down web pages into
% fragments or partials to enable PSSR.
An example of this limitation in SSR web templates was highlighted by Jeff
Atwood in 2005, who criticized Microsoft ASP.NET for loading the entire web
page into memory before sending any data to the browser~\cite{pssr2005}.
% This practice delays the display of content to the user until the server
% completes its processing, contradicting the original design intent of HTML to
% render progressively as content is received.
% Notably, both Netscape and Internet Explorer were capable of rendering partial
% HTML content from their inception.
Despite historical critiques and HTML's inherent capabilities, most Web
application frameworks, such as ASP.Net, Express.js, Spring, and others,
persistently lack support for progressive rendering, leading to the appearance
of alternative techniques leveraging client-side JavaScript. Since 2007,
various patents have addressed the PSSR issue, with Microsoft's patent enabling
the infinite scrolling technique by displaying a single page of
results~\cite{scroll2007}, and Yahoo's patent focusing on differentiating
elements based on their position relative to the visible
portion~\cite{schiller2007progressive}.

Former techniques all describe methods of progressively adding content to a web
page depending on client-side JavaScript.
Java Thymeleaf, the default SSR template engine for Spring web servers,
added support for PSSR in 2018, through the use of a specific non-blocking
Spring \texttt{ViewResolver} driver and without requiring client-side
JavaScript.
Carvalho~\cite{carvalho2023async} introduced an SSR solution that manages
multiple data models and asynchronous APIs while ensuring well-formed HTML,
PSSR, and non-blocking template resolution. That solution builds upon HtmlFlow,
a Java DSL for HTML, with an internal processing mechanism that streams HTML in
chunks as data items become available from data sources, without requiring
client-side JavaScript. Like Thymeleaf, their proposal avoids the use of
client-side JavaScript; yet, unlike Thymeleaf, it supports a wide range of
asynchronous APIs and multiple data sources, not limited to the
\texttt{Publisher} API. Nonetheless, the main counter-argument is the non-trivial
management of the \emph{\hl{resume}} callback in
continuations~\cite{von2003events,callbackhell}, which is used to linearize the
execution flow between asynchronous~calls.

\subsection{Web Framework Architectures and Approaches to PSSR}
\label{sec:web-frameworks}

In traditional thread-per-request architectures, each incoming request is
handled by a dedicated thread.~The web server maintains a thread pool from
which a thread is assigned to handle each request.~As the load increases, the
number of active threads can grow rapidly, potentially exhausting the pool.~This can lead to performance and scalability issues, as the system may become
bogged down by context switching and thread management
overhead~\cite{kant2000scalable}. In the Java ecosystem, Spring MVC is one of
the most widely used frameworks that follow this model, according to several
reports such as the Stack Overflow 2024 Developer
Survey~({\url{https://survey.stackoverflow.co/2024/technology\#1-web-frameworks-and-technologies}}, \hl{accessed on}).

On the other hand, in modern \textit{\hl{low-thread}} architectures, the server uses
a small number of threads to handle a large number of requests.
\textit{\hl{Low-thread}} servers, also known as
\textit{\hl{event-driven}}~\cite{event-driven-servers}, offer a significant
advantage in efficiently managing a high number of concurrent I/O operations
with minimal resource usage.

The \textit{\hl{non-blocking}} I/O model employed in low-thread servers is
well-suited for handling large volumes of data asynchronously~\cite{Meijer12}.
This combination of low-thread servers and asynchronous data models has
facilitated the development of highly scalable, responsive, and resilient web
applications capable of managing substantial data loads~\cite{Jin15}. The
prominence of this concept increased with the advent of Node.js in 2009, and
subsequently, various technologies adopted this approach in the Java ecosystem,
including Netty, Akka HTTP, Vert.x, and Spring WebFlux.
% Among these, Spring WebFlux stands out as the most widely used web framework in
% Java, according to surveys like JetBrains' State of Java report (2021) and
% community metrics from platforms like Github and StackOverflow.
The \textit{\hl{non-blocking}} I/O model in low-thread servers functions optimally
only when HTTP handlers avoid blocking. Therefore, HTML templates need to be
proficient in dealing with the asynchronous APIs provided by data models. While
most legacy web templates struggle with asynchronous models, DSLs for HTML face
no such limitations, leveraging all constructions available in the host
programming language. However, the unexpected intertwining of asynchronous
handlers' completion and HTML builders' execution may potentially lead to
malformed HTML with an unexpected layout~\cite{wise2024pssr}.

An alternative approach involves utilizing user-level threads, while
maintaining a blocking I/O and a synchronous programming paradigm.~On the other hand,
this approach still requires a user-level I/O subsystem capable of mitigating
system-level blocking, which is crucial for the performance of I/O-intensive
applications. This technique offers a lightweight solution for efficiently
managing a larger number of concurrent sessions by minimizing per-thread
overhead. In 2020, Karsten~\cite{karsten2020} demonstrated how this strategy
supports a synchronous programming style, effectively abstracting away the
complexities associated with managing continuations in asynchronous
programming.

Among mainstream technologies, the Kotlin programming language introduces a new
abstraction for managing coroutines and provides structured concurrency, which
ensures that coroutines are \textit{\hl{scoped}}, \textit{\hl{cancellable}}, and
\textit{\hl{coordinated}} within a well-defined
lifecycle~\cite{elizarov2021coroutines}. Although this model embraces the
\texttt{async}/\texttt{await} feature~\cite{async_await}, enabling non-blocking
routines to mimic the structure of synchronous ones, it still lacks a
user-level I/O subsystem and relies on an explicit I/O dispatcher tied to a
dedicated thread pool that frees worker threads for handling processing tasks.
Moreover, web templates using their own templating dialects, such as JSP,
Thymeleaf, or Handlebars, are unable to take advantage of such constructs, as
they typically support only a limited subset of the host library's API---most
commonly just the \texttt{Iterable} interface.

Only Java virtual threads, introduced in
JEP~444 ({\url{https://openjdk.org/jeps/444}}, \hl{accessed on}), adopt a similar approach
to Karsten's proposal, using user-mode threads to preserve a synchronous
programming model while still allowing blocking I/O operations without tying up
platform threads. When a virtual thread performs a blocking I/O operation, the
JVM intercepts the call and transparently parks the virtual thread, freeing the
underlying platform thread to perform other tasks. Once the I/O operation
completes, the virtual thread is rescheduled on a platform thread and resumes
execution. This mechanism is enabled by the JVM's integration with non-blocking
system calls at the OS level, allowing developers to write traditional
synchronous code while benefiting from the scalability of non-blocking I/O.

The primary advantage of virtual threads lies in their ability to eliminate the
need for complex asynchronous programming models. Unlike reactive or structured
concurrency approaches that require developers to adopt complex paradigms, virtual
threads enable non-blocking execution of existing synchronous APIs without code
modification. This significantly reduces the learning curve and maintenance
burden associated with asynchronous programming while maintaining comparable
performance characteristics.

\subsubsection{Spring MVC}

Spring MVC is a synchronous web framework built on the thread-per-request
model. When an HTTP request is received, it is handled by a dedicated thread
from the server’s thread pool (e.g., Tomcat, Jetty, or Undertow). This thread
is responsible for executing the full request lifecycle synchronously,
including invoking controllers, executing business logic, and rendering the
response view. Because the model is blocking, the thread remains occupied
throughout the request---even during I/O operations such as database queries or
external API calls---which can lead to performance bottlenecks under high
concurrency.

Controllers in Spring MVC typically return Java objects or models that are
processed by view resolvers in conjunction with template engines like JSP,
Thymeleaf, or FreeMarker. These templates generate the full HTML response in
one pass, which is then sent to the client only after all necessary data has
been gathered and rendered. By default, this architecture limits the ability to
perform PSSR, as traditional rendering waits for all data before sending any
output. This results in higher latency for clients and less responsive page
loads, particularly for data-intensive views. In any case, Spring MVC provides a
mechanism for response streaming via the
\texttt{StreamingResponseBody} ({\url{https://docs.spring.io/spring-framework/docs/current/javadoc-api/org/springframework/web/servlet/mvc/method/annotation/StreamingResponseBody.html}}, \hl{accessed on})
interface, introduced in Spring 4.2. When a controller method returns a
\texttt{StreamingResponseBody}, Spring writes directly to the
\texttt{HttpServletResponse} output stream, allowing the server to send parts
of the response incrementally as they become available. This is particularly
useful for:
\begin{itemize}
    \item Sending large responses without buffering the entire output in memory.
    \item Writing dynamic HTML in fragments as data is fetched or computed.
    \item Reducing Time to First Byte (TTFB) and improving perceived performance.
\end{itemize}

In the example shown in Listing \ref{lst:streaming-response-body}, HTML content is
written and flushed in discrete chunks, allowing the client to progressively
render the response as it arrives, while the server continues processing
additional data. This technique has two limitations. First, it does
not eliminate the blocking nature inherent in Spring MVC, as the handling
thread remains active during the streaming process. Second, it is constrained
by the servlet response buffer. By default, most servlet containers (e.g.,
Tomcat) use a response buffer size of approximately 8KB. Data written to the
response output stream is held in this buffer until it reaches capacity.
Consequently, if the total size of the initial HTML content is less than the
buffer threshold, the server will not transmit any data to the client until the
buffer is filled. This buffering behavior introduces a delay in sending the
first chunk of content, thereby negating the benefits of progressive rendering.
In scenarios where the rendered HTML template does not produce sufficient data
to exceed the buffer threshold early, PSSR becomes ineffective, and the client
experiences a delayed initial response.

\begin{listing}[H]
\caption{StreamingResponseBody handler in Spring MVC.}
\begin{lstlisting}[
    language=Java,
    basicstyle=\scriptsize\ttfamily,
    numbers=none,
    caption={},
    linewidth=0.986\linewidth,
    label={lst:streaming-response-body}
]
  @GetMapping("/stream")
  fun handleStream(): StreamingResponseBody? {
      return StreamingResponseBody { outputStream ->
          for (i in 0..9) {
              val htmlFragment = "<p>Chunk " + i + "</p>\n"
              outputStream.write(htmlFragment.toByteArray())
              outputStream.flush() // ensure partial response is sent
              Thread.sleep(500) // simulate delay
          }
      }
  }
\end{lstlisting}
\end{listing}

\subsubsection{Spring WebFlux}

Spring WebFlux is a reactive web framework built on a non-blocking,
event-driven architecture that enables efficient handling of a large number of
concurrent connections with minimal resource usage. Unlike traditional
servlet-based stacks that rely on a thread-per-request model, WebFlux operates
on reactive runtimes like Netty, where a small, fixed-size event loop manages
all I/O events. Incoming HTTP requests are routed to handler functions or
annotated controllers, which return reactive types such as \texttt{Mono<T>} and
\texttt{Flux<T>}---\texttt{Publisher} implementations provided by the Project
Reactor library~\cite{projectreactor}. These implementations adhere to the
Reactive Streams specification~\cite{ReactiveStreams}, enabling
interoperability with other compliant reactive stream libraries. A
\texttt{Mono} represents a single asynchronous value (or none)~\cite{promise},
while a \texttt{Flux} represents a stream of zero or more
values~\cite{rx-observable}. I/O-bound tasks such as
database queries or external API calls do not occupy threads during execution,
enabling the framework to scale gracefully under high load and maintain
responsiveness even in resource-constrained environments.

PSSR is natively supported in WebFlux via \texttt{Flux}-based controllers. When
a controller returns a \texttt{Flux<String>} or another stream of HTML
fragments, WebFlux can begin streaming content to the client as soon as the
first elements are emitted.
% This allows browsers to incrementally render parts of the response without waiting
% for the entire payload, significantly reducing Time to First Byte (TTFB) and
% improving user-perceived performance. By integrating PSSR directly into the
% reactive processing pipeline, WebFlux facilitates responsive and resilient web
% applications well-suited for modern microservices architectures.

Listing \ref{lst:pssr-webflux} shows a method that emits one HTML chunk every 500 milliseconds. As
each chunk is generated, it is immediately written to the HTTP response,
allowing the client to begin rendering content even as more data is still being
produced. Unlike \texttt{StreamingResponseBody} in Spring MVC, which is
constrained by servlet buffer thresholds and blocking semantics, WebFlux
ensures that each emitted item is pushed to the client as soon as the network
is ready, without occupying a dedicated thread. Moreover, WebFlux integrates
seamlessly with reactive data sources such as R2DBC (Reactive Relational
Database Connectivity) and reactive NoSQL drivers, making it possible to build
end-to-end non-blocking applications. HTML builders and DSLs used in
conjunction with WebFlux must be designed to accommodate reactive streams,
ensuring that view generation does not violate the non-blocking contract by
performing blocking I/O or awaiting asynchronous operations imperatively.

\begin{listing}[H]
\caption{Progressive Server-Side Rendering in Spring WebFlux.}
\begin{lstlisting}[
    language=Java,
    basicstyle=\scriptsize\ttfamily,
    numbers=none,
    caption={},
    linewidth=0.986\linewidth,
    label={lst:pssr-webflux}
]
@GetMapping(value = "/pssr", produces = MediaType.TEXT_HTML_VALUE)
public Flux<String> renderChunks() {
    return Flux.range(1, 10)
               .delayElements(Duration.ofMillis(500))
               .map(i -> "<p>Chunk " + i + "</p>\n");
}
\end{lstlisting}
\end{listing}

\subsubsection{Quarkus}

Quarkus is a modern, cloud-native Java framework that supports both imperative
and reactive programming models. It is built on top of Vert.x, which provides a
non-blocking, event-driven runtime similar to Netty. Incoming HTTP requests are
processed asynchronously on a small, fixed-size event loop thread pool,
allowing the framework to handle many concurrent connections efficiently
without blocking threads. Quarkus provides support for reactive APIs like
Mutiny~\cite{mutiny2021}, a modern alternative to Spring
Reactor~\cite{projectreactor}. Like Reactor, Mutiny complies with the Reactive
Streams specification~\cite{ReactiveStreams}. For traditional imperative-style
request handling, Quarkus provides integration with
JAX-RS~\cite{burke2013restful}, where resource methods can return synchronous
types or asynchronous constructs such as \texttt{CompletionStage}. To enable
PSSR in imperative endpoints, Quarkus leverages the JAX-RS
\texttt{StreamingOutput} interface. By returning a \texttt{StreamingOutput}
instance, developers can write parts of the HTTP response body incrementally as
data becomes available. This approach allows the server to flush HTML fragments
progressively to the client, improving the perceived responsiveness especially
for long-running or large responses. Like other streaming mechanisms
based on servlet buffers, the initial data flush may be delayed until internal
buffers (commonly around 8KB) are filled. However, Quarkus allows configuring
the buffer size to a smaller value, which can help mitigate this issue.

Listing \ref{lst:pssr-quarkus} shows how \texttt{StreamingOutput} can be used to send
partial HTML content in chunks, allowing browsers to progressively render the
response while the server continues processing. Quarkus’ architecture ensures
that the underlying event loop threads are not blocked by these writes, as the
framework offloads blocking I/O operations to worker threads. This combination
of non-blocking event-driven runtime and incremental response streaming enables
efficient PSSR even within imperative programming styles.

\vspace{-3pt}
\begin{listing}[H]
\caption{Progressive Server-Side Rendering in Quarkus using StreamingOutput.}
\begin{lstlisting}[
  language=java,
  basicstyle=\scriptsize\ttfamily,
  numbers=none,
  caption={},
  label={lst:pssr-quarkus}
]
@GET
@Path("/stream")
@Produces(MediaType.TEXT_HTML)
public StreamingOutput streamHtml() {
    return output -> {
        PrintWriter writer = new PrintWriter(output);
        for (int i = 1; i <= 5; i++) {
            writer.println("<p>Chunk " + i + "</p>");
            writer.flush(); // Flush each chunk incrementally
            Thread.sleep(500); // Simulate processing delay
        }
        writer.flush();
    };
}
\end{lstlisting}
\end{listing}

\section{Problem Statement}\label{s3}

In this section, we examine the challenges of implementing \textit{\hl{Progressive
Server-Side Rendering}} (PSSR) in modern web applications, with a focus on the
limitations of current template engine designs. Our goal is to broaden the
range of options available for PSSR, particularly within JVM-based frameworks.
% Reactive streams~\cite{ReactiveStreams}, such as those based on
% \texttt{Observable} (from RxJava)~\cite{rx-observable} and \texttt{Flow} (from
% the Java 9+ standard library~\footnote{~\url{https://openjdk.org/jeps/266}}),
% are abstractions that provide a non-blocking, asynchronous
% way to handle data streams.
% These abstractions support the incremental generation and transmission of HTML
% content as data becomes available, enabling \textit{progressive server-side
% rendering} (PSSR)~\cite{pssr2005}.
In PSSR, the server does not
wait for the entire data model to be ready before beginning to render HTML.
Instead, it processes and streams each piece of data to the client as soon as
it arrives.

Reactive types like \texttt{Observable<T>} of ReactiveX~\cite{rxjava} or Kotlin \texttt{Flow<T>}~\cite{kotlinlang}
facilitate this by representing data as a sequence of asynchronous events.
For example, a reactive stream might emit a sequence of \texttt{Presentation}
objects, each representing a talk in a conference schedule.
As each \texttt{Presentation} is emitted by the \texttt{Observable}
or \texttt{Flow}, the internal DSL-based engine---such as HtmlFlow---can render the
corresponding HTML fragment and immediately flush it to the client. This
approach is demonstrated in Listing~\ref{lst:presentation-observable}, where each
presentation is rendered asynchronously as it is emitted by an
\texttt{Observable}.
Note that the \texttt{await} builder receives an additional parameter, the
\texttt{onCompletion} callback, which is used to signal HtmlFlow that it can
proceed to render the next HTML element in the web
template~\cite{carvalho2023async}.
HtmlFlow pauses the rendering process until \texttt{onCompletion} is called,
similar to how the \emph{\hl{resume}} function works in continuations and
coroutines~\cite{coroutines_continuations}.
In Listing~\ref{lst:presentation-flow}, we show an
equivalent suspend-based implementation using Kotlin's \texttt{Flow}~\cite{wise2024pssr}.
Both examples highlight how internal DSLs can natively integrate with reactive types
to enable non-blocking, progressive rendering on the server side.

\begin{listing}[H]
\caption{\textit{\hl{HtmlFlow reactive}} presentation template in Koltin with an \texttt{Observable} model.}
\begin{lstlisting}[
    language=Kotlin,
    basicstyle=\scriptsize\ttfamily,
    numbers=none,
    caption={},
    label={lst:presentation-observable}
]
await { div, model, onCompletion ->
  model
    .doOnNext { presentation ->
      presentationFragmentAsync
        .renderAsync(presentation)
        .thenApply { frag -> div.raw(frag) }
      }
    .doOnComplete { onCompletion.finish() }
    .subscribe()
}
\end{lstlisting}
\end{listing}

\vspace{-12pt}
\begin{listing}[H]
\caption{\textit{\hl{HtmlFlow suspend}} presentation template in Kotlin with a \texttt{Flow} model.}
\begin{lstlisting}[
    language=Kotlin,
    basicstyle=\scriptsize\ttfamily,
    numbers=none,
    label={lst:presentation-flow},
    caption={},
]
suspending { model ->
  model
    .toFlowable()
    .asFlow()
    .collect { presentation ->
      presentationFragmentAsync
        .renderAsync(presentation)
        .thenApply { frag -> raw(frag) }
    }
}
\end{lstlisting}
\end{listing}

By contrast, template engines that use \textit{\hl{external}} DSLs---such as JStachio,
Thymeleaf, or Handlebars---typically define templates within static HTML
documents using custom markers, and rely on blocking interfaces like
\texttt{java.util.Iterable} or \texttt{java.util.stream.Stream}.
These interfaces require the entire data
model, to be materialized in memory before rendering can begin, which blocks server
threads during template expansion and significantly limits scalability under
high concurrency.
Some reactive libraries, such as RxJava, provide bridging mechanisms like
\texttt{Observable.blockingIterable()}, which allows asynchronous data sources
to be exposed as \texttt{Iterable} by blocking the thread until all items are available.
While useful for compatibility with traditional APIs, this approach
reintroduces blocking behavior and undermines the benefits of non-blocking
I/O---especially under high concurrency. Listing \ref{lst:presentation-jstachio}
illustrates this model using a JStachio template, where the engine performs a
blocking loop over \texttt{presentationItems}.

\begin{listing}[H]
\caption{Presentation HTML template using \textit{\hl{JStachio}}.}
\begin{lstlisting}[
  language=Kotlin,
  numbers=none,
  basicstyle=\scriptsize\ttfamily,
  caption={},
  label={lst:presentation-jstachio}
]
{{#presentationItems}}
<div class="card mb-3 shadow-sm rounded">
    <div class="card-header">
        <h5 class="card-title">
            {{title}} - {{speakerName}}
        </h5>
    </div>
    <div class="card-body">
        {{summary}}
    </div>
</div>
{{/presentationItems}}
\end{lstlisting}
\end{listing}

Despite these performance limitations, external DSLs remain popular due to
several~advantages:

\begin{enumerate}
    \item \emph{\hl{Separation of Concerns}}: HTML templates are decoupled from application logic, enabling front-end developers to contribute without modifying back-end code.
    \item \emph{\hl{Cross-Language Compatibility}}: External DSLs are portable across languages and frameworks, easing integration in multi-language environments.
    \item \emph{\hl{Familiarity}}: Many developers are comfortable with HTML syntax, lowering the barrier to entry and improving maintainability.
\end{enumerate}

These strengths make external DSLs appealing---even when they come at the cost of
blocking, synchronous rendering. However, this tradeoff becomes critical under
high concurrency, where blocking threads severely degrades throughput~\cite{wise2024pssr}.
Emerging features in the Java ecosystem, particularly \textit{\hl{virtual threads}}
introduced in Java 21 as part of Project Loom~\cite{Veen2024}, offer a promising solution to
this challenge. Virtual threads drastically reduce the overhead of blocking
operations by decoupling thread execution from OS-level threads. For this reason,
engines that rely on blocking interfaces---like those used in
external DSLs---can potentially achieve scalability levels that approach those of
non-blocking, asynchronous~engines.

% Incluir este parágrafo?
% In previous work ~\cite{PSSR-WISE2024}, Carvalho benchmarked the performance of
% HtmlFlow (an internal DSL) using suspendable templates within a Spring WebFlux
% application. HtmlFlow scaled efficiently to 128 concurrent users and delivered
% up to 4,000 requests per second. In contrast, blocking engines such as JStachio
% saturated at just 4 concurrent users, maxing out at around 400 requests per
% second---highlighting the scalability limitations of blocking models.This work
% builds upon those findings by evaluating whether virtual threads can close this
% performance gap for blocking template engines. Specifically, we aim to
% determine whether external DSLs---traditionally unsuitable for PSSR due to their
% reliance on blocking APIs---can become viable for progressive rendering when
% powered by virtual threads.

% If successful, this approach would enable developers to reap the ergonomic
% benefits of external DSLs while maintaining scalable performance, without
% adopting complex asynchronous paradigms. It would also simplify the
% implementation of PSSR across a broader range of frameworks and engines, making
% the technique more accessible to mainstream web development.

\section{Benchmark Implementation}
\label{sec:bench}

This benchmark ({\url{https://github.com/xmlet/comparing-non-blocking-progressive-ssr}}, \hl{accessed on})
is designed with a modular architecture, separating the \textit{\hl{view}} and
\textit{\hl{model}} layers from the \textit{\hl{controller}} layer ~\cite{Bucanek2009},
which allows for easy extension and integration of new template engines and
frameworks. It also includes a set of tests to ensure the correctness of
implementations and to validate the \textit{\hl{HTML}} output. It includes two
different data models, defined as \texttt{Presentation} and \texttt{Stock}, as
shown in Listing \ref{lst:stock-class} and Listing \ref{lst:presentation-class},
following the proposal of the
benchmarks~\cite{templatesbenchmark,springbenchmark}. The \texttt{Presentation}
class represents a presentation with a title, speaker name, and summary, while
the \texttt{Stock} class represents a stock with a name, URL, symbol, price,
change, and ratio.

\begin{listing}[H]
\caption{\texttt{\hl{Stock}} class.}
            \begin{lstlisting}[
  language=kotlin,
  numbers=none,
  basicstyle=\scriptsize\ttfamily,
  caption={},
  label={lst:stock-class},
]
data class Stock(
    val name: String,
    val name2: String,
    val url: String,
    val symbol: String,
    val price: Double, 
    val change: Double, 
    val ratio: Double
)
\end{lstlisting}
\end{listing}

\begin{listing}[H]
\caption{\texttt{\hl{Presentation}} class.}
            \begin{lstlisting}[
    language=kotlin,
    basicstyle=\scriptsize\ttfamily,
    numbers=none,
    caption={},
    label={lst:presentation-class},
]
data class Presentation(
    val id: Long,
    val title: String, 
    val speakerName: String,
    val summary: String
)
\end{lstlisting}
\end{listing}

The application's repository contains a list of 10 instances of the
\texttt{Presentation} class and 20 instances of the \texttt{Stock} class. Each
list is used to generate a respective HTML view. Although the instances are
kept in memory, the repository uses the \texttt{Observable} class from the
\textit{RxJava} library to interleave list items with a delay of 1 millisecond.
This delay promotes context switching and frees up the calling thread to handle
other requests in non-blocking scenarios, mimicking actual I/O operations.

By using the \texttt{blockingIterable} method of the \texttt{Observable} class,
we provide a blocking interface for template engines that do not support
asynchronous data models, while still simulating the asynchronous nature of the
data source to enable PSSR\@. Template engines that do not support non-blocking
I/O for PSSR include KotlinX, Rocker, JStachio, Pebble, Freemarker, Trimou, and
Velocity. HtmlFlow supports non-blocking I/O through suspendable templates and
asynchronous rendering, while Thymeleaf enables it using the
\texttt{ReactiveDataDriverContextVariable} in conjunction with a non-blocking
Spring \texttt{ViewResolver}.

The aforementioned blocking template engines are used in the context of Virtual
Threads or alternative coroutine dispatchers, allowing the handler thread to be
released and reused for other requests.

The Spring WebFlux core implementation uses Project Reactor to support a
reactive programming model: each method returns a \texttt{Flux<String>} as the
response body, which acts as a publisher that progressively streams the HTML
content to the client. The implementation includes four main approaches to
PSSR\@:
\begin{itemize}
      \item \textbf{\hl{Reactive}}: The template engine is used in a reactive context,
            where the HTML content rendered using the reactive programming model.
            An example of this approach is the Thymeleaf template engine
            when using the \texttt{ReactiveDataDriverContextVariable} in conjunction with a
            non-blocking Spring \texttt{ViewResolver}.
      \item \textbf{\hl{Suspendable}}: The template engine is used in a suspendable
            context, where the HTML content is rendered within the context of a suspending function.
            An example of this approach is the HtmlFlow template engine, which
            supports suspendable templates with the use of the \texttt{Flow} class from the
            Kotlin standard library.
      \item \textbf{\hl{Virtual}}: The template engine is used in a non-blocking
            context, where the HTML content is rendered within the context of Virtual
            Threads. This method is used for the template engines that do not traditionally
            support non-blocking I/O, be it either because they use external DSLs and in
            consequence only support the blocking \texttt{Iterable} interface, or
            because they do not support the asynchronous rendering of HTML~content.
      \item \textbf{\hl{Blocking}}: The template engine is used in a blocking
            context, where the HTML content is rendered using the blocking interface of the
            \texttt{\hl{Observable}} class, within the context of a OS-thread.
\end{itemize}
% It also includes methods using Kotlin coroutines and other
% asynchronous mechanisms supported by template engines, such as the
% \texttt{writeAsync} method from HtmlFlow, which enables non-blocking I/O
% using \textit{continuations}.

The Spring MVC implementation uses handlers based solely on the blocking
interface of the \texttt{Observable} class. To enable PSSR in this context, we
utilize the \texttt{StreamingResponseBody} interface, which allows the
application to write directly to the response \texttt{OutputStream} without
blocking the servlet container thread. According to the Spring documentation ({\url{https://docs.spring.io/spring-framework/docs/current/javadoc-api/org/springframework/web/servlet/mvc/method/annotation/StreamingResponseBody.html}}, \hl{accessed on}),
this class is \textit{\hl{a controller method return value type for asynchronous
request processing where the application can write directly to the response
OutputStream without holding up the Servlet container thread.}}

In Spring MVC, \texttt{StreamingResponseBody} enables asynchronous writing
relative to the request-handling thread, but the underlying I/O remains
blocking---specifically the writes to the \texttt{OutputStream}. When using
Virtual Threads, the I/O operations are more efficient when compared to
platform threads, as they are executed in the context of a lightweight thread.
Most of the computation is done in a separate thread from the one that receives
each request; we use a thread pool \texttt{TaskExecutor} to process requests,
allowing the application to scale and handle multiple clients more efficiently
as opposed to the default \texttt{TaskExecutor} implementation, which tries to
create a thread for each request.

However, the Spring MVC implementation does not effectively support PSSR for
these templates, as HTML content is not streamed progressively to the client.
This limitation occurs because the response is only flushed to the client once
the content written to the \texttt{OutputStream} exceeds the configured output
buffer size, as explained in Section~\ref{sec:web-frameworks}. Consequently,
the client receives the complete response only after the entire HTML content
has been rendered and buffered, which negates the primary benefits of
PSSR---namely, reduced time-to-first-byte and progressive content delivery.
Furthermore, Spring MVC does not provide configuration options for the response
buffer size, preventing developers from reducing it to smaller values that
would enable more frequent flushing and achieve true progressive streaming of
HTML content. This architectural constraint makes Spring MVC unsuitable for
implementing effective PSSR compared to reactive frameworks like Spring~WebFlux.

This implementation includes two main approaches to PSSR\@:
\begin{itemize}
      \item \textbf{\hl{Blocking}}: The template engine is used in a blocking context,
            where the HTML content is rendered using the blocking interface of the
            \texttt{Observable} class.
      \item \textbf{\hl{Virtual}}: The template engine is used in a non-blocking context,
            where the HTML content is rendered within the context of Virtual Threads.
\end{itemize}

The Quarkus implementation also uses handlers based on the blocking interface
of the \texttt{Observable} class. It implements the \texttt{StreamingOutput}
interface from the JAX-RS specification to enable PSSR, allowing HTML content
to be streamed to the client. While \texttt{StreamingOutput} also uses blocking
I/O, it operates on Vert.x worker threads, which prevents blocking of the event
loop. When Virtual Threads are used, the I/O operations are handled
efficiently, as they are executed in lightweight threads.

The Quarkus implementation supports PSSR for these templates by configuring the
response buffer size in the \textit{\hl{application.properties}} file. The default
buffer size is 8KB, but we reduced it to 512 bytes, which allows the response
to be sent to the client progressively as the HTML content is rendered.

This implementation includes three main approaches to PSSR\@:
\begin{itemize}
      \item \textbf{\hl{Blocking}}: The template engine is used in a blocking context,
            where the HTML content is rendered using the blocking interface of the
            \texttt{Observable} class.
      \item \textbf{\hl{Virtual}}: The template engine is used in a non-blocking context,
            where the HTML content is rendered within the context of Virtual Threads.
      \item \textbf{\hl{Reactive}}: The template engine is used in a reactive context,
            where the HTML content is rendered using the reactive programming model.
            In this case, we use the HtmlFlow template engine, which
            supports asynchronous rendering through the \texttt{writeAsync}~method.
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}\label{s5}

This section presents the evaluation results organized into four main parts:
Section~\ref{sec:environment} details the testing environment and hardware
specifications; Section~\ref{sec:presentations-results} presents scalability
results for the \texttt{Presentations} data model;
Section~\ref{sec:stocks-results} analyzes performance with the more complex
\texttt{Stocks} data model; and Section~\ref{sec:memory-analysis} provides
detailed memory consumption and resource utilization analysis.

\subsection{Environment Specifications}
\label{sec:environment}

All benchmarks were conducted on a GitHub-hosted virtual machine under \hl{GitHub} %MDPI: Please state which version of the software was used.
Actions, running \hl{Ubuntu 24.04.2 LTS with an AMD EPYC 7763 64-Core Processor
(x86\_64 architecture)} %MDPI: Please state the name of the manufacturer, city, and country from where the equipment was sourced.
 operating at 3.24 GHz, configured with 2 CPU cores and 2
threads per core, and 7.8 GB of available RAM. The system utilizes a 75 GB SSD
with ext4 filesystem, achieving 1.5 GB/s write throughput in basic I/O tests.
Network connectivity is provided through a 1500 MTU Ethernet interface. All
tests were conducted on \hl{OpenJDK 21} (Corretto build) with the G1 garbage
collector enabled by default. The JVM arguments used were: \texttt{-Xms512m
     -Xmx16g}. Each test was executed 5 times per configuration to ensure result
reliability.

All template engines were configured to use UTF-8 encoding and disable
automatic HTML escaping to ensure fair comparison and consistent output across
all template engines. Caching behaviour was left as the default for each
engine, allowing them to optimize template loading and rendering as per their
design.~These configurations ensure that all engines operate under equivalent
conditions, with template loading, encoding, and escaping behavior normalized
across implementations.

For both the Apache Bench and JMeter tests, we simulate a 1000-request warmup
period for each route with a concurrent user load of 32 users. The warmup
period is followed by the actual test period, during which we simulate 256
requests per user, scaling in increments up to 128 concurrent users.

The results are presented in the form of throughput (number of requests per
second) for each template engine, with the x-axis representing the number of
concurrent users and the y-axis representing the throughput in requests per
second.

Both Quarkus and Spring MVC implementations were configured with an 8KB output
buffer size to ensure consistency, despite Quarkus not enabling PSSR at this
size and Spring MVC not supporting PSSR for the tested templates. Testing with
a reduced 512B buffer size showed only a 0.046\% performance difference with
Rocker, indicating negligible impact. Both frameworks use unlimited platform
thread pools, enabling on-demand thread creation up to system limits for
maximum throughput and scalability observation under high concurrent load.

Since the obtained results for JMeter and Apache Bench show no significant
differences, only the JMeter results will be presented. Statistical analysis of
test configurations revealed an average absolute percentage difference of
2.83\% between the two load testing tools. While individual approach
differences ranged from $-$16.53\% to +14.66\% at the extremes, the consistent
directional bias and small magnitude of differences indicate that both tools
provide comparable performance measurements. Given this statistical equivalence
and the need for brevity, only JMeter results are presented, as they provide
representative performance characteristics across all tested configurations.

\subsection{Scalability Results for the \texttt{Presentations} Class} \label{sec:presentations-results}

\textls[-10]{The results in Figure~\ref{fig:presentations-webflux-jmeter} depict the
throughput (number of requests per second) for each template engine, with
concurrent users ranging from 1 to 128, from left to right. The benchmarks
include HtmlFlow using suspendable web templates (\textit{\hl{HtmlFlow-Susp}},
equivalent to the approach shown in Listing~\ref{lst:presentation-flow}),
Jstachio using Virtual Threads with the \texttt{Iterable} interface
(\textit{\hl{Jstachio-Virtual}}), and Thymeleaf using the reactive View Resolver
driver (Thymeleaf-Rx). \textit{\hl{Blocking}} and \textit{\hl{Virtual}} represent the
average throughput of the blocking approaches (i.e., KotlinX, Rocker, Jstachio,
Pebble, Freemarker, Trimou, HtmlFlow, and Thymeleaf) when run in the context of
a separate coroutine dispatcher or Virtual Threads, respectively.}

We show the \textit{\hl{HtmlFlow-Susp}}, \textit{\hl{Jstachio-Virtual}}, and
\textit{Thymeleaf-Rx} engines separately to observe the performance of the
non-blocking engines when using the Suspending, Virtual Thread, and Reactive
approaches. The \textit{\hl{Blocking}} and \textit{\hl{Virtual}} are aggregated due to
the similar performance of different engines when using those approaches.

\begin{figure}[H]
\vspace{-6pt}
     \includegraphics[width=0.8\textwidth]{./Graphs/presentations-webflux-jmeter.png}
     \caption{\hl{Throughput} %MDPI: The numbers in the picture are incomplete, please modify
 (requests per second) scalability results for Spring Webflux with \texttt{Presentations} class.}\label{fig:presentations-webflux-jmeter}
\end{figure}

The results in Figure~\ref{fig:presentations-webflux-jmeter} show that when
using blocking template engines with a separate coroutine dispatcher, the
engines are unable to scale effectively beyond 4 concurrent users. In contrast,
\textit{\hl{HtmlFlow-Susp}} scales up to 128 concurrent users, achieving \hl{4487} %MDPI: Commas are only used for numbers with five or more digits. We have removed them in four-digit numbers. Please confirm.
requests per second. When blocking approaches are executed in the context of
Virtual Threads---thus enabling non-blocking I/O---the engines scale up to 64
concurrent users, with Jstachio using Virtual Threads reaching \hl{3514} requests
per second. The Thymeleaf implementation using the reactive View Resolver
driver scales up to 32 concurrent users, achieving a lower maximum throughput
of \hl{2559} requests per second.

It is important to note that the differences in scalability and throughput
between the \textit{\hl{HtmlFlow-Susp}}, \textit{\hl{Jstachio-Virtual}}, and
\textit{\hl{Thymeleaf-Rx}} approaches may be influenced by the specific template
engines used, rather than the approach itself. When comparing the
\textit{\hl{Reactive}}, \textit{\hl{Suspendable}}, and \textit{\hl{Virtual}} approaches
specifically with HtmlFlow, we found that all three achieve similar
performance: HtmlFlow using a blocking approach with Virtual Threads reaches
4691 requests per second, while HtmlFlow using a reactive approach achieves
4792~requests per second.

% The results for the Spring MVC implementation, shown in
% Figure~\ref{fig:presentations-springmvc-jmeter}, compare two approaches:
% \textit{Blocking}, which uses platform threads with \texttt{StreamingResponseBody},
% and \textit{Virtual}, which uses Virtual Threads. Both approaches scale
% effectively up to 128 concurrent users, with the Virtual Thread approach
% achieving a slightly higher throughput of 9,000 requests per second. However,
% these values are slightly lower than those observed in the Spring WebFlux
% implementation.
The results for the Spring MVC implementation, shown in
Figure~\ref{fig:presentations-springmvc-jmeter}, compare two synchronous
approaches: \textit{Blocking}, which uses platform threads with
\texttt{StreamingResponseBody}, and \textit{\hl{Virtual}}, which uses Virtual
Threads. Since Spring MVC follows a thread-per-request architecture, the
asynchronous approaches---\textbf{\hl{Reactive}} and \textbf{\hl{Suspendable}}---described in
Section~\ref{sec:bench} are not applicable. Both the \textit{\hl{Blocking}} and
\textit{\hl{Virtual}} strategies scale effectively up to 32 concurrent users, with
the Virtual Threads approach achieving a slightly higher maximum throughput of
2797 requests per second, compared to 2498 requests per second for the
blocking approach. These results indicate that while Spring MVC can handle a
moderate level of concurrency, it does not reach the scalability of the
reactive or suspendable approaches available in Spring WebFlux. Furthermore,
Spring MVC does not enable Progressive Server-Side Rendering (PSSR) for the
tested templates, as previously discussed in Section~\ref{sec:bench}.

\begin{figure}[H]
\vspace{-6pt}
     \includegraphics[width=0.8\textwidth]{./Graphs/presentations-springmvc-jmeter.png}
     \caption{\hl{Throughput} %MDPI: The numbers in the picture are incomplete, please modify
 (requests per second) scalability results for Spring MVC with \texttt{Presentations} class.}\label{fig:presentations-springmvc-jmeter}
\end{figure}

The results for the Quarkus implementation, shown in
Figure~\ref{fig:presentations-quarkus-jmeter}, indicate that Quarkus handles
synchronous approaches more efficiently than Spring WebFlux. The blocking
engines scale up to 64 concurrent users, achieving up to 3744 requests per
second. When using Virtual Threads, the throughput increases even further,
reaching 4856 requests per second, allowing scalability up to 128 users. This
demonstrates that Quarkus's implementation of Virtual Threads is effective for
enabling PSSR\@, and comparable to the \textit{\hl{Suspendable}} and
\textit{\hl{Reactive}} approaches used in Spring WebFlux in terms of scalability and
throughput.

\vspace{-3pt}
\begin{figure}[H]
     \includegraphics[width=0.8\textwidth]{./Graphs/presentations-quarkus-jmeter.png}
     \caption{\hl{Throughput} %MDPI: The numbers in the picture are incomplete, please modify
 (requests per second) scalability results for Quarkus with \texttt{Presentations} class.}\label{fig:presentations-quarkus-jmeter}
\end{figure}

Additionally, \textit{\hl{HtmlFlow-Rx}}, a reactive implementation of the HtmlFlow
template engine (equivalent to the approach shown in
Listing~\ref{lst:presentation-observable}) that utilizes Quarkus's reactive
programming model, achieved a lower throughput than the the \textit{\hl{Blocking}}
and \textit{\hl{Virtual}} approaches---3088 requests per second. This demonstrates
that Quarkus's reactive programming model is effective for enabling PSSR\@,
although it does not achieve the same level of performance or scalability as
the same approach in Spring WebFlux, stagnating after 32~concurrent users.

\subsection{Scalability Results for the \texttt{Stocks} Class} \label{sec:stocks-results}

The results in Figure~\ref{fig:stocks-webflux-jmeter} use the same template
engines and approaches as the previous benchmark, but with a more complex data
model: the Stock class, which includes 20~instances and approximately two times
as many data bindings. With this data model, the scalability of the engines
remains largely unchanged; however, throughput is reduced across all engines.
Compared to the Presentation benchmark, Jstachio using virtual threads
experienced a more pronounced decrease in performance relative to the
\textit{\hl{Reactive}} and \textit{\hl{Suspending}} approaches, with Jstachio using
Virtual Threads now achieving 1509 requests per second, compared to 1750
requests per second achieved by the Thymeleaf implementation using the reactive
View Resolver driver.

\begin{figure}[H]
\vspace{-6pt}
     \includegraphics[width=0.8\textwidth]{./Graphs/stocks-webflux-jmeter.png}
     \caption{\hl{Throughput} %MDPI: The numbers in the picture are incomplete, please modify
 (requests per second) scalability results for Spring Webflux with \texttt{Stocks} class.}\label{fig:stocks-webflux-jmeter}
\end{figure}

It is again important to note that the differences in scalability and
throughput between the \textit{\hl{HtmlFlow-Susp}}, \textit{J\hl{stachio-Virtual}}, and
\textit{\hl{Thymeleaf-Rx}} approaches may be influenced by the specific template
engines used, rather than the approach itself. When comparing the
\textit{\hl{Reactive}}, \textit{\hl{Suspendable}}, and \textit{Virtual} approaches
specifically with HtmlFlow, we found that all three achieve similar
performance: HtmlFlow using a blocking approach with Virtual Threads reaches
3090 requests per second, while HtmlFlow using a reactive approach achieves
3026 requests per second. This indicates that the more pronounced decrease in
performance for Jstachio using Virtual Threads is likely due to the specific
implementation of the template engine, rather than the use of Virtual Threads
itself.

The overall throughput reduction across all engines is expected, as the Stock
class contains more data properties than the \texttt{Presentation} class,
adding overhead related to the data binding process of each template engine.

The results shown in Figure~\ref{fig:stocks-springmvc-jmeter} indicate that the
Spring MVC implementation using the blocking approach with
\texttt{StreamingResponseBody} achieves a throughput of up to 1,916 requests
per second, with no significant improvement observed when using Virtual
Threads. Both approaches scale effectively up to 64 concurrent users. Although
these approaches achieve higher throughput in Spring MVC than in Spring
WebFlux, their overall performance remains lower than that of the reactive and
suspendable approaches.

The results depicted in Figure~\ref{fig:stocks-quarkus-jmeter} show that the
Quarkus synchronous approaches scale effectively up to 128 concurrent users,
achieving performance comparable to the Spring WebFlux implementation. The
blocking approach reaches a throughput of 3019 requests per second, while the
Virtual Threads approach achieves a throughput of 3357 requests per second. In
addition to the synchronous engines, the \textit{\hl{HtmlFlow-Rx}} approach also
achieves a throughput of 1760 requests per second, indicating that this
approach achieves lower performance in Quarkus than in Spring WebFlux, where it
reached 3026 requests per second, as previously mentioned.

\begin{figure}[H]
\vspace{-3pt}
     \includegraphics[width=0.8\textwidth]{./Graphs/stocks-springmvc-jmeter.png}
     \caption{\hl{Throughput} %MDPI: The numbers in the picture are incomplete, please modify
 (requests per second) scalability results for Spring MVC with \texttt{Stocks} class.}\label{fig:stocks-springmvc-jmeter}
\end{figure}
\vspace{-6pt}
\begin{figure}[H]

     \includegraphics[width=0.8\textwidth]{./Graphs/stocks-quarkus-jmeter.png}
     \caption{\hl{Throughput} %MDPI: The numbers in the picture are incomplete, please modify
 (requests per second) scalability results for Quarkus with \texttt{Stocks}.}\label{fig:stocks-quarkus-jmeter}
\end{figure}

The results of the benchmarks show that non-blocking engines---whether using
reactive programming, Kotlin coroutines, or Java Virtual Threads---are able to
scale effectively, supporting between 32 and 128 concurrent users depending on
the approach and framework. Out of all the tested frameworks, Spring Webflux
showed itself the most effective at enabling PSSR, mostly due to its native
support for publish-subscribe interfaces, allowing for content to be streamed
as data becomes available, instead of when the response buffer is flushed.
Quarkus also enabled PSSR effectively, but it required additional configuration
of the \texttt{OutputBuffer} size to achieve the same results as Spring
Webflux. The Spring MVC implementation, on the other hand, did not enable PSSR
for the tested templates.

However, it is important to acknowledge the limitations of our chosen data
models for generalizability. The tested templates used relatively simple data
structures with limited nesting and straightforward property bindings.
Real-world applications often involve deeply nested data structures, complex
conditional logic, and iterative rendering over thousands of items. Under such
conditions, we anticipate significantly different performance characteristics:
increased memory consumption due to object traversal overhead, higher CPU
utilization for complex evaluations, and altered scalability patterns where
non-blocking approaches may become more advantageous. The performance
degradation observed with our Stock class benchmark---which merely doubled the
number of properties---suggests that higher template complexity would result in
substantially higher performance degradation. Future work should investigate
these scenarios with more realistic data models to better understand
performance boundaries and optimization strategies for complex PSSR
implementations.

\subsection{Memory Consumption and Resource Utilization Analysis}
\label{sec:memory-analysis}

\textls[-15]{This section evaluates the memory and CPU resource usage characteristics of
structured concurrency approaches, specifically comparing Virtual Threads and
suspendable~coroutines.}

Resource utilization data was collected using VisualVM during 30-s
benchmark runs, capturing detailed CPU and memory usage patterns under
sustained load conditions. Each profiling session monitored system performance
during the 1 to 128 concurrent user load test, providing comprehensive insights
into runtime behavior.

\subsubsection*{Virtual Threads vs. Structured Concurrency}

\textls[-15]{Figures~\ref{fig:cpu-susp} and \ref{fig:cpu-virt} show CPU utilization profiles
for the \texttt{HtmlFlow-Virtual} and \texttt{HtmlFlow-Susp}} implementations,
respectively. The subsequent gradual increase in CPU utilization reflects the
ramp-up of concurrent users during the benchmark. \texttt{HtmlFlow-Virtual}
exhibits a distinctive bell-shaped curve, rapidly climbing from near-zero to
peak CPU usage of around 50\% during the sustained load phase, maintaining
moderate utilization during the test period, and then dropping sharply back to
baseline. \texttt{HtmlFlow-Susp} displays a similar pattern but with a slightly
lower peak CPU utilization of approximately 42\%, showing marginally lower
intensive CPU usage during the sustained load phase. Both profiles demonstrate
comparable resource consumption patterns, with the virtual thread
implementation showing only slightly higher CPU demands. The GC activity
indicators in both profiles remain relatively low throughout the test duration,
confirming that garbage collection overhead does not significantly impact CPU
utilization in either approach.

Figures~\ref{fig:gc-susp} and~\ref{fig:gc-virt} show similar memory usage for
both approaches, with peaks around 300~MB during load. The initial spike to
450~MB in both graphs corresponds to the applications' bootstrapping phase.
Overall, these results indicate that under the tested conditions, Virtual
Threads and suspendable coroutines exhibit comparable memory~consumption.

Virtual Threads allocate a full call stack per thread, which is unmounted and
stored on the heap when suspended (e.g., during blocking I/O). In contrast,
Kotlin coroutines compile into state machines that capture only the minimal
execution state needed to resume computation. This typically allows coroutines
to scale more efficiently in terms of memory usage, especially under high
concurrency.

\begin{figure}[H]
\vspace{-3pt}
     \includegraphics[width=0.48\textwidth]{./Graphs/cpu-susp.png}
     \caption{CPU utilization profiling with HtmlFlow-Susp in Spring WebFlux during 30-s load~test.}     \label{fig:cpu-susp}
\end{figure}

\begin{figure}[H]

     \includegraphics[width=0.48\textwidth]{./Graphs/cpu-virt.png}
     \caption{CPU utilization profiling with HtmlFlow-Virtual in Spring WebFlux during 30-s load~test.}   \label{fig:cpu-virt}
\end{figure}

\begin{figure}[H]
\vspace{-6pt}
    \includegraphics[width=0.48\textwidth]{./Graphs/mem-susp.png}
     \caption{Memory utilization and garbage collection behavior with HtmlFlow-Susp in Spring WebFlux during 30-s load test.}             \label{fig:gc-susp}
\end{figure}

     \begin{figure}[H]
\vspace{-6pt}
   \includegraphics[width=0.48\textwidth]{./Graphs/mem-virt.png}
     \caption{Memory utilization and garbage collection behavior with HtmlFlow-Virtual in Spring WebFlux during 30-s load test.} \label{fig:gc-virt}
\end{figure}

\section{\highlighting{Discussion} %MDPI: To AE: This section seems to be added compared to the PDF file from susy, please check if it is correct.
}\label{s6}

Beronić et al.~\cite{9803765} compared different structured concurrency
constructs in Java and Kotlin in the context of a multithreaded HTTP server.
Their benchmark scenario differs from ours: their server awaited incoming
requests and, upon receiving one, executed a task involving object creation,
writing the object's information to a file, and returning the data as a
response. Consistent with our findings, they concluded that both Kotlin's
coroutines and Java's virtual threads offer significant performance
improvements over traditional JVM-based threads in concurrent applications.

Our results diverge from theirs regarding memory usage. They reported lower
heap usage for Java virtual threads (16–64~MB) compared to Kotlin coroutines
(52–99~MB), which contrasts with our measurements. This discrepancy may be
attributed to the different workloads: while their benchmark includes a single
I/O read-write operation per request, our experiments focus exclusively on
read-only I/O tasks.

The work of Navarro et al.~\cite{navarro2023considerations} focused on the
performance of Quarkus, which relies on Eclipse Vert.x~\cite{vertx}, itself
built on top of Netty~\cite{netty}. This represents a similar environment to
the one evaluated in our experiments with Quarkus.~Their study also emphasized
template rendering, using a blocking HTML template engine (Qute). The benchmark
they employed is the Fortunes
test ({\url{https://www.techempower.com/benchmarks}}, \hl{accessed on}), which involves
rendering a simple HTML table with only two data bindings per row.

Their results show that Quarkus with Virtual Threads outperformed the
traditional thread-per-request model under increased concurrency. Moreover,
compared to the reactive model, virtual threads demonstrated competitive
performance, particularly when executing blocking operations within template
engines.
In contrast, our experiments revealed that the reactive approach in Spring
WebFlux performs better than virtual threads in the same experiment conducted
with Quarkus, suggesting that WebFlux provides a more efficient infrastructure
for managing reactive processing flows.

The work of Šimatović et al.~\cite{vsimatovic2025evaluating} presents an
evaluation of Java Virtual Threads under different garbage collectors. They
explored three types of workloads, only one of which was based on a web
application scenario. This scenario replicated large-scale web scraping by
executing 1000 parallel tasks, each consisting of an I/O-bound operation
followed by CPU-bound string processing.

In this workload, garbage collection activity remained low across all
collectors, indicating reduced memory pressure. These findings suggest that
Virtual Threads improve memory allocation efficiency and reduce the frequency
of garbage collection cycles---particularly when used with concurrent garbage
collectors.

Our observations suggest that the choice between Virtual Threads and
suspendable coroutines should be guided by both runtime behavior and
development constraints. Virtual Threads offer the compelling advantage of
preserving synchronous programming semantics while achieving competitive
performance across moderate to high-load scenarios---a significant benefit for
teams migrating existing codebases or working with legacy template engines.
Suspendable coroutines, while requiring developers to adopt asynchronous
programming paradigms, may provide greater efficiency in scenarios involving
high concurrency, deeper call stacks, or highly dynamic workloads.

The framework-specific variations observed in our results---particularly the
performance differences between Spring WebFlux and Quarkus
implementations---suggest that the underlying reactive runtime architecture
significantly influences the effectiveness of each approach. This finding
underscores the importance of considering not only the concurrency model but
also the specific framework ecosystem when making architectural decisions for
PSSR implementations.

\section{Conclusions}\label{s7}

In recent decades, non-blocking I/O has become the standard approach for building
highly responsive and scalable web servers. However, traditional synchronous
programming models are not compatible with non-blocking APIs, which typically
rely on callback-based conventions such as
\textit{\hl{continuation-passing style}} (CPS)~\cite{scheme} or
\textit{\hl{promises}}~\cite{promise}. These approaches not only hinder
sequential readability but also increase code verbosity, making them more
error-prone.
Alternatives like the \texttt{async}/\texttt{await} idiom~\cite{async_await}
or \textit{\hl{suspend functions}}~\cite{elizarov2021coroutines} simplify
asynchronous programming by mimicking a sequential style without blocking
threads.
Recent proposals~\cite{carvalho2023async,wise2024pssr} have applied contemporary
asynchronous idioms to SSR web templates, demonstrating how they can overcome
the scalability bottlenecks present in traditional web template engines.

As an alternative, Java Virtual Threads can be applied to any blocking I/O call,
leveraging the Java runtime to transparently intercept and convert it into non-blocking I/O,
without requiring any changes to the calling code or exposing its internal complexities.
In this work, we explored how this technique can be applied to traditional
web template engines and whether it can achieve performance competitive with
reactive approaches provided by frameworks such as Thymeleaf~\cite{webflux} and HtmlFlow~\cite{htmlflow}.
Our benchmarks across Spring WebFlux, Spring MVC, and Quarkus show that synchronous
non-blocking execution using virtual threads consistently delivers performance
comparable to asynchronous non-blocking approaches under high concurrency.
These findings highlight virtual threads as a promising alternative to complex
asynchronous programming models, offering a simpler development experience
without compromising scalability or responsiveness.

% 
% All figures and tables should be cited in the main text as Figure~\ref{fig1},
% Table~\ref{tab1}, etc.
% 
% \begin{figure}[H]
% 	%\isPreprints{\centering}{} % Only used for preprints
% 	\includegraphics[width=4 cm]{Definitions/logo-mdpi}
% 	\caption{This is a figure. Schemes follow the same formatting.\label{fig1}}
% \end{figure}
% \unskip
% 
% % Example of a figure that spans the whole page width and with subfigures. The same concept works for tables, too.
% \begin{figure}[H]
% 	%\isPreprints{}{% This command is only used for ``preprints''.
% 	\begin{adjustwidth}{-\extralength}{0cm}
% 		\centering
% 		%} % If the paper is ``preprints'', please uncomment this parenthesis.
% 		\subfloat[\centering]{\includegraphics[width=7.0cm]{Definitions/logo-mdpi}}
% 		%\hfill
% 		\subfloat[\centering]{\includegraphics[width=7.0cm]{Definitions/logo-mdpi}}\\
% 		\subfloat[\centering]{\includegraphics[width=7.0cm]{Definitions/logo-mdpi}}
% 		%\hfill
% 		\subfloat[\centering]{\includegraphics[width=7.0cm]{Definitions/logo-mdpi}}
% 		%\isPreprints{}{% This command is only used for ``preprints''.
% 	\end{adjustwidth}
% 	%} % If the paper is ``preprints'', please uncomment this parenthesis.
% 	\caption{This is a wide figure. Schemes follow the same formatting. If there are multiple panels, they should be listed as: (\textbf{a}) Description of what is contained in the first panel. (\textbf{b}) Description of what is contained in the second panel. (\textbf{c}) Description of what is contained in the third panel. (\textbf{d}) Description of what is contained in the fourth panel. Figures should be placed in the main text near to the first time they are cited. A caption on a single line should be centered.\label{fig2}}
% \end{figure}
% 
% \begin{table}[H]
% 	\caption{This is a wide table.\label{tab2}}
% 	%\isPreprints{\centering}{% This command is only used for ``preprints''.
% 	\begin{adjustwidth}{-\extralength}{0cm}
% 		%} % If the paper is ``preprints'', please uncomment this parenthesis.
% 		%\isPreprints{\begin{tabularx}{\textwidth}{CCCC}}{% This command is only used for ``preprints''.
% 		\begin{tabularx}{\fulllength}{CCCC}
% 			%} % If the paper is ``preprints'', please uncomment this parenthesis.
% 			\toprule
% 			\textbf{Title 1}              & \textbf{Title 2} & \textbf{Title 3} & \textbf{Title 4} \\
% 			\midrule
% 			\multirow[m]{3}{*}{Entry 1 *} & Data             & Data             & Data             \\
% 			                              & Data             & Data             & Data             \\
% 			                              & Data             & Data             & Data             \\
% 			\midrule
% 			\multirow[m]{3}{*}{Entry 2}   & Data             & Data             & Data             \\
% 			                              & Data             & Data             & Data             \\
% 			                              & Data             & Data             & Data             \\
% 			\bottomrule
% 		\end{tabularx}
% 		%		\isPreprints{}{% This command is only used for ``preprints''.
% 	\end{adjustwidth}
% 	%} % If the paper is ``preprints'', please uncomment this parenthesis.
% 	\noindent{\footnotesize{* Tables may have a footer.}}
% \end{table}
% 
%\begin{listing}[H]
%\caption{Title of the listing}
%\rule{\columnwidth}{1pt}
%\raggedright Text of the listing. In font size footnotesize, small, or normalsize. Preferred format: left aligned and single spaced. Preferred border format: top border line and bottom border line.
%\rule{\columnwidth}{1pt}
%\end{listing}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \section{Conclusions}
% 
% This section is not mandatory, but can be added to the manuscript if the
% discussion is unusually long or complex.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\vspace{6pt}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% optional
%\supplementary{The following supporting information can be downloaded at:  \linksupplementary{s1}, Figure S1: title; Table S1: title; Video S1: title.}

% Only for journal Methods and Protocols:
% If you wish to submit a video article, please do so with any other supplementary material.
% \supplementary{The following supporting information can be downloaded at: \linksupplementary{s1}, Figure S1: title; Table S1: title; Video S1: title. A supporting video article is available at doi: link.}

% Only used for preprtints:
% \supplementary{The following supporting information can be downloaded at the website of this paper posted on \href{https://www.preprints.org/}{Preprints.org}.}

% Only for journal Hardware:
% If you wish to submit a video article, please do so with any other supplementary material.
% \supplementary{The following supporting information can be downloaded at: \linksupplementary{s1}, Figure S1: title; Table S1: title; Video S1: title.\vspace{6pt}\\
%\begin{tabularx}{\textwidth}{lll}
%\toprule
%\textbf{Name} & \textbf{Type} & \textbf{Description} \\
%\midrule
%S1 & Python script (.py) & Script of python source code used in XX \\
%S2 & Text (.txt) & Script of modelling code used to make Figure X \\
%S3 & Text (.txt) & Raw data from experiment X \\
%S4 & Video (.mp4) & Video demonstrating the hardware in use \\
%... & ... & ... \\
%\bottomrule
%\end{tabularx}
%}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\authorcontributions{Conceptualization, B.P. and F.M.C.; methodology, B.P. and F.M.C.; software, B.P. and F.M.C.; validation, B.P. and F.M.C.; formal analysis, B.P. and F.M.C.; investigation, B.P. and F.M.C.; resources, B.P. and F.M.C.; data curation, B.P. and F.M.C.; writing---original draft preparation, B.P. and F.M.C.; writing---review and editing, B.P. and F.M.C.; supervision, F.M.C.; project administration, F.M.C. All authors have read and agreed to the published version of the manuscript.}

\funding{This research received no external funding.}

\institutionalreview{This study did not involve human subjects or personal data; therefore, Institutional Review Board approval was not required.}

\dataavailability{The data presented in this study are available in Github at \url{https://github.com/xmlet/comparing-non-blocking-progressive-ssr}, \hl{accessed on}.}

% Only for journal Drones
%\durcstatement{Current research is limited to the [please insert a specific academic field, e.g., XXX], which is beneficial [share benefits and/or primary use] and does not pose a threat to public health or national security. Authors acknowledge the dual-use potential of the research involving xxx and confirm that all necessary precautions have been taken to prevent potential misuse. As an ethical responsibility, authors strictly adhere to relevant national and international laws about DURC. Authors advocate for responsible deployment, ethical considerations, regulatory compliance, and transparent reporting to mitigate misuse risks and foster beneficial outcomes.}

% Only for journal Nursing Reports
%\publicinvolvement{Please describe how the public (patients, consumers, carers) were involved in the research. Consider reporting against the GRIPP2 (Guidance for Reporting Involvement of Patients and the Public) checklist. If the public were not involved in any aspect of the research add: ``No public involvement in any aspect of this research''.}
%
%% Only for journal Nursing Reports
%\guidelinesstandards{Please add a statement indicating which reporting guideline was used when drafting the report. For example, ``This manuscript was drafted against the XXX (the full name of reporting guidelines and citation) for XXX (type of research) research''. A complete list of reporting guidelines can be accessed via the equator network: \url{https://www.equator-network.org/}.}
%
%% Only for journal Nursing Reports
%\useofartificialintelligence{Please describe in detail any and all uses of artificial intelligence (AI) or AI-assisted tools used in the preparation of the manuscript. This may include, but is not limited to, language translation, language editing and grammar, or generating text. Alternatively, please state that “AI or AI-assisted tools were not used in drafting any aspect of this manuscript”.}

\conflictsofinterest{The authors declare no conflicts of interest.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\isPreprints{}{% This command is only used for ``preprints''.
\begin{adjustwidth}{-\extralength}{0cm}
	%} % If the paper is ``preprints'', please uncomment this parenthesis.
	%\printendnotes[custom] % Un-comment to print a list of endnotes

	\reftitle{References}

	% Please provide either the correct journal abbreviation (e.g., according to the “List of Title Word Abbreviations” http://www.issn.org/services/online-services/access-to-the-ltwa/) or the full name of the journal.
	% Citations and References in Supplementary files are permitted provided that they also appear in the reference list here.

	%=====================================
	% References, variant A: external bibliography
	%=====================================
\begin{thebibliography}{999}

\bibitem[Edgar(2024)]{Edgar2024-FCP}
Edgar, M. First Contentful Paint.
\newblock In {\em Speed Metrics Guide: Choosing the Right Metrics to Use When
  Evaluating Websites}; Apress: Berkeley, CA, USA, 2024; pp. 73--91.
\newblock {\url{https://doi.org/10.1007/979-8-8688-0155-6_5}}.

\bibitem[Carvalho(2025)]{wise2024pssr}
Carvalho, F.M.
\newblock Progressive Server-Side Rendering with Suspendable Web Templates.
\newblock In \emph{Web Information Systems Engineering---WISE
  2024}; Barhamgi, M., Wang, H., Wang, X., Eds.; \hl{Springer}%MDPI: newly added, please confirm it
: Singapore,  2025; pp. 458--473.

\bibitem[Elmeleegy et~al.(2004)Elmeleegy, Chanda, Cox, and
  Zwaenepoel]{event-driven-servers}
Elmeleegy, K.; Chanda, A.; Cox, A.L.; Zwaenepoel, W.
\newblock Lazy Asynchronous I/O for Event-Driven Servers.
\newblock In Proceedings of the Annual Conference on USENIX
  Annual Technical Conference, ATEC '04, \hl{Boston, MA, USA, 27 June 27--2 July 2004;} p.~21.

\bibitem[Carvalho and Fialho(2023)]{carvalho2023async}
Carvalho, F.M.; Fialho, P.
\newblock Enhancing SSR in Low-Thread Web Servers: A Comprehensive Approach for
  Progressive Server-Side Rendering with Any Asynchronous API and Multiple Data
  Models.
\newblock In Proceedings of the 19th International
  Conference on Web Information Systems and Technologies, WEBIST '23, \hl{Rome, Italy, 15--17 November  2023.}

\bibitem[Carvalho(2017)]{htmlflow}
Carvalho, F.M.
\newblock HtmlFlow Java DSL to Write Typesafe HTML.
\newblock Technical Report, 2017. Available online: \url{https://htmlflow.org/} \hl{(accessed on).} %MDPI: Please add the access date (format: Date Month Year), e.g., accessed on 1 January 2020.

\bibitem[Fernández(2011)]{thymeleaf}
Fernández, D.
\newblock Thymeleaf.
\newblock Technical Report, 2011. Available online: \url{https://www.thymeleaf.org/} \hl{(accessed on).}

\bibitem[Fowler(2010)]{Fowler03}
\hl{Fowler, M.} %MDPI: Refs.7 and 15 are duplicated. Please remove duplicated references and rearrange all the references so that they appear in numerical order. Please ensure that there are no duplicated references.
\newblock {\em Domain Specific Languages}; Addison-Wesley Professional: \hl{Boston, MA, USA,} 2010.

\bibitem[Fowler(2002)]{Fowler02}
Fowler, M.
\newblock {\em Patterns of Enterprise Application Architecture}; Addison-Wesley
  Longman Publishing Co., Inc.: Boston, MA, USA, 2002.

\bibitem[Alur et~al.(2001)Alur, Malks, and Crupi]{Alur01}
Alur, D.; Malks, D.; Crupi, J.
\newblock {\em Core J2EE Patterns: Best Practices and Design Strategies};
  Prentice Hall PTR: Upper Saddle River, NJ, USA, 2001.

\bibitem[Parr(2004)]{Parr04}
Parr, T.J.
\newblock Enforcing Strict Model-view Separation in Template Engines.
\newblock In Proceedings of the 13th International
  Conference on World Wide Web, WWW '04, New York, NY, USA, \hl{17--20 May} 2004; pp.
  224--233.
\newblock {\url{https://doi.org/10.1145/988672.988703}}.

\bibitem[E.~Krasner and Pope(1988)]{mvc88}
Krasner, G.E.; Pope, S.
\newblock A Description of the Model-View-Controller User Interface Paradigm in
  the Smalltalk80 System.
\newblock {\em J.~Object-Oriented Program.---JOOP} {\bf 1988}, {\em
  1},~26--49.

\bibitem[{Netflix} et~al.(2015){Netflix}, {Pivotal}, {Red Hat}, {Oracle},
  {Twitter}, and {Lightbend}]{ReactiveStreams}
{Netflix}; {Pivotal}; {Red Hat}; {Oracle}; {Twitter}; {Lightbend}.
\newblock Reactive Streams Specification.
\newblock Technical Report, 2015. Available online: \url{https://www.reactive-streams.org/} \hl{(accessed on).}

\bibitem[Landin(1966)]{landin1966next}
Landin, P.J.
\newblock The next 700 programming languages.
\newblock {\em Commun.  ACM} {\bf 1966}, {\em 9},~157--166.

\bibitem[Evans and Fowler(2004)]{evans2004domain}
Evans, E.; Fowler, M.
\newblock {\em Domain-Driven Design: Tackling Complexity in the Heart of
  Software}; Addison-Wesley: \hl{Boston, MA, USA,} 2004.

\bibitem[Fowler(2010)]{dslbook}
\hl{Fowler, M.}
\newblock {\em Domain Specific Languages}, 1st ed.; Addison-Wesley
  Professional: \hl{Boston, MA, USA,} 2010.

\bibitem[Thompson(1968)]{thompson1968}
Thompson, K.
\newblock Programming Techniques: Regular Expression Search Algorithm.
\newblock {\em Commun. ACM} {\bf 1968}, {\em 11},~419–422.
\newblock {\url{https://doi.org/10.1145/363347.363387}}.

\bibitem[Resig(2007)]{resig2007pro}
Resig, J.
\newblock {\em Pro JavaScript Techniques}; Apress: \hl{New York, NY, USA,} 2007.

\bibitem[Hors et~al.(2004)Hors, Hégaret, Wood, Nicol, Robie, Champion,
  Arbortext, and Byrne]{dom}
Hors, A.L.; Hégaret, P.L.; Wood, L.; Nicol, G.; Robie, J.; Champion, M.;
  Arbortext; Byrne, S.
\newblock Document Object Model (DOM) Level 3 Core Specification.
\newblock Technical Report, 2004. Available online: \url{https://www.w3.org/TR/2004/REC-DOM-Level-3-Core-20040407/} \hl{(accessed on).}

\bibitem[Carvalho et~al.(2020)Carvalho, Duarte, and Gouesse]{carvalho2020}
Carvalho, F.M.; Duarte, L.; Gouesse, J.
\newblock Text Web Templates Considered Harmful.
\newblock In {\em Lecture Notes in Business Information Processing}; \hl{{2020};} %MDPI: We are sorry but we could not find the required information about this entry. Please add the name of the publisher and their location.
 pp. 69--95.

\bibitem[Landin(1965)]{Landin65}
Landin, P.J.
\newblock Correspondence Between ALGOL 60 and Church's Lambda-notation: Part I.
\newblock {\em Commun. ACM} {\bf 1965}, {\em 8},~89--101.
\newblock {\url{https://doi.org/10.1145/363744.363749}}.

\bibitem[Sussman and Steele(1975)]{scheme}
Sussman, G.; Steele, G.
\newblock {\em Scheme: An Interpreter for Extended Lambda Calculus}; \hl{AI Memo
  No, MIT, Artificial Intelligence Laboratory:  1975.} %MDPI: We are sorry but we could not find the required information about this entry. Please add the location of the publisher (city and country).

\bibitem[Friedman and Wise(1978)]{promise}
Friedman; Wise.
\newblock Aspects of Applicative Programming for Parallel Processing.
\newblock {\em IEEE Trans. Comput.} {\bf 1978}, {\em
  C-27},~289--296.
\newblock {\url{https://doi.org/10.1109/TC.1978.1675100}}.

\bibitem[Syme et~al.(2011)Syme, Petricek, and Lomov]{async_await}
Syme, D.; Petricek, T.; Lomov, D.
\newblock The F{\#} Asynchronous Programming Model.
\newblock In \emph{Practical Aspects of Declarative Languages};
  Rocha, R., Launchbury, J., Eds.; \hl{Springer}: Berlin/Heidelberg, Germany, 2011; pp. 175--189.

\bibitem[Elizarov et~al.(2021)Elizarov, Belyaev, Akhin, and
  Usmanov]{elizarov2021coroutines}
Elizarov, R.; Belyaev, M.; Akhin, M.; Usmanov, I.
\newblock Kotlin coroutines: Design and implementation.
\newblock In Proceedings of the 2021 ACM SIGPLAN
  International Symposium on New Ideas, New Paradigms, and Reflections on
  Programming and Software, \hl{Chicago, IL, USA, 20--22 October 2021}; pp. 68--84.

\bibitem[Meijer(2009)]{rx-observable}
Meijer, E.
\newblock Democratizing The Cloud With The .NET Reactive Framework Rx.
\newblock In Proceedings of the Internaional Softare Development Conference, \hl{Vancouver, BC, Canada, 16--24 May 2009} %MDPI: 
;  Available online: \url{https://qconsf.com/sf2009/sf2009/speaker/Erik+Meijer.html} \hl{(accessed on).} %MDPI:

\bibitem[contributors(2025)]{rxjava}
RxJava Contributors.
\newblock RX Java.
\newblock Technical Report, 2025. Available online: \url{https://github.com/ReactiveX/RxJava} \hl{(accessed on).}

\bibitem[VMWare and contributors(2025)]{projectreactor}
VMWare; Contributors.
\newblock Project Reactor.
\newblock Technical Report, 2025. Available online: \url{https://projectreactor.io/} \hl{(accessed on).}

\bibitem[Davis(2019)]{akka}
Davis, A.L. Akka HTTP and Streams.
\newblock In {\em Reactive Streams in Java: Concurrency with RxJava, Reactor,
  and Akka Streams}; Apress: Berkeley, CA, USA, 2019; pp. 105--128.

\bibitem[Ponge et~al.(2021)Ponge, Navarro, Escoffier, and
  Le~Mou\"{e}l]{mutiny2021}
Ponge, J.; Navarro, A.; Escoffier, C.; Le~Mou\"{e}l, F.
\newblock Analysing the performance and costs of reactive programming libraries
  in Java.
\newblock In Proceedings of the 8th ACM SIGPLAN
  International Workshop on Reactive and Event-Based Languages and Systems, REBLS 2021, New
  York, NY, USA, \hl{18 October} 2021; pp. 51–60.
\newblock {\url{https://doi.org/10.1145/3486605.3486788}}.

\bibitem[Breslav(2016)]{kotlinlang}
Breslav, A.
\newblock Kotlin Language Documentation.
\newblock Technical Report, 2016. Available online: \url{https://kotlinlang.org/docs/kotlin-docs.pdf} \hl{(accessed on).}

\bibitem[Vogel and Springer(2022)]{progressive-2022}
Vogel, L.; Springer, T.
\newblock User Acceptance of Modified Web Page Loading Based on Progressive
  Streaming.
\newblock In Proceedings of the International Conference on Web \hl{Engineering, 2022;} %MDPI: please add location and date for conference
  Springer:  \hl{Berlin/Heidelberg, Germany,} %We added the location of publisher. Please confirm
 2022; pp. 391--405.

\bibitem[Atwood(2005)]{pssr2005}
Atwood, J.
\newblock The Lost Art of Progressive HTML Rendering.
\newblock Technical Report, 2005.
  Available online: \url{https://blog.codinghorror.com/the-lost-art-of-progressive-html-rendering/} \hl{(accessed on).}

\bibitem[Farago et~al.(2007)Farago, Williams, Walsh, Whyte, Goel, and
  Fung]{scroll2007}
Farago, J.; Williams, H.; Walsh, J.; Whyte, N.; Goel, K.; Fung, P.
\newblock Object Search UI and Dragging Object Results\hl{,  ~23 2007.} %MDPI: please confirm whether can be deleted
\newblock US Patent Applications 11/353,787, \hl{2007}%MDPI: please add day and month
.

\bibitem[Schiller(2007)]{schiller2007progressive}
Schiller, S.
\newblock Progressive Loading.
\newblock US Patent Applications 11/364,992, \hl{2007.} %MDPI: please add day and month

\bibitem[Von~Behren et~al.(2003)Von~Behren, Condit, and Brewer]{von2003events}
Von~Behren, R.; Condit, J.; Brewer, E.
\newblock Why Events Are a Bad Idea (for $\{$High-Concurrency$\}$ Servers).
\newblock In Proceedings of the 9th Workshop on Hot Topics in Operating Systems
  (HotOS IX), \hl{Lihue, HI, USA, 18–21 May 2003.}

\bibitem[Kambona et~al.(2013)Kambona, Boix, and De~Meuter]{callbackhell}
Kambona, K.; Boix, E.G.; De~Meuter, W.
\newblock An Evaluation of Reactive Programming and Promises for Structuring
  Collaborative Web Applications.
\newblock In Proceedings of the 7th Workshop on Dynamic
  Languages and Applications, DYLA '13, New York, NY, USA, \hl{1–5 July} 2013.
\newblock {\url{https://doi.org/10.1145/2489798.2489802}}.

\bibitem[Kant and Mohapatra(2000)]{kant2000scalable}
Kant, K.; Mohapatra, P.
\newblock Scalable Internet servers: Issues and challenges.
\newblock {\em ACM SIGMETRICS Perform. Eval. Rev.} {\bf 2000}, {\em
  28},~5--8.

\bibitem[Meijer(2012)]{Meijer12}
Meijer, E.
\newblock Your Mouse is a Database.
\newblock {\em Queue} {\bf 2012}, {\em 10},~20:20--20:33.
\newblock {\url{https://doi.org/10.1145/2168796.2169076}}.

\bibitem[Jin et~al.(2015)Jin, Wah, Cheng, and Wang]{Jin15}
Jin, X.; Wah, B.W.; Cheng, X.; Wang, Y.
\newblock Significance and Challenges of Big Data Research.
\newblock {\em Big Data Res.} {\bf 2015}, {\em 2},~59--64.
\newblock {\url{https://doi.org/10.1016/j.bdr.2015.01.006}}.

\bibitem[Karsten and Barghi(2020)]{karsten2020}
Karsten, M.; Barghi, S.
\newblock User-Level Threading: Have Your Cake and Eat It Too.
\newblock {\em Proc. ACM Meas. Anal. Comput. Syst.} {\bf 2020}, {\em 4}.
\newblock {\url{https://doi.org/10.1145/3379483}}.

\bibitem[Burke(2013)]{burke2013restful}
Burke, B.
\newblock {\em RESTful Java with JAX-RS 2.0: Designing and Developing
  Distributed Web Services}; \hl{O'Reilly Media, Inc.: 2013.} %MDPI: We are sorry but we could not find the required information about this entry. Please add the location of the publisher (city and country).

\bibitem[Haynes et~al.(1984)Haynes, Friedman, and
  Wand]{coroutines_continuations}
Haynes, C.T.; Friedman, D.P.; Wand, M.
\newblock Continuations and Coroutines.
\newblock In Proceedings of the 1984 ACM Symposium on LISP
  and Functional Programming, LFP '84, \hl{Austin, TX, USA, 5--8 August 1984}; pp. 293–298.
\newblock {\url{https://doi.org/10.1145/800055.802046}}.

\bibitem[Veen and Vlijmincx(2024)]{Veen2024}
Veen, R.; Vlijmincx, D. Scoped Values.
\newblock In {\em Virtual Threads, Structured Concurrency, and Scoped Values:
  Explore Java's New Threading Model}; Apress: Berkeley, CA, USA, 2024.
\newblock {\url{https://doi.org/10.1007/979-8-8688-0500-4}}.

\bibitem[Buc(2009)]{Bucanek2009}
Model-View-Controller Pattern.
\newblock In {\em Learn Objective-C for Java Developers}; Apress: Berkeley, CA, USA, 
   2009; pp. 353--402.
\newblock {\url{https://doi.org/10.1007/978-1-4302-2370-2_20}}.

\bibitem[Bösecke(2015)]{templatesbenchmark}
Bösecke, M.
\newblock JMH Benchmark of the Most Popular Java Template Engines.
\newblock Technical Report, 2015.
  Available online: \url{https://github.com/mbosecke/template-benchmark} \hl{(accessed on).}

\bibitem[Reijn(2015)]{springbenchmark}
Reijn, J.
\newblock Comparing Template Engines for Spring MVC.
\newblock Technical Report, 2015.
 Available online: \url{https://github.com/jreijn/spring-comparing-template-engines} \hl{(accessed on).}

\bibitem[Beronić et~al.(2022)Beronić, Modrić, Mihaljević, and
  Radovan]{9803765}
Beronić, D.; Modrić, L.; Mihaljević, B.; Radovan, A.
\newblock Comparison of Structured Concurrency Constructs in Java and Kotlin---Virtual Threads and Coroutines.
\newblock In Proceedings of the 2022 45th Jubilee International Convention on
  Information, Communication and Electronic Technology (MIPRO), \hl{Opatija, Croatia, 23--27 May 2022}%MDPI: Please confirm added date and location for conference
; pp. 1466--1471.
\newblock {\url{https://doi.org/10.23919/MIPRO55190.2022.9803765}}.

\bibitem[Navarro et~al.(2023)Navarro, Ponge, Le~Mou{\"e}l, and
  Escoffier]{navarro2023considerations}
Navarro, A.; Ponge, J.; Le~Mou{\"e}l, F.; Escoffier, C.
\newblock Considerations for integrating virtual threads in a Java framework: A
  Quarkus example in a resource-constrained environment.
\newblock In Proceedings of the 17th ACM International
  Conference on Distributed and Event-based Systems, \hl{Neuchatel, Switzerland, 27--30 June 2023}; pp. 103--114.

\bibitem[Fox(2014)]{vertx}
Fox, T.
\newblock Eclipse Vert.x$^{\text{TM}}$ Reactive Applications on the JVM.
\newblock Technical Report, 2014. Available online: \url{https://vertx.io/} \hl{(accessed~on).}

\bibitem[Maurer and Wolfthal(2015)]{netty}
Maurer, N.; Wolfthal, M.
\newblock {\em Netty in Action}; \hl{Manning: 2015.} %MDPI: We are sorry but we could not find the required information about this entry. Please add the location of the publisher (city and country).

\bibitem[{\v{S}}imatovi{\'c} et~al.(2025){\v{S}}imatovi{\'c}, Markulin,
  Beroni{\'c}, and Mihaljevi{\'c}]{vsimatovic2025evaluating}
{\v{S}}imatovi{\'c}, M.; Markulin, H.; Beroni{\'c}, D.; Mihaljevi{\'c}, B.
\newblock Evaluating Memory Management and Garbage Collection Algorithms with
  Virtual Threads in High-Concurrency Java Applications.
\newblock In Proceedings of the 48th ICT and Electronics Convention MIPRO 2025. Rijeka: Croatian Society for Information, Communication and Electronic Technology–MIPRO, \hl{Opatija, Croatia, 2--6 June 2025.} %MDPI: Please confirm added conference date and location.

\bibitem[Deinum and Cosmina(2021)]{webflux}
Deinum, M.; Cosmina, I., Building Reactive Applications with Spring WebFlux.
\newblock In {\em Pro Spring MVC with WebFlux: Web Development in Spring
  Framework 5 and Spring Boot 2}; Apress: Berkeley, CA, USA, 2021; pp. 369--420.

\end{thebibliography}

	%=====================================
	% References, variant B: internal bibliography
	%=====================================

	\PublishersNote{}
	%\isPreprints{}{% This command is only used for ``preprints''.
\end{adjustwidth}
%} % If the paper is ``preprints'', please uncomment this parenthesis.
\end{document}

### 2025-08-06 docs/article/software-3702556.tex
Proofreading fixes from MDPI

@@
 Still, traditional HTML template engines often rely on blocking interfaces that hinder their use in asynchronous, non-blocking contexts required for PSSR.
-This paper analyzes how Java Virtual Threads, introduced in Java 21, enable non-blocking execution of blocking I/O operations, allowing the reuse of traditional template engines for PSSR without complex asynchronous programming models.
-We benchmark multiple engines across Spring WebFlux, Spring MVC, and Quarkus using reactive, suspendable, and virtual-thread-based approaches.~Results show that virtual threads allow blocking engines to scale comparably to those designed for non-blocking I/O, achieving high throughput and responsiveness under load.
-This demonstrates that Virtual Threads provide a compelling path to simplify the implementation of PSSR with familiar HTML templates, significantly lowering the barrier to entry while maintaining~performance.}
+This paper analyzes how Java virtual threads, introduced in Java 21, enable non-blocking execution of blocking I/O operations, allowing the reuse of traditional template engines for PSSR without complex asynchronous programming models.
+We benchmark multiple engines across Spring WebFlux, Spring MVC, and Quarkus using reactive, suspendable, and virtual thread-based approaches.~Results show that virtual threads allow blocking engines to scale comparably to those designed for non-blocking I/O, achieving high throughput and responsiveness under load.
+This demonstrates that virtual threads provide a compelling path to simplify the implementation of PSSR with familiar HTML templates, significantly lowering the barrier to entry while maintaining~performance.}
 
 % Keywords
-\keyword{web templates; Server-Side Rendering; non-blocking; Java virtual threads; asynchronous}
+\keyword{web templates; server-side rendering; non-blocking; Java virtual threads; asynchronous}
 
@@
 allowing progressive rendering as data becomes available, significantly reducing the
-\textit{\hl{Time to First Byte (TTFB)}} and improving perceived load times compared to traditional
+\textit{\hl{time-to-first-byte (TTFB)}} and improving perceived load times compared to traditional
 SSR approaches~\cite{wise2024pssr}. In this respect,
@@
 better scalability.
-By leveraging asynchronous I/O operations---such as
+By leveraging asynchronous I/O operations, such as
 database queries and API calls, servers can avoid blocking threads while
@@
 
-With the introduction of \textit{\hl{Virtual
-    Threads}}~({\url{https://openjdk.org/jeps/444}}, \hl{accessed on}%MDPI: footnote is not allowed in this journal, so we moved it to here, please confirm it, and please add the access date (format: Date Month Year), e.g., accessed on 1 January 2020.
+With the introduction of \textit{\hl{virtual
+    threads}}~({\url{https://openjdk.org/jeps/444}}, \hl{accessed on}%MDPI: footnote is not allowed in this journal, so we moved it to here, please confirm it, and please add the access date (format: Date Month Year), e.g., accessed on 1 January 2020.
 ) in Java 21, it is now
@@
 of which have been used to achieve asynchronous I/O in the Java ecosystem. As
-an alternative, we investigate whether Java’s Virtual Threads can offer
+an alternative, we investigate whether Java’s virtual threads can offer
 comparable performance while preserving the simplicity of synchronous code.
@@
 Section~\ref{s6} compares our results with those from other studies.
-Conclusions are presented in Section \ref{s7}.
+The conclusions of this work are presented in Section \ref{s7}.
 
@@
 \textit{\hl{server-side}} and \textit{\hl{client-side}} approaches, we focus solely on
-web template technologies for \textit{\hl{server-side rendering}}, as our work is
-centered on that approach. Before getting deep on each of the aforementioned
+web template technologies for \textit{\hl{Server-Side Rendering}}, as our work is
+centered on that approach. Before exploring each of the aforementioned
 characteristics, Table~\ref{table:cmplibs} presents a breakdown of mainstream
@@
 $^{(2)}$ Compile-time expression validation available via \texttt{@CheckedTemplate} and build-time metadata in Quarkus.
-$^{(3)}$ \texttt{Publisher} of Reactive Streams~\cite{ReactiveStreams}. Limited to a single model per web template.
+$^{(3)}$ \texttt{Publisher} of reactive streams~\cite{ReactiveStreams}. Limited to a single model per web template.
 $^{(4)}$ \texttt{kotlin.sequences.Sequence}
@@
 (Figure~\ref{fig:dsl-idioms}a) and \texttt{\_\_()} in \textbf{\hl{HtmlFlow}}
-(Figure~\ref{fig:dsl-idioms}c). But the \textit{\hl{nested function}} approach
+(Figure~\ref{fig:dsl-idioms}c). However, the \textit{\hl{nested function}} approach
 comes with a notable drawback: \textbf{\hl{it does not support PSSR}} because the
@@
 abstractions like the \texttt{Sequence} interface---also incompatible with
-\texttt{Iterable}. In consequence, template engines based on \textit{\hl{external}}
+\texttt{Iterable}. Consequently, template engines based on \textit{\hl{external}}
 DSLs must explicitly support each of these interfaces to accommodate the
@@
 engineers from notable companies such as Lightbend, Pivotal, and Netflix.
-The group began defining the \textit{\hl{Reactive Streams}}
+The group began defining the \textit{\hl{reactive streams}}
 specification~\cite{ReactiveStreams}, which established a standard for
@@
 The more recent Mutiny~\cite{mutiny2021}, while conforming to the
-\textit{\hl{Reactive Streams}} standard, proposes a different approach that avoids
+\textit{\hl{reactive streams}} standard, proposes a different approach that avoids
 complex optimizations, resulting in more straightforward and easier-to-maintain
@@
 
-\subsubsection{HTML safety}\label{s2.1.4}
+\subsubsection{HTML Safety}\label{s2.1.4}
 
@@
 The notion of \textit{\hl{progressive rendering}} has typically been aligned with
-\textit{\hl{client-side rendering}} (CSR), where a single HTML page with static
+\textit{\hl{Client-Side Rendering}} (CSR), where a single HTML page with static
 content is delivered upfront, while the dynamic content is fetched as the data
@@
 of the response incrementally as they become available. This is particularly
-useful for:
+useful for the following applications:
 \begin{itemize}
@@
     \item Writing dynamic HTML in fragments as data is fetched or computed.
-    \item Reducing Time to First Byte (TTFB) and improving perceived performance.
+    \item Reducing time-to-first-byte (TTFB) and improving perceived performance.
 \end{itemize}
@@
 by the servlet response buffer. By default, most servlet containers (e.g.,
-Tomcat) use a response buffer size of approximately 8KB. Data written to the
+Tomcat) use a response buffer size of approximately 8 KB. Data written to the
 response output stream is held in this buffer until it reaches capacity.
@@
 Reactor library~\cite{projectreactor}. These implementations adhere to the
-Reactive Streams specification~\cite{ReactiveStreams}, enabling
+reactive streams specification~\cite{ReactiveStreams}, enabling
 interoperability with other compliant reactive stream libraries. A
@@
 \begin{listing}[H]
-\caption{\textit{\hl{HtmlFlow reactive}} presentation template in Koltin with an \texttt{Observable} model.}
+\caption{\textit{\hl{HtmlFlow reactive}} presentation template in Kotlin with an \texttt{Observable} model.}
 \begin{lstlisting}[
@@
 These strengths make external DSLs appealing---even when they come at the cost of
-blocking, synchronous rendering. However, this tradeoff becomes critical under
+blocking synchronous rendering. However, this trade-off becomes critical under
 high concurrency, where blocking threads severely degrades throughput~\cite{wise2024pssr}.
@@
 
-The aforementioned blocking template engines are used in the context of Virtual
-Threads or alternative coroutine dispatchers, allowing the handler thread to be
+The aforementioned blocking template engines are used in the context of virtual
+threads or alternative coroutine dispatchers, allowing the handler thread to be
 released and reused for other requests.
@@
       \item \textbf{\hl{Virtual}}: The template engine is used in a non-blocking
-            context, where the HTML content is rendered within the context of Virtual
-            Threads. This method is used for the template engines that do not traditionally
+            context, where the HTML content is rendered within the context of virtual
+            threads. This method is used for the template engines that do not traditionally
             support non-blocking I/O, be it either because they use external DSLs and in
@@
             context, where the HTML content is rendered using the blocking interface of the
-            \texttt{\hl{Observable}} class, within the context of a OS-thread.
+            \texttt{\hl{Observable}} class, within the context of an OS-thread.
 \end{itemize}
@@
 blocking---specifically the writes to the \texttt{OutputStream}. When using
-Virtual Threads, the I/O operations are more efficient when compared to
+virtual threads, the I/O operations are more efficient when compared to
 platform threads, as they are executed in the context of a lightweight thread.
-Most of the computation is done in a separate thread from the one that receives
+Most of the computation is performed in a separate thread from the one that receives
 each request; we use a thread pool \texttt{TaskExecutor} to process requests,
@@
       \item \textbf{\hl{Virtual}}: The template engine is used in a non-blocking context,
-            where the HTML content is rendered within the context of Virtual Threads.
+            where the HTML content is rendered within the context of virtual threads.
 \end{itemize}
@@
 I/O, it operates on Vert.x worker threads, which prevents blocking of the event
-loop. When Virtual Threads are used, the I/O operations are handled
+loop. When virtual threads are used, the I/O operations are handled
 efficiently, as they are executed in lightweight threads.
@@
 response buffer size in the \textit{\hl{application.properties}} file. The default
-buffer size is 8KB, but we reduced it to 512 bytes, which allows the response
+buffer size is 8 KB, but we reduced it to 512 bytes, which allows the response
 to be sent to the client progressively as the HTML content is rendered.
@@
       \item \textbf{\hl{Virtual}}: The template engine is used in a non-blocking context,
-            where the HTML content is rendered within the context of Virtual Threads.
+            where the HTML content is rendered within the context of virtual threads.
       \item \textbf{\hl{Reactive}}: The template engine is used in a reactive context,
@@
 (x86\_64 architecture)} %MDPI: Please state the name of the manufacturer, city, and country from where the equipment was sourced.
- operating at 3.24 GHz, configured with 2 CPU cores and 2
+ operating at 3.24 GHz, configured with two CPU cores and two
 threads per core, and 7.8 GB of available RAM. The system utilizes a 75 GB SSD
-with ext4 filesystem, achieving 1.5 GB/s write throughput in basic I/O tests.
+with ext4 file system, achieving 1.5 GB/s write throughput in basic I/O tests.
 Network connectivity is provided through a 1500 MTU Ethernet interface. All
@@
 collector enabled by default. The JVM arguments used were: \texttt{-Xms512m
-     -Xmx16g}. Each test was executed 5 times per configuration to ensure result
+     -Xmx16g}. Each test was executed five times per configuration to ensure result
 reliability.
@@
 automatic HTML escaping to ensure fair comparison and consistent output across
-all template engines. Caching behaviour was left as the default for each
+all template engines. Caching behavior was left as the default for each
 engine, allowing them to optimize template loading and rendering as per their
@@
 
-For both the Apache Bench and JMeter tests, we simulate a 1000-request warmup
-period for each route with a concurrent user load of 32 users. The warmup
+For both the Apache Bench and JMeter tests, we simulate a 1000-request warm up
+period for each route with a concurrent user load of 32 users. The warm up
 period is followed by the actual test period, during which we simulate 256
@@
 
-Both Quarkus and Spring MVC implementations were configured with an 8KB output
+Both Quarkus and Spring MVC implementations were configured with an 8 KB output
 buffer size to ensure consistency, despite Quarkus not enabling PSSR at this
 size and Spring MVC not supporting PSSR for the tested templates. Testing with
-a reduced 512B buffer size showed only a 0.046\% performance difference with
+a reduced 512 B buffer size showed only a 0.046\% performance difference with
 Rocker, indicating negligible impact. Both frameworks use unlimited platform
@@
 equivalent to the approach shown in Listing~\ref{lst:presentation-flow}),
-Jstachio using Virtual Threads with the \texttt{Iterable} interface
-(\textit{\hl{Jstachio-Virtual}}), and Thymeleaf using the reactive View Resolver
+JStachio using virtual threads with the \texttt{Iterable} interface
+(\textit{\hl{JStachio-Virtual}}), and Thymeleaf using the reactive View Resolver
 driver (Thymeleaf-Rx). \textit{\hl{Blocking}} and \textit{\hl{Virtual}} represent the
-average throughput of the blocking approaches (i.e., KotlinX, Rocker, Jstachio,
+average throughput of the blocking approaches (i.e., KotlinX, Rocker, JStachio,
 Pebble, Freemarker, Trimou, HtmlFlow, and Thymeleaf) when run in the context of
-a separate coroutine dispatcher or Virtual Threads, respectively.}
-
-We show the \textit{\hl{HtmlFlow-Susp}}, \textit{\hl{Jstachio-Virtual}}, and
+a separate coroutine dispatcher or virtual threads, respectively.}
+
+We show the \textit{\hl{HtmlFlow-Susp}}, \textit{\hl{JStachio-Virtual}}, and
 \textit{Thymeleaf-Rx} engines separately to observe the performance of the
@@
      \caption{\hl{Throughput} %MDPI: The numbers in the picture are incomplete, please modify
- (requests per second) scalability results for Spring Webflux with \texttt{Presentations} class.}\label{fig:presentations-webflux-jmeter}
+ (requests per second) scalability results for Spring WebFlux with \texttt{Presentations} class.}\label{fig:presentations-webflux-jmeter}
 \end{figure}
@@
 using blocking template engines with a separate coroutine dispatcher, the
-engines are unable to scale effectively beyond 4 concurrent users. In contrast,
+engines are unable to scale effectively beyond four concurrent users. In contrast,
 \textit{\hl{HtmlFlow-Susp}} scales up to 128 concurrent users, achieving \hl{4487} %MDPI: Commas are only used for numbers with five or more digits.
@@
 requests per second. When blocking approaches are executed in the context of
-Virtual Threads---thus enabling non-blocking I/O---the engines scale up to 64
-concurrent users, with Jstachio using Virtual Threads reaching \hl{3514} requests
+virtual threads---thus enabling non-blocking I/O---the engines scale up to 64
+concurrent users, with JStachio using virtual threads reaching \hl{3514} requests
 per second. The Thymeleaf implementation using the reactive View Resolver
@@
 It is important to note that the differences in scalability and throughput
-between the \textit{\hl{HtmlFlow-Susp}}, \textit{\hl{Jstachio-Virtual}}, and
+between the \textit{\hl{HtmlFlow-Susp}}, \textit{\hl{JStachio-Virtual}}, and
 \textit{\hl{Thymeleaf-Rx}} approaches may be influenced by the specific template
@@
 specifically with HtmlFlow, we found that all three achieve similar
-performance: HtmlFlow using a blocking approach with Virtual Threads reaches
+performance: HtmlFlow using a blocking approach with virtual threads reaches
 4691 requests per second, while HtmlFlow using a reactive approach achieves
@@
 approaches: \textit{Blocking}, which uses platform threads with
-\texttt{StreamingResponseBody}, and \textit{\hl{Virtual}}, which uses Virtual
-Threads. Since Spring MVC follows a thread-per-request architecture, the
+\texttt{StreamingResponseBody}, and \textit{\hl{Virtual}}, which uses virtual
+threads. Since Spring MVC follows a thread-per-request architecture, the
 asynchronous approaches---\textbf{\hl{Reactive}} and \textbf{\hl{Suspendable}}---described in
@@
 \textit{\hl{Virtual}} strategies scale effectively up to 32 concurrent users, with
-the Virtual Threads approach achieving a slightly higher maximum throughput of
+the virtual threads approach achieving a slightly higher maximum throughput of
 2797 requests per second, compared to 2498 requests per second for the
@@
 engines scale up to 64 concurrent users, achieving up to 3744 requests per
-second. When using Virtual Threads, the throughput increases even further,
+second. When using virtual threads, the throughput increases even further,
 reaching 4856 requests per second, allowing scalability up to 128 users. This
-demonstrates that Quarkus's implementation of Virtual Threads is effective for
+demonstrates that Quarkus's implementation of virtual threads is effective for
 enabling PSSR\@, and comparable to the \textit{\hl{Suspendable}} and
@@
 remains largely unchanged; however, throughput is reduced across all engines.
-Compared to the Presentation benchmark, Jstachio using virtual threads
+Compared to the Presentation benchmark, JStachio using virtual threads
 experienced a more pronounced decrease in performance relative to the
-\textit{\hl{Reactive}} and \textit{\hl{Suspending}} approaches, with Jstachio using
-Virtual Threads now achieving 1509 requests per second, compared to 1750
+\textit{\hl{Reactive}} and \textit{\hl{Suspending}} approaches, with JStachio using
+virtual threads now achieving 1509 requests per second, compared to 1750
 requests per second achieved by the Thymeleaf implementation using the reactive
@@
      \caption{\hl{Throughput} %MDPI: The numbers in the picture are incomplete, please modify
- (requests per second) scalability results for Spring Webflux with \texttt{Stocks} class.}\label{fig:stocks-webflux-jmeter}
+ (requests per second) scalability results for Spring WebFlux with \texttt{Stocks} class.}\label{fig:stocks-webflux-jmeter}
 \end{figure}
@@
 specifically with HtmlFlow, we found that all three achieve similar
-performance: HtmlFlow using a blocking approach with Virtual Threads reaches
+performance: HtmlFlow using a blocking approach with virtual threads reaches
 3090 requests per second, while HtmlFlow using a reactive approach achieves
 3026 requests per second. This indicates that the more pronounced decrease in
-performance for Jstachio using Virtual Threads is likely due to the specific
-implementation of the template engine, rather than the use of Virtual Threads
+performance for JStachio using virtual threads is likely due to the specific
+implementation of the template engine, rather than the use of virtual threads
 itself.
@@
 \texttt{StreamingResponseBody} achieves a throughput of up to 1,916 requests
-per second, with no significant improvement observed when using Virtual
-Threads. Both approaches scale effectively up to 64 concurrent users. Although
+per second, with no significant improvement observed when using virtual
+threads. Both approaches scale effectively up to 64 concurrent users. Although
 these approaches achieve higher throughput in Spring MVC than in Spring
@@
 blocking approach reaches a throughput of 3019 requests per second, while the
-Virtual Threads approach achieves a throughput of 3357 requests per second. In
+virtual threads approach achieves a throughput of 3357 requests per second. In
 addition to the synchronous engines, the \textit{\hl{HtmlFlow-Rx}} approach also
@@
 The results of the benchmarks show that non-blocking engines---whether using
-reactive programming, Kotlin coroutines, or Java Virtual Threads---are able to
+reactive programming, Kotlin coroutines, or Java virtual threads---are able to
 scale effectively, supporting between 32 and 128 concurrent users depending on
-the approach and framework. Out of all the tested frameworks, Spring Webflux
+the approach and framework. Out of all the tested frameworks, Spring WebFlux
 showed itself the most effective at enabling PSSR, mostly due to its native
-support for publish-subscribe interfaces, allowing for content to be streamed
+support for publish--subscribe interfaces, allowing for content to be streamed
 as data becomes available, instead of when the response buffer is flushed.
@@
 of the \texttt{OutputBuffer} size to achieve the same results as Spring
-Webflux. The Spring MVC implementation, on the other hand, did not enable PSSR
+WebFlux. The Spring MVC implementation, on the other hand, did not enable PSSR
 for the tested templates.
@@
 \textls[-15]{This section evaluates the memory and CPU resource usage characteristics of
-structured concurrency approaches, specifically comparing Virtual Threads and
+structured concurrency approaches, specifically comparing virtual threads and
 suspendable~coroutines.}
 
-Resource utilization data was collected using VisualVM during 30-s
-benchmark runs, capturing detailed CPU and memory usage patterns under
+Resource utilization data was collected using VisualVM during 30 s benchmark runs, capturing detailed CPU and memory usage patterns under
 sustained load conditions. Each profiling session monitored system performance
@@
 450~MB in both graphs corresponds to the applications' bootstrapping phase.
-Overall, these results indicate that under the tested conditions, Virtual
-Threads and suspendable coroutines exhibit comparable memory~consumption.
-
-Virtual Threads allocate a full call stack per thread, which is unmounted and
+Overall, these results indicate that under the tested conditions, virtual
+threads and suspendable coroutines exhibit comparable memory~consumption.
+
+Virtual threads allocate a full call stack per thread, which is unmounted and
 stored on the heap when suspended (e.g., during blocking I/O). In contrast,
@@
      \includegraphics[width=0.48\textwidth]{./Graphs/cpu-susp.png}
-     \caption{CPU utilization profiling with HtmlFlow-Susp in Spring WebFlux during 30-s load~test.}     \label{fig:cpu-susp}
+     \caption{CPU utilization profiling with HtmlFlow-Susp in Spring WebFlux during 30 s load~test.}     \label{fig:cpu-susp}
 \end{figure}
@@
      \includegraphics[width=0.48\textwidth]{./Graphs/cpu-virt.png}
-     \caption{CPU utilization profiling with HtmlFlow-Virtual in Spring WebFlux during 30-s load~test.}   \label{fig:cpu-virt}
+     \caption{CPU utilization profiling with HtmlFlow-Virtual in Spring WebFlux during 30 s load~test.}   \label{fig:cpu-virt}
 \end{figure}
@@
     \includegraphics[width=0.48\textwidth]{./Graphs/mem-susp.png}
-     \caption{Memory utilization and garbage collection behavior with HtmlFlow-Susp in Spring WebFlux during 30-s load test.}             \label{fig:gc-susp}
+     \caption{Memory utilization and garbage collection behavior with HtmlFlow-Susp in Spring WebFlux during 30 s load test.}             \label{fig:gc-susp}
 \end{figure}
@@
    \includegraphics[width=0.48\textwidth]{./Graphs/mem-virt.png}
-     \caption{Memory utilization and garbage collection behavior with HtmlFlow-Virtual in Spring WebFlux during 30-s load test.} \label{fig:gc-virt}
+     \caption{Memory utilization and garbage collection behavior with HtmlFlow-Virtual in Spring WebFlux during 30 s load test.} \label{fig:gc-virt}
 \end{figure}
 
-\section{\highlighting{Discussion} %MDPI: To AE: This section seems to be added compared to the PDF file from susy, please check if it is correct.
+\section{{Discussion} 
 }\label{s6}
@@
 Beronić et al.~\cite{9803765} compared different structured concurrency
-constructs in Java and Kotlin in the context of a multithreaded HTTP server.
+constructs in Java and Kotlin in the context of a multi-threaded HTTP server.
 Their benchmark scenario differs from ours: their server awaited incoming
@@
 attributed to the different workloads: while their benchmark includes a single
-I/O read-write operation per request, our experiments focus exclusively on
+I/O read--write operation per request, our experiments focus exclusively on
 read-only I/O tasks.
@@
 
-Their results show that Quarkus with Virtual Threads outperformed the
+Their results show that Quarkus with virtual threads outperformed the
 traditional thread-per-request model under increased concurrency. Moreover,
@@
 The work of Šimatović et al.~\cite{vsimatovic2025evaluating} presents an
-evaluation of Java Virtual Threads under different garbage collectors. They
+evaluation of Java virtual threads under different garbage collectors. They
 explored three types of workloads, only one of which was based on a web
@@
 collectors, indicating reduced memory pressure. These findings suggest that
-Virtual Threads improve memory allocation efficiency and reduce the frequency
+virtual threads improve memory allocation efficiency and reduce the frequency
 of garbage collection cycles---particularly when used with concurrent garbage
@@
 
-Our observations suggest that the choice between Virtual Threads and
+Our observations suggest that the choice between virtual threads and
 suspendable coroutines should be guided by both runtime behavior and
-development constraints. Virtual Threads offer the compelling advantage of
+development constraints. Virtual threads offer the compelling advantage of
 preserving synchronous programming semantics while achieving competitive
@@
 
-As an alternative, Java Virtual Threads can be applied to any blocking I/O call,
+As an alternative, Java virtual threads can be applied to any blocking I/O call,
 leveraging the Java runtime to transparently intercept and convert it into non-blocking I/O,

### 2025-08-06 docs/article/software-3702556.tex
Fix most annotated reviews.

@@
 % Authors, for the paper (add full first names)
-\Author{\hl{Bernardo Pereira} %MDPI: 1. Please carefully check the accuracy of names and affiliations. 
- \orcidA{} and \hl{Fernando Miguel Carvalho} %MDPI: The names highlighted are different from the ones submitted online at susy.mdpi.com. Please confirm which are correct. 
+\Author{Bernardo Pereira
+ \orcidA{} and Fernando Miguel Carvalho
  *\orcidB{}}
@@
 % MDPI internal command: Authors, for metadata in PDF
-\AuthorNames{Bernardo Pereira, Miguel Carvalho}
+\AuthorNames{Bernardo Pereira, Fernando Miguel Carvalho}
 
@@
          }{
-         \AuthorCitation{\hl{Pereira}%MDPI: Please carefully check the accuracy of names
-, B.; Carvalho, F.M.}
+         \AuthorCitation{Pereira, B.; Carvalho, F.M.}
          }
@@
 % Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
-\address[1]{Polytechnical Institute of Lisbon, \hl{ISEL}%MDPI: Please add full name if possible
-, \hl{Lisbon}%MDPI: please add post code
-, Portugal; \hl{a50493@alunos.isel.pt} %MDPI: We added these email addresses here according to those submitted online at susy.mdpi.com. Please confirm.
+\address[1]{Polytechnical Institute of Lisbon, ISEL
+, Lisbon 1959-007
+, Portugal; a50493@alunos.isel.pt
 }
@@
 Modern web applications rely on different rendering strategies to optimize
-performance, user experience, and \textit{\hl{scalability}%MDPI: Please confirm if the italics are necessary; if not, please remove them. The following highlights are the same.
-}. The two most dominant
-approaches are \textit{\hl{Server-Side Rendering (SSR)}} and \textit{\hl{Client-Side Rendering (CSR)}}.
-SSR generates HTML content on the server before sending it to the client, resulting in a faster \textit{\hl{First Contentful Paint (FCP)}}~\cite{Edgar2024-FCP} and better Search Engine Optimization (SEO).
+performance, user experience, and scalability. The two most dominant
+approaches are \textit{Server-Side Rendering (SSR)} and \textit{Client-Side Rendering (CSR)}.
+SSR generates HTML content on the server before sending it to the client, resulting in a faster \textit{First Contentful Paint (FCP)}~\cite{Edgar2024-FCP} and better Search Engine Optimization (SEO).
 Nevertheless, SSR
-can increase server load and reduce \textit{\hl{throughput}} ({\hl{The} %MDPI: footnote is not allowed in this journal, so we moved it to here, please confirm it
- number of
-    requests the server can handle per second (RPS)}) since each request requires
+can increase server load and reduce \textit{throughput}
+(the number of requests the server can handle per second (RPS)) since each request requires
 additional processing before responding. In contrast, CSR shifts the rendering
@@
 JavaScript, which dynamically loads the page content. While CSR reduces the
-server’s burden, it can lead to a slower \textit{\hl{FCP}}, as users must wait for
+server’s burden, it can lead to a slower FCP, as users must wait for
 JavaScript execution before meaningful content appears.
 
-\textit{\hl{Progressive Server-Side Rendering (PSSR)}} combines benefits from both SSR and CSR
+\textit{Progressive Server-Side Rendering (PSSR)} combines benefits from both SSR and CSR
 by streaming HTML content progressively. This technique enhances user-perceived performance by
 allowing progressive rendering as data becomes available, significantly reducing the
-\textit{\hl{time-to-first-byte (TTFB)}} and improving perceived load times compared to traditional
+\textit{time-to-first-byte (TTFB)} and improving perceived load times compared to traditional
 SSR approaches~\cite{wise2024pssr}. In this respect,
@@
 
-\textit{\hl{Low-thread}} servers, also known as
-\textit{\hl{event-driven}}~\cite{event-driven-servers}, have gained prominence in
+\textit{Low-thread} servers, also known as
+\textit{event-driven}~\cite{event-driven-servers}, have gained prominence in
 contemporary web applications due to their ability to efficiently manage a large
@@
 
-With the introduction of \textit{\hl{virtual
-    threads}}~({\url{https://openjdk.org/jeps/444}}, \hl{accessed on}%MDPI: footnote is not allowed in this journal, so we moved it to here, please confirm it, and please add the access date (format: Date Month Year), e.g., accessed on 1 January 2020.
-) in Java 21, it is now
+With the introduction of \textit{virtual
+    threads}~({\url{https://openjdk.org/jeps/444}}, accessed on 6 August 2025) in Java 21, it is now
 possible to execute blocking I/O operations in a scalable, lightweight manner.
@@
 We investigate the current landscape of non-blocking PSSR, focusing on two
-primary paradigms: \textit{\hl{reactive programming}} and \textit{\hl{coroutines}}, both
+primary paradigms: \textit{reactive programming} and \textit{coroutines}, both
 of which have been used to achieve asynchronous I/O in the Java ecosystem. As
@@
 
-\textit{\hl{Web templates}} have been the most widely adopted approach for
+\textit{Web templates} have been the most widely adopted approach for
 constructing dynamic HTML pages.
-Web templates or \textit{\hl{web views}}~\cite{Fowler02,Alur01} (e.g., JSP, Handlebars,
+Web templates or \textit{web views}~\cite{Fowler02,Alur01} (e.g., JSP, Handlebars,
 or Thymeleaf), are based on HTML documents augmented with template-specific
 markers (e.g., \texttt{<\%>}, \texttt{\{\{\}\}}, or \texttt{\$\{\}}), which
-represent \textit{\hl{dynamic}} information to be replaced at runtime with the
+represent \textit{dynamic} information to be replaced at runtime with the
 results of corresponding computations, producing the final HTML page.
 The process of parsing and replacing these markers---i.e.,
-\textit{\hl{resolution}}---is the primary responsibility of the \textit{\hl{template
-  engine}}~\cite{Parr04}.
-One key characteristic of \textit{\hl{web templates}} is their ability to receive a
+\textit{resolution}---is the primary responsibility of the \textit{template
+  engine}~\cite{Parr04}.
+One key characteristic of \textit{web templates} is their ability to receive a
 \textit{context object}---equivalent to the \textit{model} in the model-view
@@
 \begin{enumerate}
-  \item \textit{\hl{Domain-specific language}} idiom
+  \item \textit{Domain-specific language} idiom
   \item Supported data model APIs
@@
 Although some of the aforementioned characteristics apply to both
-\textit{\hl{server-side}} and \textit{\hl{client-side}} approaches, we focus solely on
-web template technologies for \textit{\hl{Server-Side Rendering}}, as our work is
+\textit{server-side} and \textit{client-side} approaches, we focus solely on
+web template technologies for \textit{Server-Side Rendering}, as our work is
 centered on that approach. Before exploring each of the aforementioned
@@
     \midrule
-    \textbf{\hl{Freemarker}}
+    \textbf{Freemarker}
      & External DSL
@@
 The first half of Table~\ref{table:cmplibs} lists template engines that use
-their own templating dialects, referred to as \textit{\hl{external DSLs}}. The
-second half lists Java libraries that provide an \textit{\hl{internal DSL}},
-typically using a \textit{\hl{nested}} or \textit{\hl{chaining}} style to build HTML.
+their own templating dialects, referred to as \textit{external DSLs}. The
+second half lists Java libraries that provide an \textit{internal DSL},
+typically using a \textit{nested} or \textit{chaining} style to build HTML.
 Note that, by leveraging the host language (e.g., Clojure, Groovy, Java,
@@
 the data model and fully support all styles of data access APIs. They also
-benefit from static type checking, which helps ensure \textit{\hl{type safety}}.
+benefit from static type checking, which helps ensure \textit{type safety}.
 
@@
 
-Web templates are based on a \textit{\hl{domain-specific language}}
+Web templates are based on a \textit{domain-specific language}
 (DSL)~\cite{landin1966next}, which defines a language tailored to a specific
-\textit{\hl{domain}}~\cite{evans2004domain}---in this case, HTML for expressing web
+\textit{domain}~\cite{evans2004domain}---in this case, HTML for expressing web
 documents. The DSL constrains the template's syntax and semantics to match the
 structure and purpose of HTML.
-DSLs can be divided in two types: \textit{\hl{external}} or
-\textit{\hl{internal}}~\cite{dslbook}. \textit{\hl{External}} DSLs are languages created
+DSLs can be divided in two types: \textit{external} or
+\textit{internal}~\cite{Fowler03}. \textit{External} DSLs are languages created
 without any affiliation to a concrete programming language. An example of an
-\textit{\hl{external}} DSL is the regular expressions search
+\textit{external} DSL is the regular expressions search
 pattern~\cite{thompson1968}, since it defines its own syntax without any
-dependency of programming languages. On the other hand, an \textit{\hl{internal}} DSL
+dependency of programming languages. On the other hand, an \textit{internal} DSL
 is defined within a host programming language as a library and tends to be
@@
 
-On the other hand, \textit{\hl{internal}} DSLs for HTML allow templates to be
-defined directly within the \emph{\hl{host}} language (such as Java, Kotlin, Groovy,
+On the other hand, \textit{internal} DSLs for HTML allow templates to be
+defined directly within the \emph{host} language (such as Java, Kotlin, Groovy,
 Scala, or other general-purpose programming languages), rather than using
@@
 \begin{enumerate}
-  \item \emph{\hl{Type safety}}: Because the templates are defined with the host
+  \item \emph{Type safety}: Because the templates are defined with the host
         programming language, the compiler can check the syntax and types of the
@@
 
-  \item \emph{\hl{IDE support}}: Many modern IDEs provide code completion, syntax
+  \item \emph{IDE support}: Many modern IDEs provide code completion, syntax
         highlighting, and other features, which can make it easier to write and
@@
 
-  \item \emph{\hl{Flexibility}}: Use all the features of the host programming language
+  \item \emph{Flexibility}: Use all the features of the host programming language
         to generate HTML, can make it easier to write complex templates and reuse code.
 
-  \item \emph{\hl{Integration}}: Because the templates are defined in Java code, for
+  \item \emph{Integration}: Because the templates are defined in Java code, for
         example, you can easily integrate them with other Java code in your
@@
 be combined in a chain of calls to mimic the construction of an HTML document
-in a fluent manner. Martin Fowler~\cite{dslbook} identifies three different
-patterns for combining functions to create a DSL: (1) \textit{\hl{function
-  sequence}}; (2) \textit{\hl{nested function}}, and (3) \textit{\hl{method chaining}}, which
+in a fluent manner. Martin Fowler~\cite{Fowler03} identifies three different
+patterns for combining functions to create a DSL: (1) \textit{function
+  sequence}; (2) \textit{nested function}, and (3) \textit{method chaining}, which
 are illustrated in the snippets of Figure~\ref{fig:dsl-idioms}.
@@
 From the three examples depicted in Figure~\ref{fig:dsl-idioms}, the
-\textit{\hl{nested function}} idiom used by \textbf{\hl{ScalaTags}%MDPI: Please confirm if the bold formatting is necessary; if not, please remove it. The following highlights are the same.
-}, as shown in the
+\textit{nested function} idiom used by \textbf{ScalaTags}, as shown in the
 snippet of Figure~\ref{fig:dsl-idioms}b, is the least verbose, requiring
 fewer statements than the other two cases. This approach of combining function
-calls is also utilized by \textbf{\hl{Hiccup}} Clojure Library and \textbf{\hl{j2html}}
-Java library. One reason for verbosity in \textit{\hl{function sequence}} and
-\textit{\hl{method chaining}} approaches is the requirement of a dedicated HTML
+calls is also utilized by \textbf{Hiccup} Clojure Library and \textbf{j2html}
+Java library. One reason for verbosity in \textit{function sequence} and
+\textit{method chaining} approaches is the requirement of a dedicated HTML
 builder to emit the closing tag, exemplified by \texttt{end()} in \textbf{JATL}
-(Figure~\ref{fig:dsl-idioms}a) and \texttt{\_\_()} in \textbf{\hl{HtmlFlow}}
-(Figure~\ref{fig:dsl-idioms}c). However, the \textit{\hl{nested function}} approach
-comes with a notable drawback: \textbf{\hl{it does not support PSSR}} because the
+(Figure~\ref{fig:dsl-idioms}a) and \texttt{\_\_()} in \textbf{HtmlFlow}
+(Figure~\ref{fig:dsl-idioms}c). However, the \textit{nested function} approach
+comes with a notable drawback: \textbf{it does not support PSSR} because the
 sequence of nested functions is evaluated backward to the order in which they
@@
 
-Two other JVM libraries, \textbf{\hl{Groovy MarkupBuilder}} and
-\textbf{\hl{KotlinX.html}}, also adopt a \textit{\hl{nested function}} approach, but they
-address the backward evaluation issue by implementing \textit{\hl{lazy evaluation}}
+Two other JVM libraries, \textbf{Groovy MarkupBuilder} and
+\textbf{KotlinX.html}, also adopt a \textit{nested function} approach, but they
+address the backward evaluation issue by implementing \textit{lazy evaluation}
 of arguments~\cite{Landin65}, expressed in lambda expressions (also known as
@@
 only requires replacing parentheses with brackets for parent elements.
-\textbf{\hl{HtmlFlow}} also adopts this approach and provides a Kotlin-idiomatic API
+\textbf{HtmlFlow} also adopts this approach and provides a Kotlin-idiomatic API
 as an alternative to its original Java API.
@@
 abstractions like the \texttt{Sequence} interface---also incompatible with
-\texttt{Iterable}. Consequently, template engines based on \textit{\hl{external}}
+\texttt{Iterable}. Consequently, template engines based on \textit{external}
 DSLs must explicitly support each of these interfaces to accommodate the
@@
 
-On the other hand, template engines that employ an \textit{\hl{internal}} DSL do not
+On the other hand, template engines that employ an \textit{internal} DSL do not
 face such limitations, as they can leverage any available API within the host
@@
 language and runtime environment. Some of these approaches include
-\emph{\hl{continuation-passing style}} (CPS)~\cite{scheme},
-\textit{\hl{promises}}~\cite{promise}, async/await idiom~\cite{async_await},
-or \textit{\hl{suspend functions}}~\cite{elizarov2021coroutines}.
-
-Many \textit{\hl{general-purpose languages}} (GPLs) have embraced the
+\emph{continuation-passing style} (CPS)~\cite{scheme},
+\textit{promises}~\cite{promise}, async/await idiom~\cite{async_await},
+or \textit{suspend functions}~\cite{elizarov2021coroutines}.
+
+Many \textit{general-purpose languages} (GPLs) have embraced the
 \texttt{async}/\texttt{await} feature~\cite{async_await} enabling non-blocking
@@
 of this programming model have led to its incorporation into mainstream
-languages like \hl{C\#, JavaScript, Python, Perl, Swift, Kotlin} %MDPI: Please confirm whether this is the software used in this article. If so, please provide the version number.
-, and others,
+languages like C\#, JavaScript, Python, Perl, Swift, Kotlin, and others,
 excluding Java. However, implementing \texttt{async}/\texttt{a}wait requires
-compiler support to translate \textit{\hl{suspension points}} (i.e., \texttt{await}
+compiler support to translate \textit{suspension points} (i.e., \texttt{await}
 statements) into state machines. Most template engines operate using an
@@
 An alternative approach to asynchronous programming involves handling data as
-a sequence of asynchronous events, commonly known as a \textit{\hl{reactive stream}}.
+a sequence of asynchronous events, commonly known as a \textit{reactive stream}.
 Instead of materializing all items in memory, reactive streams emit values over
@@
 engineers from notable companies such as Lightbend, Pivotal, and Netflix.
-The group began defining the \textit{\hl{reactive streams}}
+The group began defining the \textit{reactive streams}
 specification~\cite{ReactiveStreams}, which established a standard for
@@
 The more recent Mutiny~\cite{mutiny2021}, while conforming to the
-\textit{\hl{reactive streams}} standard, proposes a different approach that avoids
+\textit{reactive streams} standard, proposes a different approach that avoids
 complex optimizations, resulting in more straightforward and easier-to-maintain
@@
 
-Another key characteristic of such DSLs is \textit{\hl{HTML safety}}, which refers
+Another key characteristic of such DSLs is \textit{HTML safety}, which refers
 to whether they produce only valid HTML conforming to a well-formed document.
 To ensure HTML safety, a DSL API should only allow combining calls to builders
-that result in valid HTML. Nonetheless, most HTML DSLs using the \textit{\hl{function
-  sequence}} or \textit{\hl{nested function}} approach cannot ensure HTML safety at
+that result in valid HTML. Nonetheless, most HTML DSLs using the \textit{function
+  sequence} or \textit{nested function} approach cannot ensure HTML safety at
 compile time.
 
-The \textit{\hl{function sequence}} approach, as illustrated in the snippet of
+The \textit{function sequence} approach, as illustrated in the snippet of
 Figure~\ref{fig:dsl-idioms}a, involves combining function calls as a sequence
 of statements, making it challenging to restrict the order of statements. In
-the \textit{\hl{nested function}} approach shown in Figure~\ref{fig:dsl-idioms}b,
+the \textit{nested function} approach shown in Figure~\ref{fig:dsl-idioms}b,
 variable-length arguments are often used to allow an undefined number of child
@@
 
-KotlinX.html, which utilizes the \textit{\hl{nested function}} idiom with lazy
+KotlinX.html, which utilizes the \textit{nested function} idiom with lazy
 evaluation, mitigates this issue through function types with a receiver. In
@@
 
-To achieve this, KotlinX.html provides HTML builders using \textit{\hl{function
-  literals with receiver}}~\cite{kotlinlang}. In Kotlin, a block of code enclosed
-in curly braces \texttt{\{...\}} is known as a \emph{\hl{lambda}}, and can be used
-as an argument to a function that expects a \emph{\hl{function literal}}. When we
+To achieve this, KotlinX.html provides HTML builders using \textit{function
+  literals with receiver}~\cite{kotlinlang}. In Kotlin, a block of code enclosed
+in curly braces \texttt{\{...\}} is known as a \emph{lambda}, and can be used
+as an argument to a function that expects a \emph{function literal}. When we
 write, for example, \texttt{body \{ div \{ hr() \} \}}, we are invoking the
@@
 HtmlFlow provides two APIs: one in idiomatic Kotlin, similar to the
-\texttt{KotlinX.html} API, and another that employs the \textit{\hl{method
-  chaining}} idiom, as illustrated in the snippet of
+\texttt{KotlinX.html} API, and another that employs the \textit{method
+  chaining} idiom, as illustrated in the snippet of
 Figure~\ref{fig:dsl-idioms}c. In this approach, the receiver object is
@@
 have been explored to deliver the initial meaningful content to the end-user as
-promptly as possible, while the remaining content \textit{\hl{progressively}} (or
+promptly as possible, while the remaining content \textit{progressively} (or
 incrementally) loads as the server streams the HTML~content.
 
-\textit{\textbf{\hl{Progressive rendering}}} and \textit{\textbf{\hl{progressive loading}}}
-encompass different concepts. The former pertains to the \textit{\hl{dynamic}} content of a
-\textit{\hl{dynamic web page}}, encompassing elements with logic and placeholders
-that are fulfilled by data from an object \textit{\hl{model}} constructed at runtime.
-On the other hand, the latter is associated with \textit{\hl{render-blocking
-  resources}} such as scripts, stylesheets, and HTML imports, which may hinder the
+\textit{\textbf{Progressive rendering}} and \textit{\textbf{progressive loading}}
+encompass different concepts. The former pertains to the \textit{dynamic} content of a
+\textit{dynamic web page}, encompassing elements with logic and placeholders
+that are fulfilled by data from an object \textit{model} constructed at runtime.
+On the other hand, the latter is associated with \textit{render-blocking
+  resources} such as scripts, stylesheets, and HTML imports, which may hinder the
 browser from rendering page content to the screen~\cite{progressive-2022}.
-The notion of \textit{\hl{progressive rendering}} has typically been aligned with
-\textit{\hl{Client-Side Rendering}} (CSR), where a single HTML page with static
+The notion of \textit{progressive rendering} has typically been aligned with
+\textit{Client-Side Rendering} (CSR), where a single HTML page with static
 content is delivered upfront, while the dynamic content is fetched as the data
@@
 which offers the advantage of not being dependent on
-\textit{\hl{render-blocking resources}}, such as the required JavaScript for CSR.
+\textit{render-blocking resources}, such as the required JavaScript for CSR.
 % \textit{Progressive server-side rendering} (PSSR) is the
@@
 \texttt{Publisher} API. Nonetheless, the main counter-argument is the non-trivial
-management of the \emph{\hl{resume}} callback in
+management of the \emph{resume} callback in
 continuations~\cite{von2003events,callbackhell}, which is used to linearize the
@@
 reports such as the Stack Overflow 2024 Developer
-Survey~({\url{https://survey.stackoverflow.co/2024/technology\#1-web-frameworks-and-technologies}}, \hl{accessed on}).
-
-On the other hand, in modern \textit{\hl{low-thread}} architectures, the server uses
+Survey~({\url{https://survey.stackoverflow.co/2024/technology\#1-web-frameworks-and-technologies}}, accessed on 6 August 2025).
+
+On the other hand, in modern \textit{low-thread} architectures, the server uses
 a small number of threads to handle a large number of requests.
-\textit{\hl{Low-thread}} servers, also known as
-\textit{\hl{event-driven}}~\cite{event-driven-servers}, offer a significant
+\textit{Low-thread} servers, also known as
+\textit{event-driven}~\cite{event-driven-servers}, offer a significant
 advantage in efficiently managing a high number of concurrent I/O operations
@@
 
-The \textit{\hl{non-blocking}} I/O model employed in low-thread servers is
+The \textit{non-blocking} I/O model employed in low-thread servers is
 well-suited for handling large volumes of data asynchronously~\cite{Meijer12}.
@@
 % community metrics from platforms like Github and StackOverflow.
-The \textit{\hl{non-blocking}} I/O model in low-thread servers functions optimally
+The \textit{non-blocking} I/O model in low-thread servers functions optimally
 only when HTTP handlers avoid blocking. Therefore, HTML templates need to be
@@
 abstraction for managing coroutines and provides structured concurrency, which
-ensures that coroutines are \textit{\hl{scoped}}, \textit{\hl{cancellable}}, and
-\textit{\hl{coordinated}} within a well-defined
+ensures that coroutines are \textit{scoped}, \textit{cancellable}, and
+\textit{coordinated} within a well-defined
 lifecycle~\cite{elizarov2021coroutines}. Although this model embraces the
@@
 Only Java virtual threads, introduced in
-JEP~444 ({\url{https://openjdk.org/jeps/444}}, \hl{accessed on}), adopt a similar approach
+JEP~444 ({\url{https://openjdk.org/jeps/444}}, accessed on 6 August 2025), adopt a similar approach
 to Karsten's proposal, using user-mode threads to preserve a synchronous
@@
 mechanism for response streaming via the
-\texttt{StreamingResponseBody} ({\url{https://docs.spring.io/spring-framework/docs/current/javadoc-api/org/springframework/web/servlet/mvc/method/annotation/StreamingResponseBody.html}}, \hl{accessed on})
+\texttt{StreamingResponseBody} ({\url{https://docs.spring.io/spring-framework/docs/current/javadoc-api/org/springframework/web/servlet/mvc/method/annotation/StreamingResponseBody.html}}, accessed on 6 August 2025)
 interface, introduced in Spring 4.2. When a controller method returns a
@@
 
-In this section, we examine the challenges of implementing \textit{\hl{Progressive
-Server-Side Rendering}} (PSSR) in modern web applications, with a focus on the
+In this section, we examine the challenges of implementing \textit{Progressive
+Server-Side Rendering} (PSSR) in modern web applications, with a focus on the
 limitations of current template engine designs. Our goal is to broaden the
@@
 HtmlFlow pauses the rendering process until \texttt{onCompletion} is called,
-similar to how the \emph{\hl{resume}} function works in continuations and
+similar to how the \emph{resume} function works in continuations and
 coroutines~\cite{coroutines_continuations}.
@@
 \begin{listing}[H]
-\caption{\textit{\hl{HtmlFlow reactive}} presentation template in Kotlin with an \texttt{Observable} model.}
+\caption{\textit{HtmlFlow reactive} presentation template in Kotlin with an \texttt{Observable} model.}
 \begin{lstlisting}[
@@
 \begin{listing}[H]
-\caption{\textit{\hl{HtmlFlow suspend}} presentation template in Kotlin with a \texttt{Flow} model.}
+\caption{\textit{HtmlFlow suspend} presentation template in Kotlin with a \texttt{Flow} model.}
 \begin{lstlisting}[
@@
 
-By contrast, template engines that use \textit{\hl{external}} DSLs---such as JStachio,
+By contrast, template engines that use \textit{external} DSLs---such as JStachio,
 Thymeleaf, or Handlebars---typically define templates within static HTML
@@
 \begin{listing}[H]
-\caption{Presentation HTML template using \textit{\hl{JStachio}}.}
+\caption{Presentation HTML template using \textit{JStachio}.}
 \begin{lstlisting}[
@@
 \begin{enumerate}
-    \item \emph{\hl{Separation of Concerns}}: HTML templates are decoupled from application logic, enabling front-end developers to contribute without modifying back-end code.
-    \item \emph{\hl{Cross-Language Compatibility}}: External DSLs are portable across languages and frameworks, easing integration in multi-language environments.
-    \item \emph{\hl{Familiarity}}: Many developers are comfortable with HTML syntax, lowering the barrier to entry and improving maintainability.
+    \item \emph{Separation of Concerns}: HTML templates are decoupled from application logic, enabling front-end developers to contribute without modifying back-end code.
+    \item \emph{Cross-Language Compatibility}: External DSLs are portable across languages and frameworks, easing integration in multi-language environments.
+    \item \emph{Familiarity}: Many developers are comfortable with HTML syntax, lowering the barrier to entry and improving maintainability.
 \end{enumerate}
@@
 high concurrency, where blocking threads severely degrades throughput~\cite{wise2024pssr}.
-Emerging features in the Java ecosystem, particularly \textit{\hl{virtual threads}}
+Emerging features in the Java ecosystem, particularly \textit{virtual threads}
 introduced in Java 21 as part of Project Loom~\cite{Veen2024}, offer a promising solution to
@@
 
-This benchmark ({\url{https://github.com/xmlet/comparing-non-blocking-progressive-ssr}}, \hl{accessed on})
-is designed with a modular architecture, separating the \textit{\hl{view}} and
-\textit{\hl{model}} layers from the \textit{\hl{controller}} layer ~\cite{Bucanek2009},
+This benchmark ({\url{https://github.com/xmlet/comparing-non-blocking-progressive-ssr}}, accessed on 6 August 2025)
+is designed with a modular architecture, separating the \textit{view} and
+\textit{model} layers from the \textit{controller} layer ~\cite{Bucanek2009},
 which allows for easy extension and integration of new template engines and
 frameworks. It also includes a set of tests to ensure the correctness of
-implementations and to validate the \textit{\hl{HTML}} output. It includes two
+implementations and to validate the HTML output. It includes two
 different data models, defined as \texttt{Presentation} and \texttt{Stock}, as
@@
 \begin{listing}[H]
-\caption{\texttt{\hl{Stock}} class.}
+\caption{\texttt{Stock} class.}
             \begin{lstlisting}[
@@
 \begin{listing}[H]
-\caption{\texttt{\hl{Presentation}} class.}
+\caption{\texttt{Presentation} class.}
             \begin{lstlisting}[
@@
 \begin{itemize}
-      \item \textbf{\hl{Reactive}}: The template engine is used in a reactive context,
+      \item \textbf{Reactive}: The template engine is used in a reactive context,
             where the HTML content rendered using the reactive programming model.
@@
             non-blocking Spring \texttt{ViewResolver}.
-      \item \textbf{\hl{Suspendable}}: The template engine is used in a suspendable
+      \item \textbf{Suspendable}: The template engine is used in a suspendable
             context, where the HTML content is rendered within the context of a suspending function.
@@
             Kotlin standard library.
-      \item \textbf{\hl{Virtual}}: The template engine is used in a non-blocking
+      \item \textbf{Virtual}: The template engine is used in a non-blocking
             context, where the HTML content is rendered within the context of virtual
@@
             because they do not support the asynchronous rendering of HTML~content.
-      \item \textbf{\hl{Blocking}}: The template engine is used in a blocking
+      \item \textbf{Blocking}: The template engine is used in a blocking
             context, where the HTML content is rendered using the blocking interface of the
-            \texttt{\hl{Observable}} class, within the context of an OS-thread.
+            \texttt{Observable} class, within the context of an OS-thread.
 \end{itemize}
@@
 blocking the servlet container thread.
-According to the Spring documentation ({\url{https://docs.spring.io/spring-framework/docs/current/javadoc-api/org/springframework/web/servlet/mvc/method/annotation/StreamingResponseBody.html}}, \hl{accessed on}),
-this class is \textit{\hl{a controller method return value type for asynchronous
+According to the Spring documentation ({\url{https://docs.spring.io/spring-framework/docs/current/javadoc-api/org/springframework/web/servlet/mvc/method/annotation/StreamingResponseBody.html}}, accessed on 6 August 2025),
+this class is \textit{a controller method return value type for asynchronous
 request processing where the application can write directly to the response
-OutputStream without holding up the Servlet container thread.}}
+OutputStream without holding up the Servlet container thread.}
 
@@
 \begin{itemize}
-      \item \textbf{\hl{Blocking}}: The template engine is used in a blocking context,
+      \item \textbf{Blocking}: The template engine is used in a blocking context,
             where the HTML content is rendered using the blocking interface of the
             \texttt{Observable} class.
-      \item \textbf{\hl{Virtual}}: The template engine is used in a non-blocking context,
+      \item \textbf{Virtual}: The template engine is used in a non-blocking context,
             where the HTML content is rendered within the context of virtual threads.
@@
 The Quarkus implementation supports PSSR for these templates by configuring the
-response buffer size in the \textit{\hl{application.properties}} file. The default
+response buffer size in the \textit{application.properties} file. The default
 buffer size is 8 KB, but we reduced it to 512 bytes, which allows the response
@@
 \begin{itemize}
-      \item \textbf{\hl{Blocking}}: The template engine is used in a blocking context,
+      \item \textbf{Blocking}: The template engine is used in a blocking context,
             where the HTML content is rendered using the blocking interface of the
             \texttt{Observable} class.
-      \item \textbf{\hl{Virtual}}: The template engine is used in a non-blocking context,
+      \item \textbf{Virtual}: The template engine is used in a non-blocking context,
             where the HTML content is rendered within the context of virtual threads.
-      \item \textbf{\hl{Reactive}}: The template engine is used in a reactive context,
+      \item \textbf{Reactive}: The template engine is used in a reactive context,
             where the HTML content is rendered using the reactive programming model.
@@
 Network connectivity is provided through a 1500 MTU Ethernet interface. All
-tests were conducted on \hl{OpenJDK 21} (Corretto build) with the G1 garbage
+tests were conducted on OpenJDK 21 (Corretto build) with the G1 garbage
 collector enabled by default. The JVM arguments used were: \texttt{-Xms512m
@@
 concurrent users ranging from 1 to 128, from left to right. The benchmarks
-include HtmlFlow using suspendable web templates (\textit{\hl{HtmlFlow-Susp}},
+include HtmlFlow using suspendable web templates (\textit{HtmlFlow-Susp},
 equivalent to the approach shown in Listing~\ref{lst:presentation-flow}),
 JStachio using virtual threads with the \texttt{Iterable} interface
-(\textit{\hl{JStachio-Virtual}}), and Thymeleaf using the reactive View Resolver
-driver (Thymeleaf-Rx). \textit{\hl{Blocking}} and \textit{\hl{Virtual}} represent the
+(\textit{JStachio-Virtual}), and Thymeleaf using the reactive View Resolver
+driver (Thymeleaf-Rx). \textit{Blocking} and \textit{Virtual} represent the
 average throughput of the blocking approaches (i.e., KotlinX, Rocker, JStachio,
@@
 
-We show the \textit{\hl{HtmlFlow-Susp}}, \textit{\hl{JStachio-Virtual}}, and
+We show the \textit{HtmlFlow-Susp}, \textit{JStachio-Virtual}, and
 \textit{Thymeleaf-Rx} engines separately to observe the performance of the
 non-blocking engines when using the Suspending, Virtual Thread, and Reactive
-approaches. The \textit{\hl{Blocking}} and \textit{\hl{Virtual}} are aggregated due to
+approaches. The \textit{Blocking} and \textit{Virtual} are aggregated due to
 the similar performance of different engines when using those approaches.
@@
 engines are unable to scale effectively beyond four concurrent users. In contrast,
-\textit{\hl{HtmlFlow-Susp}} scales up to 128 concurrent users, achieving \hl{4487} %MDPI: Commas are only used for numbers with five or more digits.
-We have removed them in four-digit numbers.
-Please confirm.
+\textit{HtmlFlow-Susp} scales up to 128 concurrent users, achieving 4487 %MDPI: Commas are only used for numbers with five or more digits. We have removed them in four-digit numbers. Please confirm.
 requests per second. When blocking approaches are executed in the context of
 virtual threads---thus enabling non-blocking I/O---the engines scale up to 64
-concurrent users, with JStachio using virtual threads reaching \hl{3514} requests
+concurrent users, with JStachio using virtual threads reaching 3514 requests
 per second. The Thymeleaf implementation using the reactive View Resolver
 driver scales up to 32 concurrent users, achieving a lower maximum throughput
-of \hl{2559} requests per second.
+of 2559 requests per second.
 
 It is important to note that the differences in scalability and throughput
-between the \textit{\hl{HtmlFlow-Susp}}, \textit{\hl{JStachio-Virtual}}, and
-\textit{\hl{Thymeleaf-Rx}} approaches may be influenced by the specific template
+between the \textit{HtmlFlow-Susp}, \textit{JStachio-Virtual}, and
+\textit{Thymeleaf-Rx} approaches may be influenced by the specific template
 engines used, rather than the approach itself. When comparing the
-\textit{\hl{Reactive}}, \textit{\hl{Suspendable}}, and \textit{\hl{Virtual}} approaches
+\textit{Reactive}, \textit{Suspendable}, and \textit{Virtual} approaches
 specifically with HtmlFlow, we found that all three achieve similar
@@
 approaches: \textit{Blocking}, which uses platform threads with
-\texttt{StreamingResponseBody}, and \textit{\hl{Virtual}}, which uses virtual
+\texttt{StreamingResponseBody}, and \textit{Virtual}, which uses virtual
 threads. Since Spring MVC follows a thread-per-request architecture, the
-asynchronous approaches---\textbf{\hl{Reactive}} and \textbf{\hl{Suspendable}}---described in
-Section~\ref{sec:bench} are not applicable. Both the \textit{\hl{Blocking}} and
-\textit{\hl{Virtual}} strategies scale effectively up to 32 concurrent users, with
+asynchronous approaches---\textbf{Reactive} and \textbf{Suspendable}---described in
+Section~\ref{sec:bench} are not applicable. Both the \textit{Blocking} and
+\textit{Virtual} strategies scale effectively up to 32 concurrent users, with
 the virtual threads approach achieving a slightly higher maximum throughput of
@@
 demonstrates that Quarkus's implementation of virtual threads is effective for
-enabling PSSR\@, and comparable to the \textit{\hl{Suspendable}} and
-\textit{\hl{Reactive}} approaches used in Spring WebFlux in terms of scalability and
+enabling PSSR\@, and comparable to the \textit{Suspendable} and
+\textit{Reactive} approaches used in Spring WebFlux in terms of scalability and
 throughput.
@@
      \includegraphics[width=0.8\textwidth]{./Graphs/presentations-quarkus-jmeter.png}
-     \caption{\hl{Throughput} %MDPI: The numbers in the picture are incomplete, please modify
+     \caption{Throughput %MDPI: The numbers in the picture are incomplete, please modify
  (requests per second) scalability results for Quarkus with \texttt{Presentations} class.}\label{fig:presentations-quarkus-jmeter}
@@
 
-Additionally, \textit{\hl{HtmlFlow-Rx}}, a reactive implementation of the HtmlFlow
+Additionally, \textit{HtmlFlow-Rx}, a reactive implementation of the HtmlFlow
 template engine (equivalent to the approach shown in
 Listing~\ref{lst:presentation-observable}) that utilizes Quarkus's reactive
-programming model, achieved a lower throughput than the the \textit{\hl{Blocking}}
-and \textit{\hl{Virtual}} approaches---3088 requests per second. This demonstrates
+programming model, achieved a lower throughput than the the \textit{Blocking}
+and \textit{Virtual} approaches---3088 requests per second. This demonstrates
 that Quarkus's reactive programming model is effective for enabling PSSR\@,
@@
 experienced a more pronounced decrease in performance relative to the
-\textit{\hl{Reactive}} and \textit{\hl{Suspending}} approaches, with JStachio using
+\textit{Reactive} and \textit{Suspending} approaches, with JStachio using
 virtual threads now achieving 1509 requests per second, compared to 1750
@@
 It is again important to note that the differences in scalability and
-throughput between the \textit{\hl{HtmlFlow-Susp}}, \textit{J\hl{stachio-Virtual}}, and
-\textit{\hl{Thymeleaf-Rx}} approaches may be influenced by the specific template
+throughput between the \textit{HtmlFlow-Susp}, \textit{Jstachio-Virtual}, and
+\textit{Thymeleaf-Rx} approaches may be influenced by the specific template
 engines used, rather than the approach itself. When comparing the
-\textit{\hl{Reactive}}, \textit{\hl{Suspendable}}, and \textit{Virtual} approaches
+\textit{Reactive}, \textit{Suspendable}, and \textit{Virtual} approaches
 specifically with HtmlFlow, we found that all three achieve similar
@@
 virtual threads approach achieves a throughput of 3357 requests per second. In
-addition to the synchronous engines, the \textit{\hl{HtmlFlow-Rx}} approach also
+addition to the synchronous engines, the \textit{HtmlFlow-Rx} approach also
 achieves a throughput of 1760 requests per second, indicating that this
@@
 they employed is the Fortunes
-test ({\url{https://www.techempower.com/benchmarks}}, \hl{accessed on}), which involves
+test ({\url{https://www.techempower.com/benchmarks}}, accessed on 6 August 2025), which involves
 rendering a simple HTML table with only two data bindings per row.
@@
 rely on callback-based conventions such as
-\textit{\hl{continuation-passing style}} (CPS)~\cite{scheme} or
-\textit{\hl{promises}}~\cite{promise}. These approaches not only hinder
+\textit{continuation-passing style} (CPS)~\cite{scheme} or
+\textit{promises}~\cite{promise}. These approaches not only hinder
 sequential readability but also increase code verbosity, making them more
@@
 Alternatives like the \texttt{async}/\texttt{await} idiom~\cite{async_await}
-or \textit{\hl{suspend functions}}~\cite{elizarov2021coroutines} simplify
+or \textit{suspend functions}~\cite{elizarov2021coroutines} simplify
 asynchronous programming by mimicking a sequential style without blocking
@@
 
-\dataavailability{The data presented in this study are available in Github at \url{https://github.com/xmlet/comparing-non-blocking-progressive-ssr}, \hl{accessed on}.}
+\dataavailability{The data presented in this study are available in Github at \url{https://github.com/xmlet/comparing-non-blocking-progressive-ssr}, accessed on 6 August 2025.}
 
@@
 \newblock In \emph{Web Information Systems Engineering---WISE
-  2024}; Barhamgi, M., Wang, H., Wang, X., Eds.; \hl{Springer}%MDPI: newly added, please confirm it
+  2024}; Barhamgi, M., Wang, H., Wang, X., Eds.; Springer%MDPI: newly added, please confirm it
 : Singapore,  2025; pp. 458--473.
@@
 \newblock In Proceedings of the Annual Conference on USENIX
-  Annual Technical Conference, ATEC '04, \hl{Boston, MA, USA, 27 June 27--2 July 2004;} p.~21.
+  Annual Technical Conference, ATEC '04, Boston, MA, USA, 27 June 27--2 July 2004; p.~21.
 
@@
 \newblock In Proceedings of the 19th International
-  Conference on Web Information Systems and Technologies, WEBIST '23, \hl{Rome, Italy, 15--17 November  2023.}
+  Conference on Web Information Systems and Technologies, WEBIST '23, Rome, Italy, 15--17 November  2023.
 
@@
 \newblock HtmlFlow Java DSL to Write Typesafe HTML.
-\newblock Technical Report, 2017. Available online: \url{https://htmlflow.org/} \hl{(accessed on).} %MDPI: Please add the access date (format: Date Month Year), e.g., accessed on 1 January 2020.
+\newblock Technical Report, 2017.
+Available online: \url{https://htmlflow.org/} accessed on 6 August 2025.
+%MDPI: Please add the access date (format: Date Month Year), e.g., accessed on 1 January 2020.
 
@@
 \newblock Thymeleaf.
-\newblock Technical Report, 2011. Available online: \url{https://www.thymeleaf.org/} \hl{(accessed on).}
+\newblock Technical Report, 2011. Available online: \url{https://www.thymeleaf.org/} accessed on 6 August 2025.
 
 \bibitem[Fowler(2010)]{Fowler03}
-\hl{Fowler, M.} %MDPI: Refs.7 and 15 are duplicated.
-Please remove duplicated references and rearrange all the references so that they appear in numerical order.
-Please ensure that there are no duplicated references.
-\newblock {\em Domain Specific Languages}; Addison-Wesley Professional: \hl{Boston, MA, USA,} 2010.
+Fowler, M.
+\newblock {\em Domain Specific Languages}; Addison-Wesley Professional: Boston, MA, USA, 2010.
 
@@
 \newblock In Proceedings of the 13th International
-  Conference on World Wide Web, WWW '04, New York, NY, USA, \hl{17--20 May} 2004; pp.
+  Conference on World Wide Web, WWW '04, New York, NY, USA, 17--20 May 2004; pp.
   224--233.
@@
 \newblock Reactive Streams Specification.
-\newblock Technical Report, 2015. Available online: \url{https://www.reactive-streams.org/} \hl{(accessed on).}
+\newblock Technical Report, 2015. Available online: \url{https://www.reactive-streams.org/} accessed on 6 August 2025.
 
@@
 \newblock {\em Domain-Driven Design: Tackling Complexity in the Heart of
-  Software}; Addison-Wesley: \hl{Boston, MA, USA,} 2004.
-
-\bibitem[Fowler(2010)]{dslbook}
-\hl{Fowler, M.}
-\newblock {\em Domain Specific Languages}, 1st ed.; Addison-Wesley
-  Professional: \hl{Boston, MA, USA,} 2010.
+  Software}; Addison-Wesley: Boston, MA, USA, 2004.
 
@@
 Resig, J.
-\newblock {\em Pro JavaScript Techniques}; Apress: \hl{New York, NY, USA,} 2007.
+\newblock {\em Pro JavaScript Techniques}; Apress: New York, NY, USA, 2007.
 
@@
 \newblock Document Object Model (DOM) Level 3 Core Specification.
-\newblock Technical Report, 2004. Available online: \url{https://www.w3.org/TR/2004/REC-DOM-Level-3-Core-20040407/} \hl{(accessed on).}
+\newblock Technical Report, 2004. Available online: \url{https://www.w3.org/TR/2004/REC-DOM-Level-3-Core-20040407/} accessed on 6 August 2025.
 
@@
 \newblock In \emph{Practical Aspects of Declarative Languages};
-  Rocha, R., Launchbury, J., Eds.; \hl{Springer}: Berlin/Heidelberg, Germany, 2011; pp. 175--189.
+  Rocha, R., Launchbury, J., Eds.; Springer: Berlin/Heidelberg, Germany, 2011; pp. 175--189.
 
@@
   International Symposium on New Ideas, New Paradigms, and Reflections on
-  Programming and Software, \hl{Chicago, IL, USA, 20--22 October 2021}; pp. 68--84.
+  Programming and Software, Chicago, IL, USA, 20--22 October 2021; pp. 68--84.
 
@@
 \newblock Democratizing The Cloud With The .NET Reactive Framework Rx.
-\newblock In Proceedings of the Internaional Softare Development Conference, \hl{Vancouver, BC, Canada, 16--24 May 2009} %MDPI: 
-;  Available online: \url{https://qconsf.com/sf2009/sf2009/speaker/Erik+Meijer.html} \hl{(accessed on).} %MDPI:
+\newblock In Proceedings of the Internaional Softare Development Conference, Vancouver, BC, Canada, 16--24 May 2009 %MDPI: 
+;  Available online: \url{https://qconsf.com/sf2009/sf2009/speaker/Erik+Meijer.html} accessed on 6 August 2025. %MDPI:
 
@@
 \newblock RX Java.
-\newblock Technical Report, 2025. Available online: \url{https://github.com/ReactiveX/RxJava} \hl{(accessed on).}
+\newblock Technical Report, 2025. Available online: \url{https://github.com/ReactiveX/RxJava} accessed on 6 August 2025.
 
@@
 \newblock Project Reactor.
-\newblock Technical Report, 2025. Available online: \url{https://projectreactor.io/} \hl{(accessed on).}
+\newblock Technical Report, 2025. Available online: \url{https://projectreactor.io/} accessed on 6 August 2025.
 
@@
   International Workshop on Reactive and Event-Based Languages and Systems, REBLS 2021, New
-  York, NY, USA, \hl{18 October} 2021; pp. 51–60.
+  York, NY, USA, 18 October 2021; pp. 51–60.
 \newblock {\url{https://doi.org/10.1145/3486605.3486788}}.
@@
 \newblock Kotlin Language Documentation.
-\newblock Technical Report, 2016. Available online: \url{https://kotlinlang.org/docs/kotlin-docs.pdf} \hl{(accessed on).}
+\newblock Technical Report, 2016. Available online: \url{https://kotlinlang.org/docs/kotlin-docs.pdf} accessed on 6 August 2025.
 
@@
   Streaming.
-\newblock In Proceedings of the International Conference on Web \hl{Engineering, 2022;} %MDPI: please add location and date for conference
-  Springer:  \hl{Berlin/Heidelberg, Germany,} %We added the location of publisher. Please confirm
+\newblock In Proceedings of the International Conference on Web Engineering, 2022; %MDPI: please add location and date for conference
+  Springer:  Berlin/Heidelberg, Germany, %We added the location of publisher. Please confirm
  2022; pp. 391--405.
@@
 \newblock Technical Report, 2005.
-  Available online: \url{https://blog.codinghorror.com/the-lost-art-of-progressive-html-rendering/} \hl{(accessed on).}
+  Available online: \url{https://blog.codinghorror.com/the-lost-art-of-progressive-html-rendering/} accessed on 6 August 2025.
 
@@
 \newblock In Proceedings of the 9th Workshop on Hot Topics in Operating Systems
-  (HotOS IX), \hl{Lihue, HI, USA, 18–21 May 2003.}
+  (HotOS IX), Lihue, HI, USA, 18–21 May 2003.
 
@@
 \newblock In Proceedings of the 7th Workshop on Dynamic
-  Languages and Applications, DYLA '13, New York, NY, USA, \hl{1–5 July} 2013.
+  Languages and Applications, DYLA '13, New York, NY, USA, 1–5 July 2013.
 \newblock {\url{https://doi.org/10.1145/2489798.2489802}}.
@@
 \newblock In Proceedings of the 1984 ACM Symposium on LISP
-  and Functional Programming, LFP '84, \hl{Austin, TX, USA, 5--8 August 1984}; pp. 293–298.
+  and Functional Programming, LFP '84, Austin, TX, USA, 5--8 August 1984; pp. 293–298.
 \newblock {\url{https://doi.org/10.1145/800055.802046}}.
@@
 \newblock Technical Report, 2015.
-  Available online: \url{https://github.com/mbosecke/template-benchmark} \hl{(accessed on).}
+  Available online: \url{https://github.com/mbosecke/template-benchmark} accessed on 6 August 2025.
 
@@
 \newblock Technical Report, 2015.
- Available online: \url{https://github.com/jreijn/spring-comparing-template-engines} \hl{(accessed on).}
+ Available online: \url{https://github.com/jreijn/spring-comparing-template-engines} accessed on 6 August 2025.
 
@@
 \newblock In Proceedings of the 17th ACM International
-  Conference on Distributed and Event-based Systems, \hl{Neuchatel, Switzerland, 27--30 June 2023}; pp. 103--114.
+  Conference on Distributed and Event-based Systems, Neuchatel, Switzerland, 27--30 June 2023; pp. 103--114.
 
@@
 \newblock Eclipse Vert.x$^{\text{TM}}$ Reactive Applications on the JVM.
-\newblock Technical Report, 2014. Available online: \url{https://vertx.io/} \hl{(accessed~on).}
+\newblock Technical Report, 2014. Available online: \url{https://vertx.io/} accessed on 6 August 2025.
 
@@
 \newblock In Proceedings of the 48th ICT and Electronics Convention MIPRO 2025.
-Rijeka: Croatian Society for Information, Communication and Electronic Technology–MIPRO, \hl{Opatija, Croatia, 2--6 June 2025.} %MDPI: Please confirm added conference date and location.
+Rijeka: Croatian Society for Information, Communication and Electronic Technology–MIPRO, Opatija, Croatia, 2--6 June 2025.
+%MDPI: Please confirm added conference date and location.
 

### 2025-08-07 docs/article/software-3702556.tex

@@
     \caption{
-    \hl{Comparison} %MDPI: 1.
-Please confirm whether the different font colors are retained.
-If so, please confirm whether additional explanation is needed.
-2.
-Please confirm whether √ and × require additional explanation.
-3.
-Please confirm whether the bold text in the table can be deleted
- of web template technologies in the Java ecosystem.
+    Comparison of web template technologies in the Java ecosystem.
 $^{(1)}$ Enforced via annotation processor that checks templates against typed interfaces at compile time.
@@
 \vspace*{6pt}
-  \caption{\hl{Utilizing} %MDPI: Please note that changes to the position/size of figures or tables may occur during the production stage.
- DSL for HTML libraries with JATL, ScalaTags, and HtmlFlow.}
+  \caption{Utilizing DSL for HTML libraries with JATL, ScalaTags, and HtmlFlow.}
   \label{fig:dsl-idioms}
@@
 
-All benchmarks were conducted on a GitHub-hosted virtual machine under \hl{GitHub} %MDPI: Please state which version of the software was used.
-Actions, running \hl{Ubuntu 24.04.2 LTS with an AMD EPYC 7763 64-Core Processor
-(x86\_64 architecture)} %MDPI: Please state the name of the manufacturer, city, and country from where the equipment was sourced.
+All benchmarks were conducted on a GitHub-hosted virtual machine under GitHub
+Actions (GitHub, Inc., San Francisco, CA, USA), running Ubuntu 24.04.2 LTS with an AMD EPYC 7763 64-Core Processor
+(Advanced Micro Devices, Inc., Santa Clara, CA, USA)
 operating at 3.24 GHz, configured with two CPU cores and two
@@
      \includegraphics[width=0.8\textwidth]{./Graphs/presentations-webflux-jmeter.png}
-     \caption{\hl{Throughput} %MDPI: The numbers in the picture are incomplete, please modify
- (requests per second) scalability results for Spring WebFlux with \texttt{Presentations} class.}\label{fig:presentations-webflux-jmeter}
+     \caption{Throughput in requests per second (\texttt{req/sec}) for Spring
+WebFlux using the \texttt{Presentation} class, showing scalability results as
+the number of concurrent users increases (1, 2, 4, 8, 16, 32, 64, and 128)
+across different rendering approaches.}
+    \label{fig:presentations-webflux-jmeter}
 \end{figure}
@@
 engines are unable to scale effectively beyond four concurrent users. In contrast,
-\textit{HtmlFlow-Susp} scales up to 128 concurrent users, achieving 4487 %MDPI: Commas are only used for numbers with five or more digits. We have removed them in four-digit numbers. Please confirm.
+\textit{HtmlFlow-Susp} scales up to 128 concurrent users, achieving 4487
 requests per second. When blocking approaches are executed in the context of
@@
      \includegraphics[width=0.8\textwidth]{./Graphs/presentations-springmvc-jmeter.png}
-     \caption{\hl{Throughput} %MDPI: The numbers in the picture are incomplete, please modify
- (requests per second) scalability results for Spring MVC with \texttt{Presentations} class.}\label{fig:presentations-springmvc-jmeter}
+     \caption{Throughput in requests per second (\texttt{req/sec}) for Spring
+MVC using the \texttt{Presentation} class, showing scalability results as
+the number of concurrent users increases (1, 2, 4, 8, 16, 32, 64, and 128)
+across different rendering approaches.}
+\label{fig:presentations-springmvc-jmeter}
 \end{figure}
@@
      \includegraphics[width=0.8\textwidth]{./Graphs/presentations-quarkus-jmeter.png}
-     \caption{Throughput %MDPI: The numbers in the picture are incomplete, please modify
- (requests per second) scalability results for Quarkus with \texttt{Presentations} class.}\label{fig:presentations-quarkus-jmeter}
+     \caption{Throughput in requests per second (\texttt{req/sec}) for Quarkus
+      using the \texttt{Presentation} class, showing scalability results as the
+      number of concurrent users increases (1, 2, 4, 8, 16, 32, 64, and 128)
+      across different rendering approaches.}
+\label{fig:presentations-quarkus-jmeter}
 \end{figure}
@@
      \includegraphics[width=0.8\textwidth]{./Graphs/stocks-webflux-jmeter.png}
-     \caption{\hl{Throughput} %MDPI: The numbers in the picture are incomplete, please modify
- (requests per second) scalability results for Spring WebFlux with \texttt{Stocks} class.}\label{fig:stocks-webflux-jmeter}
+     \caption{Throughput in requests per second (\texttt{req/sec}) for Spring
+WebFlux using the \texttt{Stock} class, showing scalability results as
+the number of concurrent users increases (1, 2, 4, 8, 16, 32, 64, and 128)
+across different rendering approaches.}
+\label{fig:stocks-webflux-jmeter}
 \end{figure}
@@
      \includegraphics[width=0.8\textwidth]{./Graphs/stocks-springmvc-jmeter.png}
-     \caption{\hl{Throughput} %MDPI: The numbers in the picture are incomplete, please modify
- (requests per second) scalability results for Spring MVC with \texttt{Stocks} class.}\label{fig:stocks-springmvc-jmeter}
+     \caption{Throughput in requests per second (\texttt{req/sec}) for Spring
+MVC using the \texttt{Stock} class, showing scalability results as
+the number of concurrent users increases (1, 2, 4, 8, 16, 32, 64, and 128)
+across different rendering approaches.}\label{fig:stocks-springmvc-jmeter}
 \end{figure}
@@
      \includegraphics[width=0.8\textwidth]{./Graphs/stocks-quarkus-jmeter.png}
-     \caption{\hl{Throughput} %MDPI: The numbers in the picture are incomplete, please modify
- (requests per second) scalability results for Quarkus with \texttt{Stocks}.}\label{fig:stocks-quarkus-jmeter}
+     \caption{Throughput in requests per second (\texttt{req/sec}) for Quarkus
+      using the \texttt{Stock} class, showing scalability results as the
+      number of concurrent users increases (1, 2, 4, 8, 16, 32, 64, and 128)
+      across different rendering approaches.}
+\label{fig:stocks-quarkus-jmeter}
 \end{figure}
@@
 \newblock In \emph{Web Information Systems Engineering---WISE
-  2024}; Barhamgi, M., Wang, H., Wang, X., Eds.; Springer%MDPI: newly added, please confirm it
+  2024}; Barhamgi, M., Wang, H., Wang, X., Eds.; Springer
 : Singapore,  2025; pp. 458--473.
@@
 \newblock HtmlFlow Java DSL to Write Typesafe HTML.
-\newblock Technical Report, 2017.
-Available online: \url{https://htmlflow.org/} accessed on 6 August 2025.
-%MDPI: Please add the access date (format: Date Month Year), e.g., accessed on 1 January 2020.
+\newblock Technical Report, 2017. Available online: \url{https://htmlflow.org/} accessed on 6 August 2025.
 
@@
 \newblock Text Web Templates Considered Harmful.  
-\newblock In {\em Lecture Notes in Business Information Processing}; \hl{{2020};} %MDPI: We are sorry but we could not find the required information about this entry.
-Please add the name of the publisher and their location.
- pp. 69--95.
+\newblock In {\em Lecture Notes in Business Information Processing}; Springer: Cham, Switzerland, 2020; pp. 69--95.
 
@@
 Sussman, G.; Steele, G.  
-\newblock {\em Scheme: An Interpreter for Extended Lambda Calculus}; \hl{AI Memo
-  No, MIT, Artificial Intelligence Laboratory:  1975.} %MDPI: We are sorry but we could not find the required information about this entry. Please add the location of the publisher (city and country).
+\newblock {\em Scheme: An Interpreter for Extended Lambda Calculus};  
+AI Memo No. 349, MIT Artificial Intelligence Laboratory: Cambridge, MA, USA, 1975.
 
@@
 \newblock Democratizing The Cloud With The .NET Reactive Framework Rx.
-\newblock In Proceedings of the Internaional Softare Development Conference, Vancouver, BC, Canada, 16--24 May 2009 %MDPI: 
-;  Available online: \url{https://qconsf.com/sf2009/sf2009/speaker/Erik+Meijer.html} accessed on 6 August 2025. %MDPI:
+\newblock In Proceedings of the Internaional Softare Development Conference, Vancouver, BC, Canada, 16--24 May 2009 
+;  Available online: \url{https://qconsf.com/sf2009/sf2009/speaker/Erik+Meijer.html} accessed on 6 August 2025.
 
@@
   Streaming.
-\newblock In Proceedings of the International Conference on Web Engineering, 2022; %MDPI: please add location and date for conference
-  Springer:  Berlin/Heidelberg, Germany, %We added the location of publisher. Please confirm
+\newblock In Proceedings of the International Conference on Web Engineering, 2022;
+  Springer:  Berlin/Heidelberg, Germany, 
  2022; pp. 391--405.
@@
 Farago, J.; Williams, H.; Walsh, J.; Whyte, N.; Goel, K.; Fung, P.
-\newblock Object Search UI and Dragging Object Results\hl{,  ~23 2007.} %MDPI: please confirm whether can be deleted
-\newblock US Patent Applications 11/353,787, \hl{2007}%MDPI: please add day and month
-.
+\newblock Object Search UI and Dragging Object Results, February 14, 2007. 
+\newblock US Patent Applications 11/353,787, 2007.
 
@@
 \newblock Progressive Loading.
-\newblock US Patent Applications 11/364,992, \hl{2007.} %MDPI: please add day and month
+\newblock US Patent Applications 11/364,992, February 26, 2007.
 
@@
 \newblock {\em RESTful Java with JAX-RS 2.0: Designing and Developing
-  Distributed Web Services}; \hl{O'Reilly Media, Inc.: 2013.} %MDPI: We are sorry but we could not find the required information about this entry.
-Please add the location of the publisher (city and country).
+  Distributed Web Services}; O'Reilly Media, Inc.: Sebastopol, CA, USA, 2013.
 
@@
 \newblock In Proceedings of the 2022 45th Jubilee International Convention on
-  Information, Communication and Electronic Technology (MIPRO), \hl{Opatija, Croatia, 23--27 May 2022}%MDPI: Please confirm added date and location for conference
+  Information, Communication and Electronic Technology (MIPRO), Opatija, Croatia, 23--27 May 2022
 ; pp. 1466--1471.
@@
 Maurer, N.; Wolfthal, M.
-\newblock {\em Netty in Action}; \hl{Manning: 2015.} %MDPI: We are sorry but we could not find the required information about this entry. Please add the location of the publisher (city and country).
+\newblock {\em Netty in Action}; Manning Publications: Shelter Island, NY, USA, 2015.
 
@@
 Rijeka: Croatian Society for Information, Communication and Electronic Technology–MIPRO, Opatija, Croatia, 2--6 June 2025.
-%MDPI: Please confirm added conference date and location.
 
